### 2014-07-16 draft.tex
Init

\documentclass[a4paper,twoside]{style/article}

\usepackage{epsfig}
\usepackage{subfigure}
\usepackage{calc}
\usepackage{amssymb}
\usepackage{amstext}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{multicol}
\usepackage{pslatex}
\usepackage{apalike}
\usepackage{style/SCITEPRESS}
\usepackage[small]{caption}

\subfigtopskip=0pt
\subfigcapskip=0pt
\subfigbottomskip=0pt

% My Packages and Commands ---
\usepackage[normalem]{ulem}
\usepackage{xcolor}
\newcommand{\rem}[1]{\textcolor{red}{\sout{#1}}}
\newcommand{\add}[1]{\textcolor{blue}{\uline{#1}}}
\newcommand{\com}[1]{\textcolor{orange}{\uline{#1}}}
% /My Packages and Commands --

\begin{document}

\title{Insight Into Lumbar Back Pain  \subtitle{What the Lumbar Spine Tells About Your Life} }

\author{\authorname{Paul Klemm\sup{1}, Sylvia Glaßer\sup{1}, Kai Lawonn\sup{1} Henry Völzke\sup{2} and Bernhard Preim\sup{1}}
\affiliation{\sup{1}Department of Simulation and Graphics, University of Magdeburg}
\affiliation{\sup{2}Greifswald}
% TODO: Mailaddy nachtragen
\email{\{paul, sylvia\}@isg.cs.uni-magdeburg.de, voelzke@anderAddy}
}

% TODO: The paper must have at least one keyword. The text must be set to 9-point font size and without the use of bold or italic font style. For more than one keyword, please use a comma as a separator. Keywords must be titlecased.
\keywords{My Keywords in Title Case}

\abstract{The abstract should summarize the contents of the paper and should contain at least 70 and at most 200 words. The text must be set to 9-point font size.}

\onecolumn \maketitle \normalsize \vfill

\section{\uppercase{Introduction}}
\label{sec:Introduction}

Our contributions are:
\begin{itemize}
	\item Analyzing back pain using image-derived variables of 2,240 subjects.
	\item Assessing the suitability of lumbar spine shape for diagnosing back pain
	\item Analyzing correlations between between image-based and socio-demographic as well as medical parameters.
\end{itemize}

\section{\uppercase{Epidemiological Background}}
\label{sec:EpidemiologicalBackground}
\noindent \com{Background. Epidemiological Workflow, focus on statistical resilience, image data hard to analyze due to the large amount of subjects, poor image quality and lack of methods}.
\subsection{Back Pain}
\com{Back pain one of the most common diseases in the western civilization; hard to analyze; Epidemiologists interested in the \emph{healthy} aging process}.

\section{\uppercase{Related Work}}
\label{sec:RelatedWork}
\noindent \com{Our own work (VIS, VMV, BVM). Sylvias paper with reference to the methods. \textbf{More information necessary here!}}

\section{\uppercase{Materials and Method}}
\label{sec:MaterialsAndMethod}

\section{\uppercase{Experiments}}
\label{sec:Experiments}

\section{\uppercase{Conclusion}}
\label{sec:Conclusion}

\section*{\uppercase{Acknowledgements}}

\noindent SHIP is part of the Community Medicine Research net of the University of Greifswald, Germany, which is funded by the Federal Ministry of Education and Research (grant no. 03ZIK012), the Ministry of Cultural Affairs as well as the Social Ministry of the Federal State of Mecklenburg-West Pomerania. Whole-body MR imaging was supported by a joint grant from Siemens Healthcare, Erlangen, Germany and the Federal State of Mecklenburg-Vorpommern. The University of Greifswald is a member of the ‘Centre of Knowledge Interchange’ program of the Siemens AG. This work was supported by the DFG Priority Program 1335: Scalable Visual Analytics.

\vfill
\bibliographystyle{style/apalike}
{\small
\bibliography{bibliography}}

\section*{\uppercase{Appendix}}

\noindent If any, the appendix should appear directly after the
references without numbering, and not on a new page. To do so please use the following command:
\textit{$\backslash$section*\{APPENDIX\}}

\vfill
\end{document}

### 2014-07-21 draft.tex
improved on the structure, added first figures
worked on the structure

@@
 \label{sec:Introduction}
+Epidemiology is the study of causation of diseases.
+%%
+Large population studies, such as the Study of Health in Pomerania (SHIP) \ref{SHIP} gather as much information as possible about participants to be assessed towards different diseases.
+%%
+These information are used to determine risk factors for diseases, helping people to make their lifestyle healthier or helping in diagnosing a disease.
+%%
+Epidemiological research is strongly hypothesis driven.
+%
+Observations made by clinicians are translated into hypothesis, which are then statistically evaluated using data variables from epidemiological studies.
 
+Modern cohort studies often comprise medical image data.
+%%
+These data are hard to analyze, since segmentation algorithms are not generally available and need to be custom-made for each body structure.
+%%
+Segmentation data it is usually analyzed by abstracting it into key figures, making it statistically comparable to non-image variables and retrieve correlations.
+%%
+
+Back pain is one of the most frequent diseases in the western civilization.
+%%
+
+Our goal is to combine data mining algorithm with data visualization to provide insight into the quality of image derived data to analyze if it acts as a risk factor for a disease.
+%\com{Combine Data Mining Algo}
+
+More than 5 variables are rarely analyzed simultaneously.
+%%
+%The variables analyzed simultaneously is limited to 
 Our contributions are:
@@
 \section{\uppercase{Epidemiological Background}}
+ \ref{VIS14}
 \label{sec:EpidemiologicalBackground}
@@
 \section{\uppercase{Materials and Method}}
+\begin{figure}[!h]
+  %\vspace{-0.2cm}
+  \centering
+  \includegraphics[width=0.5\textwidth]{figures/age-gender}
+  \caption{Age-Gender}
+  \label{fig:example1}
+ \end{figure}
 \label{sec:MaterialsAndMethod}
+
+\subsection{The Lumbar Spine Data Set}
+Our data set compiled to analyze lumbar back pain comprises of 6,753 subjects from two cohorts (4,420 from \texttt{SHIP-Trend-0} and 2,333 from \texttt{SHIP-2}).
+%%
+
+%%
+
+\subsubsection{Image Data}
+\com{From VIS'14 Paper}
+The lumbar spine was detected in the image data using a hierarchical finite element method by Rak et al.
+[32].
+This semiautomatic method requires the user to initialize the tetrahedron-based finite element models (FEM) with a click on the L3 vertebra.
+Two user-defined landmarks on the top and bottom of the L3 vertebra describe an initial model height estimation.
+The model uses a weighted sum of T1and T2-weighted MR images to detect the lumbar spine shape.
+Once registered, it captures information about the shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae [21].
+Due to incorrect initialization, strongly deformed spines, contrast differences and artifacts, the model was not able to detect lumbar spines for all subjects.
+We obtained and worked with 2,540 tetrahedron models of the lumbar spine.
+For clustering, we extracted the centerline of the lumbar spine canal, which captures information about lordosis and scoliosis (the medical terms for spine curvature) [21].
+
+\subsubsection{Non-Image Data}
+\com{Analysis Using Heterogeneous Correlation}
+The variables range from somatometric variables describing body measures to medical examinations, such as laboratory tests as well as lifestyle factors as sport activity or nutrition.
+
+\subsection{Creating a Decision Tree using C5.0}
+\com{C4.5 Algorithm for creating decision trees; }
+
+\subsection{Parameter Assessment}
+
+\section{\uppercase{Experiments}}
+\label{sec:Experiments}
+\subsection{Results}
+\com{We can discriminate back pain using non-image data, but not with image-derived parameter}
 

### 2014-07-24 draft.tex
Reworked Structure with latest results

@@
 	\item Analyzing correlations between between image-based and socio-demographic as well as medical parameters.
+	\item Identification of most important variables using data mining methods \com{noch schoen schreiben}
+	\item semiquantitative Auswertmöglichkeiten werden dem User oft zur verfügung gestellt
+	\item Techniken sind als Teil des IVA-Frameworkes zu verstehen
 \end{itemize}
@@
 
+Risk factors for are not completely known, thus epidemiologists search for new correlations in cohort studies.
+
 \section{\uppercase{Related Work}}
@@
 
-\section{\uppercase{Materials and Method}}
+\section{The Lumbar Spine Data Set}
 \begin{figure}[!h]
@@
 
-\subsection{The Lumbar Spine Data Set}
 Our data set compiled to analyze lumbar back pain comprises of 6,753 subjects from two cohorts (4,420 from \texttt{SHIP-Trend-0} and 2,333 from \texttt{SHIP-2}).
@@
 %%
+\subsection{Non-Image Data}
+\com{Analysis Using Heterogeneous Correlation}
+The variables range from somatometric variables describing body measures to medical examinations, such as laboratory tests as well as lifestyle factors as sport activity or nutrition.
 
-\subsubsection{Image Data}
+\com{Es wurde mit einem weiten Spektrum Daten gesammelt, vom denen man glaubt, dass sie irgendwas mit der Wirbelsäulenverformung und dem Alterungsprozess zu tun haben.
+It is an active research area to determine these correlations.}
+
+\subsection{Image Data}
 \com{From VIS'14 Paper}
@@
 
-\subsubsection{Non-Image Data}
-\com{Analysis Using Heterogeneous Correlation}
-The variables range from somatometric variables describing body measures to medical examinations, such as laboratory tests as well as lifestyle factors as sport activity or nutrition.
+%\subsection{Creating a Decision Tree using C5.0}
+%\com{C4.5 Algorithm for creating decision trees; }
 
-\subsection{Creating a Decision Tree using C5.0}
-\com{C4.5 Algorithm for creating decision trees; }
-
-\subsection{Parameter Assessment}
+%\subsection{Parameter Assessment}
+%Based on first results
 
@@
 \label{sec:Experiments}
-\subsection{Results}
+
+\subsection{Preliminary Results}
+\begin{itemize}
+	\item Direkter Vergleich
+	\item Verteilung des Means
+	\item Korrelationsmatrix
+	\item PCA
+\end{itemize}
+Weil das nicht ausreichend war, haben wir einen C5.0 Baum erstellt, um alle potentiellen Zusammenhänge zwischen bildbezogenen und Nichtbildbezogenen Parametern herzustellen - Was kann man mit den Bilddaten erklären und was nicht?
+
+\subsection{Experimental Settings and Results}
+Decision Tree erklaeren, ebenfalls Umrechnen in die Dummy-Variablen. Was sind die Zielvariablen
+\begin{itemize}
+\item Altersgruppierung
+\item Male/Female
+\item Gewicht (BMI - dünn, normal, dick)
+\end{itemize}
+% Creating a Decision Tree using C5.0
+
+%\com{C4.5 Algorithm for creating decision trees; }
+
+\section{\uppercase{Evaluation of Decision Trees}}
+\label{sec:Experiments}
+\begin{itemize}
+\item Vorgehen mit Scatterplot erklären
+\item Beschreiben der Ergebnisse
+\item Beschreiben des interaktiven Systems zum Anzeigen der Decision Trees
+\end{itemize}
+
 \com{We can discriminate back pain using non-image data, but not with image-derived parameter}
-
-\section{\uppercase{Experiments}}
-\label{sec:Experiments}
+Welche parameter können wir gut unterscheiden
 

### 2014-07-28 draft.tex
Wrote first draft of epidemiological background chapter

@@
 %%
-These data are hard to analyze, since segmentation algorithms are not generally available and need to be custom-made for each body structure.
-%%
-Segmentation data it is usually analyzed by abstracting it into key figures, making it statistically comparable to non-image variables and retrieve correlations.
+\com{Short summary}
 %%
@@
 \section{\uppercase{Epidemiological Background}}
- \ref{VIS14}
 \label{sec:EpidemiologicalBackground}
-\noindent \com{Background. Epidemiological Workflow, focus on statistical resilience, image data hard to analyze due to the large amount of subjects, poor image quality and lack of methods}.
-\subsection{Back Pain}
-\com{Back pain one of the most common diseases in the western civilization; hard to analyze; Epidemiologists interested in the \emph{healthy} aging process}.
+%\noindent \com{Background. Epidemiological Workflow, focus on statistical resilience, image data hard to analyze due to the large amount of subjects, poor image quality and lack of methods}.
+\noindent Epidemiology is the study of dissemination, causes and results of health-related states and events.
+%%
+Epidemiological reasoning relies on a strict statistically driven workflow \cite{Fletcher}:
+%%
+\begin{itemize}
+%%
+	\item Physicians formulate hypotheses based on observations made in their clinical practice.
+%%
+	\item To assess a hypothesis, epidemiologists compile a list of variables depicting it.
+%%
+	\item Statistical methods, such as regression analysis, assess the association of selected variables with the investigated disease.
+%%
+\end{itemize}
+%%
+Mutually dependent variables make this analysis challenging.
+%%
+Many diseases, such as different cancer types, are more likely with increasing age.
+%%
+When for example analyzing influences of nutrition to prostate cancer, the results need to be age-normalized.
+%%
+Age acts as an \emph{confounder} for prostate cancer.
+%%
+These \emph{confounding} variables are often hard to find.
+%%
+Statistical correlation does not imply causation--epidemiologists need to assess the medical soundness of the statistical results.
 
-Risk factors for are not completely known, thus epidemiologists search for new correlations in cohort studies.
+\subsection{Lower Back Pain}
+% \com{Back pain one of the most common diseases in the western civilization; hard to analyze; Epidemiologists interested in the \emph{healthy} aging process}.
+%
+% Risk factors for are not completely known, thus epidemiologists search for new correlations in cohort studies.
+Lower back pain is one of the most frequent diseases in the western civilization.
+%%
+Epidemiologists assume associations between lumbar back pain and lifestyle factors.
+%%
+These include nutrition, sporting activities and body posture at work.
+%%
+The exact causes as well as particularly vulnerable risk groups are not known.
+%%
+Interaction between risk factors are also not known.
+%%
+
+Epidemiologists want to characterize the healthy aging process of the spine.
+%%
+To achieve this, they have to analyze the lumbar spine shape as well as the mentioned lifestyle factors.
+%% -------------------
+\subsection{Epidemiological Data}
+Epidemiological data can stem from a wide range of studies.
+%%
+The study type depends on the condition of interest.
+%%
+Most common are case-control studies, analyzing one specific disease and its influences.
+%%
+We focus on data from large scale cohort studies.
+%%
+These studies aim to collect as much data as possible for each subject.
+%%
+As a result, these data can be analyzed regarding many diseases and conditions.
+%%
+
+Epidemiological data is heterogenous and incomplete.
+%%
+Women-specific question can only be acquired for female subjects, data about a disease treatment only for subjects suffering from this condition.
+%%
+Therefore, statistical analysis has to take missing data into account.
+%%
+
+Epidemiologists acquire data using a wide range of techniques, such as medical examinations, self reported questionnaires or genetic examinations.
+%%
+This yields a heterogenous information space.
+%%
+To compare these data, information reduction techniques are applied.
+%%
+For example, continuous data, such as age is often discretized into age-bins.
+%%
+Every information reduction can introduce a bias to the data, since it reflects an assumption about the data.
+%%
+Using age to divide subjects into \emph{young} and \emph{old} categories can distort statistical results.
+%%
+This distortion is reduced with increasing number of discretization steps.
+%%
+
+Modern cohort studies often comprise medical image data.
+%%
+These data are hard to analyze, since segmentation algorithms are not generally available and need to be custom-made for each body structure.
+%%
+Segmentation data is usually analyzed by abstracting it into key figures, such as diameters or distances.
+%%
+These numeric values can be compared with non-image variables to retrieve correlations.
+%%
 

### 2014-07-28 draft.tex
Added Information on deriving image data

@@
 Statistical correlation does not imply causation--epidemiologists need to assess the medical soundness of the statistical results.
-
-\subsection{Lower Back Pain}
-% \com{Back pain one of the most common diseases in the western civilization; hard to analyze; Epidemiologists interested in the \emph{healthy} aging process}.
-%
-% Risk factors for are not completely known, thus epidemiologists search for new correlations in cohort studies.
-Lower back pain is one of the most frequent diseases in the western civilization.
-%%
-Epidemiologists assume associations between lumbar back pain and lifestyle factors.
-%%
-These include nutrition, sporting activities and body posture at work.
-%%
-The exact causes as well as particularly vulnerable risk groups are not known.
-%%
-Interaction between risk factors are also not known.
-%%
-
-Epidemiologists want to characterize the healthy aging process of the spine.
-%%
-To achieve this, they have to analyze the lumbar spine shape as well as the mentioned lifestyle factors.
-%% -------------------
+%%------------------------------
 \subsection{Epidemiological Data}
@@
 These numeric values can be compared with non-image variables to retrieve correlations.
+%%--------------------------------
+\subsection{Lower Back Pain}
+%%
+The lower (\emph{lumbar}) spine is the most stressed part of the spine.
+%%
+Lower back pain is one of the most frequent diseases in the western civilization.
+%%
+Epidemiologists assume associations between lumbar back pain and lifestyle factors.
+%%
+These include nutrition, sporting activities and body posture at work.
+%%
+The exact causes as well as particularly vulnerable risk groups are not known.
+%%
+Potential \emph{confounding} effects are also subject of current research.
 %%
 
+Epidemiologists want to characterize the healthy aging process of the spine.
+%%
+To achieve this, they have to analyze the lumbar spine shape as well as the mentioned lifestyle factors.
+%% -------------------------------
 \section{\uppercase{Related Work}}
@@
 \label{sec:MaterialsAndMethod}
-
-Our data set compiled to analyze lumbar back pain comprises of 6,753 subjects from two cohorts (4,420 from \texttt{SHIP-Trend-0} and 2,333 from \texttt{SHIP-2}).
+%%
+Our approach allows to analyze many variables simultaneously.
+%%
+Therefore, epidemiologists compiled the data set with a with a wide range of variables possibly correlating with lumbar back pain.
+%%
+The data set comprises of 6,753 subjects from two cohorts (4,420 from \texttt{SHIP-Trend-0} and 2,333 from \texttt{SHIP-2}).
+%%
+%It is an active research area to determine these correlations.
 %%
@@
 \subsection{Non-Image Data}
-\com{Analysis Using Heterogeneous Correlation}
+%\com{Analysis Using Heterogeneous Correlation}
 The variables range from somatometric variables describing body measures to medical examinations, such as laboratory tests as well as lifestyle factors as sport activity or nutrition.
 
-\com{Es wurde mit einem weiten Spektrum Daten gesammelt, vom denen man glaubt, dass sie irgendwas mit der Wirbelsäulenverformung und dem Alterungsprozess zu tun haben.
-It is an active research area to determine these correlations.}
-
 \subsection{Image Data}
+\begin{figure}[!h]
+  %\vspace{-0.2cm}
+  \centering
+  \includegraphics[width=0.5\textwidth]{figures/centerline}
+  \caption{Centerline}
+  \label{fig:centerline}
+\end{figure}
 \com{From VIS'14 Paper}
@@
 For clustering, we extracted the centerline of the lumbar spine canal, which captures information about lordosis and scoliosis (the medical terms for spine curvature) [21].
+%%
+
+We have to assess the model accuracy to extract key figures from it.
+%%
+The detection model depicts the vertebrae positions, but lacks detailed information about their volume.
+%%
+It captures reliable information about spine canal curvature.
+%%
+In a previous work \cite{KlemmVMV13, KlemmVIS14}, we extracted a centerline representation of the lumbar spine canal from the detection model (Fig.~\ref{fig:centerline} (a)).
+%%
+Using the Frenet formulas \cite{Frenet}, we extracted the following metrics from the model (Fig.~\ref{fig:centerline} (b)):
+%%
+%Thus, we extracted the following metrics from the model:
+%%
+%\paragraph{Curvature}
+\begin{itemize}
+	% \item The curvature of the centerline is calculated using the summed mean curvature between adjacent points:
+% %%
+% $\sum \frac{curvature_i}{curvature_all}$.
+	\item \emph{Curvature} is calculated as weighted sum of curvature between all adjacent points describing the centerline: $\sum \frac{curvature_i}{curvature_all}$.
+%%
+	\item \emph{Torsion} (deviation of a curve from the course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum \frac{torsion_i}{torsion_all}$.
+%%
+\end{itemize}
+%%
+These figures are also extracted in the sagital, coronal and transversal projection of the model.
+%%
+We assess the information gain of each dimension using these projections.
+%%
+This gives us a total of 8 image-derived parameters.
 

### 2014-07-29 draft.tex
added stuff on experiments chapter

@@
 	\item \emph{Torsion} (deviation of a curve from the course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum \frac{torsion_i}{torsion_all}$.
+	\item \emph{Curvature angle} is defined by the middle point of the spine canal centerline as \emph{vertex} and the line between middle point and top/bottom point as \emph{sides}.
 %%
@@
 %%
-These figures are also extracted in the sagital, coronal and transversal projection of the model.
+These figures are also extracted in the sagittal, coronal and transversal projection of the model.
 %%
@@
 %%
-This gives us a total of 8 image-derived parameters.
+This gives us a total of 9 image-derived parameters.
 
@@
 \label{sec:Experiments}
+Spine shape is confounded by several somatometric variables.
+%%
+Larger people have also a longer spine and its shape is more straight compared to smaller subjects.
+%%
+Since men are on average taller than women, gender is another confounder.
+%%
+Large body weight increases the spine load, resulting in a bend shape.
+%%
+All these variables need to be taken into account, when correlating spine curvature and torsion with non-image parameter.
+%%
+Since the gender confounder mainly encodes body-height, we decided to divide subjects into body-size groups.
+%%
+To avoid small outlier groups, epidemiologists recommend using quartiles to discretize metric variables \cite{KlemmVIS14}.
+\subsection{Preliminary Results}
+\begin{figure}[!h]
+  %\vspace{-0.2cm}
+  \centering
+  \includegraphics[width=0.5\textwidth]{figures/image-parameter-range}
+  \caption{image-parameter-range}
+  \label{fig:image-parameter-range}
+\end{figure}
+%%
+\noindent At first we assessed the image parameter.
+%%
+Figure~\ref{fig:image-parameter-range} (right) shows the range of each variable as box plot.
+%%
+The projections to the transversal planes attract attention as they have many outliers.
+%%
+We conclude that curvature is not as reliable on the transversal plane as it is on the other planes.
+%%--------------------------------
+\subsubsection{Correlation Matrix}
+%%
+We calculated an association matrix to to assess correlations between the image parameters.
+%%
+Curvature, curvature angle and torsion correlate with their planar projections.
+%%
+Also the mean curvature and the curvature angle correlate by a factor of -0.89.
+%%
+Torsion and curvature/curvature angle do not correlate.
+%%--------------------------------------------------------
+\subsubsection{Correlation of Image Parameters With Back Pain}
+%%
+As first experiment we correlated the shape parameter with the binary back pain indicator.
+%%
+No statistical significant correlation could be observed through all subject groups.
+\com{Hier weitermachen. GGPLOT erklaeren.}
 
-\subsection{Preliminary Results}
 \begin{itemize}

### 2014-07-31 draft.tex
Worked on C5 chapter

@@
 \usepackage{xcolor}
+\usepackage[]{algorithm2e}
 \newcommand{\rem}[1]{\textcolor{red}{\sout{#1}}}
@@
 
-\abstract{The abstract should summarize the contents of the paper and should contain at least 70 and at most 200 words. The text must be set to 9-point font size.}
+\abstract{The abstract should summarize the contents of the paper and should contain at least 70 and at most 200 words.
+The text must be set to 9-point font size.
+Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
+Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
+Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
+Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.}
 
@@
 % $\sum \frac{curvature_i}{curvature_all}$.
-	\item \emph{Curvature} is calculated as weighted sum of curvature between all adjacent points describing the centerline: $\sum \frac{curvature_i}{curvature_all}$.
+	\item \emph{Curvature} is calculated as weighted sum of curvature between all adjacent points describing the centerline: $\sum \frac{curvature_i}{curvature_{all}}$.
 %%
-	\item \emph{Torsion} (deviation of a curve from the course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum \frac{torsion_i}{torsion_all}$.
+	\item \emph{Torsion} (deviation of a curve from the course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum \frac{torsion_i}{torsion_{all}}$.
 	\item \emph{Curvature angle} is defined by the middle point of the spine canal centerline as \emph{vertex} and the line between middle point and top/bottom point as \emph{sides}.
@@
 \label{sec:Experiments}
-Spine shape is confounded by several somatometric variables.
+\noindent Spine shape is confounded by several somatometric variables.
 %%
@@
 \subsection{Preliminary Results}
-\begin{figure}[!h]
+\begin{figure*}[htb]
   %\vspace{-0.2cm}
   \centering
-  \includegraphics[width=0.5\textwidth]{figures/image-parameter-range}
+  \includegraphics[width=1.0\textwidth]{figures/image-parameter-range}
   \caption{image-parameter-range}
   \label{fig:image-parameter-range}
-\end{figure}
+\end{figure*}
 %%
-\noindent At first we assessed the image parameter.
+\noindent As first experiment we correlated the shape parameter with the binary back pain indicator using a pairs plot \cite{ggally}.
+%%
+Pairs plots visualize continuous and categorical variables pairwise in a matrix view.
+%%
+Scatterplots display pairs of continuous variables, bar charts for continuous variable as a function of a categorical variable and box plots pairs of categorical variables.
+%%
+Since our image-derived variables are metric, their pairwise combinations are visualized using scatterplots.
+%%
+The combination of the image variables with back pain is visualized as histogram at the left side of the matrix and as box plot on the right side.
 %%
@@
 \subsubsection{Correlation Matrix}
-%%
 We calculated an association matrix to to assess correlations between the image parameters.
 %%
-Curvature, curvature angle and torsion correlate with their planar projections.
+The pearson correlation coefficients between the numeric variables are depicted right of the main diagonal of Figure~\ref{fig:image-parameter-range}.
+%%
+Curvature, curvature angle and torsion correlate strongly with their planar projections.
 %%
@@
 %%
-Torsion and curvature/curvature angle do not correlate.
+%Torsion and curvature/curvature angle do not correlate.
+Torsion does not correlate with any other image-derived variable.
 %%--------------------------------------------------------
 \subsubsection{Correlation of Image Parameters With Back Pain}
+Figure~\ref{fig:image-parameter-range} shows the distribution for all subjects.
 %%
-As first experiment we correlated the shape parameter with the binary back pain indicator.
+We created a pairs plot for subject groups derived by dividing them into quantile groups of body-size.
 %%
 No statistical significant correlation could be observed through all subject groups.
-\com{Hier weitermachen. GGPLOT erklaeren.}
+%%
+The box plots show no difference between subjects with and without back pain.
 
+\subsubsection{Assessing the Information Gain Using the PCA}
+%%
+To determine the information gain per image-derived variables, we calculated a PCA and compared the loadings per dimension.
+%%
+The first three principal components explain 75\% of variance in the image-derived variables.
+%%
+The first principal component explains 47\% of the variance and weights primarily \emph{mean curvature}, \emph{curvature sagittal}, \emph{curvature angle} and \emph{curvature angle sagittal}.
+%%
+The second component, adding 16\% of variance, weigths  \emph{curvature coronal} and \emph{curvature angle coronal}.
+%%
+The third component explains 12\% of variance and weights \emph{torsion} and \emph{curvature transverse}.
+%%
+This supports our prior conclusion about the low information gain of the transverse planes.
+%%
+\emph{Torsion} also adds little variance to the information space.
+%% ------------------------------------------
+\subsubsection{Heterogenous Correlations}
+%%
+We then expanded our focus on correlations of image-derived parameters with all other non-image variables.
+%%
+We applied a heterogenous correlation technique to derive correlations between all variables is the data set.
+%%
+The method uses the following correlation metrics for the different type combination:
+%%
 \begin{itemize}
-	\item Direkter Vergleich
-	\item Verteilung des Means
-	\item Korrelationsmatrix
-	\item PCA
+%%
+\item \emph{pearson product-moment} for two continuous variables, 
+\item \emph{polyserial} correlation for one continuous and one categorical variable, and
+\item \emph{polychoric} correlations for two categorical variables.
 \end{itemize}
-Weil das nicht ausreichend war, haben wir einen C5.0 Baum erstellt, um alle potentiellen Zusammenhänge zwischen bildbezogenen und Nichtbildbezogenen Parametern herzustellen - Was kann man mit den Bilddaten erklären und was nicht?
+%%
+All correlation values are scaled between 0 - no correlation and 1 - identical.
+%%
+Some variables are too sparse for calculating correlations, for example \com{some variables from calculation here}.
+%%
+We display the resulting \emph{contingency matrix} using a heat map, mapping correlation values to color brightness with white for 0 and dark blue for 1 \cite{Klemm2014VIS}.
+%%
+We calculated the contingency matrix for all size groups and looked for correlations between image- and non-image derived variables.
+%%
+The resulting contingency matrix shows no strong correlation with any of the parameter.
+%%
+Only weak correlations could be found for \emph{Mean Curvature} with \emph{gender} (0.42), \emph{body size} (0.39) and \emph{number of born children} (0.29).
+%%
+One surprising result was the small correlation with \emph{torsion}, which correlated with almost no variables (p values between 0 and 0.05) and \emph{parkinson} (0.24).
+%%
+This observation brought us to the conclusion to incorporate more sophisticated data mining techniques to assess the influence of the image-derived parameters.
+
+% \subsubsection{Diese Themen hier unterbringen}
+% \begin{itemize}
+% 	\item Direkter Vergleich
+% 	\item Verteilung des Means
+% 	\item Korrelationsmatrix
+% 	\item PCA
+% \end{itemize}
+% Weil das nicht ausreichend war, haben wir einen C5.0 Baum erstellt, um alle potentiellen Zusammenhänge zwischen bildbezogenen und Nichtbildbezogenen Parametern herzustellen - Was kann man mit den Bilddaten erklären und was nicht?
+%
+% HetCor hier noch mit unterbringen.
+Damit kann man eine Korrelationsmatrix unabhängig vom Datentyp berechnen, um eine große Korrellationsmatrix zu erhalten (Supplementary Material).
+Hier können die Einträge in den Zeilen der Bildparameter angeschaut werden, um zu schauen, ob man eine Korrellation mit bestimmten Parametern finden kann!
+Das müsste auch wieder für alle Größengruppen passieren!
+(https://stats.stackexchange.com/questions/101016/pca-and-component-scores-based-on-a-mix-of-continuous-binary-and-categorical-va)
 
@@
 \section{\uppercase{Evaluation of Decision Trees}}
-\label{sec:Experiments}
+\label{sec:DecisionTrees}
+%%
+\noindent As described before, correlation coefficients fail to infer back pain status based on lumbar spine canal curvature and torsion.
+%%
+We rely on predictive classification trained to obtain a complex rule set on how combination of the image-parameter explain non-image variables.
+%%
+Decision trees are a popular classification method in data mining for creating predictive models.
+%%
+Leaves of a decision tree represent class labels, branches represent feature conjunctions leading to the class labels.
+%%
+Decision trees are easy to understand and to read.
+%%
+They work with numerical as well as categorical data.
+%%
+This allows epidemiological domain experts to interpret the results without having deep knowledge about the algorithm creating the tree.
+%%
+Readability is only granted for small trees, complex structures with many branches are not desirable.
+%%
+\com{Quelle}
+
+\subsection{The C4.5/C5.0 Algorithm}
+Using a trained data set, the C4.5 algorithm builds decision trees based on information entropy.
+%%
+Such a calculation requires a numeric or categorical target variable.
+%%
+The algorithm then tries to find a decision tree, which divides the samples using the input variables just like the target variables.
+%%
+This means, that every node in the tree represents the attribute which splits the data most efficient into the target subset.
+%%
+The pseudocode for the algorithm is defined in Algorithm~\ref{alg:c4.5} \cite{C4.5Algorithm}.
+\begin{algorithm}[]
+ Check for base cases\;
+ \For{each attribute $a$}{
+  Find the normalized information gain ratio from splitting on $a$\;
+ }
+ Let $a_{best}$ be the attribute with the highest normalized information gain\;
+ Create a decision node that splits on $a_{best}$\;
+ Recurse on the sublists obtained by splitting on $a_{best}$, and add those nodes as children of node\;
+ \caption{Building a decision tree using the C4.5 Algorithm}
+ \label{alg:c4.5}
+\end{algorithm}
+C5.0 is developed to produce smaller decision trees than C4.5 and improve the execution time.
+%%
+We use the \texttt{R} implementation of C5.0 \cite{c5.0lassification}.
+%%
+% \begin{enumerate}
+% %%
+% \item Check for base cases
+% %%
+% \item For each attribute $a$
+% %%
+% \begin{enumerate}
+% %%
+% \item Find the normalized information gain ratio from splitting on $a$
+% %%
+% \end{enumerate}
+% %%
+% \item Let $a_{best}$ be the attribute with the highest normalized information gain
+% %%
+% \item Create a decision node that splits on $a_{best}$
+% %%
+% \item Recurse on the sublists obtained by splitting on $a_{best}$, and add those nodes as children of node
+% %%
+% \end{enumerate}
+%%
+Categorical attributes with more levels are biased with more information gain in a decision tree \cite{deng2011bias}.
+%%
+Creating dummy variables bypasses this problem.
+%%
+
+The actual use for the resulting tree is the classification of new observations (subjects).
+%%
+Yet, we are interested in the decision rules and the classification accuracy.
+
+\subsection{Interactive Display of Decision Trees}
+\com{Alle Variablen einzeln durchgehen ist nervig, man braucht ein interaktives System hierfür!}
+\com{Describe Shiny App!}
+
+\subsection{Das hier unterbringen}
 \begin{itemize}
-\item Vorgehen mit Scatterplot erklären
-\item Beschreiben der Ergebnisse
-\item Beschreiben des interaktiven Systems zum Anzeigen der Decision Trees
+\item \com{Vorgehen mit Scatterplot erklären}
+\item \com{Beschreiben der Ergebnisse}
+\item \com{Beschreiben des interaktiven Systems zum Anzeigen der Decision Trees}
 \end{itemize}
 
-\com{We can discriminate back pain using non-image data, but not with image-derived parameter}
-Welche parameter können wir gut unterscheiden
+\com{We can discriminate back pain using non-image data, but not with image-derived parameter. Welche parameter können wir gut unterscheiden}
 

### 2014-08-01 draft.tex
continued main section

@@
 %%
-Some variables are too sparse for calculating correlations, for example \com{some variables from calculation here}.
+Some variables are too sparse for calculating correlations, for example \com{TODO: some variables from calculation here; TODO: Image?, eventuell interactiven Tile-Plot}.
 %%
@@
 \subsection{Interactive Display of Decision Trees}
-\com{Alle Variablen einzeln durchgehen ist nervig, man braucht ein interaktives System hierfür!}
-\com{Describe Shiny App!}
+We have to create a decision tree for every non-image variable to see which one can be explained by image-derived variables.
+%%
+Since we have 134 non-image variables, the calculation yields a corresponding number of trees.
+%%
+Dividing the subjects into four groups using body-size, the number increases to 402 trees.
+%%
+To make the results cognitive feasible, we have to abstract the results of the classification.
+%%------------
+\subsubsection{Visualization of Classification Results}
+%%
+We follow the visual information-seeking mantra and want to provide an overview first, then details-on-demand \ref{shneiderman1996}.
+%%
+The optimal classification uses a few rules to precisely characterize the target variable.
+%%
+Therefore, we are interested in \emph{small trees} with a \emph{low classification error rate}.
+%%
+The error rate has to be judged depending on the number of classes of the target variable.
+%%
+An error rate of 50\% for a target variable with two manifestations makes the classifier equal to throwing a coin.
+%%
+The same error rate is acceptable a target variable with more than two manifestations.
+%%
+Picking a value for a variable of four manifestations with an error rate of 50\% is better than the random selection (error rate of 75\%).
+%%
+The following scales the error rate so that 0 denotes perfect classification and 1 is equal to random selection.
+%%
+\begin{equation}
+error_{normalized} = \frac{error}{100 - \frac{1}{class number}}
+\end{equation}
+%%
+This allows for comparability of error rates between variables.
+%%
+Abstracting the classification results into two continuos variables allows us to visualize the result of 134 classifications in one scatterplot.
+%%
+By mapping the group affiliation (in our case body-size) to color or shape, we can even plot all 402 results in a single visualization (Fig.~\ref{fig:decision-tree} a).
+%%
+\com{ToDo: Dummy-Variables}
+%%------------
+\subsubsection{Interaction With the Visualization}
+%%
+The described visualization provides a good overview over the classification results.
+%%
+We however still want to be able to look display \emph{details-on-demand} \ref{shneiderman1996} and examinate a decision tree in detail.
+%%
+This is realized by clicking on a entry on the visualization, which then displays the corresponding decision tree in detail (Fig.~\ref{fig:decision-tree} b).
+%%
+This allows for sequentially analyzing the interactions.
+%%------------
+\subsubsection{Implementation}
+All analysis are carried out using \texttt{R}, a popular programming language for statistical calculations and visualizations.
+%%
+The interactive visualizations are realized using the \texttt{ggvis}\footnote{GGVIS Footnote} package.
+%%
+As opposed to the standard \texttt{R} plots, \texttt{ggvis} allows to adjust visualization parameters using UI controls, such as sliders.
+%%
+We implemented our application as web based prototype using \texttt{R Shiny}\footnote{R Shiny footnote}.
+%%
+The package combines the power of \texttt{R} with advantages of web-based applications.
+%%
+It allows us to quickly exchange results with our collaborating epidemiological partners.
+%%
+They can try out the technique without installing anything.
+%%
+Exchanging software becomes as easy as exchanging a hyperlink.
+%%---------
+\subsection{Results}
 
-\subsection{Das hier unterbringen}
-\begin{itemize}
-\item \com{Vorgehen mit Scatterplot erklären}
-\item \com{Beschreiben der Ergebnisse}
-\item \com{Beschreiben des interaktiven Systems zum Anzeigen der Decision Trees}
-\end{itemize}
+\subsection{Themen hier unterbringen}
 

### 2014-08-13 draft.tex
Started Results part

@@
 \begin{equation}
-error_{normalized} = \frac{error}{100 - \frac{1}{class number}}
+error_{normalized} = \frac{error}{100 - \frac{100}{class number}}
 \end{equation}
@@
 %%
-\com{ToDo: Dummy-Variables}
+
+Many follow-up variables are sparse, such as medication of diabetes or reason of early retirement.
+%%
+The classification algorithm may produce higher accuracy for variables with less subjects due to the small sample size.
+%%
+This makes these results less reliable.
+%%
+We map the number of subjects associated with a variable to point diameter in the scatter plot.
+%%
+This allows to instantly assess the reliability of the result.
+%%
+\subsubsection{Dummy Variables}
+Dummy variables abstract a categorical variable with multiple manifestations into several binary variables.
+%%
+Each binary variables encodes the presence of a manifestation.
+%%
+For example, a pain indicator variable ranging from \emph{1 - no pain} to \emph{4 - large pain} is subdivided into four binary variables (\emph{No pain - Yes/No} to \emph{Large Pain - Yes/no}).
+%%
+One subject can only have one of these variables set to true.
+%%
+This is useful for our classification, because it allows to determine which manifestation of a variable can be described best using the image data parameter.
+%%\com{ToDo: Dummy-Variables}
 %%------------
@@
 %%
-We however still want to be able to look display \emph{details-on-demand} \ref{shneiderman1996} and examinate a decision tree in detail.
+We however still want to be able to look display \emph{details-on-demand} \ref{shneiderman1996} and examine a decision tree in detail.
 %%
@@
 \subsection{Results}
+We ran the analysis using different subject groups:
+\begin{enumerate}
+	%%
+	\item All subjects,
+	%%
+	\item subdivision into \emph{male} and \emph{female},
+	%%
+	\item subdivision into \emph{age quantiles}, yielding the groups (21, 43.8] (43.8, 54] (54, 64] (64, 90],
+	%%
+	\item subdivision into \emph{size quantiles}, yielding the groups (139, 164], (164, 171], (171, 177], (177, 202].
+	%%
+\end{enumerate}
+
+\subsubsection{All Variables}
+% None of the pain indicators can be reliably described using the image-based parameter.
+None of the pain indicators can be described reliably using the image-based parameter.
+%%
+The perfect categorization of the \emph{dialysis} parameter was conspicuous.
+%%
+This is due to the subject distribution among the manifestation.
+%%
+Only two subjects receive a dialysis.
+%%
+The good discrimination is not meaningful.
+%%
+Gender can be discriminated with an error rate of 31,5\% using 7 rules and incorporates only curvature and curvature angle variables.
+%%
+Surprisingly high is the classification of medication for high blood pressure.
+%%
+1,058 subjects are classified with an error rate of 23,8\% solely based on coronal mean curvature.
+%%
+Other features well classified include marital status, foot pain and number of born children.
+%%
+
+The dummy variable analysis yields fewer results.
+%%
+It shows that subjects between 64 and 90 years of age can be detected with 23\% error rate by using the curvature angle and coronal mean curvature.
+%%
+Subjects with body size between 177 cm and 202 cm can classified with an error rate of 22,1\%. solely using mean curvature.
+%%
+\subsubsection{Gender Groups}
+\subsubsection{Age Groups}
+\subsubsection{Size Groups}
+%%
+Many previously described results can be associated to be confounded with subject size.
+%%
+Differences in the gender analysis are mostly due to the average height difference between males and females.
+%%
+For example, large subjects are already characterized by their rather straight spine.
+%
+The question is, whether the inter-group spine-variability parameter is enough for predicting other parameter or not.
+%%
+Dividing subjects into height groups potentially highlights classifications not confounded by body-height.
+%%
 
@@
 
-\com{We can discriminate back pain using non-image data, but not with image-derived parameter. Welche parameter können wir gut unterscheiden}
+\com{We can discriminate back pain using non-image data, but not with image-derived parameter. Welche parameter koennen wir gut unterscheiden}
+
+\com{Allgmeine Analyse der Daten mit einbringen.
+Hierfuer statistisch Auswertung machen, welche Parameter am meisten verwendet wurden bei der Kategorisierung.
+Nur Parameter verwenden, die genauer sind als 0.5\%, der Entscheidungsbaum kleiner als 15 ist und das Feature mindestens 100 Probanden enthält }
 

### 2014-08-15 draft.tex
worked on results section, nearly done

@@
 %%
-\com{Quelle}
+Too many branches also impose to the data.
+\com{Quelle Decision Tree}
 
@@
 %%
-The error rate has to be judged depending on the number of classes of the target variable.
+The two measures form the axis for a \emph{scatter plot} of the classification results.
+
+\paragraph{The Error Term. }
+Normally the error rate for a classification is calculated with $error = \frac{n_{classifiedCorrectly}} {n}$ where $n$ is the number of subjects.
 %%
-An error rate of 50\% for a target variable with two manifestations makes the classifier equal to throwing a coin.
+This metric says nothing about the discriminative power of each manifestation.
 %%
-The same error rate is acceptable a target variable with more than two manifestations.
+If for example a variable indicating a disease is negative for 90\% of the subject and the classifier simply assigns all subjects to \emph{not ill} the error metric would yield an error of 10\%, even though it is very bad.
 %%
-Picking a value for a variable of four manifestations with an error rate of 50\% is better than the random selection (error rate of 75\%).
+Therefore the discriminative power of each dimension in the equation
+%error = 1 - \frac{\[ \sum_{m=0}^M \] \frac{M_{classifiedCorrectly}}{M_all}}{number of dimensions}
+\begin{equation}
+error = 1 - \frac{\sum_{m=0}^M \frac{m_{classifiedCorrectly}}{m_{all}}}{number_{dimensions}}
+\end{equation}
+where $M$ represents each manifestation of a variable.
 %%
-The following scales the error rate so that 0 denotes perfect classification and 1 is equal to random selection.
-%%
-\begin{equation}
-error_{normalized} = \frac{error}{100 - \frac{100}{class number}}
-\end{equation}
+The error is scaled to denote perfect classification with 0 and 1 is equal to random selection.
+% The error rate has to be judged depending on the number of classes of the target variable.
+% %%
+% An error rate of 50\% for a target variable with two manifestations makes the classifier equal to throwing a coin.
+% %%
+% The same error rate is acceptable a target variable with more than two manifestations.
+% %%
+% Picking a value for a variable of four manifestations with an error rate of 50\% is better than the random selection (error rate of 75\%).
+% %%
+% The following scales the error rate so that 0 denotes perfect classification and 1 is equal to random selection.
+% %%
+% \begin{equation}
+% error_{normalized} = \frac{error}{100 - \frac{100}{class number}}
+% \end{equation}
 %%
 This allows for comparability of error rates between variables.
+
+\paragraph{Attribute Mapping}
+The scatter plot axis are defined by tree size and the previously described error metric.
 %%
-Abstracting the classification results into two continuos variables allows us to visualize the result of 134 classifications in one scatterplot.
+This allows us to visualize a multitude of classification results in one plot.
 %%
-By mapping the group affiliation (in our case body-size) to color or shape, we can even plot all 402 results in a single visualization (Fig.~\ref{fig:decision-tree} a).
+Lets say we want to classify and compare the same variables for different subject groups, for example male and female subjects.
 %%
+We can support this by color coding the data points according to group affiliation.
 
@@
 This allows to instantly assess the reliability of the result.
+%%
+\com{ToDo: Manche Variablen rausgeschmissen, weil sie zu wenig Probanden enthielten, oder Auspraegungen nur duenn besetzt waren!}
+
+\com{ToDo: SQRT Skalierung der Treesize-Dimension, da hauptsaechlich geringe Groessen von interesse sind, ausreisser aber diesen Bereich schwer durchschaubar machen}
+% %%
+%
+% %%
+% Abstracting the classification results into two continuos variables allows us to visualize the result of 134 classifications in one scatter plot.
+% %%
+% By mapping the group affiliation (in our case body-size) to color or shape, we can even plot all 402 results in a single visualization (Fig.~\ref{fig:decision-tree} a).
 %%
@@
 %%
-This allows for sequentially analyzing the interactions.
+This allows for sequentially analyzing the classifications.
 %%------------
@@
 \end{enumerate}
+We plotted each group twice.
+%%
+The first plot shows all original variables.
+%%
+The second plot shows all categorical variables transformed into dichotomous dummy variables.
 
 \subsubsection{All Variables}
+%%
+The vast majority of parameters can not be described well using the classifier.
+%%
+This is reflected in the large amount of variables classified with a error rate above $0,6$.
+%%
+%The first impression of the resulting scatter plots quickly shows a very high amount of variables above an error rate of $0,6$.
+%%
+
 % None of the pain indicators can be reliably described using the image-based parameter.
@@
 %%
-The perfect categorization of the \emph{dialysis} parameter was conspicuous.
+The only variable reliably classified in this group is gender.% with an error rate of 31,5\% using a decision tree of seven rules.
+%
+It can be discriminated with an error rate of $0,31$ using 7 rules and incorporates only curvature and curvature angle variables.
+%
+%It mostly incorporates \emph{mean curvature}, \emph{sagittal curvature angle} und sagittal \emph{mean curvature}.
+%The perfect categorization of the \emph{dialysis} parameter was conspicuous.
 %%
-This is due to the subject distribution among the manifestation.
+%This is due to the subject distribution among the manifestation.
 %%
-Only two subjects receive a dialysis.
+%Only two subjects receive a dialysis.
 %%
-The good discrimination is not meaningful.
-%%
-Gender can be discriminated with an error rate of 31,5\% using 7 rules and incorporates only curvature and curvature angle variables.
+%The good discrimination is not meaningful.
 %%
@@
 %%
-1,058 subjects are classified with an error rate of 23,8\% solely based on coronal mean curvature.
+1,058 subjects are classified with an error rate of $0,47$ solely based on coronal mean curvature.
 %%
-Other features well classified include marital status, foot pain and number of born children.
+Almost all subjects who are medicated ($796/1,058$) were correctly classified, the vast majority of non-medicated subjects ($262/1,058$) are false-positive classified.
 %%
+Therefore the classifier is not as useful as the error-rate would indicate.
+%%This is however due to the high amount of people, who are medicated (68\%) and who are almost all correctly class
+%%
+%Other features well classified include marital status, foot pain and number of born children.
+%%
+The four body size groups could be characterized with an error rate $0,48$, but the decision tree comprises of 71 rules and imposes overfitting.
 
-The dummy variable analysis yields fewer results.
+The analysis of the dummy variable yields similar results like the blood pressure medication.
 %%
-It shows that subjects between 64 and 90 years of age can be detected with 23\% error rate by using the curvature angle and coronal mean curvature.
+Variables, such subjects sized 139-164 cm, between 64 and 90 years of age or nutrition related parameters are dominantly populated by one manifestation.
 %%
-Subjects with body size between 177 cm and 202 cm can classified with an error rate of 22,1\%. solely using mean curvature.
+The classifier neglects the other group and yields a error rate below $0,5$.
+%The dummy variable analysis yields fewer results.
+%%
+%It shows that subjects between 64 and 90 years of age can be detected with 23\% error rate by using the curvature angle and coronal mean curvature.
+%%
+%Subjects with body size between 177 cm and 202 cm can classified with an error rate of 22,1\%. solely using mean curvature.
 %%
 \subsubsection{Gender Groups}
+%%
+Classification using groups divided by gender do not produce satisfying results.
+%%
+Only hypothyroidism could be described for male subjects with an error rate of $0,24$ for 110 subjects using the \emph{mean curvature} and \emph{curvature angle}.
+%%
+Since there are only 30 male subjects diagnosed with hypothyroidism, reducing the statistical power of the result.
+%%
+The dummy variable analysis showed that female subjects of 139-164 cm body height could be discriminated using the mean curvature and curvature angle, with an error of $0,38$.
+
+%\com{Potentially skip these due to the confounding of body size}
 \subsubsection{Age Groups}
+%%
+Gender could be described for each age group using \emph{mean curvature} and \emph{curvature angle}.
+%%
+The accuracy varies between $0,35$ ($43,8-54$ years of age, 3 decision rules) to $0,27$ ($21-43,8$ years of age, 6 decision rules).
+%%
+Many variables, such as body size can be described with an error rate of $0,35$ to $0,45$ but only using large decision trees with over 20 rules.
+%Body-size can be determined with an error rate of $0,36$ to $0,39$ for subject
+% age groups, but also imposes over fitting to the data yielding a decision tree of over 30 rules.
+%%
+Notable is also the increasing accuracy with increasing decision tree size.
+%%
+For subjects between $43,8-54$ years of age, the classifier discriminated subjects with thyroid nodules with an error rate of $0,32$.
+
+The dummy variable analysis shows many results using a decision tree with one rule based on \emph{mean curvature} with accuracy between $0,4$ and $0,5$.
+%%
+It shows that mean curvature can be used to predict dependencies to variables, such as \emph{high blood pressure}, \emph{hypothyroidism} or \emph{nutrition}.
+
 \subsubsection{Size Groups}
@@
 %%
+
+\com{Schmerzparameter mit drin, aber nur mit komplexen bäumen}
+
+\com{- Kinderanzahl: 630 cases, 3,7\% Fehlerrrate, liegt aber daran, dass die meisten Probanden kein Kind haben.
+Die Einteilung ist aktuell auch unsinnig, das erste Quantil hat (-.0004,0.8] Kinder.
+Die Probanden werden nicht sauber erkannt.}
+
+\com{Manchmal kann es passieren, dass die Quantile fuer eine Variable gleich sind - z.B.
+bei Dauer Jahre Schwerarbeit, weil viele Probanden hier eben eine 0 haben.
+Deswegen kann das Ding nicht in vier Qunatile unterteilt werden, die ersten 3 sind 0.
+Deswegen wurden nur die einzigartigen Unterscheidungen genommen (statt \emph{quantile(myVar)} wurde \emph{unique(quantile(myVar))} verwendet.)}
 

### 2014-08-16 draft.tex
Filled a few blanks, improved error rate description

@@
 %%
-This metric says nothing about the discriminative power of each manifestation.
+The metric usually works well for normally distributed variables.
+%%
+It distorts the result for other distribution types.
 %%
@@
 %%
-Therefore the discriminative power of each dimension in the equation
+Our error term therefore incorporates the discriminative power of each manifestation and is denoted as follows:
+%%
+%Therefore the discriminative power of each dimension in the equation
 %error = 1 - \frac{\[ \sum_{m=0}^M \] \frac{M_{classifiedCorrectly}}{M_all}}{number of dimensions}
@@
 \end{equation}
-where $M$ represents each manifestation of a variable.
+$M$ represents the set of manifestation of each variable.
 %%
@@
 %%
-This allows for comparability of error rates between variables.
+It allows for comparability of error rates between variables with different manifestation count.
 
@@
 %%
+
+We apply a square root scale for the tree size axis to highlight data points with few decision rules.
+%%
+Outlier results with very large decision trees distort otherwise the resulting plot.
+%%
+
+\com{Needs to be moved}
+\add{We removed variables with less than 100 subjects, since they have to few subjects to be statistically significant.}
 \com{ToDo: Manche Variablen rausgeschmissen, weil sie zu wenig Probanden enthielten, oder Auspraegungen nur duenn besetzt waren!}
 
-\com{ToDo: SQRT Skalierung der Treesize-Dimension, da hauptsaechlich geringe Groessen von interesse sind, ausreisser aber diesen Bereich schwer durchschaubar machen}
+%\com{ToDo: SQRT Skalierung der Treesize-Dimension, da hauptsaechlich geringe Groessen von interesse sind, ausreisser aber diesen Bereich schwer durchschaubar machen}
 % %%

### 2014-08-18 draft.tex
Size Result section finished

@@
 %%
-The metric usually works well for normally distributed variables.
+The metric usually works well for variables with uniform distribution.
 %%
@@
 %%
+\paragraph{Large Decision Trees. }
+Back-pain associated variables can be explained for various size-groups, but we could not extract universal rules.
+%%
+Radiating back pain could be described with error rate of $0,39$ using 23 rules for subjects between $171-177 cm$ body size using torsion and mean-curvature.
+%%
+For subjects sized $177-202 cm$ the accuracy drops to $0,47$ using 20  decision rules.
+%%
+There are several decision trees for laboratory values, such as alanine aminotransferase value in the blood can be described with a high precision of $0,4$ ($139-164 cm$) to $0,36$ ($164 - 171 cm$).
+%%
+Similar values can be observed for cholesterol or age.
+%%
+Due to the large decision trees, these results are not usable and impose overfitting to the data.
+%%
+\paragraph{Small Decision Trees. }
+%%
+The dummy variables show several variables described using only one decision rule with accuracy between $0,42$ to $0,47$.
+%%
+Most of these variables have a dominant manifestation and the classifier shows a low detection precision for the second manifestation.
+%%
+These variables include nutrition parameter, thyroid disorder and social problems induced by back pain.
 
-\com{Schmerzparameter mit drin, aber nur mit komplexen bäumen}
+%The resulting trees consist either only one rule or a rich set of them (>15).
+%Either they divide the variables using only one decision rule, or a large set of rules (> 15).
 
-\com{- Kinderanzahl: 630 cases, 3,7\% Fehlerrrate, liegt aber daran, dass die meisten Probanden kein Kind haben.
-Die Einteilung ist aktuell auch unsinnig, das erste Quantil hat (-.0004,0.8] Kinder.
-Die Probanden werden nicht sauber erkannt.}
+\subsubsection{Concluding Remarks}
+Except for gender in the age-plot, no variable could be explained with reasonable accuracy for age or body size.
 
@@
 Hierfuer statistisch Auswertung machen, welche Parameter am meisten verwendet wurden bei der Kategorisierung.
-Nur Parameter verwenden, die genauer sind als 0.5\%, der Entscheidungsbaum kleiner als 15 ist und das Feature mindestens 100 Probanden enthält }
+Nur Parameter verwenden, die genauer sind als 0.5\%, der Entscheidungsbaum kleiner als 15 ist und das Feature mindestens 100 Probanden enthält}
+
+\com{Weiterhin interessant die Beschreibung rueckenschmerzassoziitierter Parameter, bei einigen zeigt sich, dass Torsion sehr dominant vertreten ist.}
+
+\com{Algorithmus schreiben, der eng zusammenliegende Parameter fuer alle Probanden sucht.}
 

### 2014-08-18 draft.tex
wrote first part of conclusion

@@
 
-\subsubsection{Concluding Remarks}
-Except for gender in the age-plot, no variable could be explained with reasonable accuracy for age or body size.
-
 \com{Manchmal kann es passieren, dass die Quantile fuer eine Variable gleich sind - z.B.
@@
 \label{sec:Conclusion}
+The presented results indicate that the parameters torsion, curvature and curvature angle are not enough to characterize pathological changes in the lumbar spine.
+%%
+They are well suited to characterize subject gender due to their height difference using mean curvature and mean curvature angle.
+%%
+The reason for this lies in the higher curvature of smaller subjects and the significant difference of average body size between males and females.
+%%
+Gender was characterized well in the analysis of all subjects as well as subjects divided by age.
+
+Subjects grouped using body height have a similar spine shape and a lower variance.
+%%
+The remaining information within these groups is not enough to characterize back pain related variables.
+%%
+Certain subgroups, however, can be classified for various size groups, such as nutrition or psychological parameters.
+%%
+
+Another observation during the analysis are very large decision trees achieving low error scores.
+%%
+Their complexity however impose overfitting to the data and they are not suitable for extracting universal rules.
+
+\subsection{Proposed Features}
+
+\subsection{Applicability of the Method}
+
+Except for gender in the age-plot, no variable could be explained with reasonable accuracy for age or body size.
+\com{Vorgestellte Parameter nicht ausreichend.}
+\com{Weitere Parameter wichtig - z.B. Dicke der Wirbelkoerper, Durchmesser des Spinalkanals, Form der gesamten Wirbelsaeule, Cobb-Winkel}
 

### 2014-08-18 draft.tex
added 'proposed features' section

@@
 %%
-Their complexity however impose overfitting to the data and they are not suitable for extracting universal rules.
+Their complexity impose overfitting to the data and they are not suitable for extracting universal rules.
+%%
+\subsection{Proposed Features}
+\begin{figure}[!h]
+  %\vspace{-0.2cm}
+  \centering
+  \includegraphics[width=0.5\textwidth]{figures/placeholder}
+  \caption{Proposed features.
+(a) dented vertebraes are a sign of heavy stress and bone absorption.
+(b) Spine canal thickness is associated with herniated disks.
+(c) the spine canal shape can be used to characterize scoliosis and lordosis as well as the \emph{Cobb-Angles} \com{TODO CITE Cobb Angle}.
+}
+  \label{fig:proposedFeatures}
+\end{figure}
+%%
+\noindent Pathological deformation of the lumbar spine is usually the last stage and is associated with very strong lower back pain \com{Zitat Frau Hegenscheid fragen}.
+%%
+We have to capture earlier signs of pathological change in order to derive better predictive models.
+%%
 
-\subsection{Proposed Features}
+A very early sign of pathological deformation is the \emph{bone resorption in the center of a vertebra}.
+%%
+They change from being block-shaped to be dented in the center (Fig~\ref{fig:proposedFeatures}~(a)).
+%%
+This information can be obtained by segmenting the whole vertebra or the top and bottom point if each vertebra center.
+%%
+Another valuable variable would be the \emph{spine canal thickness} (Fig~\ref{fig:proposedFeatures}~(b).
+%%
+Low spine canal thickness can be an indicator for an impending herniated disk.
+%%
+Both surface texture of the vertebra and thickness of the spine canal are used to diagnose herniated disks.
+%%
+The \emph{overall spine canal shape} is also of interested, since scoliosis and lordosis can be characterized more precise by deriving the cobb angles from this shape (Fig~\ref{fig:proposedFeatures}~(c).
+%%
 

### 2014-08-18 draft.tex
Wrote 'Applicability of Method' section

@@
 Exchanging software becomes as easy as exchanging a hyperlink.
+%%
+\com{TODO: Auf Eignung von RMarkdown als Report eingehen - ergebnisse koennen schnell publiziert und mit neuen Daten durchgetestet werden.}
 %%---------
@@
 (b) Spine canal thickness is associated with herniated disks.
-(c) the spine canal shape can be used to characterize scoliosis and lordosis as well as the \emph{Cobb-Angles} \com{TODO CITE Cobb Angle}.
+(c) the spine canal shape can be used to characterize scoliosis and lordosis as well as the \emph{Cobb-Angles} \cite{Cobb}.
 }
@@
 %%
-The \emph{overall spine canal shape} is also of interested, since scoliosis and lordosis can be characterized more precise by deriving the cobb angles from this shape (Fig~\ref{fig:proposedFeatures}~(c).
+The \emph{overall spine canal shape} is also of interested, since scoliosis and lordosis can be characterized more precise by deriving the Cobb angles from this shape (Fig~\ref{fig:proposedFeatures}~(c).
+%% ------------------------------------
+\subsection{Applicability of the Method}
+%%
+Classification methods based on decision trees showed to be useful to assess the discriminative power of a variable set.
+%%
+Their ability to assess variable combination makes them more powerful than correlation coefficients calculated for each variable.
+%%
+This advantage comes with a much more complex output, making their results harder to assess and to abstract.
+%%
+Our method to plot derived metrics and custom tailored error measures proved to be effective.
+%%
+Huge result spaces could be navigated fast using our brushable visualization.
+%%
+Therefore the method is applicable not only for deriving information based on image data, but on all potential target variables.
 %%
 
-\subsection{Applicability of the Method}
-
-Except for gender in the age-plot, no variable could be explained with reasonable accuracy for age or body size.
-\com{Vorgestellte Parameter nicht ausreichend.}
-\com{Weitere Parameter wichtig - z.B. Dicke der Wirbelkoerper, Durchmesser des Spinalkanals, Form der gesamten Wirbelsaeule, Cobb-Winkel}
+As described in a prior work \cite{Klemm2014}, the suggestion of potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
+%%
+We can achieve this also using the presented method by trying to describe several (dichotomous) target values with a decision tree constructed from all available data.
+%%
+This allows both to assess the discriminative power of the data set regarding the target variables, as well as the most important parameters for the classification.
+%Except for gender in the age-plot, no variable could be explained with reasonable accuracy for age or body size.
 

### 2014-08-18 draft.tex
Summary and Future Work

@@
 \label{sec:Conclusion}
-The presented results indicate that the parameters torsion, curvature and curvature angle are not enough to characterize pathological changes in the lumbar spine.
+\noindent The presented results indicate that the parameters torsion, curvature and curvature angle are not enough to characterize pathological changes in the lumbar spine.
 %%
@@
 \subsection{Proposed Features}
+\label{sec:ProposedFeatures}
 \begin{figure}[!h]
@@
 
+\subsection{Summary and Future Work}
+%%
+\noindent We showed, that torsion, curvature and curvature angle of the lumbar spine at the presented precision are not enough to characterize lumbar back pain in the SHIP-Data set.
+%%
+Our method made it possible to assess their discriminative power, which is largely limited for separating male and female subjects, nutrition variables, as well as different disease indicators.
+%%
+The C4.5 algorithm showed to be an effective tool for evaluating a set of derived metrics regarding their suitability to classify non-image parameters.
+%%
+Over-fitting to the data indicated by complex decision trees have to be taken into account into the analysis as well.
+%%
+
+In our future work, we will focus on more precise models for extracting measures described in Section~\ref{sec:ProposedFeatures}.
+%%
+As another focus, we want to include the method into existing visual analytics methods designed for analyzing shape information for epidemiological data \cite{Klemm2014}.
+%%
+
+Combining the power of statistical analysis, visual analytics and advanced data mining techniques is essential for analyzing increasingly complex heterogenous population data.
+%%
+These methods do not aim to replace the traditional epidemiological workflow, but rather intervenes at the weak points of standard statistical methods.
+
 \section*{\uppercase{Acknowledgements}}

### 2014-08-25 draft.tex
included results figure
Added revised introduction from Sylvia

@@
 Epidemiology is the study of causation of diseases.
+For this purpose, large population studies, such as the \emph{Study of Health in Pomerania} (SHIP) [??] gather as much information as possible about participants to be assessed towards different diseases.
 %%
-Large population studies, such as the Study of Health in Pomerania (SHIP) \ref{SHIP} gather as much information as possible about participants to be assessed towards different diseases.
+These information are used to determine risk factors for diseases, help people to improve their lifestyle and their health or to support diagnosis of widespread diseases.
 %%
-These information are used to determine risk factors for diseases, helping people to make their lifestyle healthier or helping in diagnosing a disease.
+Epidemiological research is strongly hypothesis-driven.
 %%
-Epidemiological research is strongly hypothesis driven.
-%
-Observations made by clinicians are translated into hypothesis, which are then statistically evaluated using data variables from epidemiological studies.
-
-Modern cohort studies often comprise medical image data.
+More precisely, observations made by clinicians are translated into hypotheses, which are then statistically evaluated using data variables from epidemiological studies.
 %%
-\com{Short summary}
+As a result, the hypothesis will be accepted or rejected.
 %%
 
+%For epidemiological research, cohort studies are acquired.
+For epidemiological research, cohort studies are carried out.
+%%
+These studies often comprise medical image data, e.g., tomographic image data sets, as well as general patient-specific attributes like gender, age and weight.
+%%
 Back pain is one of the most frequent diseases in the western civilization. 
 %%
+Epidemiological analysis of back pain aim to characterize the disease by determining risk factors.
+%Many epidemiological-motivated approaches aim at improved characterization of this disease with respect to patient-specific attributes.
+%Since back pain is one of the most frequent diseases in the western civilization, many epidemiological-motivated approaches aim at improved characterization of this disease with respect to patient-specific attributes.
+%%
+Although the shape and constitution of the spine, especially the lumbar spine, plays an important role for the presence of back pain, no automatic classification approach to characterize back pain based on lumbar back spine attributes has been established yet.
+%%
+Furthermore, previous analyses mainly focused on a single parameter or a small subset of parameters [Quelle].
+%%
 
-Our goal is to combine data mining algorithm with data visualization to provide insight into the quality of image derived data to analyze if it acts as a risk factor for a disease.
-%\com{Combine Data Mining Algo}
+In this paper, we present a thorough analysis of the image-derived data from lumbar back spine datasets and analyze their discriminating power for different characteristics of the disease.
+%%
+Therefore, we combine data mining algorithms with data visualization techniques.
+%%
+%Then, interactive visual analysis provides insight into the quality of the image-derived data as well as their potential for possible back pain risk factors.
+Then, interactive visual analysis provides insight into the quality of the image-derived data as well as their potential for being back pain risk factors.
+%%
+\com{Da in der Auflistung oefter vom IVA-Framework die Rede ist: Wollen wir hierzu noch was schreiben, z.B.
+dass das im VIS-Paper definiert wurde und was wir darunter verstehen; sowie dass diese Arbeit als Erweiterung dieses Frameworks gedacht ist?}
 
-More than 5 variables are rarely analyzed simultaneously.
-%%
-%The variables analyzed simultaneously is limited to 
 Our contributions are:
 \begin{itemize}
-	\item Analyzing back pain using image-derived variables of 2,240 subjects.
-	\item Assessing the suitability of lumbar spine shape for diagnosing back pain
-	\item Analyzing correlations between between image-based and socio-demographic as well as medical parameters.
-	\item Identification of most important variables using data mining methods \com{noch schoen schreiben}
-	\item semiquantitative Auswertmöglichkeiten werden dem User oft zur verfügung gestellt
-	\item Techniken sind als Teil des IVA-Frameworkes zu verstehen
+\item An IVA-framework for back pain analysis based on image-derived variables of 2,240 subjects \com{Ist diskutabel, so richtig interaktiv ist es nicht. IVA wurde auch noch nicht vorher definiert.}
+%%
+\item The identification and evaluation of lumbar spine shape properties for back pain diagnosis
+%%
+\item Detection of correlations between image-based and socio-demographic as well as medical parameters for an improved hypothesis generation
+%%
+\item Identification of the most important variables via data mining methods (including a novel semi-quantitative evaluation concept) embedded in our IVA-framework \com{Wieder: IVA wurde noch nicht definiert}
+%%
 \end{itemize}
+
+% -- Old Introduction
+% Epidemiology is the study of causation of diseases.
+% %%
+% Large population studies, such as the Study of Health in Pomerania (SHIP) \ref{SHIP} gather as much information as possible about participants to be assessed towards different diseases.
+% %%
+% These information are used to determine risk factors for diseases, helping people to make their lifestyle healthier or helping in diagnosing a disease.
+% %%
+% Epidemiological research is strongly hypothesis driven.
+% %
+% Observations made by clinicians are translated into hypothesis, which are then statistically evaluated using data variables from epidemiological studies.
+%
+% Modern cohort studies often comprise medical image data.
+% %%
+% \com{Short summary}
+% %%
+%
+% Back pain is one of the most frequent diseases in the western civilization.
+% %%
+%
+% Our goal is to combine data mining algorithm with data visualization to provide insight into the quality of image derived data to analyze if it acts as a risk factor for a disease.
+% %\com{Combine Data Mining Algo}
+%
+% More than 5 variables are rarely analyzed simultaneously.
+% %%
+% %The variables analyzed simultaneously is limited to
+% Our contributions are:
+% \begin{itemize}
+% 	\item Analyzing back pain using image-derived variables of 2,240 subjects.
+% 	\item Assessing the suitability of lumbar spine shape for diagnosing back pain
+% 	\item Analyzing correlations between between image-based and socio-demographic as well as medical parameters.
+% 	\item Identification of most important variables using data mining methods \com{noch schoen schreiben}
+% 	\item semiquantitative Auswertmöglichkeiten werden dem User oft zur verfügung gestellt
+% 	\item Techniken sind als Teil des IVA-Frameworkes zu verstehen
+% \end{itemize}
 
@@
 \subsection{Results}
+\begin{figure*}[p!]
+  %\vspace{-0.2cm}
+  \centering
+  \includegraphics[width=1.0\textwidth]{figures/results}
+  \caption{myResults}
+  \label{fig:results}
+\end{figure*}
 We ran the analysis using different subject groups:

### 2014-08-25 draft.tex
Results figure now fits its canvas
filling blank spots

@@
   \includegraphics[width=1.0\textwidth]{figures/image-parameter-range}
-  \caption{image-parameter-range}
+  \caption{%%
+A generalized pairs plot of all image-derived variables \cite{GPLOMS} colored by presence or absence of back pain.
+%%
+Pairwise combinations of image-derived variables are visualized using scatter plots left of the matrix diagonal and their correlation with back pain is denoted right of the matrix diagonal.
+%%
+The box plots (left) and histograms (bottom) display the distribution of each image-variable encoded with back pain.
+%%
+No correlation with back pain are present can be seen in this plot.
+}
   \label{fig:image-parameter-range}
@@
 %%
-\noindent As first experiment we correlated the shape parameter with the binary back pain indicator using a pairs plot \cite{ggally}.
+\noindent As first experiment we correlated the shape parameter with the binary back pain indicator using a generalized pairs plot \cite{GPLOMS}.
 %%
@@
 %%
-The pearson correlation coefficients between the numeric variables are depicted right of the main diagonal of Figure~\ref{fig:image-parameter-range}.
+The pearson correlation coefficients between the numeric variables are depicted right of the matrix diagonal in Figure~\ref{fig:image-parameter-range}.
 %%
@@
 \subsubsection{Correlation of Image Parameters With Back Pain}
-Figure~\ref{fig:image-parameter-range} shows the distribution for all subjects.
+Figure~\ref{fig:image-parameter-range} shows the distribution of all image-parameters \add{as generalized pairs plot}.
 %%
-We created a pairs plot for subject groups derived by dividing them into quantile groups of body-size.
+\rem{We created a generalized pairs plot for subject groups derived by dividing them into quantile groups of body-size.}
 %%
@@
 \end{itemize}
+\com{TODO: Abbildung einer Korrelationsmatrix? Waere sehr gross!}
 %%
@@
 %%
-Some variables are too sparse for calculating correlations, for example \com{TODO: some variables from calculation here; TODO: Image?, eventuell interactiven Tile-Plot}.
+Some variables are too sparse for calculating correlations, for example \emph{treatment of diabetes}, or \emph{medication against high blood pressure}.
 %%
@@
 %%
-The resulting contingency matrix shows no strong correlation with any of the parameter.
+The resulting contingency matrices are available as supplementary material and show no strong correlation with any of the parameter.
 %%
@@
 \subsubsection{Visualization of Classification Results}
+\label{subsec:VisualizationOfClassificationResults}
 %%
@@
 %%
-We map the number of subjects associated with a variable to point diameter in the scatter plot.
+We removed variables with less than 100 subjects, since they have to few subjects to be statistically significant.
+%%
+Furthermore, we map the number of subjects associated with a variable to point diameter in the scatter plot.
 %%
@@
 %%
-
-\com{Needs to be moved}
-\add{We removed variables with less than 100 subjects, since they have to few subjects to be statistically significant.}
-\com{ToDo: Manche Variablen rausgeschmissen, weil sie zu wenig Probanden enthielten, oder Auspraegungen nur duenn besetzt waren!}
 
@@
 %%
-We implemented our application as web based prototype using \texttt{R Shiny}\footnote{R Shiny footnote}.
+In order to make the train of thought comprehensible, we used \texttt{RMarkdown}, which allows to make reports by combining \texttt{R} with the \texttt{Markdown} syntax.
 %%
-The package combines the power of \texttt{R} with advantages of web-based applications.
+We used \texttt{R Shiny}\footnote{R Shiny footnote} to make the report available as dynamic web application.
 %%
-It allows us to quickly exchange results with our collaborating epidemiological partners.
+It allows to combine both the ability to create comprehensible reports as well as UI-controls, to adjust visible data and parameters in the plot to allow the viewer to analyze the data.
 %%
-They can try out the technique without installing anything.
+Furthermore, calculations based on a prior data selection can be redone within the report.
+%We implemented our application as web based prototype using \texttt{R Shiny}\footnote{R Shiny footnote} in combination with \texttt{RMarkdown}.
 %%
-Exchanging software becomes as easy as exchanging a hyperlink.
+%The package combines the power of \texttt{R} with advantages of web-based applications.
 %%
-\com{TODO: Auf Eignung von RMarkdown als Report eingehen - ergebnisse koennen schnell publiziert und mit neuen Daten durchgetestet werden.}
+The web-based approach allows us to quickly exchange results with our collaborating epidemiological partners.
+%%
+They can try out the technique without installing any software.
+%%
+Exchanging the prototype becomes as easy as exchanging a hyperlink.
 %%---------
@@
   \centering
-  \includegraphics[width=1.0\textwidth]{figures/results}
-  \caption{myResults}
+  \includegraphics[width=0.8\textwidth]{figures/results}
+  \caption{
+%%
+Scatterplot visualization of all C5.0 classification results.
+%%
+The x-axis shows the number of decisions of the underlying model, the y-axis the classification error rate (see Section~\ref{subsec:VisualizationOfClassificationResults}).
+%%
+The left scatterplot shows the results for all variables, either metrics expressed via their quantiles, or categorical.
+%%
+The right scatterplot displays the dummy variables derived from the original variables.
+%%
+Group affiliation of a data point is color coded:
+%%
+no group (a), subdivision into male and female subjects (b), quartiles of age (c) and quartiles of body size (d).
+%%
+%(a) displays all subjects, (b) is divided into male and female subjects, (c) is categorized into quantiles of age and (d) shows the classification of subjects divided by quantiles of body size.
+%%
+The number of subjects represented in a variable is denoted using the dot diameter.
+%%
+We only display variables with an error rate below $0,5$, results above this threshold are inaccurate.
+%%
+A interactive versions of the plot (see supplemental material) makes all data points clickable, showing the corresponding decision tree in a tool tip.
+%%
+}
   \label{fig:results}
@@
 %%
+The starting age of smoking could be characterized well with an accuracy between $0,25$ to $0,3$ for all age groups except of $21,1-43,8$ years of age.
+%%
+The result however is probably over fitted to the data due to tree sizes between $16$ and $23$.
+%%
+\com{Feedback der Greifswalder notwendig, warum das so sein koennte.}
+%%
+
 Many variables, such as body size can be described with an error rate of $0,35$ to $0,45$ but only using large decision trees with over 20 rules.
@@
 
-\com{Manchmal kann es passieren, dass die Quantile fuer eine Variable gleich sind - z.B.
+%TODO: Muss das wirklich noch rein?
+% \com{Manchmal kann es passieren, dass die Quantile fuer eine Variable gleich sind - z.B.
 bei Dauer Jahre Schwerarbeit, weil viele Probanden hier eben eine 0 haben.
@@
 
-\subsection{Themen hier unterbringen}
-
-\com{We can discriminate back pain using non-image data, but not with image-derived parameter. Welche parameter koennen wir gut unterscheiden}
-
-\com{Allgmeine Analyse der Daten mit einbringen.
-Hierfuer statistisch Auswertung machen, welche Parameter am meisten verwendet wurden bei der Kategorisierung.
-Nur Parameter verwenden, die genauer sind als 0.5\%, der Entscheidungsbaum kleiner als 15 ist und das Feature mindestens 100 Probanden enthält}
-
-\com{Weiterhin interessant die Beschreibung rueckenschmerzassoziitierter Parameter, bei einigen zeigt sich, dass Torsion sehr dominant vertreten ist.}
-
-\com{Algorithmus schreiben, der eng zusammenliegende Parameter fuer alle Probanden sucht.}
+\com{Auswertung der Anteile der Parameter an den Ergebnissen (Torsion, Curvature, etc. - welches ist der dominanteste Parameter?)}
 

### 2014-08-26 draft.tex
Rewrote Image Data section

@@
 \end{figure}
-\com{From VIS'14 Paper}
-The lumbar spine was detected in the image data using a hierarchical finite element method by Rak et al.
-[32].
-This semiautomatic method requires the user to initialize the tetrahedron-based finite element models (FEM) with a click on the L3 vertebra.
-Two user-defined landmarks on the top and bottom of the L3 vertebra describe an initial model height estimation.
-The model uses a weighted sum of T1and T2-weighted MR images to detect the lumbar spine shape.
-Once registered, it captures information about the shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae [21].
-Due to incorrect initialization, strongly deformed spines, contrast differences and artifacts, the model was not able to detect lumbar spines for all subjects.
-We obtained and worked with 2,540 tetrahedron models of the lumbar spine.
-For clustering, we extracted the centerline of the lumbar spine canal, which captures information about lordosis and scoliosis (the medical terms for spine curvature) [21].
+\com{Image Data}
+Four trained technicians acquired the medical image data for each subject in a standardized way using a 1.5~Tesla MRI scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany).
 %%
+The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
+%%
+
+A hierarchical finite element method was used to detect the lumbar spine in the MRI data \cite{Rak2013}.
+%%
+The tetrahedron-based finite element model (FEM) is initialized using three user-placed land marks.
+%%
+A click on the L3 vertebra center gives the model the initial position, two clicks on the top and bottom of the vertebra describes an initial model height estimation.
+%%
+The registered model captures information about the lumbar spine canal shape and the position of the L1-L5 vertebrae \cite{Klemm2013VMV} (Fig.~\ref{fig:centerline}~a).
+%%
+The detection fails for several subjects due imprecise initialization, imaging artifacts or strongly deformed spines.
+%%
+$2,540$ lumbar spine models were obtained and constitute the foundation of this work.
+%%
+%%
+% \com{From VIS'14 Paper}
+% The lumbar spine was detected in the image data using a hierarchical finite element method by Rak et al. [32].
+% %%
+% This semiautomatic method requires the user to initialize the tetrahedron-based finite element models (FEM) with a click on the L3 vertebra.
+% %%
+% Two user-defined landmarks on the top and bottom of the L3 vertebra describe an initial model height estimation.
+% %%
+% The model uses a weighted sum of T1 and T2-weighted MR images to detect the lumbar spine shape.
+% %%
+% Once registered, it captures information about the shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae [21].
+% %%
+% Due to incorrect initialization, strongly deformed spines, contrast differences and artifacts, the model was not able to detect lumbar spines for all subjects.
+% %%
+% We obtained and worked with 2,540 tetrahedron models of the lumbar spine.
+% %%
+% For clustering, we extracted the centerline of the lumbar spine canal, which captures information about lordosis and scoliosis (the medical terms for spine curvature) [21].
+% %%
 
@@
 %%
-In a previous work \cite{KlemmVMV13, KlemmVIS14}, we extracted a centerline representation of the lumbar spine canal from the detection model (Fig.~\ref{fig:centerline} (a)).
+In a previous work \cite{Klemm2013VMV,Klemm2014VIS}, we extracted a centerline representation of the lumbar spine canal from the detection model (Fig.~\ref{fig:centerline} (a)).
 %%
@@
 %%
-To avoid small outlier groups, epidemiologists recommend using quartiles to discretize metric variables \cite{KlemmVIS14}.
+To avoid small outlier groups, epidemiologists recommend using quartiles to discretize metric variables \cite{Klemm2014VIS}.
 \subsection{Preliminary Results}
@@
 
-As described in a prior work \cite{Klemm2014}, the suggestion of potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
+As described in a prior work \cite{Klemm2014VIS}, the suggestion of potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
 %%
@@
 %%
-As another focus, we want to include the method into existing visual analytics methods designed for analyzing shape information for epidemiological data \cite{Klemm2014}.
+As another focus, we want to include the method into existing visual analytics methods designed for analyzing shape information for epidemiological data \cite{Klemm2014VIS}.
 %%

### 2014-08-26 draft.tex
Caption of centerline figure done

@@
   \includegraphics[width=0.5\textwidth]{figures/centerline}
-  \caption{Centerline}
+  \caption{
+	(a) left: Tetrahedron-based finite element model (FEM) of the lumbar spine, capturing the L1-L5 vertebrae as well as the lumbar spine canal.
+	%%
+	(a) right: FEM aligned to a MRI data set.
+	%%
+	The purple dashed line represents the centerline, describing the lumbar spine canal with 93 points.
+	%%
+	(b) Using the Frenet formulas, we extract the curvature, torsion and the curvature angle from the lumbar spine canal center line.
+	}
   \label{fig:centerline}
@@
 %%
-The tetrahedron-based finite element model (FEM) is initialized using three user-placed land marks.
+The tetrahedron-based finite element model is initialized using three user-placed land marks.
 %%
@@
 %%
-In a previous work \cite{Klemm2013VMV,Klemm2014VIS}, we extracted a centerline representation of the lumbar spine canal from the detection model (Fig.~\ref{fig:centerline} (a)).
+In a previous work \cite{Klemm2013VMV,Klemm2014VIS}, we extracted a centerline representation of the lumbar spine canal from the detection model (Fig.~\ref{fig:centerline}~a).
 %%
-Using the Frenet formulas \cite{Frenet}, we extracted the following metrics from the model (Fig.~\ref{fig:centerline} (b)):
+Using the Frenet formulas \cite{Frenet}, we extracted the following metrics from the model (Fig.~\ref{fig:centerline}~b):
 %%

### 2014-08-26 draft.tex
Added Centerline figure

@@
 %%
-	\item \emph{Torsion} (deviation of a curve from the course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum \frac{torsion_i}{torsion_{all}}$.
+	\item \emph{Torsion} (deviation of a curve from its current course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum \frac{torsion_i}{torsion_{all}}$.
 	\item \emph{Curvature angle} is defined by the middle point of the spine canal centerline as \emph{vertex} and the line between middle point and top/bottom point as \emph{sides}.

### 2014-08-26 draft.tex
Figure centerline with proper label

@@
 	%%
-	The purple dashed line represents the centerline, describing the lumbar spine canal with 93 points.
-	%%
-	(b) Using the Frenet formulas, we extract the curvature, torsion and the curvature angle from the lumbar spine canal center line.
+	The purple dashed line represents the centerline, describing the lumbar spine canal with 92 points.
+	%%
+	(b) Using the Frenet formulas, we extract the weighted sum of curvature and torsion for all 92 points (white dashes).
+	%%
+	The curvature angle is calculated using the top, bottom and middle point.
+	%%
+	We also extracted the variables for each projection axis which to assess their information gain.
+	%(b) Using the Frenet formulas, we extract the curvature, torsion and the curvature angle from the lumbar spine canal center line.
 	}

### 2014-08-26 draft.tex
Revised age-gender figure and added caption

@@
 \section{The Lumbar Spine Data Set}
-\begin{figure}[!h]
-  %\vspace{-0.2cm}
-  \centering
-  \includegraphics[width=0.5\textwidth]{figures/age-gender}
-  \caption{Age-Gender}
-  \label{fig:example1}
-\end{figure}
 \label{sec:MaterialsAndMethod}
 %%
-Our approach allows to analyze many variables simultaneously.
+\noindent Our approach allows to analyze many variables simultaneously.
 %%
@@
 \label{sec:Experiments}
+\begin{figure}[!h]
+  %\vspace{-0.2cm}
+  \centering
+  \includegraphics[width=0.5\textwidth]{figures/age-gender}
+  \caption{
+	%%
+	Correlation of age and gender regarding the lumbar spine size.
+	%%
+	The bar chart shows subjects divided into four different age groups (x-axis) and their subject number (y-axis).
+	%%
+	Each bar contains the mean lumbar spine of the respective group.
+	%%
+	The shape color encodes the distance to the overall male (top chart) or female (bottom chart) mean shape.
+	%%
+	The dark gray share of each bar encodes the portion of male (top chart) or female (bottom chart) subjects.
+	%%
+	Women have a higher live expectance than men, hence their higher share in the old age group.
+	%%
+	Also, women are on average smaller than men, hence the larger shape similarity with older subjects.
+	%%
+	Due to bone erosion, older people are also smaller on average.
+	}
+  \label{fig:age-gender}
+\end{figure}
 \noindent Spine shape is confounded by several somatometric variables.
@@
 %%
-Since men are on average taller than women, gender is another confounder.
+Since men are on average taller than women and people of old age shrink due to bone erosion, gender and age are also confounder (Fig.~\ref{fig:age-gender}).
 %%
@@
 (https://stats.stackexchange.com/questions/101016/pca-and-component-scores-based-on-a-mix-of-continuous-binary-and-categorical-va)
-
-\subsection{Experimental Settings and Results}
-Decision Tree erklaeren, ebenfalls Umrechnen in die Dummy-Variablen. Was sind die Zielvariablen
-\begin{itemize}
-\item Altersgruppierung
-\item Male/Female
-\item Gewicht (BMI - dünn, normal, dick)
-\end{itemize}
 % Creating a Decision Tree using C5.0
@@
 %%
-They change from being block-shaped to be dented in the center (Fig~\ref{fig:proposedFeatures}~(a)).
+They change from being block-shaped to be dented in the center (Fig~\ref{fig:proposedFeatures}~a).
 %%
@@
 %%
-Another valuable variable would be the \emph{spine canal thickness} (Fig~\ref{fig:proposedFeatures}~(b).
+Another valuable variable would be the \emph{spine canal thickness} (Fig~\ref{fig:proposedFeatures}~b).
 %%
@@
 %%
-The \emph{overall spine canal shape} is also of interested, since scoliosis and lordosis can be characterized more precise by deriving the Cobb angles from this shape (Fig~\ref{fig:proposedFeatures}~(c).
+The \emph{overall spine canal shape} is also of interested, since scoliosis and lordosis can be characterized more precise by deriving the Cobb angles from this shape (Fig~\ref{fig:proposedFeatures}~c).
 %% ------------------------------------

### 2014-08-27 draft.tex
added draft of proposed-features figure
Revised centerline and age-gender figure

@@
 	%%
-	The curvature angle is calculated using the top, bottom and middle point.
+	The curvature angle ($\alpha$) is calculated using the top, bottom and middle point.
 	%%
@@
 	\item \emph{Torsion} (deviation of a curve from its current course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum \frac{torsion_i}{torsion_{all}}$.
-	\item \emph{Curvature angle} is defined by the middle point of the spine canal centerline as \emph{vertex} and the line between middle point and top/bottom point as \emph{sides}.
+	\item \emph{Curvature angle $\alpha$} is defined by the middle point of the spine canal centerline as \emph{vertex} and the line between middle point and top/bottom point as \emph{sides}.
 %%
@@
 %%
-\item \emph{pearson product-moment} for two continuous variables, 
-\item \emph{polyserial} correlation for one continuous and one categorical variable, and
-\item \emph{polychoric} correlations for two categorical variables.
+\item \emph{Pearson product-moment} for two continuous variables, 
+\item \emph{Polyserial} correlation for one continuous and one categorical variable, and
+\item \emph{Polychoric} correlations for two categorical variables.
 \end{itemize}
@@
 \paragraph{The Error Term. }
-Normally the error rate for a classification is calculated with $error = \frac{n_{classifiedCorrectly}} {n}$ where $n$ is the number of subjects.
+Normally the error rate for a classification is calculated with $error = \frac{classifiedCorrectly_{n}} {n}$ where $n$ is the number of subjects.
 %%
@@
 \begin{equation}
-error = 1 - \frac{\sum_{m=0}^M \frac{m_{classifiedCorrectly}}{m_{all}}}{number_{dimensions}}
+error = 1 - \frac{\sum_{m=0}^M \frac{classifiedCorrectly_{m}}{all_{m}}}{number_{dimensions}}
 \end{equation}
@@
 %%
-The accuracy varies between $0,35$ ($43,8-54$ years of age, 3 decision rules) to $0,27$ ($21-43,8$ years of age, 6 decision rules).
-%%
-The starting age of smoking could be characterized well with an accuracy between $0,25$ to $0,3$ for all age groups except of $21,1-43,8$ years of age.
+The accuracy varies between $0.35$ ($43.8-54$ years of age, 3 decision rules) to $0.27$ ($21-43.8$ years of age, 6 decision rules).
+%%
+The starting age of smoking could be characterized well with an accuracy between $0.25$ to $0.3$ for all age groups except of $21.1-43.8$ years of age.
 %%
@@
 
-Many variables, such as body size can be described with an error rate of $0,35$ to $0,45$ but only using large decision trees with over 20 rules.
+Many variables, such as body size can be described with an error rate of $0.35$ to $0.45$ but only using large decision trees with over 20 rules.
 %Body-size can be determined with an error rate of $0,36$ to $0,39$ for subject
@@
 %%
-For subjects between $43,8-54$ years of age, the classifier discriminated subjects with thyroid nodules with an error rate of $0,32$.
-
-The dummy variable analysis shows many results using a decision tree with one rule based on \emph{mean curvature} with accuracy between $0,4$ and $0,5$.
+For subjects between $43.8-54$ years of age, the classifier discriminated subjects with thyroid nodules with an error rate of $0.32$.
+
+The dummy variable analysis shows many results using a decision tree with one rule based on \emph{mean curvature} with accuracy between $0.4$ and $0.5$.
 %%
@@
 %%
-Radiating back pain could be described with error rate of $0,39$ using 23 rules for subjects between $171-177 cm$ body size using torsion and mean-curvature.
-%%
-For subjects sized $177-202 cm$ the accuracy drops to $0,47$ using 20  decision rules.
-%%
-There are several decision trees for laboratory values, such as alanine aminotransferase value in the blood can be described with a high precision of $0,4$ ($139-164 cm$) to $0,36$ ($164 - 171 cm$).
+Radiating back pain could be described with error rate of $0.39$ using 23 rules for subjects between $171-177~cm$ body size using torsion and mean-curvature.
+%%
+For subjects sized $177-202~cm$ the accuracy drops to $0.47$ using 20  decision rules.
+%%
+There are several decision trees for laboratory values, such as alanine aminotransferase value in the blood can be described with a high precision of $0.4$ ($139-164~cm$) to $0.36$ ($164 - 171~cm$).
 %%
@@
   \includegraphics[width=0.5\textwidth]{figures/placeholder}
-  \caption{Proposed features.
+  \caption{
+	Proposed features.
+	%%
 	(a) dented vertebraes are a sign of heavy stress and bone absorption.
+	%%
 	(b) Spine canal thickness is associated with herniated disks.
+	%%
 	(c) the spine canal shape can be used to characterize scoliosis and lordosis as well as the \emph{Cobb-Angles} \cite{Cobb}.
+	%%
 	}

### 2014-08-27 draft.tex
Revised Proposed Features figure and its label

@@
   \centering
-  \includegraphics[width=0.5\textwidth]{figures/placeholder}
+  \includegraphics[width=0.5\textwidth]{figures/proposed-features}
   \caption{
@@
 	%%
-	(b) Spine canal thickness is associated with herniated disks.
-	%%
-	(c) the spine canal shape can be used to characterize scoliosis and lordosis as well as the \emph{Cobb-Angles} \cite{Cobb}.
+	(b, c) the spine canal shape can be used to characterize scoliosis (curvature sagittal) and lordosis (curvature coronal) as well as the \emph{Cobb-Angles} \cite{Cobb}.
+	%%
+	(c) Spine canal thickness is associated with herniated disks.
 	%%
@@
 %%
-Another valuable variable would be the \emph{spine canal thickness} (Fig~\ref{fig:proposedFeatures}~b).
+Another valuable variable would be the \emph{spine canal thickness} (Fig~\ref{fig:proposedFeatures}~c).
 %%
@@
 %%
-The \emph{overall spine canal shape} is also of interested, since scoliosis and lordosis can be characterized more precise by deriving the Cobb angles from this shape (Fig~\ref{fig:proposedFeatures}~c).
+The \emph{overall spine canal shape} is also of interested, since scoliosis and lordosis can be characterized more precise by deriving the Cobb angles from this shape (Fig~\ref{fig:proposedFeatures}~b,~c).
 %% ------------------------------------

### 2014-08-27 draft.tex
added revised Results figure
begun related work section

@@
 \noindent \com{Our own work (VIS, VMV, BVM). Sylvias paper with reference to the methods. \textbf{More information necessary here!}}
+\com{From VIS Paper}
+
+ToDo Hier weitermachen
+Steenwijk
+Turkey (Skeweness-Paper)
+Angelelli
+Abschließend Uniqueness beschreiben
+
+\paragraph{Visual Analysis of Heterogenous Non-Image Data.}
+Zhang et al. \cite{Zhang2012} provide a web-based system for analyzing subject groups with linked views and batch-processing capabilities for categorizing new subject entries into the data set.
+%
+Their definition of a cohort differs from the understanding of the term in an epidemiological context by denoting every parameter-divided subject group as individual cohort.
+%
+Due to the short paper length, detail is missing on the data types and their algorithms of identifying similar subjects or whether they employ statistical measures.
+%
+We employ the idea of adding variables via drag and drop into a canvas area.
+
+Generalized Pairs Plots (\texttt{GPLOM´S}) are an information visualization technique comparing heterogenous variables pairwise using a plot-matrix grouped by type \cite{GPLOMS, Francois2013}.
+%
+They are useful to gain an overview over numerous variables and their distributions.
+%
+Histograms, bar charts, scatter plots and heat maps are used to visualize variable combinations with regard to their type.
+%
+The resulting matrix provides an \emph{overview visualization}, but requires a lot of screen space for many variables (127 in our application scenario).
+%
+%Brushing, linking and filtering are feasible with \texttt{GPLOMS}, but have limitations such as making only one category brushable at a time.
+%
+%We applied this technique to our data and it shows promising potential for simultaneously visualizing many different variables. %but does not fit in the scope of this paper.
+%
+We incorporate the idea of adaptive type-dependent visualizations.
+%\add{The inspiration on the chosen visualization techniques stems from this publication.}
+%
+Dai et al.
+\cite{Dai2005} explored risk factors by incorporating choropleth maps of epidemiological variables (e.g., mortality rates in a region) with parallel coordinates, bar charts and scatter plots with integrated regression lines.
+%
+Their findings yielded a \emph{Concept Map}, which linked cancer-related associations via graph edges.
+%
+While their goal to identify possible risk factors using socio-economic and health data is similar to ours, they focus on iteratively refining defined hypotheses and on geographical data.
+%
+We employ the use of small multiples for incorporating heterogenous data types for comparability.
+%
+Chui et al. \cite{Chui2011} visualized associations in time-dependent epidemiological data using time-series plots highlighting risk factor differences in age and gender.
+%
+%The connection of time series plots with histograms and heat maps allows for fast insight into complex relationships and 
+While the work shows how different visualization techniques provide insight into these data sets, it focuses on the time aspect, which is not present for our data.
+
+\paragraph{Prior Work.}
+We visualized lumbar spine variabilities based on a semi-automatic shape detection algorithm of 490 participants of the \texttt{SHIP-2} cohort \cite{Klemm2013VMV}.
+%
+Hierarchical agglomerative clustering divided the population into shape-related groups.
+%
+As proof of concept, a relation between the size of the segmented shape and the measured size of the subjects was shown.
+%
+This work focuses on incorporating these derived data as new variables, enabling to include it into the hypothesis validation and generation process.
+%
+When applying clustering techniques to the non-image data it was found that \texttt{k-Prototypes} and \texttt{DBSCAN} are appropriate, but are strongly dependent on the chosen variables and distance measures \cite{Klemm2014BVM}.
+%
+Niemann et al. \cite{Niemann2014} presented an interactive data mining tool for the assessment of risk factors of hepatic steatosis, the fatty liver disease.
+%
+Association rules created by data mining methods can be analyzed interactively with their tool and highlight potentially overlooked variables.
 
@@
   \centering
-  \includegraphics[width=0.8\textwidth]{figures/results}
+  \includegraphics[width=0.82\textwidth]{figures/results}
   \caption{

### 2014-08-28 draft.tex
Worked on Related Work section

@@
 \documentclass[a4paper,twoside]{style/article}
-
 \usepackage{epsfig}
@@
 \usepackage[]{algorithm2e}
+\usepackage[utf8]{inputenc}
 \newcommand{\rem}[1]{\textcolor{red}{\sout{#1}}}
@@
 \label{sec:RelatedWork}
-\noindent \com{Our own work (VIS, VMV, BVM). Sylvias paper with reference to the methods. \textbf{More information necessary here!}}
-\com{From VIS Paper}
-
-ToDo Hier weitermachen
-Steenwijk
-Turkey (Skeweness-Paper)
-Angelelli
-Abschließend Uniqueness beschreiben
-
+\com{Steenwijk, 
+Turkey (Skeweness-Paper), 
+Angelelli, 
+Klemm2014
+Abschließend Uniqueness beschreiben}
+
+\paragraph{Decision Rule Driven Analysis of Medical Data.}
+Closest to our work is the work of Gla{\ss}er \cite{Glasser2013} and Niemann et al. \cite{Niemann2014}.
+%%
+Related work in this field often focuses on clinical diagnosis.
+%%
+Gla{\ss}er et al. \cite{Glasser2013} use parameters derived from DCE-MRI image data to classify breast tumors.
+%%
+Their parameters are derived from clustering of bloodflow in the tumorous tissue.
+%%
+These features trains a decision-tree classifier, concluding that the extracted features alone are not sufficient for tumor type classification.
+%%
+Niemann et al. \cite{Niemann2014} assess hepatic steatosis (fatty liver disease) risk factors using decision trees.
+%%
+They presented an interactive data mining tool, which can analyze association rules and highlight interesting relations, which may trigger new hypothesis about the data.
+%%
+The presented paper either validate the significance of image derived parameter or derive new relationships by training decision trees to explain target variables.
+%%
+
+Pinhero et al. \cite{Pinheiro} applied association rule mining to create high-confidence decision rules about the mortality rate in liver cancer data.
+%%
+Decision rules are easily readable and understandable and are therefore well suited for classifying medical data.
+%%
+The results can be easily communicated to domain experts with little knowledge about the algorithm behind the results.
+%%
+Once these rules, however, get too complex, analyzing these decision rules can get complex.
+%%
+
+Sekhavat and Hoeber \cite{Sekhavat} propose a interactive visualization techniques for decision trees, where all decisions are displayed in a heat map
+%%
+The visualization translates the decision tree into a heat map, where each cell represents the connectivity between a initial value and an target value.
+%%
+The resulting interactive visualization is well suited for analyzing a decision tree, but not for multiple trees.
+%%
 \paragraph{Visual Analysis of Heterogenous Non-Image Data.}
+\com{FROM VIS PAPER}
 Zhang et al. \cite{Zhang2012} provide a web-based system for analyzing subject groups with linked views and batch-processing capabilities for categorizing new subject entries into the data set.
@@
 
-Generalized Pairs Plots (\texttt{GPLOM´S}) are an information visualization technique comparing heterogenous variables pairwise using a plot-matrix grouped by type \cite{GPLOMS, Francois2013}.
+Generalized Pairs Plots (\texttt{GPLOM'S}) are an information visualization technique comparing heterogenous variables pairwise using a plot-matrix grouped by type \cite{GPLOMS, Francois2013}.
 %
@@
 
-\paragraph{Prior Work.}
-We visualized lumbar spine variabilities based on a semi-automatic shape detection algorithm of 490 participants of the \texttt{SHIP-2} cohort \cite{Klemm2013VMV}.
-%
-Hierarchical agglomerative clustering divided the population into shape-related groups.
-%
-As proof of concept, a relation between the size of the segmented shape and the measured size of the subjects was shown.
-%
-This work focuses on incorporating these derived data as new variables, enabling to include it into the hypothesis validation and generation process.
-%
-When applying clustering techniques to the non-image data it was found that \texttt{k-Prototypes} and \texttt{DBSCAN} are appropriate, but are strongly dependent on the chosen variables and distance measures \cite{Klemm2014BVM}.
-%
-Niemann et al. \cite{Niemann2014} presented an interactive data mining tool for the assessment of risk factors of hepatic steatosis, the fatty liver disease.
-%
-Association rules created by data mining methods can be analyzed interactively with their tool and highlight potentially overlooked variables.
-
+\paragraph{SHIP Data Analysis.}
+Klemm et al. analyzed the variability of the lumbar spine data of 490 \texttt{SHIP-2} subjects \cite{Klemm2013VMV}.
+%%
+They incorporate hierarchical agglomerative clustering to derive shape groups, yielding several groups of average shape and several outlier cluster.
+%%
+The extracted lumbar spine shape was correlated with subject size.
+%%
+Out data extraction process is based on the detection algorithm and lumbar spine canal extraction presented in this paper.
+%Our work is based on data extracted in the same manner as described in this paper.
+%We visualized lumbar spine variabilities based on a semi-automatic shape detection algorithm of 490 participants of the \texttt{SHIP-2} cohort \cite{Klemm2013VMV}.
+%
+%Hierarchical agglomerative clustering divided the population into shape-related groups.
+%
+%As proof of concept, a relation between the size of the segmented shape and the measured size of the subjects was shown.
+%
+%This work focuses on incorporating these derived data as new variables, enabling to include it into the hypothesis validation and generation process.
+%
+Clustering techniques applied to non-image data are strongly dependent on the chosen variables and distance measures.
+%%
+Klemm et al. found \texttt{k-Prototypes} and \texttt{DBSCAN} appropriate for this task \cite{Klemm2014BVM}.
+%Analyzing non-image data using clustering techniques 
+%When applying clustering techniques to the non-image data it was found that \texttt{k-Prototypes} and \texttt{DBSCAN} are appropriate, but are strongly dependent on the chosen variables and distance measures \cite{Klemm2014BVM}.
+%
 \section{The Lumbar Spine Data Set}

### 2014-08-29 draft.tex
finished related work section, still has to be revised

--- 2014-08-27 draft.tex
@@
 \label{sec:RelatedWork}
-\noindent \com{Our own work (VIS, VMV, BVM). Sylvias paper with reference to the methods. \textbf{More information necessary here!}}
-
+%%
+\paragraph{Visual Analysis of Image and non-Image Data}
+%%
+Klemm et al. \cite{Klemm2014VIS} propose an extension to the existing epidemiological workflow using Interactive Visual Analysis (IVA).
+%%
+The workflow consists of an iterative sequence of group selection, variable selection and visualization.
+%%
+It aims to trigger \emph{hypothesis generation} by providing visualizations able to concurrently analyzing multiple heterogenous variables at once using correlation measures.
+%%
+We incorporate the IVA approach by concurrently displaying complex relationships between variables using decision trees.
+%%
+Steenwijk et al. \cite{Steenwijk} propose a coordinated linked view system for both image-related and non-image data.
+%%
+It incorporates multiple plot types, such as scatter plots, parallel coordinates and time plots with brushing and linking facilities to successively analyze a data set.
+%%
+Similar to our work, they quantify image data to map them into the same information space as non-image data.
+%%
+Turkay et al. \cite{Turkay} follow a similar approach by deriving descriptive data metrics from image data.
+%%
+Their proposed \emph{deviation plot} shows distribution-specific measures of a variable, such as skewness or inter-quantile-range, making dimensions comparable.
+%%
+This approach aims to trigger \emph{hypothesis generation} by outlining tendencies between these dimensions.
+
+\paragraph{Decision Rule Driven Analysis of Medical Data.}
+Closest to our work is the work of Gla{\ss}er \cite{Glasser2013} and Niemann et al. \cite{Niemann2014}.
+%%
+Related work in this field often focuses on clinical diagnosis.
+%%
+Gla{\ss}er et al. \cite{Glasser2013} use parameters derived from DCE-MRI image data to classify breast tumors.
+%%
+Their parameters are derived from clustering of bloodflow in the tumorous tissue.
+%%
+These features trains a decision-tree classifier, concluding that the extracted features alone are not sufficient for tumor type classification.
+%%
+Niemann et al. \cite{Niemann2014} assess hepatic steatosis (fatty liver disease) risk factors using decision trees.
+%%
+They presented an interactive data mining tool, which can analyze association rules and highlight interesting relations, which may trigger new hypothesis about the data.
+%%
+The presented paper either validate the significance of image derived parameter or derive new relationships by training decision trees to explain target variables.
+%%
+
+Pinhero et al. \cite{Pinheiro} applied association rule mining to create high-confidence decision rules about the mortality rate in liver cancer data.
+%%
+Decision rules are easily readable and understandable and are therefore well suited for classifying medical data.
+%%
+The results can be easily communicated to domain experts with little knowledge about the algorithm behind the results.
+%%
+Once these rules, however, get too complex, analyzing these decision rules can get complex.
+%%
+Sekhavat and Hoeber \cite{Sekhavat} propose a interactive visualization techniques for decision trees, where all decisions are displayed in a heat map
+%%
+The visualization translates the decision tree into a heat map, where each cell represents the connectivity between a initial value and an target value.
+%%
+The resulting interactive visualization is well suited for analyzing a decision tree, but not for multiple trees.
+%%
+\paragraph{Visual Analysis of Heterogenous Non-Image Data.}
+%%
+Zhang et al. \cite{Zhang} analyze subject groups in a web-based linked view system.
+%%
+The resulting decision rules aim to categorize new subjects as they are added to the data.
+%%
+They define a cohort as parameter-divided subject group, differing from the epidemiological understanding of a cohort.
+%%
+The paper lacks details about handling of missing data, the definition of similarity or which statistical measures are used.
+%%
+Generalized Pairs Plots (\texttt{GPLOM'S}) visualize heterogenous variables in a plot matrix \cite{GPLOMS, Francois2013}.
+%%
+The plot depends on the type of variables, which are pairwise visualized.
+%%
+The plot matrix is well suited for an overview visualization, but takes up much screen space and is therefore only suitable for few variables at once (see Fig.~\ref{image-parameter-range}).
+%%
+Dai et al. \cite{Dai} defined a \emph{Concept Map} linking cancer-associated features by incorporating choropleth maps with bar charts, parallel coordinates and scatter plots with regression lines.
+%%
+The \emph{Concept Map} is iteratively refined using primarily the underlying geographical data.
+%%
+Our work does not contain geographical information of any kind.
+%%----------------------------
+\paragraph{SHIP Data Analysis.}
+Klemm et al. analyzed the lumbar spine variability of of 490 \texttt{SHIP-2} subjects \cite{Klemm2013VMV}.
+%%
+They incorporate hierarchical agglomerative clustering to derive shape groups, yielding several groups of average shape and several outlier cluster.
+%%
+The extracted lumbar spine shape was correlated with subject size.
+%%
+Out data extraction process is based on the detection algorithm and lumbar spine canal extraction presented in this paper.
+%%
+Clustering techniques applied to non-image data are strongly dependent on the chosen variables and distance measures.
+%%
+Klemm et al. found \texttt{k-Prototypes} and \texttt{DBSCAN} appropriate for this task \cite{Klemm2014BVM}.
+%%
+\\
+Unique for our work is the combination of an IVA-approach by observing interesting feature relations in context of image-derived features using multiple decision trees for one visualization.
+%%
+We abstract the decision tree results similar to \cite{Turkay} to make them comparable in a overview visualization.
+%%
 \section{The Lumbar Spine Data Set}

### 2014-08-29 draft.tex
Reviised results figure

@@
 \end{figure}
-\com{Image Data}
 Four trained technicians acquired the medical image data for each subject in a standardized way using a 1.5~Tesla MRI scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany).
@@
   \centering
-  \includegraphics[width=0.82\textwidth]{figures/results}
+  \includegraphics[width=0.95\textwidth]{figures/results}
   \caption{

### 2014-08-29 draft.tex
Revvised Paper

@@
 \subsection{Epidemiological Data}
-Epidemiological data can stem from a wide range of studies.
+Epidemiological data originates from a wide range of studies.
 %%
@@
 %%
-Women-specific question can only be acquired for female subjects, data about a disease treatment only for subjects suffering from this condition.
-%%
-Therefore, statistical analysis has to take missing data into account.
+For example women-specific questions or data about a disease treatment only for subjects suffering from this condition.
+%%
+Statistical analysis has to take missing data into account.
 %%
@@
 %%
-The lower (\emph{lumbar}) spine is the most stressed part of the spine.
+The lower (\emph{lumbar}) spine is the most stressed spine part.
 %%
@@
 %%
-These include nutrition, sporting activities and body posture at work.
+These include nutrition, sporting activities and body posture combined with physical stress at work.
 %%
@@
 %%
-Klemm et al. \cite{Klemm2014VIS} propose an extension to the existing epidemiological workflow using Interactive Visual Analysis (IVA).
-%%
-The workflow consists of an iterative sequence of group selection, variable selection and visualization.
+Klemm et al. \cite{Klemm2014VIS} propose an extension to the previously described epidemiological workflow using \emph{Interactive Visual Analysis} (IVA).
+%%
+The workflow consists of an iterative sequence of \emph{group selection}, \emph{variable selection} and \emph{visualization}.
 %%
@@
 %%
-It incorporates multiple plot types, such as scatter plots, parallel coordinates and time plots with brushing and linking facilities to successively analyze a data set.
-%%
-Similar to our work, they quantify image data to map them into the same information space as non-image data.
+It incorporates multiple plot types, such as scatter plots, parallel coordinates and time plots with brushing and linking facilities.
+%%
+Similar to our work, they quantify image data project it into the non-image data information space.
 %%
@@
 %%
-Their proposed \emph{deviation plot} shows distribution-specific measures of a variable, such as skewness or inter-quantile-range, making dimensions comparable.
-%%
-This approach aims to trigger \emph{hypothesis generation} by outlining tendencies between these dimensions.
-
+Their proposed \emph{deviation plot} shows distribution-specific measures of a variable, such as skewness or inter-quantile-range, making variables comparable.
+%%
+This approach aims to trigger \emph{hypothesis generation} by outlining tendencies between these variables.
+%%
+\paragraph{Visual Analysis of Heterogenous Non-Image Data.}
+%%
+Zhang et al. \cite{Zhang} analyze subject groups in a web-based linked view system.
+%%
+The resulting decision rules aim to categorize new subjects as they are added to the data.
+%%
+They define a cohort as parameter-divided subject group, differing from the epidemiological understanding of the term.
+%%
+The paper lacks details about handling of missing data, the definition of similarity or which statistical measures are applied.
+%%
+Generalized Pairs Plots (\texttt{GPLOM'S}) visualize heterogenous variables in a plot matrix \cite{GPLOMS,Francois}.
+%%
+The plot depends on the type of variables, which are pairwise visualized.
+%%
+The plot matrix is well suited for a overview visualization, but takes up much screen space and is therefore only suitable for few variables at once (see Fig.~\ref{fig:image-parameter-range}).
+%%
+Dai et al. \cite{Dai} defined a \emph{Concept Map}, linking cancer-associated features by incorporating choropleth maps with bar charts, parallel coordinates and scatter plots with regression lines.
+%%
+The \emph{Concept Map} is iteratively refined using primarily the underlying geographical data.
+%%
+%Our work, however, does not contain geographical information of any kind.
+%%----------------------------
 \paragraph{Decision Rule Driven Analysis of Medical Data.}
-Closest to our work is the work of Gla{\ss}er \cite{Glasser2013} and Niemann et al. \cite{Niemann2014}.
-%%
 Related work in this field often focuses on clinical diagnosis.
 %%
-Gla{\ss}er et al. \cite{Glasser2013} use parameters derived from DCE-MRI image data to classify breast tumors.
-%%
-Their parameters are derived from clustering of bloodflow in the tumorous tissue.
-%%
-These features trains a decision-tree classifier, concluding that the extracted features alone are not sufficient for tumor type classification.
+Closest to ours is the work of Gla{\ss}er \cite{Glasser2013} and Niemann et al. \cite{Niemann2014}.
+%%
+Gla{\ss}er et al. \cite{Glasser2013} use variables derived from DCE-MRI image capturing the blood flow in the tumorous tissue.
+%%
+They aim to classify breast tumors.
+%%
+The extracted variables train a decision-tree classifier, concluding that the extracted features alone are not sufficient for tumor type classification.
 %%
@@
 %%
-The presented paper either validate the significance of image derived parameter or derive new relationships by training decision trees to explain target variables.
-%%
+We combine ideas of both papers by validating the significance of image derived parameter aiming to derive new relationships by training decision trees to explain target variables.
+%%----------------------------
 
@@
 %%
-Decision rules are easily readable and understandable and are therefore well suited for classifying medical data.
+Decision rules are easily readable and are well suited for classifying medical data. \com{Das frueher erwaehnen}
 %%
@@
 %%
-Once these rules, however, get too complex, analyzing these decision rules can get complex.
-%%
-Sekhavat and Hoeber \cite{Sekhavat} propose a interactive visualization techniques for decision trees, where all decisions are displayed in a heat map
+Once these rules, get too complex, analyzing these decision rules becomes complex.
+%%
+Sekhavat and Hoeber \cite{Sekhavat} propose a interactive visualization technique for decision trees, where all decisions are displayed in a heat map.
 %%
@@
 %%
-The resulting interactive visualization is well suited for analyzing a decision tree, but not for multiple trees.
-%%
-\paragraph{Visual Analysis of Heterogenous Non-Image Data.}
-%%
-Zhang et al. \cite{Zhang} analyze subject groups in a web-based linked view system.
-%%
-The resulting decision rules aim to categorize new subjects as they are added to the data.
-%%
-They define a cohort as parameter-divided subject group, differing from the epidemiological understanding of a cohort.
-%%
-The paper lacks details about handling of missing data, the definition of similarity or which statistical measures are used.
-%%
-Generalized Pairs Plots (\texttt{GPLOM'S}) visualize heterogenous variables in a plot matrix \cite{GPLOMS, Francois2013}.
-%%
-The plot depends on the type of variables, which are pairwise visualized.
-%%
-The plot matrix is well suited for an overview visualization, but takes up much screen space and is therefore only suitable for few variables at once (see Fig.~\ref{image-parameter-range}).
-%%
-Dai et al. \cite{Dai} defined a \emph{Concept Map} linking cancer-associated features by incorporating choropleth maps with bar charts, parallel coordinates and scatter plots with regression lines.
-%%
-The \emph{Concept Map} is iteratively refined using primarily the underlying geographical data.
-%%
-Our work does not contain geographical information of any kind.
-%%----------------------------
+The resulting interactive visualization is well suited for analyzing single, but not multiple decision trees.
+%%
 \paragraph{SHIP Data Analysis.}
@@
 %%
-\\
+\\\\
 Unique for our work is the combination of an IVA-approach by observing interesting feature relations in context of image-derived features using multiple decision trees for one visualization.
 %%
-We abstract the decision tree results similar to \cite{Turkay} to make them comparable in a overview visualization.
+We abstract the decision tree results similar to \cite{Turkay}, making them comparable in a overview visualization.
 %%
@@
 The variables range from somatometric variables describing body measures to medical examinations, such as laboratory tests as well as lifestyle factors as sport activity or nutrition.
+\com{Ein paar Details hierzu noch - Datentyp, missing data, aufeinander aufbauende parameter}
 
@@
 %%
-A click on the L3 vertebra center gives the model the initial position, two clicks on the top and bottom of the vertebra describes an initial model height estimation.
+A click on the L3 vertebra center initialized the model, two clicks on the top and bottom of the vertebra describes an initial height estimation.
 %%
@@
 %%
-Using the Frenet formulas \cite{Frenet}, we extracted the following metrics from the model (Fig.~\ref{fig:centerline}~b):
+Using the Frenet formulas \cite{Frenet}, we calculated the following metrics from the model (Fig.~\ref{fig:centerline}~b):
 %%
@@
 %%
-Larger people have also a longer spine and its shape is more straight compared to smaller subjects.
+Larger people have also a longer spine and its shape is more straight.
 %%
@@
 %%
-\noindent As first experiment we correlated the shape parameter with the binary back pain indicator using a generalized pairs plot \cite{GPLOMS}.
-%%
-Pairs plots visualize continuous and categorical variables pairwise in a matrix view.
-%%
-Scatterplots display pairs of continuous variables, bar charts for continuous variable as a function of a categorical variable and box plots pairs of categorical variables.
+\noindent As first experiment we correlated the shape parameter with the binary back pain indicator using \texttt{GPLOM's} \cite{GPLOMS}.
+%%
+%Pairs plots visualize continuous and categorical variables pairwise in a matrix view.
+%%
+%Scatterplots display pairs of continuous variables, bar charts for continuous variable as a function of a categorical variable and box plots pairs of categorical variables.
 %%
@@
 %%
-Curvature, curvature angle and torsion correlate strongly with their planar projections.
-%%
-Also the mean curvature and the curvature angle correlate by a factor of -0.89.
+\emph{Curvature}, \emph{curvature angle} and \emph{torsion} correlate strongly with their planar projections.
+%%
+Also the mean curvature and the curvature angle correlate by a factor of $-0.89$.
 %%
@@
 %%
-All correlation values are scaled between 0 - no correlation and 1 - identical.
+All correlation values are scaled between \texttt{0 - no correlation} and \texttt{1 - identical}.
 %%
@@
 %%
-This observation brought us to the conclusion to incorporate more sophisticated data mining techniques to assess the influence of the image-derived parameters.
+
+These observations brought us to the conclusion to incorporate more sophisticated data mining techniques to assess the influence of the image-derived parameters.
 
@@
 %%
-Leaves of a decision tree represent class labels, branches represent feature conjunctions leading to the class labels.
+Leafs of a decision tree represent class labels, branches represent feature conjunctions leading to the class labels.
 %%
@@
 %%
-Too many branches also impose to the data.
+Too many branches also impose overfitting to the data.
 \com{Quelle Decision Tree}
@@
 %%
-We follow the visual information-seeking mantra and want to provide an overview first, then details-on-demand \ref{shneiderman1996}.
+We follow the visual information-seeking mantra and want to provide an overview first, then details-on-demand \cite{shneiderman1996}.
 %%
@@
 
-Many follow-up variables are sparse, such as medication of diabetes or reason of early retirement.
+Many follow-up variables are sparse, such as \emph{medication of diabetes} or \emph{reason of early retirement}.
 %%
@@
 %%
-Outlier results with very large decision trees distort otherwise the resulting plot.
+Outlier results with very large decision trees would otherwise distort the resulting plot.
 %%
@@
 \subsubsection{Dummy Variables}
-Dummy variables abstract a categorical variable with multiple manifestations into several binary variables.
+Dummy variables convert a categorical variable with multiple manifestations into several binary variables.
 %%
@@
 %%
-We however still want to be able to look display \emph{details-on-demand} \ref{shneiderman1996} and examine a decision tree in detail.
-%%
-This is realized by clicking on a entry on the visualization, which then displays the corresponding decision tree in detail (Fig.~\ref{fig:decision-tree} b).
+We however still want to be able to look display \emph{details-on-demand} \cite{shneiderman1996} and examine a decision tree in detail.
+%%
+This is realized by clicking on a entry on the visualization, which then displays the corresponding decision tree in detail.
 %%

### 2014-09-01 draft.tex
revised version from Bernhard

@@
 \label{sec:Introduction}
-Epidemiology is the study of causation of diseases.
+%%Epidemiology is the study of causation of diseases.
+Epidemiology is the study of dissemination, causes and results of health-related states and events.
+%%
 For this purpose, large population studies, such as the \emph{Study of Health in Pomerania} (SHIP) [??] gather as much information as possible about participants to be assessed towards different diseases.
@@
 %%
-These studies often comprise medical image data, e.g., tomographic image data sets, as well as general patient-specific attributes like gender, age and weight.
+Since recently, these studies often comprise medical image data, e.g., tomographic image data sets, as well as general patient-specific attributes, such as gender, age and weight.
 %%
@@
 %%
-Epidemiological analysis of back pain aim to characterize the disease by determining risk factors.
+Epidemiological analysis of back pain aims to characterize the disease by determining risk factors.
 %Many epidemiological-motivated approaches aim at improved characterization of this disease with respect to patient-specific attributes.
@@
 %%
-Although the shape and constitution of the spine, especially the lumbar spine, plays an important role for the presence of back pain, no automatic classification approach to characterize back pain based on lumbar back spine attributes has been established yet.
+Although the shape and constitution of the spine, especially the lumbar spine, plays an important role for the presence of back pain, no automatic classification approach to characterize back pain based on lumbar back spine attributes has been established.
 %%
@@
 %%
-Therefore, we combine data mining algorithms with data visualization techniques.
+For this purpose, we combine data mining algorithms with data visualization techniques.
 %%
@@
 %\noindent \com{Background. Epidemiological Workflow, focus on statistical resilience, image data hard to analyze due to the large amount of subjects, poor image quality and lack of methods}.
-\noindent Epidemiology is the study of dissemination, causes and results of health-related states and events.
-%%
-Epidemiological reasoning relies on a strict statistically driven workflow \cite{Fletcher}:
+\noindent Epidemiological reasoning relies on a strict statistically driven workflow \cite{Fletcher}:
 %%
@@
 %%
-Age acts as an \emph{confounder} for prostate cancer.
-%%
-These \emph{confounding} variables are often hard to find.
+Age acts as a \emph{confounder} for prostate cancer.
+%%
+%These \emph{confounding} variables are often hard to find.
 %%
@@
 %%
-Most common are case-control studies, analyzing one specific disease and its influences.
+Most common are case-control studies, analyzing one specific disease and its influences on the human body.
 %%
@@
 %%
-For example women-specific questions or data about a disease treatment only for subjects suffering from this condition.
+For example women-specific questions or data about a disease treatment only subjects suffering from this condition.
 %%
@@
 %%
-These data are hard to analyze, since segmentation algorithms are not generally available and need to be custom-made for each body structure.
+These data are hard to analyze.
+%%
+Individual diagnosis or manual segmentation of each body structure by an radiologists is tedious, costly and comprises little reproducibility.
+%%
+Segmentation algorithms are not generally available and need to be custom-made for each body structure.
 %%
@@
 %%
-It aims to trigger \emph{hypothesis generation} by providing visualizations able to concurrently analyzing multiple heterogenous variables at once using correlation measures.
+It aims to trigger \emph{hypothesis generation} by providing visualizations able to concurrently analyze multiple heterogenous variables at once using correlation measures.
 %%
@@
 %%
-Similar to our work, they quantify image data project it into the non-image data information space.
+Similar to our work, they quantify image data and project it into the non-image data information space.
 %%
@@
 %%
-The plot matrix is well suited for a overview visualization, but takes up much screen space and is therefore only suitable for few variables at once (see Fig.~\ref{fig:image-parameter-range}).
-%%
-Dai et al. \cite{Dai} defined a \emph{Concept Map}, linking cancer-associated features by incorporating choropleth maps with bar charts, parallel coordinates and scatter plots with regression lines.
+The plot matrix is well suited for an overview visualization, but takes up much screen space and is therefore only suitable for few variables at once (see Fig.~\ref{fig:image-parameter-range}).
+%%
+Dai and Gahegan \cite{Dai} defined a \emph{Concept Map}, linking cancer-associated features by incorporating choropleth maps with bar charts, parallel coordinates and scatter plots with regression lines.
 %%
@@
 %%
-Gla{\ss}er et al. \cite{Glasser2013} use variables derived from DCE-MRI image capturing the blood flow in the tumorous tissue.
+Gla{\ss}er et al. \cite{Glasser2013} use variables derived from DCE-MRI data capturing the perfusion in the tumorous tissue.
 %%
@@
 %%
-We combine ideas of both papers by validating the significance of image derived parameter aiming to derive new relationships by training decision trees to explain target variables.
+We combine ideas of both papers by validating the significance of image-derived parameter aiming to derive new relationships by training decision trees to explain target variables.
 %%----------------------------
 
-Pinhero et al. \cite{Pinheiro} applied association rule mining to create high-confidence decision rules about the mortality rate in liver cancer data.
+Pinhero et al.
+\cite{Pinheiro} applied association rule mining to create high-confidence decision rules based on variables such as age, gender and geographic location about the mortality rate in liver cancer data.
 %%
@@
 %%
-Sekhavat and Hoeber \cite{Sekhavat} propose a interactive visualization technique for decision trees, where all decisions are displayed in a heat map.
-%%
-The visualization translates the decision tree into a heat map, where each cell represents the connectivity between a initial value and an target value.
+Sekhavat and Hoeber \cite{Sekhavat} propose an interactive visualization technique for decision trees, where all decisions are displayed in a heat map.
+%%
+The visualization translates the decision tree into a heat map, where each cell represents the connectivity between an initial value and an target value.
 %%
@@
 %%
-Out data extraction process is based on the detection algorithm and lumbar spine canal extraction presented in this paper.
+The data extraction process is based on the detection algorithm and lumbar spine canal extraction presented in this paper.
 %%
@@
 %%
-Klemm et al. found \texttt{k-Prototypes} and \texttt{DBSCAN} appropriate for this task \cite{Klemm2014BVM}.
+%Klemm et al. found \texttt{k-Prototypes} and \texttt{DBSCAN} appropriate for this task \cite{Klemm2014BVM}.
 %%
@@
 %%
-We abstract the decision tree results similar to \cite{Turkay}, making them comparable in a overview visualization.
+We abstract the decision tree results similar to \cite{Turkay}, making them comparable in an overview visualization.
 %%
@@
 %%
-The data set comprises of 6,753 subjects from two cohorts (4,420 from \texttt{SHIP-Trend-0} and 2,333 from \texttt{SHIP-2}).
+The data set comprises of 6.753 subjects from two cohorts (4.420 from \texttt{SHIP-Trend-0} and 2.333 from \texttt{SHIP-2}).
 %%
@@
 %\com{Analysis Using Heterogeneous Correlation}
-The variables range from somatometric variables describing body measures to medical examinations, such as laboratory tests as well as lifestyle factors as sport activity or nutrition.
+The variables range from somatometric variables describing body measures to medical examinations, such as laboratory tests as well as lifestyle factors as sporting activity or nutrition.
 \com{Ein paar Details hierzu noch - Datentyp, missing data, aufeinander aufbauende parameter}
@@
 	%%
-	We also extracted the variables for each projection axis which to assess their information gain.
+	We also extracted the variables for each projection axis to assess their information gain.
 	%(b) Using the Frenet formulas, we extract the curvature, torsion and the curvature angle from the lumbar spine canal center line.
@@
 \end{figure}
-Four trained technicians acquired the medical image data for each subject in a standardized way using a 1.5~Tesla MRI scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany).
+Four trained radiology technicians acquired the medical image data for each subject in a standardized way using a 1.5~Tesla MRI scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany).
 %%
@@
 %%
-The tetrahedron-based finite element model is initialized using three user-placed land marks.
+The tetrahedron-based finite element model is initialized using three user-placed landmarks.
 %%
@@
 %%
-In a previous work \cite{Klemm2013VMV,Klemm2014VIS}, we extracted a centerline representation of the lumbar spine canal from the detection model (Fig.~\ref{fig:centerline}~a).
+In a previous work \cite{Klemm2013VMV}, we extracted a centerline representation of the lumbar spine canal from the detection model (Fig.~\ref{fig:centerline}~a).
 %%
@@
 %%
-Since men are on average taller than women and people of old age shrink due to bone erosion, gender and age are also confounder (Fig.~\ref{fig:age-gender}).
+Since men are on average taller than women and people of old age shrink due to bone erosion, gender and age are also confounders (Fig.~\ref{fig:age-gender}).
 %%
@@
 %%
-Since the gender confounder mainly encodes body-height, we decided to divide subjects into body-size groups.
+Since the gender confounder mainly encodes body-size, we decided to divide subjects into body-size groups.
 %%
@@
   \caption{%%
-A generalized pairs plot of all image-derived variables \cite{GPLOMS} colored by presence or absence of back pain.
+A generalized pairs plot of all image-derived variables colored by presence or absence of back pain.
 %%
@@
 \subsubsection{Correlation Matrix}
-We calculated an association matrix to to assess correlations between the image parameters.
+We calculated an association matrix to assess correlations between the image parameters.
 %%
@@
 %%
-No statistical significant correlation could be observed through all subject groups.
+No statistically significant correlation could be observed through all subject groups.
 %%
@@
 %%
-The first three principal components explain 75\% of variance in the image-derived variables.
+The first three principal components explain 75\% of the variance in the image-derived variables.
 %%
@@
 %%
-The second component, adding 16\% of variance, weigths  \emph{curvature coronal} and \emph{curvature angle coronal}.
-%%
-The third component explains 12\% of variance and weights \emph{torsion} and \emph{curvature transverse}.
+The second component, adding 16\% of the variance, weigths  \emph{curvature coronal} and \emph{curvature angle coronal}.
+%%
+The third component explains 12\% of the variance and weights \emph{torsion} and \emph{curvature transverse}.
 %%
@@
 %%
-We rely on predictive classification trained to obtain a complex rule set on how combination of the image-parameter explain non-image variables.
+We rely on predictive classification trained to obtain a complex rule set on how combinations of the image-parameter explain non-image variables.
 %%
@@
 %%
-This allows epidemiological domain experts to interpret the results without having deep knowledge about the algorithm creating the tree.
+This allows epidemiologists to interpret the results without having deep knowledge about the algorithm creating the tree.
 %%
@@
 %%
-This means, that every node in the tree represents the attribute which splits the data most efficient into the target subset.
+This means, that every node in the tree represents the attribute which splits the data most efficiently into the target subset.
 %%
@@
 \subsection{Interactive Display of Decision Trees}
-We have to create a decision tree for every non-image variable to see which one can be explained by image-derived variables.
+We have to create a decision tree for every non-image variable to analyze which one can be explained by image-derived variables.
 %%
@@
 %%
-Dividing the subjects into four groups using body-size, the number increases to 402 trees.
-%%
-To make the results cognitive feasible, we have to abstract the results of the classification.
+Dividing the subjects into four groups using body size, the number increases to 402 trees.
+%%
+%To make the results cognitive feasible, we have to abstract the results of the classification.
+We have to abstract the results of the classification to keep the mental effort of interpreting the data low.
 %%------------
@@
 %%
-We follow the visual information-seeking mantra and want to provide an overview first, then details-on-demand \cite{shneiderman1996}.
+%We follow the visual information-seeking mantra and want to provide an overview first, then details-on-demand \cite{shneiderman1996}.
+We follow the Visual Analytics mantra by analyzing first, show the important and analyze further \cite{Keim}.
+%%
+A first analysis step was performed by applying the classification algorithm to the data.
 %%
@@
 %%
-If for example a variable indicating a disease is negative for 90\% of the subject and the classifier simply assigns all subjects to \emph{not ill} the error metric would yield an error of 10\%, even though it is very bad.
+If for example a variable indicating a disease is negative for 90\% of the subjects and the classifier simply assigns all subjects to \emph{not ill} the error metric would yield an error of 10\%, even though it is very bad.
 %%
@@
 %%
+%%
+We consider a result rated below $0.25$ a good classification.
+%%
 It allows for comparability of error rates between variables with different manifestation count.
@@
 %%
-Lets say we want to classify and compare the same variables for different subject groups, for example male and female subjects.
+If we for example classify and compare the same variables for different subject groups, for example male and female subjects.
 %%
@@
 %%
-We removed variables with less than 100 subjects, since they have to few subjects to be statistically significant.
+We removed variables with less than 100 subjects, since they have too few subjects to be statistically significant.
 %%
@@
 %%
-This is realized by clicking on a entry on the visualization, which then displays the corresponding decision tree in detail.
+This is realized by clicking on an entry on the visualization, which then displays the corresponding decision tree in detail.
 %%
@@
 \subsubsection{Implementation}
-All analysis are carried out using \texttt{R}, a popular programming language for statistical calculations and visualizations.
+All analysis are carried out using \texttt{R}, a widely used programming language for statistical calculations and visualizations.
 %%
@@
 %%
-In order to make the train of thought comprehensible, we used \texttt{RMarkdown}, which allows to make reports by combining \texttt{R} with the \texttt{Markdown} syntax.
+In order to make the train of thought comprehensible, we used \texttt{RMarkdown}, which allows to create reports by combining \texttt{R} with the \texttt{Markdown} syntax.
 %%
@@
 %%
-They can try out the technique without installing any software.
+They can use the technique without installing any software.
 %%
@@
 %%
-This is reflected in the large amount of variables classified with a error rate above $0,6$.
+This is reflected in the large amount of variables classified with an error rate above $0,6$.
 %%
@@
 % None of the pain indicators can be reliably described using the image-based parameter.
-None of the pain indicators can be described reliably using the image-based parameter.
-%%
-The only variable reliably classified in this group is gender.% with an error rate of 31,5\% using a decision tree of seven rules.
+None of the pain indicators can be described reliably using the image-based parameters.
+%%
+The only variable reliably classified in this group is gender. % with an error rate of 31,5\% using a decision tree of seven rules.
 %
 It can be discriminated with an error rate of $0,31$ using 7 rules and incorporates only curvature and curvature angle variables.
+%%
+The distinctness lies in the average difference in body size between male and female subjects.
 %
@@
 %%
-Surprisingly high is the classification of medication for high blood pressure.
-%%
-1,058 subjects are classified with an error rate of $0,47$ solely based on coronal mean curvature.
+\emph{Medication for high blood pressure} is classified for 1,058 subjects with an error rate of $0,47$ solely based on coronal mean curvature.
+%%
+%1,058 subjects are classified with an error rate of $0,47$ solely based on coronal mean curvature.
 %%
@@
 %%
-Therefore the classifier is not as useful as the error-rate would indicate.
+Therefore the classifier is not as useful.
 %%This is however due to the high amount of people, who are medicated (68\%) and who are almost all correctly class
@@
 %%
-The four body size groups could be characterized with an error rate $0,48$, but the decision tree comprises of 71 rules and imposes overfitting.
+The four body size groups could be characterized with an error rate $0,48$, but the decision tree comprises 71 rules and imposes overfitting.
 
@@
 %%
-Since there are only 30 male subjects diagnosed with hypothyroidism, reducing the statistical power of the result.
-%%
-The dummy variable analysis showed that female subjects of 139-164 cm body height could be discriminated using the mean curvature and curvature angle, with an error of $0,38$.
+Since there are only 30 male subjects diagnosed with hypothyroidism, the statistical power of the result is reduced.
+%%
+The dummy variable analysis showed that female subjects of 139-164 cm body size could be discriminated using the mean curvature and curvature angle, with an error of $0,38$.
 
@@
 %%
-Gender could be described for each age group using \emph{mean curvature} and \emph{curvature angle}.
+Gender could be classified for each age group using \emph{mean curvature} and \emph{curvature angle}.
 %%
@@
 %%
-The starting age of smoking could be characterized well with an accuracy between $0.25$ to $0.3$ for all age groups except of $21.1-43.8$ years of age.
+The starting age of smoking could be characterized well with an error rate between $0.25$ to $0.3$ for all age groups except of $21.1-43.8$ years of age.
 %%
@@
 %%
-Many previously described results can be associated to be confounded with subject size.
-%%
-Differences in the gender analysis are mostly due to the average height difference between males and females.
+Many previously described results are confounded with subject size.
+%%
+Differences between male and female subjects are explained by to the average body size difference.
 %%
@@
 %%
-Dividing subjects into height groups potentially highlights classifications not confounded by body-height.
+Dividing subjects into body size groups potentially highlights classifications not confounded by body size.
 %%
 \paragraph{Large Decision Trees. }
-Back-pain associated variables can be explained for various size-groups, but we could not extract universal rules.
-%%
-Radiating back pain could be described with error rate of $0.39$ using 23 rules for subjects between $171-177~cm$ body size using torsion and mean-curvature.
+Back pain-associated variables can be explained for various size-groups, but we could not extract universal rules.
+%%
+Radiating back pain could be described with error rate of $0.39$ using 23 rules for subjects between $171-177~cm$ body size using \emph{torsion} and \emph{mean curvature}.
 %%
@@
 %%
-There are several decision trees for laboratory values, such as alanine aminotransferase value in the blood can be described with a high precision of $0.4$ ($139-164~cm$) to $0.36$ ($164 - 171~cm$).
+There are several decision trees for laboratory values, such as alanine aminotransferase value (relevant for diagnosis of lover or gallbladder illness) in the blood can be described with a high precision of $0.4$ ($139-164~cm$) to $0.36$ ($164 - 171~cm$).
 %%
@@
 %%
-They are well suited to characterize subject gender due to their height difference using mean curvature and mean curvature angle.
-%%
-The reason for this lies in the higher curvature of smaller subjects and the significant difference of average body size between males and females.
+They are well suited to characterize subject gender due to their body size difference using mean curvature and mean curvature angle.
+%%
+This is due to the higher curvature of smaller subjects and the significant difference of average body size between males and females.
 %%
@@
 
-Subjects grouped using body height have a similar spine shape and a lower variance.
-%%
-The remaining information within these groups is not enough to characterize back pain related variables.
-%%
-Certain subgroups, however, can be classified for various size groups, such as nutrition or psychological parameters.
-%%
-
-Another observation during the analysis are very large decision trees achieving low error scores.
+Subjects grouped using body size have a similar spine shape and a lower variance.
+%%
+The remaining information within these groups is not enough to characterize back pain-related variables.
+%%
+%Certain parameters can be classified in 
+For some body size groups, certain nutrition or psychological parameters can be predicted with a comparatively high error rate.
+%Certain subgroups, however, can be classified for various size groups, such as nutrition or psychological parameters.
+%%
+
+Another observation during the analysis is very large decision trees achieving low error scores.
 %%
@@
 %% ------------------------------------
-\subsection{Applicability of the Method}
+\subsection{Applicability}
 %%
@@
 %%
-\noindent We showed, that torsion, curvature and curvature angle of the lumbar spine at the presented precision are not enough to characterize lumbar back pain in the SHIP-Data set.
+\noindent We showed, that torsion, curvature and curvature angle of the lumbar spine at the presented precision are not enough to characterize lumbar back pain in the SHIP data set.
 %%
@@
 
-\noindent SHIP is part of the Community Medicine Research net of the University of Greifswald, Germany, which is funded by the Federal Ministry of Education and Research (grant no.
+\noindent \com{TODO: Marko hinzufuegen} SHIP is part of the Community Medicine Research net of the University of Greifswald, Germany, which is funded by the Federal Ministry of Education and Research (grant no.
 03ZIK012), the Ministry of Cultural Affairs as well as the Social Ministry of the Federal State of Mecklenburg-West Pomerania.

### 2014-09-01 draft.tex
Rough draft of abstract

@@
 
-\abstract{The abstract should summarize the contents of the paper and should contain at least 70 and at most 200 words.
+% \abstract{
+% The abstract should summarize the contents of the paper and should contain at least 70 and at most 200 words.
 The text must be set to 9-point font size.
@@
 Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.}
+\abstract{ 
+% Motivation
+Epidemiology is the study of causation of diseases.
+%%
+Subject groups (a cohort) is analyzed to correlate lifestyle and medical variables with diseases.
+%%
+Most recently, large cohort studies also acquire medical image data.
+%% Problem statement
+We correlate image-derived variables from the lumbar spine with lower back pain.
+%% Approach
+Our analysis is based on detection masks from a semi-automatic model based detection of 2,540 lumbar spines.
+%%
+From the detection masks, we derive shape-describing variables, such as course, curvature and torsion of the spine canal.
+%%
+We classify non-image parameter using decision trees created using the C4.5/C5.0 algorithm.
+%%
+The analysis yields a decision tree per non-image parameter.
+%%
+We make these trees comparable by introducing tree size and a classification error rate metrics.
+%%
+A interactive scatterplot visualization allows to compare the decision trees.
+%%
+%\\\\
+% Results
+% Conclusions
+We conclude that the metrics extracted from the detection masks are only sufficient for classification of lifestyle factors, such as gender or nutrition.
+%%
+The presented approach is a valuable extension to the interactive visual analysis of epidemiological data and not limited to image-parameter analysis.
+}
 
@@
 %%
-We however still want to be able to look display \emph{details-on-demand} \cite{shneiderman1996} and examine a decision tree in detail.
+We still want to be able to display \emph{details-on-demand} \cite{shneiderman1996} and examine a decision tree in detail.
 %%

### 2014-09-01 draft.tex
Revised abstract

@@
 %%
-Subject groups (a cohort) is analyzed to correlate lifestyle and medical variables with diseases.
+Subject groups (a cohort) are analyzed to correlate lifestyle and medical properties with diseases.
 %%
@@
 %% Problem statement
-We correlate image-derived variables from the lumbar spine with lower back pain.
+%%
+We want to assess the predictive quality of image-derived variables for lower back pain
+%%
+%We correlate image-derived variables from the lumbar spine with lower back pain.
 %% Approach
-Our analysis is based on detection masks from a semi-automatic model based detection of 2,540 lumbar spines.
-%%
-From the detection masks, we derive shape-describing variables, such as course, curvature and torsion of the spine canal.
-%%
-We classify non-image parameter using decision trees created using the C4.5/C5.0 algorithm.
+Our analysis is based on masks from a semi-automatic model based detection of 2,540 lumbar spines.
+%%
+These masks are quantified using shape-describing variables, such as spine canal course, -curvature and -torsion.
+%From the detection masks, we derive shape-describing variables, such as course, curvature and torsion of the spine canal.
+%%
+Using the C4.5/C5.0 algorithm we created decision trees to predict non-image parameter, such as pain indicators or occupational circumstances.
+%We classify non-image parameter using decision trees created using the C4.5/C5.0 algorithm.
 %%
@@
 %%
-We make these trees comparable by introducing tree size and a classification error rate metrics.
+%We make these trees comparable in a scatter plot using the \emph{tree size} and a \emph{classification error rate} metric.
+We make these trees comparable by using the \emph{tree size} and a \emph{classification error rate} metric.
 %%
 A interactive scatterplot visualization allows to compare the decision trees.
+%%
+The comparative analysis was performed on subjects divided by age, gender and body size to assess the influences of these variables.
 %%
@@
 %%
-The presented approach is a valuable extension to the interactive visual analysis of epidemiological data and not limited to image-parameter analysis.
+The presented approach is a valuable extension to the interactive visual analysis of epidemiological data and is not limited to image-parameter analysis.
 }

### 2014-09-02 draft.tex
fixed switched plots in results figure

@@
 %%
-There are several decision trees for laboratory values, such as alanine aminotransferase value (relevant for diagnosis of lover or gallbladder illness) in the blood can be described with a high precision of $0.4$ ($139-164~cm$) to $0.36$ ($164 - 171~cm$).
+There are several decision trees for laboratory values, such as alanine aminotransferase value (relevant for diagnosis of lover or gallbladder illness) in the blood can be described with an error rate of $0.4$ ($139-164~cm$) to $0.36$ ($164 - 171~cm$).
 %%

### 2014-09-03 draft.tex
Revised Abstract with Sylvia

--- 2014-09-01 draft.tex
@@
 \abstract{ 
-% Motivation
-Epidemiology is the study of causation of diseases.
-%%
-Subject groups (a cohort) is analyzed to correlate lifestyle and medical variables with diseases.
-%%
-Most recently, large cohort studies also acquire medical image data.
-%% Problem statement
-We correlate image-derived variables from the lumbar spine with lower back pain.
-%% Approach
-Our analysis is based on detection masks from a semi-automatic model based detection of 2,540 lumbar spines.
-%%
-From the detection masks, we derive shape-describing variables, such as course, curvature and torsion of the spine canal.
-%%
-We classify non-image parameter using decision trees created using the C4.5/C5.0 algorithm.
-%%
-The analysis yields a decision tree per non-image parameter.
-%%
-We make these trees comparable by introducing tree size and a classification error rate metrics.
-%%
-A interactive scatterplot visualization allows to compare the decision trees.
-%%
-%\\\\
-% Results
-% Conclusions
-We conclude that the metrics extracted from the detection masks are only sufficient for classification of lifestyle factors, such as gender or nutrition.
-%%
-The presented approach is a valuable extension to the interactive visual analysis of epidemiological data and not limited to image-parameter analysis.
+The goal of epidemiology is to provide insight into disease causations.
+%%
+Hence, subject groups (cohorts) are analyzed to correlate the subjects' varying lifestyles and medical properties with diseases.
+%%
+Recently, these cohort studies comprise medical image data as well.
+%%
+We assess the predictive quality of image-derived variables for lower back pain.
+%%
+The evaluation is embedded in an interactive visual analysis (IVA) framework that is based on 2,540 segmented lumbar spine data sets.
+%%
+The segmentation results are evaluated and quantified by employing shape-describing variables, such as the spine canal course, its curvature and torsion.
+%%
+We analyze the predictive power of shape-describing variables for non-image variables, e.g., pain indicators.
+%%
+Therefore, we automatically extract a decision tree classifier for each non-image variable.
+%%
+Furthermore, we provide an IVA technique to compare the classifiers with interactive scatterplot visualizations.
+%%
+As a first result, we conclude that image-based variables are only sufficient for classification of lifestyle factors.
+%%
+A correlation between lumbar spine shape and lower back pain disease was not existent in the automatically trained classifiers.
+%%
+However, the presented approach is a valuable extension for the IVA of epidemiological data.
+%%
+Hence, correlations between non-image parameters were successfully detected and characterized.
 }

### 2014-09-03 draft.tex
numbers now properly delimited by points

@@
 %%
-We only display variables with an error rate below $0,5$, results above this threshold are inaccurate.
+We only display variables with an error rate below $0.5$, results above this threshold are inaccurate.
 %%
@@
 %%
-This is reflected in the large amount of variables classified with an error rate above $0,6$.
-%%
-%The first impression of the resulting scatter plots quickly shows a very high amount of variables above an error rate of $0,6$.
+This is reflected in the large amount of variables classified with an error rate above $0.6$.
+%%
+%The first impression of the resulting scatter plots quickly shows a very high amount of variables above an error rate of $0.6$.
 %%
@@
 %%
-The only variable reliably classified in this group is gender. % with an error rate of 31,5\% using a decision tree of seven rules.
+The only variable reliably classified in this group is gender. % with an error rate of 31.5\% using a decision tree of seven rules.
 %
-It can be discriminated with an error rate of $0,31$ using 7 rules and incorporates only curvature and curvature angle variables.
+It can be discriminated with an error rate of $0.31$ using 7 rules and incorporates only curvature and curvature angle variables.
 %%
@@
 %%
-\emph{Medication for high blood pressure} is classified for 1,058 subjects with an error rate of $0,47$ solely based on coronal mean curvature.
+\emph{Medication for high blood pressure} is classified for 1,058 subjects with an error rate of $0.47$ solely based on coronal mean curvature.
 %%
@@
 %%
-The four body size groups could be characterized with an error rate $0,48$, but the decision tree comprises 71 rules and imposes overfitting.
+The four body size groups could be characterized with an error rate $0.48$, but the decision tree comprises 71 rules and imposes overfitting.
 
@@
 %%
-The classifier neglects the other group and yields a error rate below $0,5$.
+The classifier neglects the other group and yields a error rate below $0.5$.
 %The dummy variable analysis yields fewer results.
@@
 %%
-Only hypothyroidism could be described for male subjects with an error rate of $0,24$ for 110 subjects using the \emph{mean curvature} and \emph{curvature angle}.
+Only hypothyroidism could be described for male subjects with an error rate of $0.24$ for 110 subjects using the \emph{mean curvature} and \emph{curvature angle}.
 %%
@@
 %%
-The dummy variable analysis showed that female subjects of 139-164 cm body size could be discriminated using the mean curvature and curvature angle, with an error of $0,38$.
+The dummy variable analysis showed that female subjects of 139-164 cm body size could be discriminated using the mean curvature and curvature angle, with an error of $0.38$.
 
@@
 %%
-The dummy variables show several variables described using only one decision rule with accuracy between $0,42$ to $0,47$.
+The dummy variables show several variables described using only one decision rule with accuracy between $0.42$ to $0.47$.
 %%

### 2014-09-03 draft.tex
Filled in blanks in Document heading

@@
 
-\author{\authorname{Paul Klemm\sup{1}, Sylvia Glaßer\sup{1}, Kai Lawonn\sup{1} Henry Völzke\sup{2} and Bernhard Preim\sup{1}}
+\author{\authorname{Paul Klemm\sup{1}, Sylvia Glaßer\sup{1}, Kai Lawonn\sup{1} Henry Völzke\sup{2}, Katrin Hegenscheid\sup{3} and Bernhard Preim\sup{1}}
 \affiliation{\sup{1}Department of Simulation and Graphics, University of Magdeburg}
-\affiliation{\sup{2}Greifswald}
-% TODO: Mailaddy nachtragen
-\email{\{paul, sylvia\}@isg.cs.uni-magdeburg.de, voelzke@anderAddy}
+\affiliation{\sup{2} Institut for Community Medicine, University of Greifswald}
+\affiliation{\sup{3} Institute for Diagnostic Radiology and Neuro-Radiology, University of Greifswald}
+\email{\{paul, sylvia\}@isg.cs.uni-magdeburg.de, \{katrin.hegenscheid, voelzke\}@uni-greifswald.de}
 }
 
-% TODO: The paper must have at least one keyword.
-The text must be set to 9-point font size and without the use of bold or italic font style.
-For more than one keyword, please use a comma as a separator.
-Keywords must be titlecased.
-\keywords{My Keywords in Title Case}
-
-% \abstract{
-% The abstract should summarize the contents of the paper and should contain at least 70 and at most 200 words.
-The text must be set to 9-point font size.
-Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.}
+\keywords{Epidemiology, Interactive Visual Analysis}
+
+% \abstract{The abstract should summarize the contents of the paper and should contain at least 70 and at most 200 words.}
 \abstract{ 

### 2014-09-03 draft.tex
Most of the variables  no italic

@@
 \end{itemize}
-
-% -- Old Introduction
-% Epidemiology is the study of causation of diseases.
-% %%
-% Large population studies, such as the Study of Health in Pomerania (SHIP) \ref{SHIP} gather as much information as possible about participants to be assessed towards different diseases.
-% %%
-% These information are used to determine risk factors for diseases, helping people to make their lifestyle healthier or helping in diagnosing a disease.
-% %%
-% Epidemiological research is strongly hypothesis driven.
-% %
-% Observations made by clinicians are translated into hypothesis, which are then statistically evaluated using data variables from epidemiological studies.
-%
-% Modern cohort studies often comprise medical image data.
-% %%
-% \com{Short summary}
-% %%
-%
-% Back pain is one of the most frequent diseases in the western civilization.
-% %%
-%
-% Our goal is to combine data mining algorithm with data visualization to provide insight into the quality of image derived data to analyze if it acts as a risk factor for a disease.
-% %\com{Combine Data Mining Algo}
-%
-% More than 5 variables are rarely analyzed simultaneously.
-% %%
-% %The variables analyzed simultaneously is limited to
-% Our contributions are:
-% \begin{itemize}
-% 	\item Analyzing back pain using image-derived variables of 2,240 subjects.
-% 	\item Assessing the suitability of lumbar spine shape for diagnosing back pain
-% 	\item Analyzing correlations between between image-based and socio-demographic as well as medical parameters.
-% 	\item Identification of most important variables using data mining methods \com{noch schoen schreiben}
-% 	\item semiquantitative Auswertmöglichkeiten werden dem User oft zur verfügung gestellt
-% 	\item Techniken sind als Teil des IVA-Frameworkes zu verstehen
-% \end{itemize}
-
+%% ---------------------------------------------
 \section{\uppercase{Epidemiological Background}}
 \label{sec:EpidemiologicalBackground}
-%\noindent \com{Background. Epidemiological Workflow, focus on statistical resilience, image data hard to analyze due to the large amount of subjects, poor image quality and lack of methods}.
+%%
 \noindent Epidemiological reasoning relies on a strict statistically driven workflow \cite{Fletcher}:
@@
 %%
-no group (a), subdivision into male and female subjects (b), quartiles of age (c) and quartiles of body size (d).
+no group (a), subdivision into \emph{male and female subjects} (b), quartiles of \emph{age} (c) and quartiles of \emph{body size} (d).
 %%
@@
 
-The analysis of the dummy variable yields similar results like the blood pressure medication.
-%%
-Variables, such subjects sized 139-164 cm, between 64 and 90 years of age or nutrition related parameters are dominantly populated by one manifestation.
+The analysis of the dummy variable yields similar results like the \emph{blood pressure medication}.
+%%
+Variables, such as subjects sized 139-164 cm, between 64 and 90 years of age or nutrition related parameters are dominantly populated by one manifestation.
 %%
@@
 %%
-Only hypothyroidism could be described for male subjects with an error rate of $0.24$ for 110 subjects using the \emph{mean curvature} and \emph{curvature angle}.
-%%
-Since there are only 30 male subjects diagnosed with hypothyroidism, the statistical power of the result is reduced.
-%%
-The dummy variable analysis showed that female subjects of 139-164 cm body size could be discriminated using the mean curvature and curvature angle, with an error of $0.38$.
+Only \emph{hypothyroidism} could be described for male subjects with an error rate of $0.24$ for 110 subjects using the \emph{mean curvature} and \emph{curvature angle}.
+%%
+Since there are only 30 male subjects diagnosed with \emph{hypothyroidism}, the statistical power of the result is reduced.
+%%
+The dummy variable analysis showed that female subjects of \emph{139-164~cm body size} could be discriminated using the \emph{mean curvature} and \emph{curvature angle}, with an error of $0.38$.
 
@@
 %%
-The accuracy varies between $0.35$ ($43.8-54$ years of age, 3 decision rules) to $0.27$ ($21-43.8$ years of age, 6 decision rules).
-%%
-The starting age of smoking could be characterized well with an error rate between $0.25$ to $0.3$ for all age groups except of $21.1-43.8$ years of age.
+The accuracy varies between $0.35$ (\emph{$43.8-54$ years of age}, 3 decision rules) to $0.27$ (\emph{$21-43.8$ years of age}, 6 decision rules).
+%%
+The starting age of smoking could be characterized well with an error rate between $0.25$ to $0.3$ for all \emph{age} groups except of \emph{$21.1-43.8$ years of age}.
 %%
@@
 
-Many variables, such as body size can be described with an error rate of $0.35$ to $0.45$ but only using large decision trees with over 20 rules.
+Many variables, such as \emph{body size} can be described with an error rate of $0.35$ to $0.45$ but only using large decision trees with over 20 rules.
 %Body-size can be determined with an error rate of $0,36$ to $0,39$ for subject
@@
 %%
-For subjects between $43.8-54$ years of age, the classifier discriminated subjects with thyroid nodules with an error rate of $0.32$.
+For subjects between \emph{$43.8-54$ years of age}, the classifier discriminated subjects with thyroid nodules with an error rate of $0.32$.
 
@@
 %%
-Differences between male and female subjects are explained by to the average body size difference.
+Differences between \emph{male} and \emph{female} subjects are explained by to the average \emph{body size} difference.
 %%
@@
 %%
-Dividing subjects into body size groups potentially highlights classifications not confounded by body size.
+Dividing subjects into \emph{body size} groups potentially highlights classifications not confounded by \emph{body size}.
 %%
 \paragraph{Large Decision Trees. }
-Back pain-associated variables can be explained for various size-groups, but we could not extract universal rules.
-%%
-Radiating back pain could be described with error rate of $0.39$ using 23 rules for subjects between $171-177~cm$ body size using \emph{torsion} and \emph{mean curvature}.
-%%
-For subjects sized $177-202~cm$ the accuracy drops to $0.47$ using 20  decision rules.
-%%
-There are several decision trees for laboratory values, such as alanine aminotransferase value (relevant for diagnosis of lover or gallbladder illness) in the blood can be described with an error rate of $0.4$ ($139-164~cm$) to $0.36$ ($164 - 171~cm$).
-%%
-Similar values can be observed for cholesterol or age.
+Back pain-associated variables can be explained for various \emph{size}-groups, but we could not extract universal rules.
+%%
+Radiating back pain could be described with error rate of $0.39$ using 23 rules for subjects between \emph{$171-177~cm$ body size} using \emph{torsion} and \emph{mean curvature}.
+%%
+For subjects sized \emph{$177-202~cm$} the accuracy drops to $0.47$ using 20 decision rules.
+%%
+There are several decision trees for laboratory values, such as \emph{alanine aminotransferase} value (relevant for diagnosis of liver or gallbladder illness) in the blood can be described with an error rate of $0.4$ (\emph{$139-164~cm$}) to $0.36$ (\emph{$164 - 171~cm$}).
+%%
+Similar values can be observed for \emph{cholesterol} or \emph{age}.
 %%
@@
 %%
-These variables include nutrition parameter, thyroid disorder and social problems induced by back pain.
+These variables include \emph{nutrition} parameter, \emph{thyroid disorder} and \emph{social problems induced by back pain}.
 
@@
 \label{sec:Conclusion}
-\noindent The presented results indicate that the parameters torsion, curvature and curvature angle are not enough to characterize pathological changes in the lumbar spine.
-%%
-They are well suited to characterize subject gender due to their body size difference using mean curvature and mean curvature angle.
-%%
-This is due to the higher curvature of smaller subjects and the significant difference of average body size between males and females.
-%%
-Gender was characterized well in the analysis of all subjects as well as subjects divided by age.
-
-Subjects grouped using body size have a similar spine shape and a lower variance.
+\noindent The presented results indicate that the parameters \emph{torsion}, \emph{curvature} and \emph{curvature angle} are not enough to characterize pathological changes in the lumbar spine.
+%%
+They are well suited to characterize subject gender due to their body size difference using \emph{mean curvature} and \emph{mean curvature angle}.
+%%
+This is due to the higher curvature of smaller subjects and the significant difference of average \emph{body size} between \emph{males} and \emph{females}.
+%%
+Gender was characterized well in the analysis of all subjects as well as subjects divided by \emph{age}.
+
+Subjects grouped using \underline{body size} have a similar spine shape and a lower variance.
 %%
@@
 %Certain parameters can be classified in 
-For some body size groups, certain nutrition or psychological parameters can be predicted with a comparatively high error rate.
+For some \emph{body size} groups, certain \emph{nutrition} or \emph{psychological} parameters can be predicted with a comparatively high error rate.
 %Certain subgroups, however, can be classified for various size groups, such as nutrition or psychological parameters.
@@
 %%
-\noindent We showed, that torsion, curvature and curvature angle of the lumbar spine at the presented precision are not enough to characterize lumbar back pain in the SHIP data set.
-%%
-Our method made it possible to assess their discriminative power, which is largely limited for separating male and female subjects, nutrition variables, as well as different disease indicators.
+\noindent We showed, that \emph{torsion}, \emph{curvature} and \emph{curvature angle} of the lumbar spine at the presented precision are not enough to characterize lumbar back pain in the SHIP data set.
+%%
+Our method made it possible to assess their discriminative power, which is largely limited for separating \emph{male} and \emph{female} subjects, \emph{nutrition} variables, as well as different disease indicators.
 %%

### 2014-09-03 draft.tex
Moved cite commands

@@
 %%
-Klemm et al. \cite{Klemm2014VIS} propose an extension to the previously described epidemiological workflow using \emph{Interactive Visual Analysis} (IVA).
+Klemm et al. propose an extension to the previously described epidemiological workflow using \emph{Interactive Visual Analysis} (IVA) \cite{Klemm2014VIS}.
 %%
@@
 %%
-Steenwijk et al. \cite{Steenwijk} propose a coordinated linked view system for both image-related and non-image data.
+Steenwijk et al. propose a coordinated linked view system for both image-related and non-image data \cite{Steenwijk}.
 %%
@@
 %%
-Turkay et al. \cite{Turkay} follow a similar approach by deriving descriptive data metrics from image data.
+Turkay et al. follow a similar approach by deriving descriptive data metrics from image data \cite{Turkay}.
 %%
@@
 %%
-Zhang et al. \cite{Zhang} analyze subject groups in a web-based linked view system.
+Zhang et al. analyze subject groups in a web-based linked view system \cite{Zhang}.
 %%
@@
 %%
-Dai and Gahegan \cite{Dai} defined a \emph{Concept Map}, linking cancer-associated features by incorporating choropleth maps with bar charts, parallel coordinates and scatter plots with regression lines.
+Dai and Gahegan defined a \emph{Concept Map}, linking cancer-associated features by incorporating choropleth maps with bar charts, parallel coordinates and scatter plots with regression lines \cite{Dai}.
 %%
@@
 %%
-Gla{\ss}er et al. \cite{Glasser2013} use variables derived from DCE-MRI data capturing the perfusion in the tumorous tissue.
+Gla{\ss}er et al. use variables derived from DCE-MRI data capturing the perfusion in the tumorous tissue \cite{Glasser2013}.
 %%
@@
 %%
-Niemann et al. \cite{Niemann2014} assess hepatic steatosis (fatty liver disease) risk factors using decision trees.
+Niemann et al. assess hepatic steatosis (fatty liver disease) risk factors using decision trees \cite{Niemann2014}.
 %%
@@
 Pinhero et al.
-\cite{Pinheiro} applied association rule mining to create high-confidence decision rules based on variables such as age, gender and geographic location about the mortality rate in liver cancer data.
+applied association rule mining to create high-confidence decision rules based on variables such as age, gender and geographic location about the mortality rate in liver cancer data \cite{Pinheiro}.
 %%
@@
 %%
-Sekhavat and Hoeber \cite{Sekhavat} propose an interactive visualization technique for decision trees, where all decisions are displayed in a heat map.
+Sekhavat and Hoeber propose an interactive visualization technique for decision trees, where all decisions are displayed in a heat map \cite{Sekhavat}.
 %%
@@
 %%
-In a previous work \cite{Klemm2013VMV}, we extracted a centerline representation of the lumbar spine canal from the detection model (Fig.~\ref{fig:centerline}~a).
-%%
-Using the Frenet formulas \cite{Frenet}, we calculated the following metrics from the model (Fig.~\ref{fig:centerline}~b):
+In a previous work (\cite{Klemm2013VMV}), we extracted a centerline representation of the lumbar spine canal from the detection model (Fig.~\ref{fig:centerline}~a).
+%%
+Using the Frenet formulas, we calculated the following metrics from the model (Fig.~\ref{fig:centerline}~b) \cite{Frenet}:
 %%
@@
 
-As described in a prior work \cite{Klemm2014VIS}, the suggestion of potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
+As described in a prior work (\cite{Klemm2014VIS}), the suggestion of potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
 %%

### 2014-09-03 draft.tex
Crushed todo list

@@
   \centering
-  \includegraphics[width=0.5\textwidth]{figures/centerline}
+  \includegraphics[width=0.475\textwidth]{figures/centerline}
   \caption{
@@
 	%%
-	(b) Using the Frenet formulas, we extract the weighted sum of curvature and torsion for all 92 points (white dashes).
+	(b) Using the Frenet frame, we extract the weighted sum of curvature and torsion for all 92 points (white dashes).
 	%%
@@
 It captures reliable information about spine canal curvature.
-%%
-In a previous work (\cite{Klemm2013VMV}), we extracted a centerline representation of the lumbar spine canal from the detection model (Fig.~\ref{fig:centerline}~a).
-%%
-Using the Frenet formulas, we calculated the following metrics from the model (Fig.~\ref{fig:centerline}~b) \cite{Frenet}:
+%% BLIND
+%In a previous work (\cite{Klemm2013VMV}), we extracted a centerline representation of the lumbar spine canal from the detection model (Fig.~\ref{fig:centerline}~a).
+The centerline representation of the lumbar spine canal was extracted from the detection model (Fig.~\ref{fig:centerline}~a) using the approach of \cite{Klemm2013VMV}.
+%%
+Using the Frenet frame, we calculated the following metrics from the model (Fig.~\ref{fig:centerline}~b) \cite{Frenet}:
 %%
@@
 % $\sum \frac{curvature_i}{curvature_all}$.
-	\item \emph{Curvature} is calculated as weighted sum of curvature between all adjacent points describing the centerline: $\sum \frac{curvature_i}{curvature_{all}}$.
-%%
-	\item \emph{Torsion} (deviation of a curve from its current course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum \frac{torsion_i}{torsion_{all}}$.
+	\item \emph{Normalized Curvature} is calculated as weighted sum of curvature between all adjacent points describing the centerline: $\sum_{i=1}^I \frac{curvature_i}{i}$.
+We refer to total curvature from now on as \emph{curvature}.
+%%
+	\item \emph{Normalized Torsion} (deviation of a curve from its current course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum_{i=1}^I \frac{torsion_i}{i}$.
+We refer to total torsion from now on as \emph{torsion}
 	\item \emph{Curvature angle $\alpha$} is defined by the middle point of the spine canal centerline as \emph{vertex} and the line between middle point and top/bottom point as \emph{sides}.
@@
   \centering
-  \includegraphics[width=0.5\textwidth]{figures/age-gender}
+  \includegraphics[width=0.475\textwidth]{figures/age-gender}
   \caption{
@@
   \centering
-  \includegraphics[width=0.5\textwidth]{figures/proposed-features}
+  \includegraphics[width=0.475\textwidth]{figures/proposed-features}
   \caption{
@@
 
-As described in a prior work (\cite{Klemm2014VIS}), the suggestion of potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
+% BLIND
+%As described in a prior work (\cite{Klemm2014VIS}), the suggestion of potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
+Suggesting potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology \cite{Klemm2014VIS}.
 %%

### 2014-09-03 draft.tex
minor stuff, non-image data, sources

@@
 %%
-For this purpose, large population studies, such as the \emph{Study of Health in Pomerania} (SHIP) [??] gather as much information as possible about participants to be assessed towards different diseases.
+For this purpose, large population studies, such as the \emph{Study of Health in Pomerania} (SHIP) \cite{SHIP} gather as much information as possible about participants to be assessed towards different diseases.
 %%
@@
 %%
-Since recently, these studies often comprise medical image data, e.g., tomographic image data sets, as well as general patient-specific attributes, such as gender, age and weight.
-%%
-Back pain is one of the most frequent diseases in the western civilization. 
+Since recently, these studies often comprise medical image data, e.g., tomographic image data sets, as well as general patient-specific attributes, such as \emph{gender}, \emph{age} and \emph{weight}.
+%%
+Back pain is one of the most frequent diseases in the western civilization and gains importance due to the aging society.
 %%
@@
 %%
-Furthermore, previous analyses mainly focused on a single parameter or a small subset of parameters [Quelle].
+Furthermore, previous analyses mainly focused on a single parameter or a small subset of parameters.
 %%
@@
 %%
-\com{Da in der Auflistung oefter vom IVA-Framework die Rede ist: Wollen wir hierzu noch was schreiben, z.B.
-dass das im VIS-Paper definiert wurde und was wir darunter verstehen; sowie dass diese Arbeit als Erweiterung dieses Frameworks gedacht ist?}
-
+The workflow is part of an interactive visual analysis (IVA) framework, constructed to enhance the epidemiological workflow with iterative data driven techniques to derive insight \cite{Klemm2014VIS}.
+%%
 Our contributions are:
 \begin{itemize}
-\item An IVA-framework for back pain analysis based on image-derived variables of 2,240 subjects \com{Ist diskutabel, so richtig interaktiv ist es nicht. IVA wurde auch noch nicht vorher definiert.}
+\item An IVA workflow for back pain analysis based on image-derived variables of 2,240 subjects
 %%
@@
 %%
-\item Identification of the most important variables via data mining methods (including a novel semi-quantitative evaluation concept) embedded in our IVA-framework \com{Wieder: IVA wurde noch nicht definiert}
+\item Identification of the most important variables via data mining methods (including a novel semi-quantitative evaluation concept) embedded in our IVA-framework
 %%
@@
 %%
-Closest to ours is the work of Gla{\ss}er \cite{Glasser2013} and Niemann et al. \cite{Niemann2014}.
+Closest to ours is the work of Gla{\ss}er \cite{Glasser2013} and Niemann et al. \cite{Niemann2014}, who both use decision rules for their analysis.
+%%
+Decision rules are easily readable and are well suited for classifying medical data.
+%%
+The results can be easily communicated to domain experts with little knowledge about the algorithm behind the results.
+%%
+The effort of analyzing the decision rules increases with decision rule complexity.
 %%
@@
 applied association rule mining to create high-confidence decision rules based on variables such as age, gender and geographic location about the mortality rate in liver cancer data \cite{Pinheiro}.
-%%
-Decision rules are easily readable and are well suited for classifying medical data. \com{Das frueher erwaehnen}
-%%
-The results can be easily communicated to domain experts with little knowledge about the algorithm behind the results.
-%%
-Once these rules, get too complex, analyzing these decision rules becomes complex.
 %%
@@
 The variables range from somatometric variables describing body measures to medical examinations, such as laboratory tests as well as lifestyle factors as sporting activity or nutrition.
-\com{Ein paar Details hierzu noch - Datentyp, missing data, aufeinander aufbauende parameter}
+%%
+Without the 9 image-derived parameter, the data set comprises of 134 variables:
+\begin{itemize}
+	\item 21 metric variables, mostly describing somatometric variables and markers retrieved from blood analysis
+	\item 113 categorical variables divided into
+	\begin{itemize}
+		\item 43 dichotomous variables, mostly indicating the presence of a disease
+		\item 70 variables with more than two levels, indicating pain levels, nutrition and social factors, such as marital or retirement status
+	\end{itemize}
+\end{itemize}
+%%
+Many variables comprise missing data.
+%%
+Some variables are follow-up questions, covering the treatment of a disease.
+%%
+Others are exclusive to subgroups, such as women-specific variables, e.g. menstrual status.
+%%
+Sparse variables are statistically less resilient and need to be treated with care in the analysis.
 
@@
 \subsubsection{Correlation of Image Parameters With Back Pain}
-Figure~\ref{fig:image-parameter-range} shows the distribution of all image-parameters \add{as generalized pairs plot}.
-%%
-\rem{We created a generalized pairs plot for subject groups derived by dividing them into quantile groups of body-size.}
+Figure~\ref{fig:image-parameter-range} shows the distribution of all image-parameters as generalized pairs plot.
 %%
@@
 \end{itemize}
-\com{TODO: Abbildung einer Korrelationsmatrix? Waere sehr gross!}
 %%
@@
 %%
-Too many branches also impose overfitting to the data.
-\com{Quelle Decision Tree}
+Too many branches also impose overfitting to the data \cite{DecisionTree}.
 

### 2014-09-08 draft.tex
revised results plot, now contains BMI

@@
 
-\author{\authorname{Paul Klemm\sup{1}, Sylvia Glaßer\sup{1}, Kai Lawonn\sup{1} Henry Völzke\sup{2}, Katrin Hegenscheid\sup{3} and Bernhard Preim\sup{1}}
+\author{\authorname{Paul Klemm\sup{1}, Sylvia Glaßer\sup{1}, Kai Lawonn\sup{1}, Marko Rak\sup{1} Henry Völzke\sup{2}, Katrin Hegenscheid\sup{3} and Bernhard Preim\sup{1}}
 \affiliation{\sup{1}Department of Simulation and Graphics, University of Magdeburg}
@@
 %%
+Each subject's model is mapped to a common reference frame using procrustes analysis to compensate for differences in translation, 
+rotation and scaling.
+%%
 The detection fails for several subjects due imprecise initialization, imaging artifacts or strongly deformed spines.
@@
 $2,540$ lumbar spine models were obtained and constitute the foundation of this work.
-%%
 %%
@@
 
+%\com{Marko, kannst du mal bitte schauen, ob das hier so korrekt ist, bzw. ob man das konkreter formulieren kann?}
 We have to assess the model accuracy to extract key figures from it.
@@
 The centerline representation of the lumbar spine canal was extracted from the detection model (Fig.~\ref{fig:centerline}~a) using the approach of \cite{Klemm2013VMV}.
+
 %%
@@
 %%
-\com{Feedback der Greifswalder notwendig, warum das so sein koennte.}
+\com{Warum?}
 %%
@@
 
-%TODO: Muss das wirklich noch rein?
 % \com{Manchmal kann es passieren, dass die Quantile fuer eine Variable gleich sind - z.B.
@@
 
-\com{Auswertung der Anteile der Parameter an den Ergebnissen (Torsion, Curvature, etc. - welches ist der dominanteste Parameter?)}
+%TODO: Muss das wirklich noch rein?
+%\com{Auswertung der Anteile der Parameter an den Ergebnissen (Torsion, Curvature, etc. - welches ist der dominanteste Parameter?)}
 
@@
 %%
-\noindent Pathological deformation of the lumbar spine is usually the last stage and is associated with very strong lower back pain \com{Zitat Frau Hegenscheid fragen}.
+\noindent Pathological deformation of the lumbar spine is usually the last stage and is associated with very strong lower back pain \com{Referenz?}.
 %%
@@
 
-\noindent \com{TODO: Marko hinzufuegen} SHIP is part of the Community Medicine Research net of the University of Greifswald, Germany, which is funded by the Federal Ministry of Education and Research (grant no.
+\noindent SHIP is part of the Community Medicine Research net of the University of Greifswald, Germany, which is funded by the Federal Ministry of Education and Research (grant no.
 03ZIK012), the Ministry of Cultural Affairs as well as the Social Ministry of the Federal State of Mecklenburg-West Pomerania.

### 2014-09-09 draft.tex
Removed section about Age; Added section about BMI

@@
 %%
-no group (a), subdivision into \emph{male and female subjects} (b), quartiles of \emph{age} (c) and quartiles of \emph{body size} (d).
+%no group (a), subdivision into \emph{male and female subjects} (b), quartiles of \emph{age} (c) and quartiles of \emph{body size} (d).
+no group (a), subdivision into \emph{male and female subjects} (b), quartiles of \emph{Body-Mass-Index} (\emph{BMI}) (c) and quartiles of \emph{body size} (d).
 %%
@@
 	%%
-	\item subdivision into \emph{age quantiles}, yielding the groups (21, 43.8] (43.8, 54] (54, 64] (64, 90],
+	\item subdivision into \emph{Body-Mass-Index quantiles} ($BMI = \frac{m}{l^2}$ where $m$ is the \emph{body mass} in kilogram and $l$ is the \emph{body size} in meter), yielding the groups (17, 24.7] (24.7, 27.4] (27.4, 30.5] (30.5, 48],
 	%%
@@
 \end{enumerate}
+%%
+%The \emph{BMI} is calculated using the formula $BMI = \frac{m}{l^2}$ where $m$ is the \emph{body mass} in kilogram and $l$ is the \emph{body size} in meter.
+%%
 We plotted each group twice.
@@
 %\com{Potentially skip these due to the confounding of body size}
-\subsubsection{Age Groups}
-%%
-Gender could be classified for each age group using \emph{mean curvature} and \emph{curvature angle}.
-%%
-The accuracy varies between $0.35$ (\emph{$43.8-54$ years of age}, 3 decision rules) to $0.27$ (\emph{$21-43.8$ years of age}, 6 decision rules).
-%%
-The starting age of smoking could be characterized well with an error rate between $0.25$ to $0.3$ for all \emph{age} groups except of \emph{$21.1-43.8$ years of age}.
-%%
-The result however is probably over fitted to the data due to tree sizes between $16$ and $23$.
+\subsubsection{Body-Mass-Index Groups}
+%%
+Gender could be classified for each \emph{BMI} group using \emph{mean curvature} and \emph{curvature angle}.
+%%
+The accuracy varies between $0.31$ (\emph{BMI} of \emph{$30.5-48$}, 4 decision rules) to $0.39$ (\emph{BMI} of \emph{$24.7-27.4$}, 5 decision rules).
+%%
+The starting age of smoking could be characterized well with an error rate between $0.25$ to $0.32$ for all \emph{BMI} groups except of subjects with BMI of \emph{$30.5-48$}.
+%%
+The result however is probably over fitted to the data due to tree sizes between $14$ and $16$.
 %%
@@
 
-Many variables, such as \emph{body size} can be described with an error rate of $0.35$ to $0.45$ but only using large decision trees with over 20 rules.
-%Body-size can be determined with an error rate of $0,36$ to $0,39$ for subject
-% age groups, but also imposes over fitting to the data yielding a decision tree of over 30 rules.
+Some variables, such as \emph{body size} can be described with an error rate of $0.3$ to $0.36$ but only using large decision trees with over 20 rules.
 %%
@@
 %%
-For subjects between \emph{$43.8-54$ years of age}, the classifier discriminated subjects with thyroid nodules with an error rate of $0.32$.
-
-The dummy variable analysis shows many results using a decision tree with one rule based on \emph{mean curvature} with accuracy between $0.4$ and $0.5$.
-%%
-It shows that mean curvature can be used to predict dependencies to variables, such as \emph{high blood pressure}, \emph{hypothyroidism} or \emph{nutrition}.
+Using mostly \emph{mean curvature} and \emph{curvature angle}, the leg pain level can be predicted using $14$ rules with an error rate of $0.46$ for obese subjects (\emph{BMI} higher than $30$).
+%%
+The dichotomous variable whether the subject has felt pain in the last seven days can also be predicted for this group using the same features.
+%%
+The resulting tree consists of $8$ rules and has an error of $0.35$.
+%%
+Obese subjects are prone to back and leg pain due to a more stressed lumbar spine.
+%%
+How much the spine is deformed under the heavy stress seems to directly influence the pain levels for these subjects.
+
+The dummy variable analysis shows many results using a decision tree with one rule based on \emph{mean curvature} or \emph{curvature angle} with accuracy between $0.35$ and $0.47$.
+%%
+%It shows that mean curvature can be used to predict dependencies to variables, such as \emph{high blood pressure}, \emph{hypothyroidism} or \emph{nutrition}.
+
+% \subsubsection{Age Groups}
+% %%
+% Gender could be classified for each age group using \emph{mean curvature} and \emph{curvature angle}.
+% %%
+% The accuracy varies between $0.35$ (\emph{$43.8-54$ years of age}, 3 decision rules) to $0.27$ (\emph{$21-43.8$ years of age}, 6 decision rules).
+% %%
+% The starting age of smoking could be characterized well with an error rate between $0.25$ to $0.3$ for all \emph{age} groups except of \emph{$21.1-43.8$ years of age}.
+% %%
+% The result however is probably over fitted to the data due to tree sizes between $16$ and $23$.
+% %%
+% \com{Warum?}
+% %%
+%
+% Many variables, such as \emph{body size} can be described with an error rate of $0.35$ to $0.45$ but only using large decision trees with over 20 rules.
+% %%
+% Notable is also the increasing accuracy with increasing decision tree size.
+% %%
+% For subjects between \emph{$43.8-54$ years of age}, the classifier discriminated subjects with thyroid nodules with an error rate of $0.32$.
+%
+% The dummy variable analysis shows many results using a decision tree with one rule based on \emph{mean curvature} with accuracy between $0.4$ and $0.5$.
+% %%
+% It shows that mean curvature can be used to predict dependencies to variables, such as \emph{high blood pressure}, \emph{hypothyroidism} or \emph{nutrition}.
 
@@
 %%
-Gender was characterized well in the analysis of all subjects as well as subjects divided by \emph{age}.
+Gender was characterized well in the analysis of all subjects as well as subjects divided by \emph{BMI}.
 

### 2014-09-09 draft.tex
Revised part to new IVA tool

@@
 %%
-Dividing the subjects into four groups using body size, the number increases to 402 trees.
+Further subdivision, e.g. by quantiles of \emph{body size} increases the number to 402 trees.
+%Dividing the subjects into four groups using body size, the number increases to 402 trees.
 %%
@@
 %%
-We removed variables with less than 100 subjects, since they have too few subjects to be statistically significant.
+%We removed variables with less than 100 subjects, since they have too few subjects to be statistically significant.
+Therefore we provide a way to adjust the minimal number of subjects for each variable using a slider input.
+%%
+The initial value is empirically set to $100$, marking a good tradeoff between sparse variables and statistical informative value.
 %%
@@
 This allows for sequentially analyzing the classifications.
+
+We also provide controls for adjusting the maximum classification error and minimum subject count for a variable.
+%%
+This gives the user control to abstract or refine the displayed information.
+%%
+Selecting a variable using a drop down menu allows to user to select the variable used for subject subdivisions, e.g. \emph{gender} or \emph{disease treatment type}.
+%%
+Categorization using metric variables, such as \emph{body size}, is done using their quantiles.
+%%
+This allows to assess the influences of a variable to the classification process.
 %%------------
@@
 %%
-The interactive visualizations are realized using the \texttt{ggvis}\footnote{GGVIS Footnote} package.
+The interactive visualizations are realized using the \texttt{ggvis}\footnote{Developed by RStudio, Inc; \texttt{ggvis.rstudio.com}} package.
 %%
@@
 %%
-We used \texttt{R Shiny}\footnote{R Shiny footnote} to make the report available as dynamic web application.
+We used \texttt{R Shiny}\footnote{Developed by RStudio, Inc; \texttt{shiny.rstudio.com}} to make the report available as dynamic web application.
 %%

### 2014-09-10 draft.tex
Added Web Reference to supplementary material

@@
 \usepackage[utf8]{inputenc}
+\usepackage{url}
 \newcommand{\rem}[1]{\textcolor{red}{\sout{#1}}}
@@
 %%
-The resulting contingency matrices are available as supplementary material and show no strong correlation with any of the parameter.
+The resulting contingency matrices are available as supplementary material and show no strong correlation with any of the parameter (see \url{blind.dnsalias.com}).
 %%
@@
 Exchanging the prototype becomes as easy as exchanging a hyperlink.
+%%
+The prototype is available under the following link: \url{blind.dnsalias.com}
 %%---------
@@
 
-\noindent If any, the appendix should appear directly after the
-references without numbering, and not on a new page. To do so please use the following command:
-\textit{$\backslash$section*\{APPENDIX\}}
+\noindent The IVA tool and all shown plots presented in this paper are available as supplementary material under the following link: \url{blind.dnsalias.com}.
 

### 2014-09-11 draft.tex
Added Reference to Supplementary Material at the introduction

@@
 \end{itemize}
+%%
+The analysis, as well as the presented IVA tool is available as supplementary material in a interactive \texttt{R-Markdown}\footnote{Developed by RStudio, Inc; \texttt{rmarkdown.rstudio.com}} document (\url{blind.dnsalias.com}).
+%%
+%The IVA tool and all shown plots presented in this paper, except of Fig.~\ref{fig:age-gender}, are available as supplementary material under the following link: \url{blind.dnsalias.com}.
 %% ---------------------------------------------
@@
 \emph{Torsion} also adds little variance to the information space.
+%%
+The complete analysis can be found under the experiments section of the supplementary material (\url{blind.dnsalias.com})
 %% ------------------------------------------
@@
 %%
-We applied a heterogenous correlation technique to derive correlations between all variables is the data set.
+We applied a heterogenous correlation technique to derive correlations between all variables in the data set.
 %%
@@
 
-\section*{\uppercase{Appendix}}
-
-\noindent The IVA tool and all shown plots presented in this paper are available as supplementary material under the following link: \url{blind.dnsalias.com}.
+%\section*{\uppercase{Appendix}}
+%\noindent The IVA tool and all shown plots presented in this paper are available as supplementary material under the following link: \url{blind.dnsalias.com}.
 

### 2014-09-12 draft.tex
Revised Version of Bernhard

@@
 %%
-Epidemiological research is strongly hypothesis-driven.
+So far, epidemiological research is strongly hypothesis-driven.
 %%
@@
 %For epidemiological research, cohort studies are acquired.
-For epidemiological research, cohort studies are carried out.
+Cohort studies are an instrument of epidemiological research.
+%For epidemiological research, cohort studies are carried out.
 %%
@@
 %%
-Individual diagnosis or manual segmentation of each body structure by an radiologists is tedious, costly and comprises little reproducibility.
+Individual diagnosis or manual segmentation of each body structure by radiologists is tedious, costly and comprises little reproducibility.
 %%
@@
 %%
+A survey on image-centric cohort studies and strategies to analyze the resulting data is given in \cite{Preim2014}.
+%%
 \paragraph{Visual Analysis of Heterogenous Non-Image Data.}
@@
 %%
-Closest to ours is the work of Gla{\ss}er \cite{Glasser2013} and Niemann et al. \cite{Niemann2014}, who both use decision rules for their analysis.
+Closest to ours is the work of Gla{\ss}er et al. and Niemann et al., who both use decision rules for their analysis \cite{Glasser2013,Niemann2014}.
 %%
@@
 %%
-The data extraction process is based on the detection algorithm and lumbar spine canal extraction presented in this paper.
+Our data extraction process is based on the detection algorithm and lumbar spine canal extraction presented by \cite{Klemm2013VMV}.
 %%
@@
 \\\\
-Unique for our work is the combination of an IVA-approach by observing interesting feature relations in context of image-derived features using multiple decision trees for one visualization.
+Unique for our work is the combination of data mining techniques with an IVA-approach by observing interesting feature relations in context of image-derived features using multiple decision trees for one visualization.
 %%
@@
 %%
-Each subject's model is mapped to a common reference frame using procrustes analysis to compensate for differences in translation, 
+Each subjects' model is mapped to a common reference frame using procrustes analysis to compensate for differences in translation, 
 rotation and scaling.
 %%
-The detection fails for several subjects due imprecise initialization, imaging artifacts or strongly deformed spines.
+The detection fails for several subjects due to imprecise initialization, imaging artifacts or strongly deformed spines.
 %%
@@
 	\item \emph{Normalized Torsion} (deviation of a curve from its current course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum_{i=1}^I \frac{torsion_i}{i}$.
-We refer to total torsion from now on as \emph{torsion}
+We refer to total torsion from now on as \emph{torsion}.
+%%
 	\item \emph{Curvature angle $\alpha$} is defined by the middle point of the spine canal centerline as \emph{vertex} and the line between middle point and top/bottom point as \emph{sides}.
@@
 %%
-Readability is only granted for small trees, complex structures with many branches are not desirable.
+Readability is only granted for small trees. %, complex structures with many branches are not desirable.
 %%
@@
 \paragraph{Attribute Mapping}
+%%
 The scatter plot axis are defined by tree size and the previously described error metric.
@@
 %%
-If we for example classify and compare the same variables for different subject groups, for example male and female subjects.
-%%
-We can support this by color coding the data points according to group affiliation.
+%If we for example classify and compare the same variables for different subject groups, for example male and female subjects.
+Classification and comparison of variables for subject groups (e.g. male and female subjects) in one plot can be achieved by color coding group affiliation on the data points.
+%Classification of the same variable 
+%%
+%We can support this by color coding the data points according to group affiliation.
 
@@
 %%
-Each binary variables encodes the presence of a manifestation.
+Each binary variable encodes the presence of a manifestation.
 %%
@@
 %%
-Our method to plot derived metrics and custom tailored error measures proved to be effective.
+Our method to plot derived metrics and custom-tailored error measures proved to be effective.
 %%

### 2014-09-12 draft.tex
Revised Bernhard done

@@
 Pinhero et al.
-applied association rule mining to create high-confidence decision rules based on variables such as age, gender and geographic location about the mortality rate in liver cancer data \cite{Pinheiro}.
+applied association rule mining to create high-confidence decision rules based on variables, such as age, gender and geographic location about the mortality rate in liver cancer data \cite{Pinheiro}.
 %%
@@
 %%
-Therefore, we are interested in \emph{small trees} with a \emph{low classification error rate}.
+Therefore, we are interested in \emph{small trees} with a \emph{low classification error}.
 %%
@@
 \paragraph{The Error Term. }
-Normally the error rate for a classification is calculated with $error = \frac{classifiedCorrectly_{n}} {n}$ where $n$ is the number of subjects.
+Normally the error for a classification is calculated with $error = \frac{classifiedCorrectly_{n}} {n}$ where $n$ is the number of subjects.
 %%
@@
 %%
-The x-axis shows the number of decisions of the underlying model, the y-axis the classification error rate (see Section~\ref{subsec:VisualizationOfClassificationResults}).
+The x-axis shows the number of decisions of the underlying model, the y-axis the classification error (see Section~\ref{subsec:VisualizationOfClassificationResults}).
 %%
@@
 %%
-We only display variables with an error rate below $0.5$, results above this threshold are inaccurate.
+We only display variables with an error below $0.5$, results above this threshold are inaccurate.
 %%
@@
 %%
-This is reflected in the large amount of variables classified with an error rate above $0.6$.
+This is reflected in the large amount of variables classified with an error above $0.6$.
 %%
@@
 %
-It can be discriminated with an error rate of $0.31$ using 7 rules and incorporates only curvature and curvature angle variables.
+It can be discriminated with an error of $0.31$ using 7 rules and incorporates only curvature and curvature angle variables.
 %%
@@
 %%
-\emph{Medication for high blood pressure} is classified for 1,058 subjects with an error rate of $0.47$ solely based on coronal mean curvature.
+\emph{Medication for high blood pressure} is classified for 1,058 subjects with an error of $0.47$ solely based on coronal mean curvature.
 %%
@@
 %%
-The four body size groups could be characterized with an error rate $0.48$, but the decision tree comprises 71 rules and imposes overfitting.
+The four body size groups could be characterized with an error $0.48$, but the decision tree comprises 71 rules and imposes overfitting.
 
@@
 %%
-The classifier neglects the other group and yields a error rate below $0.5$.
+The classifier neglects the other group and yields a error below $0.5$.
 %The dummy variable analysis yields fewer results.
@@
 %%
-Only \emph{hypothyroidism} could be described for male subjects with an error rate of $0.24$ for 110 subjects using the \emph{mean curvature} and \emph{curvature angle}.
+Only \emph{hypothyroidism} could be described for male subjects with an error of $0.24$ for 110 subjects using the \emph{mean curvature} and \emph{curvature angle}.
 %%
@@
 %%
-The accuracy varies between $0.31$ (\emph{BMI} of \emph{$30.5-48$}, 4 decision rules) to $0.39$ (\emph{BMI} of \emph{$24.7-27.4$}, 5 decision rules).
-%%
-The starting age of smoking could be characterized well with an error rate between $0.25$ to $0.32$ for all \emph{BMI} groups except of subjects with BMI of \emph{$30.5-48$}.
+The error varies between $0.31$ (\emph{BMI} of \emph{$30.5-48$}, 4 decision rules) to $0.39$ (\emph{BMI} of \emph{$24.7-27.4$}, 5 decision rules).
+%%
+The starting age of smoking could be characterized well with an error between $0.25$ to $0.32$ for all \emph{BMI} groups except of subjects with BMI of \emph{$30.5-48$}.
 %%
@@
 
-Some variables, such as \emph{body size} can be described with an error rate of $0.3$ to $0.36$ but only using large decision trees with over 20 rules.
-%%
-Notable is also the increasing accuracy with increasing decision tree size.
-%%
-Using mostly \emph{mean curvature} and \emph{curvature angle}, the leg pain level can be predicted using $14$ rules with an error rate of $0.46$ for obese subjects (\emph{BMI} higher than $30$).
+Some variables, such as \emph{body size} can be described with an error of $0.3$ to $0.36$ but only using large decision trees with over 20 rules.
+%%
+Notable is also the decreasing error with increasing decision tree size.
+%%
+Using mostly \emph{mean curvature} and \emph{curvature angle}, the leg pain level can be predicted using $14$ rules with an error of $0.46$ for obese subjects (\emph{BMI} higher than $30$).
 %%
@@
 
-The dummy variable analysis shows many results using a decision tree with one rule based on \emph{mean curvature} or \emph{curvature angle} with accuracy between $0.35$ and $0.47$.
+The dummy variable analysis shows many results using a decision tree with one rule based on \emph{mean curvature} or \emph{curvature angle} with an error between $0.35$ and $0.47$.
 %%
@@
 %%
-Radiating back pain could be described with error rate of $0.39$ using 23 rules for subjects between \emph{$171-177~cm$ body size} using \emph{torsion} and \emph{mean curvature}.
-%%
-For subjects sized \emph{$177-202~cm$} the accuracy drops to $0.47$ using 20 decision rules.
-%%
-There are several decision trees for laboratory values, such as \emph{alanine aminotransferase} value (relevant for diagnosis of liver or gallbladder illness) in the blood can be described with an error rate of $0.4$ (\emph{$139-164~cm$}) to $0.36$ (\emph{$164 - 171~cm$}).
+Radiating back pain could be described with error of $0.39$ using 23 rules for subjects between \emph{$171-177~cm$ body size} using \emph{torsion} and \emph{mean curvature}.
+%%
+For subjects sized \emph{$177-202~cm$} the error increases to $0.47$ using 20 decision rules.
+%%
+There are several decision trees for laboratory values, such as \emph{alanine aminotransferase} value (relevant for diagnosis of liver or gallbladder illness) in the blood can be described with an error of $0.4$ (\emph{$139-164~cm$}) to $0.36$ (\emph{$164 - 171~cm$}).
 %%
@@
 %%
-The dummy variables show several variables described using only one decision rule with accuracy between $0.42$ to $0.47$.
+The dummy variables show several variables described using only one decision rule with an error between $0.42$ to $0.47$.
 %%
@@
 %%
-%Certain parameters can be classified in 
-For some \emph{body size} groups, certain \emph{nutrition} or \emph{psychological} parameters can be predicted with a comparatively high error rate.
-%Certain subgroups, however, can be classified for various size groups, such as nutrition or psychological parameters.
+For some \emph{body size} groups, certain \emph{nutrition} or \emph{psychological} parameters can be predicted with a comparatively high error.
 %%
@@
 \section*{\uppercase{Acknowledgements}}
-
-\noindent SHIP is part of the Community Medicine Research net of the University of Greifswald, Germany, which is funded by the Federal Ministry of Education and Research (grant no.
+\noindent Omitted due to blind review.
+%\noindent SHIP is part of the Community Medicine Research net of the University of Greifswald, Germany, which is funded by the Federal Ministry of Education and Research (grant no.
 03ZIK012), the Ministry of Cultural Affairs as well as the Social Ministry of the Federal State of Mecklenburg-West Pomerania.

### 2014-09-12 draft.tex
Minor changes to title

@@
 
-\title{Insight Into Lumbar Back Pain  \subtitle{What the Lumbar Spine Tells About Your Life} }
+%\title{Insight Into Lumbar Back Pain  \subtitle{What the Lumbar Spine Tells About Your Life} }
+\title{Interactive Visual Analysis of Lumbar Back Pain  \subtitle{What the Lumbar Spine Tells About Your Life} }
 
 \author{\authorname{Paul Klemm\sup{1}, Sylvia Glaßer\sup{1}, Kai Lawonn\sup{1}, Marko Rak\sup{1} Henry Völzke\sup{2}, Katrin Hegenscheid\sup{3} and Bernhard Preim\sup{1}}
-\affiliation{\sup{1}Department of Simulation and Graphics, University of Magdeburg}
-\affiliation{\sup{2} Institut for Community Medicine, University of Greifswald}
-\affiliation{\sup{3} Institute for Diagnostic Radiology and Neuro-Radiology, University of Greifswald}
-\email{\{paul, sylvia\}@isg.cs.uni-magdeburg.de, \{katrin.hegenscheid, voelzke\}@uni-greifswald.de}
+\affiliation{\sup{1}Department of Simulation and Graphics, University of Magdeburg, Germany}
+\affiliation{\sup{2} Institute for Community Medicine, University of Greifswald, Germany}
+\affiliation{\sup{3} Institute for Diagnostic Radiology and Neuro-Radiology, University of Greifswald, Germany}
+\email{\{klemm, glasser\}@isg.cs.uni-magdeburg.de}
 }
 
-\keywords{Epidemiology, Interactive Visual Analysis}
+\keywords{Epidemiology, Interactive Visual Analysis, Classification, Multi Modal Data}
 
@@
 %%
-The lower (\emph{lumbar}) spine is the most stressed spine part.
-%%
-Lower back pain is one of the most frequent diseases in the western civilization.
+%The lower (\emph{lumbar}) spine is the most stressed spine part .
+%%
+Lower back pain is one of the most frequent diseases in the western civilization \cite{Hoy2010}.
 %%
@@
 %%
-\com{Warum?}
-%%
 
@@
 %%
-How much the spine is deformed under the heavy stress seems to directly influence the pain levels for these subjects.
+The stress induced spine deformation seems to influence the pain levels for these subjects directly.
+%How much the spine is deformed under the heavy stress seems to directly influence the pain levels for these subjects.
 
@@
 
-Subjects grouped using \underline{body size} have a similar spine shape and a lower variance.
+Subjects grouped using \emph{body size} have a similar spine shape and a lower variance.
 %%
@@
 %%
-\noindent Pathological deformation of the lumbar spine is usually the last stage and is associated with very strong lower back pain \com{Referenz?}.
+\noindent Pathological deformation of the lumbar spine is usually the last stage and is associated with very strong lower back pain.
 %%

### 2014-09-16 draft.tex
First round of Sylvias revisions

@@
 %%
-Hence, subject groups (cohorts) are analyzed to correlate the subjects' varying lifestyles and medical properties with diseases.
-%%
-Recently, these cohort studies comprise medical image data as well.
+Hence, subject groups (cohorts) are analyzed to correlate the subjects' varying lifestyles, their medical properties and diseases.
+%%
+Recently, these cohort studies comprise medical image data.
 %%
@@
 %%
-The evaluation is embedded in an interactive visual analysis (IVA) framework that is based on 2,540 segmented lumbar spine data sets.
+Therefore, an interactive visual analysis (IVA) framework was created and tested with 2,540 segmented lumbar spine data sets.
 %%
@@
 %%
-For this purpose, large population studies, such as the \emph{Study of Health in Pomerania} (SHIP) \cite{SHIP} gather as much information as possible about participants to be assessed towards different diseases.
-%%
-These information are used to determine risk factors for diseases, help people to improve their lifestyle and their health or to support diagnosis of widespread diseases.
+For this purpose, large population studies, such as the \emph{Study of Health in Pomerania} (SHIP) \cite{SHIP}, gather as much information as possible about participants to be assessed towards different diseases.
+%%
+These information are used to determine risk factors for diseases, help people to improve their lifestyle and their health or to support the diagnosis of widespread diseases.
 %%
@@
 %%
-Since recently, these studies often comprise medical image data, e.g., tomographic image data sets, as well as general patient-specific attributes, such as \emph{gender}, \emph{age} and \emph{weight}.
-%%
-Back pain is one of the most frequent diseases in the western civilization and gains importance due to the aging society.
+Nowadays, these studies often comprise medical image data, e.g., tomographic image data sets, as well as general patient-specific attributes, such as \emph{gender}, \emph{age} and \emph{weight}.
+%%
+Back pain is one of the most frequent diseases in the western civilization and gains importance due to the population aging.
 %%
@@
 %%
-Although the shape and constitution of the spine, especially the lumbar spine, plays an important role for the presence of back pain, no automatic classification approach to characterize back pain based on lumbar back spine attributes has been established.
-%%
-Furthermore, previous analyses mainly focused on a single parameter or a small subset of parameters.
-%%
-
-In this paper, we present a thorough analysis of the image-derived data from lumbar back spine datasets and analyze their discriminating power for different characteristics of the disease.
+Although the shape and constitution of the spine, especially the lumbar spine, plays an important role for the presence of back pain, an automatic classification approach for characterization of back pain based on lumbar back spine attributes is still missing.
+%%
+Furthermore, previous analyses mainly focused on a single parameter (e.g. \emph{BMI}) or a small subset of parameters.
+%%
+
+In this paper, we present a thorough analysis of the image-derived data of a cohort study including lumbar back spine datasets.
+%%
+We analyze their discriminating power for different characteristics of back pain.
 %%
@@
 %%
-The workflow is part of an interactive visual analysis (IVA) framework, constructed to enhance the epidemiological workflow with iterative data driven techniques to derive insight \cite{Klemm2014VIS}.
+The workflow is part of an interactive visual analysis (IVA) framework, constructed to enhance the epidemiological workflow with iterative data driven techniques.
 %%
@@
 \begin{itemize}
-\item An IVA workflow for back pain analysis based on image-derived variables of 2,240 subjects
-%%
-\item The identification and evaluation of lumbar spine shape properties for back pain diagnosis
-%%
-\item Detection of correlations between image-based and socio-demographic as well as medical parameters for an improved hypothesis generation
-%%
-\item Identification of the most important variables via data mining methods (including a novel semi-quantitative evaluation concept) embedded in our IVA-framework
+\item An IVA workflow for back pain analysis based on image-derived variables of 2,240 subjects,
+%%
+\item The identification and evaluation of lumbar spine shape properties for back pain diagnosis,
+%%
+\item The detection of correlations between image-based and socio-demographic as well as medical parameters for an improved hypothesis generation,
+%%
+\item The identification of the most important variables via data mining methods (including a novel semi-quantitative evaluation concept) embedded in our IVA-framework
 %%
@@
 %%
-\noindent Epidemiological reasoning relies on a strict statistically driven workflow \cite{Fletcher}:
+\noindent Epidemiological reasoning relies on a strict statistically-driven workflow \cite{Fletcher}:
 %%
@@
 %%
-	\item To assess a hypothesis, epidemiologists compile a list of variables depicting it.
+	\item To prove a hypothesis, epidemiologists compile a list of variables depicting it.
 %%
@@
 %%
-When for example analyzing influences of nutrition to prostate cancer, the results need to be age-normalized.
+For example, the analysis of the influence of nutrition on prostate cancer is strongly age dependent and requires therefore age-normalization.
+%%
+%When for example analyzing influences of nutrition to prostate cancer, the results need to be age-normalized.
 %%
@@
 %%------------------------------
-\subsection{Epidemiological Data}
+%\subsection{Epidemiological Data}
+\subsection{Challenges of Epidemiological Data Analysis}
 Epidemiological data originates from a wide range of studies.
@@
 %%
-Most common are case-control studies, analyzing one specific disease and its influences on the human body.
+Most common are case-control studies analyzing one specific disease and its influence on the human body.
 %%
@@
 %%
-These studies aim to collect as much data as possible for each subject.
+These studies aim at collecting as much data as possible for each subject.
 %%
@@
 
-Epidemiological data is heterogenous and incomplete.
-%%
-For example women-specific questions or data about a disease treatment only subjects suffering from this condition.
-%%
-Statistical analysis has to take missing data into account.
-%%
-
-Epidemiologists acquire data using a wide range of techniques, such as medical examinations, self reported questionnaires or genetic examinations.
+Epidemiological data are heterogenous and incomplete.
+%%
+For example, women-specific questions or data about a disease treatment only affects subjects suffering from this condition.
+%%
+Therefore, statistical analysis has to take missing data into account.
+%%
+
+Epidemiological data acquisition includes a variety of techniques, such as medical examinations, self reported questionnaires or genetic examinations.
+%Epidemiologists acquire data using a wide range of techniques, such as medical examinations, self reported questionnaires or genetic examinations.
 %%
@@
 %%
-To compare these data, information reduction techniques are applied.
+To compare these data, information reduction techniques are necessary.
 %%
@@
 %%
-Every information reduction can introduce a bias to the data, since it reflects an assumption about the data.
+Every information reduction can introduce a bias to the data, since it proposes an assumption about the data.
 %%
@@
 %%
-Segmentation algorithms are not generally available and need to be custom-made for each body structure.
+Segmentation algorithms are not generally available and need to be carefully adapted for each body structure.
 %%
@@
 %%
-\paragraph{Visual Analysis of Image and non-Image Data}
-%%
-Klemm et al. propose an extension to the previously described epidemiological workflow using \emph{Interactive Visual Analysis} (IVA) \cite{Klemm2014VIS}.
+\paragraph{Visual Analysis of Medical Image and Non-Image Data}
+%%
+%Klemm et al. propose an extension to the previously described epidemiological workflow using \emph{Interactive Visual Analysis} (IVA) \cite{Klemm2014VIS}.
+\cite{Klemm2014VIS} propose an extension to the previously described epidemiological workflow using \emph{Interactive Visual Analysis} (IVA).
 %%
@@
 %%
-We incorporate the IVA approach by concurrently displaying complex relationships between variables using decision trees.
-%%
-Steenwijk et al. propose a coordinated linked view system for both image-related and non-image data \cite{Steenwijk}.
+%% TODO BLIND "We incorporate their" -> We incorporate the ...
+We incorporate their IVA approach by concurrently displaying complex relationships between variables via decision trees.
+%%
+\cite{Steenwijk} propose a coordinated linked view system for both image-related and non-image data.
 %%
@@
 %%
-Turkay et al. follow a similar approach by deriving descriptive data metrics from image data \cite{Turkay}.
+\cite{Turkay} follow a similar approach by deriving descriptive data metrics from image data.
 %%
@@
 %%
-The paper lacks details about handling of missing data, the definition of similarity or which statistical measures are applied.
+The described method lacks details about handling of missing data, the definition of similarity or the choice of the statistical measures.
 %%
@@
 %%
-The plot matrix is well suited for an overview visualization, but takes up much screen space and is therefore only suitable for few variables at once (see Fig.~\ref{fig:image-parameter-range}).
-%%
-Dai and Gahegan defined a \emph{Concept Map}, linking cancer-associated features by incorporating choropleth maps with bar charts, parallel coordinates and scatter plots with regression lines \cite{Dai}.
+\texttt{GPLOM'S} are well suited for an overview visualization, but take up much screen space and are therefore only suitable for few variables at once (see example in Fig.~\ref{fig:image-parameter-range} later on).
+%%
+\cite{Dai} define a \emph{Concept Map}, linking cancer-associated features by incorporating choropleth maps with bar charts, parallel coordinates and scatter plots with regression lines.
 %%
@@
 %%----------------------------
-\paragraph{Decision Rule Driven Analysis of Medical Data.}
+\paragraph{Decision Rule-Driven Analysis of Medical Data.}
 Related work in this field often focuses on clinical diagnosis.
 %%
-Closest to ours is the work of Gla{\ss}er et al. and Niemann et al., who both use decision rules for their analysis \cite{Glasser2013,Niemann2014}.
-%%
-Decision rules are easily readable and are well suited for classifying medical data.
-%%
-The results can be easily communicated to domain experts with little knowledge about the algorithm behind the results.
-%%
-The effort of analyzing the decision rules increases with decision rule complexity.
+Closest to ours is the work of \cite{Glasser2013} and \cite{Niemann2014}.
+%%
+Both use decision trees for their analyses.
+%%
+Decision trees are easily readable and are well suited for classifying medical data.
+%%
+%The results can be easily communicated to domain experts with little knowledge about the algorithm behind the results.
+%%
+The effort of analyzing the decision trees increases with decision tree complexity.
 %%
@@
 %%
-They aim to classify breast tumors.
-%%
-The extracted variables train a decision-tree classifier, concluding that the extracted features alone are not sufficient for tumor type classification.
+To classify breast tumors, they train a decision-tree classifier, concluding that the extracted kinetic and morphological features alone are not sufficient for tumor type classification.
 %%
@@
 %%
-They presented an interactive data mining tool, which can analyze association rules and highlight interesting relations, which may trigger new hypothesis about the data.
-%%
-We combine ideas of both papers by validating the significance of image-derived parameter aiming to derive new relationships by training decision trees to explain target variables.
+They present an interactive data mining tool, which can analyze association rules and highlight interesting relations, which may trigger new hypothesis about the data.
+%%
+We combine both ideas by validating the significance of image-derived parameters.
+We aim at deriving new relationships by training decision trees to explain target variables.
 %%----------------------------
@@
 %%
-The visualization translates the decision tree into a heat map, where each cell represents the connectivity between an initial value and an target value.
+The visualization translates the decision tree into a heat map, where each cell represents the connectivity between an initial value and a target value.
 %%
@@
 \paragraph{SHIP Data Analysis.}
-Klemm et al. analyzed the lumbar spine variability of of 490 \texttt{SHIP-2} subjects \cite{Klemm2013VMV}.
-%%
-They incorporate hierarchical agglomerative clustering to derive shape groups, yielding several groups of average shape and several outlier cluster.
+Klemm et al. analyzed the lumbar spine variability of 490 \texttt{SHIP-2} subjects \cite{Klemm2013VMV}.
+%%
+They incorporate hierarchical agglomerative clustering to derive shape groups, yielding several groups of average shape and several outlier clusters.
 %%
@@
 \\\\
-Unique for our work is the combination of data mining techniques with an IVA-approach by observing interesting feature relations in context of image-derived features using multiple decision trees for one visualization.
+Unique for our work is the combination of data mining techniques with an IVA-approach by observing interesting feature relations in the context of image-derived features using multiple decision trees for one visualization.
 %%
@@
 %%
-\noindent Our approach allows to analyze many variables simultaneously.
-%%
-Therefore, epidemiologists compiled the data set with a with a wide range of variables possibly correlating with lumbar back pain.
-%%
-The data set comprises of 6.753 subjects from two cohorts (4.420 from \texttt{SHIP-Trend-0} and 2.333 from \texttt{SHIP-2}).
+\noindent Our approach allows the simultaneous analysis of many variables.
+%%
+Therefore, epidemiologists compiled the data set with a wide range of variables possibly correlating with lumbar back pain.
+%%
+The data set comprises 6,753 subjects from two cohorts (4,420 from \texttt{SHIP-Trend-0} and 2,333 from \texttt{SHIP-2}) including non-image and image-derived data.
 %%
@@
 %\com{Analysis Using Heterogeneous Correlation}
-The variables range from somatometric variables describing body measures to medical examinations, such as laboratory tests as well as lifestyle factors as sporting activity or nutrition.
-%%
-Without the 9 image-derived parameter, the data set comprises of 134 variables:
+The non-image variables range from somatometric variables describing body measures to medical examinations, such as laboratory tests as well as lifestyle factors, e.g. sporting activity or nutrition.
+%%
+%Without the 9 image-derived parameter, the data set comprises of 134 variables:
+The data set comprises of 134 variables (and 9 additional image-derived variables):
 \begin{itemize}
@@
 	\begin{itemize}
-		\item 43 dichotomous variables, mostly indicating the presence of a disease
+		\item 43 dichotomous (binary) variables, mostly indicating the presence of a disease, e.g. pancreatitis or high blood pressure
 		\item 70 variables with more than two levels, indicating pain levels, nutrition and social factors, such as marital or retirement status
@@
 %%
-Sparse variables are statistically less resilient and need to be treated with care in the analysis.
-
-\subsection{Image Data}
+Sparse variables are statistically less resilient and need to be treated with care during the analysis.
+
+\subsection{Image-Derived Data}
 \begin{figure}[!h]
@@
 	%%
-	(a) right: FEM aligned to a MRI data set.
+	(a) right: FEM aligned to an MRI data set.
 	%%
@@
 \end{figure}
-Four trained radiology technicians acquired the medical image data for each subject in a standardized way using a 1.5~Tesla MRI scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany).
-%%
-The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
+Four trained radiologists acquired the medical image data for each subject in a standardized way using a 1.5~Tesla Magnetic Resonance Imaging (MRI) scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany).
+%%
+The spine protocol consisted of a sagittal $T_1$-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) and a sagittal $T_2$-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
 %%
@@
 %%
-The tetrahedron-based finite element model is initialized using three user-placed landmarks.
-%%
-A click on the L3 vertebra center initialized the model, two clicks on the top and bottom of the vertebra describes an initial height estimation.
-%%
-The registered model captures information about the lumbar spine canal shape and the position of the L1-L5 vertebrae \cite{Klemm2013VMV} (Fig.~\ref{fig:centerline}~a).
+The tetrahedron-based finite element model is initialized using three user-defined landmarks.
+%%
+A click on the L3 vertebra center initialized the model, two clicks on the top and bottom of the vertebra describes an initial height estimation (Fig.~\ref{fig:centerline}~a).
+%%
+The registered model captures information about the lumbar spine canal shape and the position of the L1-L5 vertebrae.
 %%
@@
 %%
-$2,540$ lumbar spine models were obtained and constitute the foundation of this work.
+$2,540$ out of $6,753$ lumbar spine models were obtained and constitute the foundation of this work.
 %%
@@
 	\item \emph{Normalized Curvature} is calculated as weighted sum of curvature between all adjacent points describing the centerline: $\sum_{i=1}^I \frac{curvature_i}{i}$.
-We refer to total curvature from now on as \emph{curvature}.
+We refer to total curvature as \emph{curvature}.
 %%
@@
 %%
-These figures are also extracted in the sagittal, coronal and transversal projection of the model.
+These metrics are also extracted in the sagittal, coronal and transversal projection of the model.
 %%
@@
 %%
-This gives us a total of 9 image-derived parameters.
+This yields a total of 9 image-derived parameters.
+%%
+In the next section, we present the experiments we conducted to assess the influence of the lumbar spine shape to lower back pain.
 
@@
 	%%
-	Due to bone erosion, older people are also smaller on average.
+	%Due to bone erosion, older people are also smaller on average.
 	}
@@
 %%
-All these variables need to be taken into account, when correlating spine curvature and torsion with non-image parameter.
+When correlating spine curvature and torsion with non-image parameter, all these variables need to be taken into account
 %%
@@
 %%
-To avoid small outlier groups, epidemiologists recommend using quartiles to discretize metric variables \cite{Klemm2014VIS}.
+To avoid small outlier groups, epidemiologists recommend using quartiles to discretize metric variables.
 \subsection{Preliminary Results}
@@
   \caption{%%
-A generalized pairs plot of all image-derived variables colored by presence or absence of back pain.
-%%
-Pairwise combinations of image-derived variables are visualized using scatter plots left of the matrix diagonal and their correlation with back pain is denoted right of the matrix diagonal.
-%%
-The box plots (left) and histograms (bottom) display the distribution of each image-variable encoded with back pain.
-%%
-No correlation with back pain are present can be seen in this plot.
+A generalized pairs plot of all image-derived variables colored by presence (red) or absence (turquoise) of back pain.
+%%
+Pairwise combinations of image-derived variables are visualized via scatter plots left of the matrix diagonal and their correlation with back pain is denoted right of the matrix diagonal.
+%%
+The box plots (right) and histograms (bottom) display the distribution of each image-variable encoded with back pain.
+%%
+No correlations with back pain are present as it can be seen in this plot.
 }
@@
 %%
-\noindent As first experiment we correlated the shape parameter with the binary back pain indicator using \texttt{GPLOM's} \cite{GPLOMS}.
+\noindent As a first experiment we correlated the shape parameter with the binary back pain indicator using \texttt{GPLOM's} \cite{GPLOMS}.
 %%
@@
 %%
-The pearson correlation coefficients between the numeric variables are depicted right of the matrix diagonal in Figure~\ref{fig:image-parameter-range}.
-%%
-\emph{Curvature}, \emph{curvature angle} and \emph{torsion} correlate strongly with their planar projections.
+The pearson correlation coefficients between the numeric variables are depicted right of the matrix diagonal (Fig.~\ref{fig:image-parameter-range}).
+%%
+\emph{Curvature}, \emph{curvature angle} and \emph{torsion} strongly correlate with their planar projections.
 %%
@@
 
-\subsubsection{Assessing the Information Gain Using the PCA}
-%%
-To determine the information gain per image-derived variables, we calculated a PCA and compared the loadings per dimension.
+\subsubsection{Assessing the Information Gain Using the Principal Component Analysis}
+%%
+To determine the information gain per image-derived variables, we calculated a Principal Component Analysis (PCA) and compared the loadings per dimension.
+%%
+The PCA results in a linear projection of the parameter space ordering their components by maximum variance.
 %%
@@
 %%
-The method uses the following correlation metrics for the different type combination:
+The method uses the following correlation metrics for the different type combinations:
 %%
@@
 %%
-All correlation values are scaled between \texttt{0 - no correlation} and \texttt{1 - identical}.
+All correlation values are scaled between \texttt{0 - no correlation} and \texttt{1 - perfect correlation}.
 %%
@@
 %%
-We display the resulting \emph{contingency matrix} using a heat map, mapping correlation values to color brightness with white for 0 and dark blue for 1 \cite{Klemm2014VIS}.
+These variables have to be omitted in the analysis, since they are not statistically resilient.
+%%
+We display the resulting \emph{contingency matrix} using a heat map, mapping correlation values to color brightness with white for 0 and dark blue for 1, as presented in \cite{Klemm2014VIS}.
 %%
@@
 %%
-Decision trees are a popular classification method in data mining for creating predictive models.
-%%
-Leafs of a decision tree represent class labels, branches represent feature conjunctions leading to the class labels.
+Decision trees are a popular classification method in data mining to create predictive models.
+%%
+Leafs of a decision tree represent class labels, branches represent decision trees.
 %%
@@
 \subsection{The C4.5/C5.0 Algorithm}
-Using a trained data set, the C4.5 algorithm builds decision trees based on information entropy.
+The C4.5 algorithm builds decision trees based on information entropy on a data set.
 %%
@@
 %%
-The algorithm then tries to find a decision tree, which divides the samples using the input variables just like the target variables.
-%%
-This means, that every node in the tree represents the attribute which splits the data most efficiently into the target subset.
+%The algorithm then tries to find a decision tree, which divides the samples using the input variables just like the target variables.
+%%
+%This means, that every node in the tree represents the attribute which splits the data most efficiently into the target subset.
 %%
@@
 %%
-We use the \texttt{R} implementation of C5.0 \cite{c5.0lassification}.
+We use the \texttt{R} implementation of C5.0 \cite{c5.0classification}.
 %%
@@
 %%
-
-The actual use for the resulting tree is the classification of new observations (subjects).
-%%
-Yet, we are interested in the decision rules and the classification accuracy.
+%
+%The actual use for the resulting tree is the classification of new observations (subjects).
+%%
+%Yet, we are interested in the decision rules and the classification accuracy.
+%%
+In the following analysis, we strongly focus on the complexity of decision trees and the classification accuracy.
 
@@
 %%
-Since we have 134 non-image variables, the calculation yields a corresponding number of trees.
+Since we have 134 non-image variables, the calculation yields the same amount of trees.
 %%
@@
 We have to abstract the results of the classification to keep the mental effort of interpreting the data low.
+%%
+This is included in our IVA framework.
 %%------------
@@
 %%
-The two measures form the axis for a \emph{scatter plot} of the classification results.
+The two measures form the axes for a \emph{scatter plot} of the classification results.
+%%
+The scatter plot is our central element for the interactive analysis of decision trees.
 
 \paragraph{The Error Term. }
-Normally the error for a classification is calculated with $error = \frac{classifiedCorrectly_{n}} {n}$ where $n$ is the number of subjects.
+Normally the error for a classification is calculated with $error = \frac{correctlyClassified_{n}} {n}$ where $n$ is the number of subjects.
 %%
@@
 %%
-If for example a variable indicating a disease is negative for 90\% of the subjects and the classifier simply assigns all subjects to \emph{not ill} the error metric would yield an error of 10\%, even though it is very bad.
+For example, if a variable indicates a disease is negative for 90\% of the subjects and the classifier simply assigns all subjects to \emph{not ill} the error metric would yield an error of 10\%, even though it is very bad.
 %%
@@
 \begin{equation}
-error = 1 - \frac{\sum_{m=0}^M \frac{classifiedCorrectly_{m}}{all_{m}}}{number_{dimensions}}
+error = 1 - \frac{\sum_{m=0}^M \frac{correctlyClassified_{m}}{m}}{number_{dimensions}}
 \end{equation}
@@
 %%
-For example, a pain indicator variable ranging from \emph{1 - no pain} to \emph{4 - large pain} is subdivided into four binary variables (\emph{No pain - Yes/No} to \emph{Large Pain - Yes/no}).
+For example, a pain indicator variable ranging from \emph{1 - no pain} to \emph{4 - large pain} is subdivided into four binary variables. %(\emph{No pain - Yes/No} to \emph{Large Pain - Yes/no}).
 %%
@@
 %%
-This is realized by clicking on an entry on the visualization, which then displays the corresponding decision tree in detail.
+This is realized by clicking on an entry on the visualization, which then interactively displays the corresponding decision tree in detail.
 %%
@@
 %%
-Selecting a variable using a drop down menu allows to user to select the variable used for subject subdivisions, e.g. \emph{gender} or \emph{disease treatment type}.
+Selecting a variable using a drop down menu allows the user to select the variable used for subject subdivisions, e.g. \emph{gender} or \emph{disease treatment type}.
 %%
@@
 %%
-As opposed to the standard \texttt{R} plots, \texttt{ggvis} allows to adjust visualization parameters using UI controls, such as sliders.
+As opposed to the standard \texttt{R} plots, \texttt{ggvis} allows to adjust visualization parameters using user interface controls, such as sliders.
 %%
@@
 %%
-The prototype is available under the following link: \url{blind.dnsalias.com}
+The prototype is available under the following link: \url{blind.dnsalias.com}.
 %%---------
@@
 %%
-A interactive versions of the plot (see supplemental material) makes all data points clickable, showing the corresponding decision tree in a tool tip.
+%A interactive versions of the plot (see supplemental material) makes all data points clickable, showing the corresponding decision tree in a tool tip.
+The interactive plot (see supplemental material) has clickable data points, displaying corresponding decision tree in a tool tip.
+%A interactive versions of the plot (see supplemental material) makes all data points clickable, showing the corresponding decision tree in a tool tip.
 %%
@@
 \end{figure*}
-We ran the analysis using different subject groups:
+\com{ToDo: Eigene Section daraus machen?}. 
+In this section, we show which non-image variables can be described using the 9 image derived variables.
+%%
+We also create subject groups to assess the influence of variables affecting the lumbar spine shape.
+%%
+We carry out the analysis using different subject groups:
 \begin{enumerate}
@@
 %%
-The vast majority of parameters can not be described well using the classifier.
+The vast majority of non-image variables can not be described well using the classifier.
 %%
@@
 %
-It can be discriminated with an error of $0.31$ using 7 rules and incorporates only curvature and curvature angle variables.
+It can be discriminated with an error of $0.31$ using 7 rules and incorporates only \emph{curvature} and \emph{curvature angle} related variables.
 %%
@@
 %%
-\emph{Medication for high blood pressure} is classified for 1,058 subjects with an error of $0.47$ solely based on coronal mean curvature.
+\emph{Medication for high blood pressure} is classified for 1,058 subjects with an error of $0.47$ solely based on \emph{coronal mean curvature}.
 %%
@@
 %%
-Almost all subjects who are medicated ($796/1,058$) were correctly classified, the vast majority of non-medicated subjects ($262/1,058$) are false-positive classified.
-%%
-Therefore the classifier is not as useful.
+Almost all subjects who are medicated ($796$ of $1,058$) were correctly classified, the vast majority of non-medicated subjects ($262$ of $1,058$) are false-positive classified, yielding a poor quality of the classifier w.r.t epidemiological research.
 %%This is however due to the high amount of people, who are medicated (68\%) and who are almost all correctly class
@@
 
-The analysis of the dummy variable yields similar results like the \emph{blood pressure medication}.
+The analysis of the dummy variables yields similar results like the \emph{blood pressure medication}.
 %%

### 2014-09-16 draft.tex
Further revisions

@@
 %%
-Scatterplot visualization of all C5.0 classification results.
+Interactive scatterplot visualization of all C5.0 classification results.
 %%
@@
 %%
-The classifier neglects the other group and yields a error below $0.5$.
+The classifier neglects the other group and yields an error below $0.5$.
 %The dummy variable analysis yields fewer results.
@@
 %%
-The result however is probably over fitted to the data due to tree sizes between $14$ and $16$.
+The result however is probably overfitted to the data due to tree sizes between $14$ and $16$.
 %%
@@
 %%
-Notable is also the decreasing error with increasing decision tree size.
+%Notable is also the decreasing error with increasing decision tree size.
 %%
@@
 %%
-Differences between \emph{male} and \emph{female} subjects are explained by to the average \emph{body size} difference.
+Differences between \emph{male} and \emph{female} subjects are explained by the average \emph{body size} difference.
 %%
@@
 %
-The question is, whether the inter-group spine-variability parameter is enough for predicting other parameter or not.
+The question is, whether the inter-group spine-variability parameter is enough for predicting other parameters or not.
 %%
@@
 %%
-There are several decision trees for laboratory values, such as \emph{alanine aminotransferase} value (relevant for diagnosis of liver or gallbladder illness) in the blood can be described with an error of $0.4$ (\emph{$139-164~cm$}) to $0.36$ (\emph{$164 - 171~cm$}).
+There are several decision trees for laboratory values, e.g.
+\emph{alanine aminotransferase} value (relevant for diagnosis of liver or gallbladder illness) in the blood can be described with an error of $0.4$ (\emph{$139-164~cm$}) to $0.36$ (\emph{$164 - 171~cm$}).
 %%
@@
 \label{sec:Conclusion}
-\noindent The presented results indicate that the parameters \emph{torsion}, \emph{curvature} and \emph{curvature angle} are not enough to characterize pathological changes in the lumbar spine.
+\noindent The presented results indicate that the image based variables \emph{torsion}, \emph{curvature} and \emph{curvature angle} are not enough to characterize pathological changes in the lumbar spine.
 %%

### 2014-09-16 draft.tex
Began to restructure last section

@@
 
-Another observation during the analysis is very large decision trees achieving low error scores.
+Another observation during the analysis is that large decision trees achieve low error scores.
 %%
 Their complexity impose overfitting to the data and they are not suitable for extracting universal rules.
+%%
+\paragraph{Applicability.}
+%%
+Classification methods based on decision trees showed to be useful to assess the discriminative power of a variable set.
+%%
+Their ability to assess variable combination makes them more powerful than correlation coefficients calculated for each variable.
+%%
+This advantage comes with a much more complex output, making their results harder to assess and to abstract.
+%%
+Our method to plot derived metrics and custom-tailored error measures proved to be effective.
+%%
+Huge result spaces could be navigated fast using our brushable visualization.
+%%
+Therefore the method is applicable not only for deriving information based on image data, but on all potential target variables.
+%%
+
+% BLIND
+%As described in a prior work (\cite{Klemm2014VIS}), the suggestion of potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
+Suggesting potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
+%%
+We can achieve this also using the presented method by trying to describe several (dichotomous) target values with a decision tree constructed from all available data.
+%%
+This allows both to assess the discriminative power of the data set regarding the target variables, as well as the most important parameters for the classification.
+%Except for gender in the age-plot, no variable could be explained with reasonable accuracy for age or body size.
+\paragraph{Predictive Power of Image-Derived Variables.}
+We showed, that \emph{torsion}, \emph{curvature} and \emph{curvature angle} of the lumbar spine at the presented precision are not enough to characterize lumbar back pain in the SHIP data set.
+%%
+Our method made it possible to assess their discriminative power, which is largely limited for separating \emph{male} and \emph{female} subjects, \emph{nutrition} variables, as well as different disease indicators.
+%%
+The C4.5 algorithm showed to be an effective tool for evaluating a set of derived metrics regarding their suitability to classify non-image parameters.
+%%
+Over-fitting to the data indicated by complex decision trees have to be taken into account into the analysis as well.
 %%
@@
 %%
-The \emph{overall spine canal shape} is also of interested, since scoliosis and lordosis can be characterized more precise by deriving the Cobb angles from this shape (Fig~\ref{fig:proposedFeatures}~b,~c).
+The \emph{overall spine canal shape} is also of interest, since scoliosis and lordosis can be characterized more precise by deriving the Cobb angles from this shape (Fig~\ref{fig:proposedFeatures}~b,~c).
 %% ------------------------------------
-\subsection{Applicability}
-%%
-Classification methods based on decision trees showed to be useful to assess the discriminative power of a variable set.
-%%
-Their ability to assess variable combination makes them more powerful than correlation coefficients calculated for each variable.
-%%
-This advantage comes with a much more complex output, making their results harder to assess and to abstract.
-%%
-Our method to plot derived metrics and custom-tailored error measures proved to be effective.
-%%
-Huge result spaces could be navigated fast using our brushable visualization.
-%%
-Therefore the method is applicable not only for deriving information based on image data, but on all potential target variables.
-%%
-
-% BLIND
-%As described in a prior work (\cite{Klemm2014VIS}), the suggestion of potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
-Suggesting potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology \cite{Klemm2014VIS}.
-%%
-We can achieve this also using the presented method by trying to describe several (dichotomous) target values with a decision tree constructed from all available data.
-%%
-This allows both to assess the discriminative power of the data set regarding the target variables, as well as the most important parameters for the classification.
-%Except for gender in the age-plot, no variable could be explained with reasonable accuracy for age or body size.
-
-\subsection{Summary and Future Work}
-%%
-\noindent We showed, that \emph{torsion}, \emph{curvature} and \emph{curvature angle} of the lumbar spine at the presented precision are not enough to characterize lumbar back pain in the SHIP data set.
-%%
-Our method made it possible to assess their discriminative power, which is largely limited for separating \emph{male} and \emph{female} subjects, \emph{nutrition} variables, as well as different disease indicators.
-%%
-The C4.5 algorithm showed to be an effective tool for evaluating a set of derived metrics regarding their suitability to classify non-image parameters.
-%%
-Over-fitting to the data indicated by complex decision trees have to be taken into account into the analysis as well.
+\subsection{Future Work}
+%% Contributions
+We provide a method for comparative analysis of decision trees independent of the variable manifestation count using interactive scatter plots.
+%%
+We applied the technique to derive insight into the predictive power of 9 image-derived variables for 134 non-image variables.
+%%TODO HIER WEITER
+%We carried out an analysis on subjects divided by \emph{gender}, \emph{BMI}, and \emph{body size} to assess their influence on the lumbar spine.
+%To assess the influence of parameters which strongly influence the spine shape, we applied the technique to.
 %%

### 2014-09-17 draft.tex
Sylvias revision complete

@@
 %%
-Hence, correlations between non-image parameters were successfully detected and characterized.
+Hence, correlations between non-image variables were successfully detected and characterized.
 }
@@
 %%
-Furthermore, previous analyses mainly focused on a single parameter (e.g. \emph{BMI}) or a small subset of parameters.
+Furthermore, previous analyses mainly focused on a single variable (e.g. \emph{BMI}) or a small subset of variables.
 %%
@@
 %%
-\item The detection of correlations between image-based and socio-demographic as well as medical parameters for an improved hypothesis generation,
+\item The detection of correlations between image-based and socio-demographic as well as medical variables for an improved hypothesis generation,
 %%
@@
 %%
-They define a cohort as parameter-divided subject group, differing from the epidemiological understanding of the term.
+They define a cohort as variable-divided subject group, differing from the epidemiological understanding of the term.
 %%
@@
 %%
-We combine both ideas by validating the significance of image-derived parameters.
+We combine both ideas by validating the significance of image-derived variables.
 We aim at deriving new relationships by training decision trees to explain target variables.
@@
 %%
-\section{The Lumbar Spine Data Set}
+\section{\uppercase{The Lumbar Spine Data Set}}
 \label{sec:MaterialsAndMethod}
@@
 %%
-This yields a total of 9 image-derived parameters.
+This yields a total of 9 image-derived variables.
 %%
@@
 %%
-When correlating spine curvature and torsion with non-image parameter, all these variables need to be taken into account
+When correlating spine curvature and torsion with non-image variable, all these variables need to be taken into account
 %%
@@
 %%
-\noindent As a first experiment we correlated the shape parameter with the binary back pain indicator using \texttt{GPLOM's} \cite{GPLOMS}.
+\noindent As a first experiment we correlated the shape variable with the binary back pain indicator using \texttt{GPLOM's} \cite{GPLOMS}.
 %%
@@
 \subsubsection{Correlation Matrix}
-We calculated an association matrix to assess correlations between the image parameters.
+We calculated an association matrix to assess correlations between the image variables.
 %%
@@
 %%--------------------------------------------------------
-\subsubsection{Correlation of Image Parameters With Back Pain}
-Figure~\ref{fig:image-parameter-range} shows the distribution of all image-parameters as generalized pairs plot.
+\subsubsection{Correlation of Image Variables With Back Pain}
+Figure~\ref{fig:image-parameter-range} shows the distribution of all image-variables as generalized pairs plot.
 %%
@@
 %%
-The PCA results in a linear projection of the parameter space ordering their components by maximum variance.
+The PCA results in a linear projection of the variable space ordering their components by maximum variance.
 %%
@@
 %%
-We then expanded our focus on correlations of image-derived parameters with all other non-image variables.
+We then expanded our focus on correlations of image-derived variables with all other non-image variables.
 %%
@@
 %%
-The resulting contingency matrices are available as supplementary material and show no strong correlation with any of the parameter (see \url{blind.dnsalias.com}).
+The resulting contingency matrices are available as supplementary material and show no strong correlation with any of the variable (see \url{blind.dnsalias.com}).
 %%
@@
 
-These observations brought us to the conclusion to incorporate more sophisticated data mining techniques to assess the influence of the image-derived parameters.
+These observations brought us to the conclusion to incorporate more sophisticated data mining techniques to assess the influence of the image-derived variables.
 
@@
 %%
-We rely on predictive classification trained to obtain a complex rule set on how combinations of the image-parameter explain non-image variables.
+We rely on predictive classification trained to obtain a complex rule set on how combinations of the image-variable explain non-image variables.
 %%
@@
 %%
-This is useful for our classification, because it allows to determine which manifestation of a variable can be described best using the image data parameter.
+This is useful for our classification, because it allows to determine which manifestation of a variable can be described best using the image data variable.
 %%\com{ToDo: Dummy-Variables}
@@
 %%
-As opposed to the standard \texttt{R} plots, \texttt{ggvis} allows to adjust visualization parameters using user interface controls, such as sliders.
+As opposed to the standard \texttt{R} plots, \texttt{ggvis} allows to adjust visualization variables using user interface controls, such as sliders.
 %%
@@
 %%
-It allows to combine both the ability to create comprehensible reports as well as UI-controls, to adjust visible data and parameters in the plot to allow the viewer to analyze the data.
+It allows to combine both the ability to create comprehensible reports as well as UI-controls, to adjust visible data and variables in the plot to allow the viewer to analyze the data.
 %%
@@
 %%---------
-\subsection{Results}
+\section{\uppercase{Results}}
 \begin{figure*}[p!]
@@
 \end{figure*}
-\com{ToDo: Eigene Section daraus machen?}. 
-In this section, we show which non-image variables can be described using the 9 image derived variables.
+\noindent In this section, we show which non-image variables can be described using the 9 image derived variables.
 %%
@@
 
-\subsubsection{All Variables}
+\subsection{All Variables}
 %%
@@
 % None of the pain indicators can be reliably described using the image-based parameter.
-None of the pain indicators can be described reliably using the image-based parameters.
+None of the pain indicators can be described reliably using the image-based variables.
 %%
@@
 %%
-Variables, such as subjects sized 139-164 cm, between 64 and 90 years of age or nutrition related parameters are dominantly populated by one manifestation.
+Variables, such as subjects sized 139-164 cm, between 64 and 90 years of age or nutrition related variables are dominantly populated by one manifestation.
 %%
@@
 %%
-\subsubsection{Gender Groups}
+\subsection{Gender Groups}
 %%
@@
 %\com{Potentially skip these due to the confounding of body size}
-\subsubsection{Body-Mass-Index Groups}
+\subsection{Body-Mass-Index Groups}
 %%
@@
 
-\subsubsection{Size Groups}
+\subsection{Size Groups}
 %%
@@
 %
-The question is, whether the inter-group spine-variability parameter is enough for predicting other parameters or not.
+The question is, whether the inter-group spine-variability variable is enough for predicting other variables or not.
 %%
@@
 %%
-These variables include \emph{nutrition} parameter, \emph{thyroid disorder} and \emph{social problems induced by back pain}.
+These variables include \emph{nutrition}, \emph{thyroid disorder} and \emph{social problems induced by back pain}.
 
@@
 %%
-For some \emph{body size} groups, certain \emph{nutrition} or \emph{psychological} parameters can be predicted with a comparatively high error.
+For some \emph{body size} groups, certain \emph{nutrition} or \emph{psychological} variables can be predicted with a comparatively high error.
 %%
@@
 %%
-This allows both to assess the discriminative power of the data set regarding the target variables, as well as the most important parameters for the classification.
+This allows both to assess the discriminative power of the data set regarding the target variables, as well as the most important variables for the classification.
 %Except for gender in the age-plot, no variable could be explained with reasonable accuracy for age or body size.
@@
 %%
-The C4.5 algorithm showed to be an effective tool for evaluating a set of derived metrics regarding their suitability to classify non-image parameters.
+The C4.5 algorithm showed to be an effective tool for evaluating a set of derived metrics regarding their suitability to classify non-image variables.
 %%
@@
 %%
-\subsection{Proposed Features}
-\label{sec:ProposedFeatures}
-\begin{figure}[!h]
-  %\vspace{-0.2cm}
-  \centering
-  \includegraphics[width=0.475\textwidth]{figures/proposed-features}
-  \caption{
-	Proposed features.
-	%%
-	(a) dented vertebraes are a sign of heavy stress and bone absorption.
-	%%
-	(b, c) the spine canal shape can be used to characterize scoliosis (curvature sagittal) and lordosis (curvature coronal) as well as the \emph{Cobb-Angles} \cite{Cobb}.
-	%%
-	(c) Spine canal thickness is associated with herniated disks.
-	%%
-	}
-  \label{fig:proposedFeatures}
-\end{figure}
-%%
-\noindent Pathological deformation of the lumbar spine is usually the last stage and is associated with very strong lower back pain.
-%%
-We have to capture earlier signs of pathological change in order to derive better predictive models.
-%%
-
-A very early sign of pathological deformation is the \emph{bone resorption in the center of a vertebra}.
-%%
-They change from being block-shaped to be dented in the center (Fig~\ref{fig:proposedFeatures}~a).
-%%
-This information can be obtained by segmenting the whole vertebra or the top and bottom point if each vertebra center.
-%%
-Another valuable variable would be the \emph{spine canal thickness} (Fig~\ref{fig:proposedFeatures}~c).
-%%
-Low spine canal thickness can be an indicator for an impending herniated disk.
-%%
-Both surface texture of the vertebra and thickness of the spine canal are used to diagnose herniated disks.
-%%
-The \emph{overall spine canal shape} is also of interest, since scoliosis and lordosis can be characterized more precise by deriving the Cobb angles from this shape (Fig~\ref{fig:proposedFeatures}~b,~c).
+% \subsection{Proposed Features}
+% \label{sec:ProposedFeatures}
+% \begin{figure}[!h]
+%   %\vspace{-0.2cm}
+%   \centering
+%   \includegraphics[width=0.475\textwidth]{figures/proposed-features}
+%   \caption{
+% 	Proposed features.
+% 	%%
+% 	(a) dented vertebraes are a sign of heavy stress and bone absorption.
+% 	%%
+% 	(b, c) the spine canal shape can be used to characterize scoliosis (curvature sagittal) and lordosis (curvature coronal) as well as the \emph{Cobb-Angles} \cite{Cobb}.
+% 	%%
+% 	(c) Spine canal thickness is associated with herniated disks.
+% 	%%
+% 	}
+%   \label{fig:proposedFeatures}
+% \end{figure}
+% %%
+% \noindent Pathological deformation of the lumbar spine is usually the last stage and is associated with very strong lower back pain.
+% %%
+% We have to capture earlier signs of pathological change in order to derive better predictive models.
+% %%
+%
+% A very early sign of pathological deformation is the \emph{bone resorption in the center of a vertebra}.
+% %%
+% They change from being block-shaped to be dented in the center (Fig~\ref{fig:proposedFeatures}~a).
+% %%
+% This information can be obtained by segmenting the whole vertebra or the top and bottom point if each vertebra center.
+% %%
+% Another valuable variable would be the \emph{spine canal thickness} (Fig~\ref{fig:proposedFeatures}~c).
+% %%
+% Low spine canal thickness can be an indicator for an impending herniated disk.
+% %%
+% Both surface texture of the vertebra and thickness of the spine canal are used to diagnose herniated disks.
+% %%
+% The \emph{overall spine canal shape} is also of interest, since scoliosis and lordosis can be characterized more precise by deriving the Cobb angles from this shape (Fig~\ref{fig:proposedFeatures}~b,~c).
 %% ------------------------------------
-\subsection{Future Work}
+\paragraph{Future Work.}
+\com{ToDo: Passen die Contributions hier wirklich?}
 %% Contributions
@@
 We applied the technique to derive insight into the predictive power of 9 image-derived variables for 134 non-image variables.
-%%TODO HIER WEITER
-%We carried out an analysis on subjects divided by \emph{gender}, \emph{BMI}, and \emph{body size} to assess their influence on the lumbar spine.
-%To assess the influence of parameters which strongly influence the spine shape, we applied the technique to.
-%%
-
-In our future work, we will focus on more precise models for extracting measures described in Section~\ref{sec:ProposedFeatures}.
+%%
+The analysis was carried out for subjects divided by \emph{gender}, \emph{BMI}, and \emph{body size} to assess their influence on the lumbar spine shape.
+
+%In our future work, we will focus on more precise models for extracting measures described in Section~\ref{sec:ProposedFeatures}.
+In our future work, we will focus on more precise models for extracting measures.
+%%
+Dented vertebras are a early sign of a pathological deformation, so we want to capture this by segmenting the top and bottom point of each vertebra center.
+%%
+Another interesting metric is the spine canal thickness over the whole spine, capturing early signs of herniated disc disease.
 %%

### 2014-09-17 draft.tex
Minor Changes + ToDo for categorization

@@
 \author{\authorname{Paul Klemm\sup{1}, Sylvia Glaßer\sup{1}, Kai Lawonn\sup{1}, Marko Rak\sup{1} Henry Völzke\sup{2}, Katrin Hegenscheid\sup{3} and Bernhard Preim\sup{1}}
-\affiliation{\sup{1}Department of Simulation and Graphics, University of Magdeburg, Germany}
+\affiliation{\sup{1} Department of Simulation and Graphics, University of Magdeburg, Germany}
 \affiliation{\sup{2} Institute for Community Medicine, University of Greifswald, Germany}
@@
 %%
-The analysis, as well as the presented IVA tool is available as supplementary material in a interactive \texttt{R-Markdown}\footnote{Developed by RStudio, Inc; \texttt{rmarkdown.rstudio.com}} document (\url{blind.dnsalias.com}).
+The analysis, as well as the presented IVA tool is available as supplementary material in a interactive \texttt{R Markdown}\footnote{Developed by RStudio, Inc; \texttt{rmarkdown.rstudio.com}} document (\url{blind.dnsalias.com}).
 %%

### 2014-09-17 draft.tex
Revised version

@@
 %%
-Therefore, an interactive visual analysis (IVA) framework was created and tested with 2,540 segmented lumbar spine data sets.
+Therefore, an Interactive Visual Analysis (IVA) framework was created and tested with 2,540 segmented lumbar spine data sets.
 %%
@@
 %%Epidemiology is the study of causation of diseases.
-Epidemiology is the study of dissemination, causes and results of health-related states and events.
+\noindent Epidemiology is the study of dissemination, causes and results of health-related states and events.
 %%
@@
 %%
-These information are used to determine risk factors for diseases, help people to improve their lifestyle and their health or to support the diagnosis of widespread diseases.
+These information are used to determine risk factors for diseases, help people to live healthier or to support the diagnosis of widespread diseases.
 %%
@@
 
-In this paper, we present a thorough analysis of the image-derived data of a cohort study including lumbar back spine datasets.
-%%
-We analyze their discriminating power for different characteristics of back pain.
+We present a thorough analysis of the image-derived data of a cohort study including lumbar spine datasets.
+%%
+We analyze their discriminative power for different back pain characteristics.
 %%
@@
 %Then, interactive visual analysis provides insight into the quality of the image-derived data as well as their potential for possible back pain risk factors.
-Then, interactive visual analysis provides insight into the quality of the image-derived data as well as their potential for being back pain risk factors.
-%%
-The workflow is part of an interactive visual analysis (IVA) framework, constructed to enhance the epidemiological workflow with iterative data driven techniques.
+Then, Interactive Visual Analysis (IVA) provides insight into the quality of the image-derived data as well as their potential for being back pain risk factors.
+%%
+The workflow is part of an IVA framework, constructed to enhance the epidemiological workflow with iterative data driven techniques.
 %%
@@
 %%
-For example, the analysis of the influence of nutrition on prostate cancer is strongly age dependent and requires therefore age-normalization.
+For example, the effect of nutrition on prostate cancer requires age-normalization due to the increased incidence for older subjects.
 %%
@@
 %%
-We focus on data from large scale cohort studies.
+We focus on large scale cohort study data.
 %%
@@
 %%
-As a result, these data can be analyzed regarding many diseases and conditions.
+These data can be analyzed regarding many diseases and conditions.
 %%
@@
 %%
-For example, continuous data, such as age is often discretized into age-bins.
-%%
-Every information reduction can introduce a bias to the data, since it proposes an assumption about the data.
-%%
-Using age to divide subjects into \emph{young} and \emph{old} categories can distort statistical results.
+For example, continuous data, such as age, is often discretized into quantile bins.
+%%
+%Every information reduction can introduce a bias to the data, since it proposes an assumption about the data.
+Every information reduction proposes an assumption about the data and can therefore introduce a bias.
+%%
+For example, using age to divide subjects into \emph{young} and \emph{old} categories can distort statistical results.
 %%
@@
 %%
-These include nutrition, sporting activities and body posture combined with physical stress at work.
-%%
-The exact causes as well as particularly vulnerable risk groups are not known.
+These include nutrition, physical activity and body posture combined with physical stress at work.
+%%
+The exact causes as well as particularly vulnerable risk groups are unknown.
 %%
@@
 %Klemm et al. propose an extension to the previously described epidemiological workflow using \emph{Interactive Visual Analysis} (IVA) \cite{Klemm2014VIS}.
-\cite{Klemm2014VIS} propose an extension to the previously described epidemiological workflow using \emph{Interactive Visual Analysis} (IVA).
+\cite{Klemm2014VIS} propose an extension to the previously described epidemiological workflow using IVA.
 %%
@@
 %%
-The effort of analyzing the decision trees increases with decision tree complexity.
-%%
-Gla{\ss}er et al. use variables derived from DCE-MRI data capturing the perfusion in the tumorous tissue \cite{Glasser2013}.
+The effort of analyzing decision trees increases with their complexity.
+%%
+\cite{Glasser2013} use variables derived from DCE-MRI data capturing the perfusion in the tumorous tissue.
 %%
@@
 %%
-Niemann et al. assess hepatic steatosis (fatty liver disease) risk factors using decision trees \cite{Niemann2014}.
+\cite{Niemann2014} assess hepatic steatosis (fatty liver disease) risk factors using decision trees.
 %%
@@
 
-Pinhero et al.
-applied association rule mining to create high-confidence decision rules based on variables, such as age, gender and geographic location about the mortality rate in liver cancer data \cite{Pinheiro}.
-%%
-Sekhavat and Hoeber propose an interactive visualization technique for decision trees, where all decisions are displayed in a heat map \cite{Sekhavat}.
-%%
-The visualization translates the decision tree into a heat map, where each cell represents the connectivity between an initial value and a target value.
+\cite{Pinheiro} applied association rule mining to create high-confidence decision rules based on variables, such as age, gender and geographic location about the mortality rate in liver cancer data.
+%%
+\cite{Sekhavat} propose an interactive visualization technique for decision trees, where all decisions are displayed in a heat map.
+%%
+The visualization translates the decision tree into a heat map, where each cell represents the connectivity between an initial and a target value.
 %%
@@
 \paragraph{SHIP Data Analysis.}
-Klemm et al. analyzed the lumbar spine variability of 490 \texttt{SHIP-2} subjects \cite{Klemm2013VMV}.
+\cite{Klemm2013VMV} analyzed the lumbar spine variability of 490 \texttt{SHIP-2} subjects.
 %%
@@
 %%
-Clustering techniques applied to non-image data are strongly dependent on the chosen variables and distance measures.
-%%
-%Klemm et al. found \texttt{k-Prototypes} and \texttt{DBSCAN} appropriate for this task \cite{Klemm2014BVM}.
+Clustering techniques applied to non-image data are strongly dependent on the chosen variables and distance measures \cite{Klemm2014BVM}.
+%%
+%Klemm et al. found \texttt{k-Prototypes} and \texttt{DBSCAN} appropriate for this task.
 %%
@@
 %Without the 9 image-derived parameter, the data set comprises of 134 variables:
-The data set comprises of 134 variables (and 9 additional image-derived variables):
+The data set comprises of 134 variables (and 9 additional image-derived variables described in the next subsection):
 \begin{itemize}
@@
 %%
-A click on the L3 vertebra center initialized the model, two clicks on the top and bottom of the vertebra describes an initial height estimation (Fig.~\ref{fig:centerline}~a).
+A click on the L3 vertebra center initializes the model, two clicks on the top and bottom of the vertebra describes an initial height estimation (Fig.~\ref{fig:centerline}~a).
 %%
@@
 
-\section{\uppercase{Experiments}}
+\section{\uppercase{Experiments and Preliminary Results}}
 \label{sec:Experiments}
@@
 %%
-Larger people have also a longer spine and its shape is more straight.
+Larger people have also a longer spine with a straighter shape.
 %%
@@
 %%
-When correlating spine curvature and torsion with non-image variable, all these variables need to be taken into account
-%%
-Since the gender confounder mainly encodes body-size, we decided to divide subjects into body-size groups.
+All these variables need to be taken into account, when spine curvature and torsion is correlated with non-image variables.
+%%
+Since the gender confounder mainly encodes body size, we decided to divide subjects into body size groups.
 %%
 To avoid small outlier groups, epidemiologists recommend using quartiles to discretize metric variables.
-\subsection{Preliminary Results}
 \begin{figure*}[htb]
@@
 %%
-Pairwise combinations of image-derived variables are visualized via scatter plots left of the matrix diagonal and their correlation with back pain is denoted right of the matrix diagonal.
+Pairwise combinations of image-derived variables are visualized via scatter plots on the left of the matrix diagonal.
+%%
+Their correlation with back pain is denoted right of the matrix diagonal.
 %%
@@
 %%
-\noindent As a first experiment we correlated the shape variable with the binary back pain indicator using \texttt{GPLOM's} \cite{GPLOMS}.
+\paragraph{GPLOM Analysis.}
+As a first experiment we correlated the shape variable with the dichotomous back pain indicator using \texttt{GPLOM's} \cite{GPLOMS}.
 %%
@@
 %%--------------------------------
-\subsubsection{Correlation Matrix}
+\paragraph{Correlation Matrix.}
 We calculated an association matrix to assess correlations between the image variables.
@@
 %%
-Also the mean curvature and the curvature angle correlate by a factor of $-0.89$.
+Also the \emph{mean curvature} and \emph{curvature angle} correlate by a factor of $-0.89$.
 %%
 %Torsion and curvature/curvature angle do not correlate.
-Torsion does not correlate with any other image-derived variable.
+\emph{Torsion} does not correlate with any other image-derived variable.
 %%--------------------------------------------------------
-\subsubsection{Correlation of Image Variables With Back Pain}
+\paragraph{Correlation of Image Variables With Lumbar Back Pain.}
 Figure~\ref{fig:image-parameter-range} shows the distribution of all image-variables as generalized pairs plot.
@@
 
-\subsubsection{Assessing the Information Gain Using the Principal Component Analysis}
-%%
-To determine the information gain per image-derived variables, we calculated a Principal Component Analysis (PCA) and compared the loadings per dimension.
+\paragraph{Assessing the Information Gain Using the Principal Component Analysis.}
+%%
+To determine the information gain per image-derived variable, we calculated a Principal Component Analysis (PCA) and compared the dimension loadings.
 %%
@@
 %% ------------------------------------------
-\subsubsection{Heterogenous Correlations}
+\paragraph{Heterogenous Correlations.}
 %%
@@
 %%
-We display the resulting \emph{contingency matrix} using a heat map, mapping correlation values to color brightness with white for 0 and dark blue for 1, as presented in \cite{Klemm2014VIS}.
+We display the resulting \emph{contingency matrix} using a heat map, encoding correlation values with color brightness with white for $0$ and dark blue for $1$, as presented in \cite{Klemm2014VIS}.
 %%
@@
 %%
-Only weak correlations could be found for \emph{Mean Curvature} with \emph{gender} (0.42), \emph{body size} (0.39) and \emph{number of born children} (0.29).
-%%
-One surprising result was the small correlation with \emph{torsion}, which correlated with almost no variables (p values between 0 and 0.05) and \emph{parkinson} (0.24).
-%%
-
-These observations brought us to the conclusion to incorporate more sophisticated data mining techniques to assess the influence of the image-derived variables.
+Only weak correlations could be found for \emph{mean curvature} with \emph{gender} (0.42), \emph{body size} (0.39) and \emph{number of born children} (0.29).
+%%
+One surprising result was the small correlation of \emph{torsion} with \emph{parkinson} (0.24).
+%%
+Other than that, \emph{torsion} correlated with almost no variables (p values between 0 and 0.05).
+%%
+
+These observations brought us to the decision to incorporate sophisticated data mining techniques to assess the influence of the image-derived variables.
 
@@
 %%
-\noindent As described before, correlation coefficients fail to infer back pain status based on lumbar spine canal curvature and torsion.
+\noindent As described before, correlation coefficients fail to infer back pain status based on lumbar spine canal \emph{curvature} and \emph{torsion}.
 %%
@@
 %%
-Leafs of a decision tree represent class labels, branches represent decision trees.
+%Leafs of a decision tree represent class labels, branches represent decision trees.
+Leafs of a decision tree represent class labels, branches represent feature conjunctions leading to the class labels.
 %%
@@
 %%
-This allows epidemiologists to interpret the results without having deep knowledge about the algorithm creating the tree.
+Epidemiologists can interpret the results without having deep knowledge about the classification algorithm.
 %%
@@
 \subsection{The C4.5/C5.0 Algorithm}
+\label{sec:C5.0}
 The C4.5 algorithm builds decision trees based on information entropy on a data set.
@@
 %%
-Creating dummy variables bypasses this problem.
+%Creating dummy variables bypasses this problem.
+Creating a dummy variable by converting each manifestation of a categorical variable into a dichotomous variable bypasses this problem.
 %%
@@
 %We follow the visual information-seeking mantra and want to provide an overview first, then details-on-demand \cite{shneiderman1996}.
-We follow the Visual Analytics mantra by analyzing first, show the important and analyze further \cite{Keim}.
+We follow the Visual Analytics mantra of analyzing first, show the important and analyze further \cite{Keim}.
 %%
@@
 %%
-The two measures form the axes for a \emph{scatter plot} of the classification results.
+The two measures form the axes for a scatter plot of the classification results.
 %%
@@
 %%
-For example, if a variable indicates a disease is negative for 90\% of the subjects and the classifier simply assigns all subjects to \emph{not ill} the error metric would yield an error of 10\%, even though it is very bad.
+For example, if a variable indicating a disease is negative for 90\% of the subjects and the classifier simply assigns all subjects to \emph{not ill} the error metric would yield an error of 10\%, even though it is very bad.
 %%
@@
 %%
-The error is scaled to denote perfect classification with 0 and 1 is equal to random selection.
+The error is scaled to denote perfect classification with 0; 1 is equal to random selection.
 % The error rate has to be judged depending on the number of classes of the target variable.
@@
 %%
-This allows to instantly assess the reliability of the result.
+This allows instant reliability assessment of the result.
 %%
@@
 % %%
-% By mapping the group affiliation (in our case body-size) to color or shape, we can even plot all 402 results in a single visualization (Fig.~\ref{fig:decision-tree} a).
+% By mapping the group affiliation (in our case body size) to color or shape, we can even plot all 402 results in a single visualization (Fig.~\ref{fig:decision-tree} a).
 %%
 \subsubsection{Dummy Variables}
-Dummy variables convert a categorical variable with multiple manifestations into several binary variables.
-%%
-Each binary variable encodes the presence of a manifestation.
-%%
-For example, a pain indicator variable ranging from \emph{1 - no pain} to \emph{4 - large pain} is subdivided into four binary variables. %(\emph{No pain - Yes/No} to \emph{Large Pain - Yes/no}).
+Dummy variables convert a categorical variable with multiple manifestations into several dichotomous variables.
+%%
+Each dichotomous variable encodes the presence of a manifestation.
+%%
+For example, a pain indicator variable ranging from \emph{1 - no pain} to \emph{4 - large pain} is subdivided into four dichotomous variables. %(\emph{No pain - Yes/No} to \emph{Large Pain - Yes/no}).
 %%
@@
 %%
-This allows for sequentially analyzing the classifications.
+This allows to sequentially analyze the classifications.
 
@@
 %%
-Selecting a variable using a drop down menu allows the user to select the variable used for subject subdivisions, e.g. \emph{gender} or \emph{disease treatment type}.
-%%
-Categorization using metric variables, such as \emph{body size}, is done using their quantiles.
+Selecting a variable using a drop down menu allows the user to select the variable used for subject subdivisions, e.g. \emph{gender} or \emph{employment status}.
+%%
+%Categorization using metric variables, such as \emph{body size}, is done using their quantiles.
+Metric variables, such as \emph{body size} are discretized using their quantiles.
 %%
@@
 %%
-It allows to combine both the ability to create comprehensible reports as well as UI-controls, to adjust visible data and variables in the plot to allow the viewer to analyze the data.
+It allows to combine the power of static \texttt{R Markdown} reports with dynamically parameterized \texttt{ggvis} plots.
+%It allows the combination of the ability to create comprehensible reports as well as adjusting parameters using user interface controls, to adjust visible data and variables in the plot to allow the viewer to analyze the data.
 %%
@@
 %%
-The web-based approach allows us to quickly exchange results with our collaborating epidemiological partners.
+The web-based approach allows us to quickly exchange results with our collaborating epidemiological experts.
 %%
@@
 %%
-The prototype is available under the following link: \url{blind.dnsalias.com}.
+The prototype is available under \url{blind.dnsalias.com}.
 %%---------
@@
 %no group (a), subdivision into \emph{male and female subjects} (b), quartiles of \emph{age} (c) and quartiles of \emph{body size} (d).
-no group (a), subdivision into \emph{male and female subjects} (b), quartiles of \emph{Body-Mass-Index} (\emph{BMI}) (c) and quartiles of \emph{body size} (d).
+no group (a), subdivision into \emph{male} and \emph{female} subjects (b), quartiles of \emph{Body-Mass-Index} (\emph{BMI}) (c) and quartiles of \emph{body size} (d).
 %%
@@
 	%%
-	\item subdivision into \emph{male} and \emph{female},
-	%%
-	\item subdivision into \emph{Body-Mass-Index quantiles} ($BMI = \frac{m}{l^2}$ where $m$ is the \emph{body mass} in kilogram and $l$ is the \emph{body size} in meter), yielding the groups (17, 24.7] (24.7, 27.4] (27.4, 30.5] (30.5, 48],
-	%%
-	\item subdivision into \emph{size quantiles}, yielding the groups (139, 164], (164, 171], (171, 177], (177, 202].
+	\item subdivision into \emph{males} and \emph{females},
+	%%
+	\item subdivision by \emph{Body-Mass-Index quantiles} ($BMI = \frac{m}{l^2}$ where $m$ is the \emph{body mass} in kilogram and $l$ is the \emph{body size} in meter), yielding the groups (17, 24.7] (24.7, 27.4] (27.4, 30.5] (30.5, 48],
+	%%
+	\item subdivision by \emph{size quantiles}, yielding the groups (139, 164], (164, 171], (171, 177], (177, 202].
 	%%
@@
 %%
-The second plot shows all categorical variables transformed into dichotomous dummy variables.
+The second plot shows all categorical variables transformed into dichotomous dummy variables (recall subsection~\ref{sec:C5.0}).
 
@@
 %%
-The distinctness lies in the average difference in body size between male and female subjects.
+The distinctness lies in the average difference in \emph{body size} between \emph{male} and \emph{female} subjects.
 %
@@
 %%
-The four body size groups could be characterized with an error $0.48$, but the decision tree comprises 71 rules and imposes overfitting.
-
-The analysis of the dummy variables yields similar results like the \emph{blood pressure medication}.
-%%
-Variables, such as subjects sized 139-164 cm, between 64 and 90 years of age or nutrition related variables are dominantly populated by one manifestation.
+The four \emph{body size} groups could be characterized with an error $0.48$, but the decision tree comprises 71 rules and imposes overfitting.
+
+The analysis of the dummy variables yields a result similar to the \emph{blood pressure medication}.
+%%
+Variables, such as subjects sized 139-164 cm, between 64 and 90 years of \emph{age} or \emph{nutrition} related variables are dominantly populated by one manifestation.
 %%
@@
 %%
-Classification using groups divided by gender do not produce satisfying results.
+Classification using groups divided by \emph{gender} do not produce satisfying results.
 %%
@@
 %%
-Gender could be classified for each \emph{BMI} group using \emph{mean curvature} and \emph{curvature angle}.
+\emph{Gender} could be classified for each \emph{BMI} group using \emph{mean curvature} and \emph{curvature angle}.
 %%
@@
 %%
-The starting age of smoking could be characterized well with an error between $0.25$ to $0.32$ for all \emph{BMI} groups except of subjects with BMI of \emph{$30.5-48$}.
+The starting age of smoking could be characterized well with an error between $0.25$ to $0.32$ for all \emph{BMI} groups except of subjects with \emph{BMI} of \emph{$30.5-48$}.
 %%
@@
 %%
-Using mostly \emph{mean curvature} and \emph{curvature angle}, the leg pain level can be predicted using $14$ rules with an error of $0.46$ for obese subjects (\emph{BMI} higher than $30$).
-%%
-The dichotomous variable whether the subject has felt pain in the last seven days can also be predicted for this group using the same features.
+Using mostly \emph{mean curvature} and \emph{curvature angle}, the \emph{leg pain level} can be predicted using $14$ rules with an error of $0.46$ for obese subjects (\emph{BMI} higher than $30$).
+%%
+The dichotomous variable whether the subject has felt \emph{pain in the last seven days} can also be predicted for this group using the same features.
 %%
@@
 %%
-Obese subjects are prone to back and leg pain due to a more stressed lumbar spine.
+Obese subjects are prone to \emph{back} and \emph{leg pain} due to a more stressed lumbar spine.
 %%
@@
 %%
-Many previously described results are confounded with subject size.
-%%
-Differences between \emph{male} and \emph{female} subjects are explained by the average \emph{body size} difference.
+Many previously described results are confounded by \emph{subject size}.
+%%
+Differences between \emph{male} and \emph{female} subjects can be explained by the average \emph{body size} difference.
 %%
@@
 %
-The question is, whether the inter-group spine-variability variable is enough for predicting other variables or not.
+The question is, whether the inter-group spine variability variable is enough for predicting other variables or not.
 %%
@@
 \paragraph{Large Decision Trees. }
-Back pain-associated variables can be explained for various \emph{size}-groups, but we could not extract universal rules.
-%%
-Radiating back pain could be described with error of $0.39$ using 23 rules for subjects between \emph{$171-177~cm$ body size} using \emph{torsion} and \emph{mean curvature}.
+\emph{Back pain}-associated variables can be explained for various \emph{size}-groups, but we could not extract universal rules.
+%%
+\emph{Radiating back pain} could be described with error of $0.39$ using 23 rules for subjects between \emph{$171-177~cm$ body size} using \emph{torsion} and \emph{mean curvature}.
 %%
@@
 %%
-They are well suited to characterize subject gender due to their body size difference using \emph{mean curvature} and \emph{mean curvature angle}.
-%%
-This is due to the higher curvature of smaller subjects and the significant difference of average \emph{body size} between \emph{males} and \emph{females}.
-%%
-Gender was characterized well in the analysis of all subjects as well as subjects divided by \emph{BMI}.
+They are well suited to characterize subject \emph{gender} due to their \emph{body size} difference using \emph{mean curvature} and \emph{mean curvature angle}.
+%%
+This is due to the higher curvature of smaller subjects and the significant difference of average \emph{body size} between males and females.
+%%
+\emph{Gender} was characterized well in the analysis of all subjects as well as subjects divided by \emph{BMI}.
 
@@
 %%
-Classification methods based on decision trees showed to be useful to assess the discriminative power of a variable set.
-%%
-Their ability to assess variable combination makes them more powerful than correlation coefficients calculated for each variable.
-%%
-This advantage comes with a much more complex output, making their results harder to assess and to abstract.
+Classification methods based on decision trees have proven to be useful for assessing the discriminative power of a variable set.
+%%
+Their ability to consider variable combination makes them more powerful than correlation coefficients calculated for each variable.
+%%
+This advantage comes with a much more complex output, the results are harder to assess and to abstract.
 %%
@@
 %%
-Huge result spaces could be navigated fast using our brushable visualization.
+Huge result spaces could be navigated fast using our interactive visualization.
 %%
@@
 %%
-Our method made it possible to assess their discriminative power, which is largely limited for separating \emph{male} and \emph{female} subjects, \emph{nutrition} variables, as well as different disease indicators.
+Our method made it possible to assess their discriminative power, which is largely limited for separating male and female subjects, \emph{nutrition} variables, as well as different disease indicators.
 %%
@@
 %%
-Over-fitting to the data indicated by complex decision trees have to be taken into account into the analysis as well.
+Over-fitting to the data indicated by complex decision trees have to be taken into account as well.
 %%
@@
 %%
-
-Combining the power of statistical analysis, visual analytics and advanced data mining techniques is essential for analyzing increasingly complex heterogenous population data.
+\\\\
+Combining the power of statistical analysis, visual analytics and data mining techniques is essential for analyzing increasingly complex heterogenous population data.
 %%

### 2014-09-18 draft.tex
Minor changes towards blind version

@@
 
-\author{\authorname{Paul Klemm\sup{1}, Sylvia Glaßer\sup{1}, Kai Lawonn\sup{1}, Marko Rak\sup{1} Henry Völzke\sup{2}, Katrin Hegenscheid\sup{3} and Bernhard Preim\sup{1}}
-\affiliation{\sup{1} Department of Simulation and Graphics, University of Magdeburg, Germany}
-\affiliation{\sup{2} Institute for Community Medicine, University of Greifswald, Germany}
-\affiliation{\sup{3} Institute for Diagnostic Radiology and Neuro-Radiology, University of Greifswald, Germany}
-\email{\{klemm, glasser\}@isg.cs.uni-magdeburg.de}
-}
+% TODO: BLIND Version
+\author{Blind}
+% \author{\authorname{Paul Klemm\sup{1}, Sylvia Glaßer\sup{1}, Kai Lawonn\sup{1}, Marko Rak\sup{1} Henry Völzke\sup{2}, Katrin Hegenscheid\sup{3} and Bernhard Preim\sup{1}}
+% \affiliation{\sup{1} Department of Simulation and Graphics, University of Magdeburg, Germany}
+% \affiliation{\sup{2} Institute for Community Medicine, University of Greifswald, Germany}
+% \affiliation{\sup{3} Institute for Diagnostic Radiology and Neuro-Radiology, University of Greifswald, Germany}
+% \email{\{klemm, glasser\}@isg.cs.uni-magdeburg.de}
+% }
 
@@
 \paragraph{Future Work.}
-\com{ToDo: Passen die Contributions hier wirklich?}
+%\com{ToDo: Passen die Contributions hier wirklich?}
 %% Contributions

### 2014-09-29 draft.tex
Revised Version from Anja

@@
 %%
-We analyze the predictive power of shape-describing variables for non-image variables, e.g., pain indicators.
+We analyze the predictive power of shape-describing variables for non-image variables, e.g. pain indicators.
 %%
@@
 %%
-Back pain is one of the most frequent diseases in the western civilization and gains importance due to the population aging.
+Back pain is one of the most frequent diseases in the western civilization and gains importance due to the population aging \cite{Hoy2010}.
 %%
@@
 %%
-	\item To prove a hypothesis, epidemiologists compile a list of variables depicting it.
+	\item Epidemiologists compile a list of variables depicting a hypothesis to prove it.
 %%
@@
 %%
-Many diseases, such as different cancer types, are more likely with increasing age.
+Many diseases, such as different cancer types, are more likely with increasing age \cite{Fletcher}.
 %%
@@
 %%
-These data can be analyzed regarding many diseases and conditions.
+The collected data can be analyzed regarding many diseases and conditions.
 %%
@@
 %%
-These data are hard to analyze.
+The analysis of these data is challenging.
+%These data are hard to analyze.
 %%
@@
 %%
-Zhang et al. analyze subject groups in a web-based linked view system \cite{Zhang}.
+\cite{Zhang} analyze subject groups in a web-based linked view system.
 %%
@@
 	%%
-	Women have a higher live expectance than men, hence their higher share in the old age group.
-	%%
-	Also, women are on average smaller than men, hence the larger shape similarity with older subjects.
+%	Women have a higher live expectance than men, hence their higher share in the old age group.
+	%%
+%	Also, women are on average smaller than men, hence the larger shape similarity with older subjects.
 	%%
@@
 \end{figure}
+%%
 \noindent Spine shape is confounded by several somatometric variables.
@@
 %%
+Women have a higher live expectance than men, hence their higher share in the old age group.
+%%
+Also, women are on average smaller than men, hence the larger shape similarity with older subjects.
+%%
+%Due to bone erosion, older people are also smaller on average.
+%%
 Large body weight increases the spine load, resulting in a bend shape.
@@
 To avoid small outlier groups, epidemiologists recommend using quartiles to discretize metric variables.
+%%
+\paragraph{GPLOM Analysis.}
+As a first experiment we correlated the shape variable with the dichotomous back pain indicator using \texttt{GPLOM's} \cite{GPLOMS}.
+%%
+%Pairs plots visualize continuous and categorical variables pairwise in a matrix view.
+%%
+%Scatterplots display pairs of continuous variables, bar charts for continuous variable as a function of a categorical variable and box plots pairs of categorical variables.
+%%
+Since our image-derived variables are metric, their pairwise combinations are visualized using scatterplots.
+%%
+The combination of the image variables with back pain is visualized as histogram at the left side of the matrix and as box plot on the right side.
+%%
+Figure~\ref{fig:image-parameter-range} (right) shows the range of each variable as box plot.
+%%
+The projections to the transversal planes attract attention as they have many outliers.
+%%
+We conclude that curvature is not as reliable on the transversal plane as it is on the other planes.
+%%--------------------------------
+\paragraph{Correlation Matrix.}
 \begin{figure*}[htb]
@@
 \end{figure*}
-%%
-\paragraph{GPLOM Analysis.}
-As a first experiment we correlated the shape variable with the dichotomous back pain indicator using \texttt{GPLOM's} \cite{GPLOMS}.
-%%
-%Pairs plots visualize continuous and categorical variables pairwise in a matrix view.
-%%
-%Scatterplots display pairs of continuous variables, bar charts for continuous variable as a function of a categorical variable and box plots pairs of categorical variables.
-%%
-Since our image-derived variables are metric, their pairwise combinations are visualized using scatterplots.
-%%
-The combination of the image variables with back pain is visualized as histogram at the left side of the matrix and as box plot on the right side.
-%%
-Figure~\ref{fig:image-parameter-range} (right) shows the range of each variable as box plot.
-%%
-The projections to the transversal planes attract attention as they have many outliers.
-%%
-We conclude that curvature is not as reliable on the transversal plane as it is on the other planes.
-%%--------------------------------
-\paragraph{Correlation Matrix.}
 We calculated an association matrix to assess correlations between the image variables.
@@
 \paragraph{The Error Term. }
-Normally the error for a classification is calculated with $error = \frac{correctlyClassified_{n}} {n}$ where $n$ is the number of subjects.
+Normally the error for a classification is calculated with $error = \frac{correctlyClassified_{n}} {total_{n}}$ where $n$ is the number of subjects.
 %%
@@
 \begin{equation}
-error = 1 - \frac{\sum_{m=0}^M \frac{correctlyClassified_{m}}{m}}{number_{dimensions}}
+error = 1 - \frac{\sum_{m=0}^M \frac{correctlyClassified_{m}}{total_{m}}}{number_{dimensions}}
 \end{equation}
@@
 %%
-We consider a result rated below $0.25$ a good classification.
+%We consider a result rated below $0.5$ an result of interest, a value below $0.25$ represents a good classification.
+We consider a result rated $0.5$ or below good enough to display it in the visualization, a value below $0.25$ represents a good classification.
 %%
@@
 %%
-The error varies between $0.31$ (\emph{BMI} of \emph{$30.5-48$}, 4 decision rules) to $0.39$ (\emph{BMI} of \emph{$24.7-27.4$}, 5 decision rules).
+The error varies between $0.31$ (\emph{BMI} of \emph{$30.5-48$}. 4 decision rules) to $0.39$ (\emph{BMI} of \emph{$24.7-27.4$}, 5 decision rules).
 %%
@@
 %
-The question is, whether the inter-group spine variability variable is enough for predicting other variables or not.
+The question is, whether the inter-group spine variability parameter is enough for predicting other variables or not.
 %%
@@
 %%
-This advantage comes with a much more complex output, the results are harder to assess and to abstract.
+This advantage comes with a much more complex output, the results are more challenging to assess and to abstract.
 %%
@@
 These methods do not aim to replace the traditional epidemiological workflow, but rather intervenes at the weak points of standard statistical methods.
+%%
+Our method provides a novel way to gain insight into these complex data sets and amplifies hypotheses generation.
 

### 2014-09-29 draft.tex
First revised Version from Henry

@@
 %%
-We assess the predictive quality of image-derived variables for lower back pain.
+%We assess the predictive quality of image-derived variables for lower back pain.
+We assess the relation between image-derived variables of the lumbar spine with lower back pain in a cross-sectional study.
+%We assess the predictive quality of image-derived variables for lower back pain.
 %%
@@
 %%
-We analyze the predictive power of shape-describing variables for non-image variables, e.g. pain indicators.
+%We analyze the predictive power of shape-describing variables for non-image variables, e.g. pain indicators.
+We analyze mutual dependencies among shape-describing variables and non-image variables, e.g. pain indicators.
 %%
@@
 %%
-As a first result, we conclude that image-based variables are only sufficient for classification of lifestyle factors.
-%%
-A correlation between lumbar spine shape and lower back pain disease was not existent in the automatically trained classifiers.
+As a first result, we conclude that image-based variables are only sufficient for classification of lifestyle factors within the data set.
+%%
+A correlation between lumbar spine shape and lower back pain was not existent in the automatically trained classifiers.
 %%
@@
 %%
-Hence, correlations between non-image variables were successfully detected and characterized.
+Hence, associations between non-image variables were successfully detected and characterized.
 }
@@
 %%
-These information are used to determine risk factors for diseases, help people to live healthier or to support the diagnosis of widespread diseases.
+This information is used to determine risk factors for diseases, help people to live healthier or to support the diagnosis of widespread diseases.
 %%
@@
 %%
-Back pain is one of the most frequent diseases in the western civilization and gains importance due to the population aging \cite{Hoy2010}.
-%%
-Epidemiological analysis of back pain aims to characterize the disease by determining risk factors.
+Back pain is one of the most frequent complaints in the western civilization and gains importance due to the population aging \cite{Hoy2010}.
+%%
+Epidemiological analysis of back pain aims to determine risk factors.
 %Many epidemiological-motivated approaches aim at improved characterization of this disease with respect to patient-specific attributes.
@@
 %%
-Although the shape and constitution of the spine, especially the lumbar spine, plays an important role for the presence of back pain, an automatic classification approach for characterization of back pain based on lumbar back spine attributes is still missing.
+Although the shape and constitution of the spine, especially the lumbar spine, plays an important role for back pain, an automatic classification approach for characterization of back pain based on lumbar back spine attributes is still missing.
 %%
@@
 
-We present a thorough analysis of the image-derived data of a cohort study including lumbar spine datasets.
-%%
-We analyze their discriminative power for different back pain characteristics.
+We present a analysis of the image-derived data of a cohort study including lumbar spine datasets.
+%%
+We extract associations between spine shape and back pain characteristics.
+%We analyze their discriminative power for different back pain characteristics.
 %%
@@
 %%
-%Then, interactive visual analysis provides insight into the quality of the image-derived data as well as their potential for possible back pain risk factors.
-Then, Interactive Visual Analysis (IVA) provides insight into the quality of the image-derived data as well as their potential for being back pain risk factors.
+%Then, Interactive Visual Analysis (IVA) provides insight into the quality of the image-derived data as well as their potential as back pain risk factors.
+Then, Interactive Visual Analysis (IVA) highlights mutual dependencies between image-derived data and back pain related variables.
 %%
@@
 %%
-\item The identification and evaluation of lumbar spine shape properties for back pain diagnosis,
-%%
-\item The detection of correlations between image-based and socio-demographic as well as medical variables for an improved hypothesis generation,
+\item The identification of lumbar spine shape properties associated with back pain,
+%\item The identification and evaluation of lumbar spine shape properties associated with back pain,
+%%
+\item The detection of associations between image-based, socio-demographic and medical variables for hypothesis generation,
 %%
@@
 %%
-Age acts as a \emph{confounder} for prostate cancer.
+Age acts as a \emph{risk factor} for prostate cancer.
 %%
@@
 %%
-For example, women-specific questions or data about a disease treatment only affects subjects suffering from this condition.
+%For example, women-specific questions or data about a disease treatment only affects subjects suffering from this condition.
+For example, data about a disease treatment only affects subjects suffering from this condition.
 %%
@@
 %%
-For example, continuous data, such as age, is often discretized into quantile bins.
+For example, continuous data, such as age, is often discretized into quantile bins for visualizing age-dependent disease frequencies.
 %%
@@
 %%
-Lower back pain is one of the most frequent diseases in the western civilization \cite{Hoy2010}.
+Lower back pain is one of the most frequent complaints in the western civilization \cite{Hoy2010}.
 %%
@@
 %%
-Potential \emph{confounding} effects are also subject of current research.
-%%
-
-Epidemiologists want to characterize the healthy aging process of the spine.
+%Potential \emph{confounding} effects are also subject of current research.
+%%
+
+%Epidemiologists want to characterize the healthy aging process of the spine.
+Epidemiologists want to characterize the relation between the aging and degenerative process of the spine \cite{Szpalski2005}. %healthy aging process of the spine.
 %%
@@
 %%
-Clustering techniques applied to non-image data are strongly dependent on the chosen variables and distance measures \cite{Klemm2014BVM}.
+%Clustering techniques applied to non-image data are strongly dependent on the chosen variables and distance measures \cite{Klemm2014BVM}.
+Clustering techniques are strongly dependent on the chosen variables and distance measures \cite{Klemm2014BVM}.
 %%
@@
 %%
-\noindent Spine shape is confounded by several somatometric variables.
+\noindent Spine shape is possibly associated with several somatometric variables.
 %%
@@
 %%
-Since men are on average taller than women and people of old age shrink due to bone erosion, gender and age are also confounders (Fig.~\ref{fig:age-gender}).
+Since men are on average taller than women and people of old age shrink due to bone erosion, gender and age are also risk factors (Fig.~\ref{fig:age-gender}).
 %%
@@
 %%
-All these variables need to be taken into account, when spine curvature and torsion is correlated with non-image variables.
-%%
-Since the gender confounder mainly encodes body size, we decided to divide subjects into body size groups.
-%%
-To avoid small outlier groups, epidemiologists recommend using quartiles to discretize metric variables.
+To assess their influence, we take them into account when spine curvature and torsion is correlated with non-image variables.
+%All these variables need to be taken into account, when spine curvature and torsion is correlated with non-image variables.
+%%
+Since the gender encodes body size, we decided to divide subjects into body size groups.
+%%
+Discretizing metric variables using quartiles avoids small outlier groups.
+%To avoid small outlier groups, epidemiologists recommend using quartiles to discretize metric variables.
 %%
@@
 %%
-Epidemiologists can interpret the results without having deep knowledge about the classification algorithm.
+%Epidemiologists can interpret the results without having deep knowledge about the classification algorithm.
 %%
@@
 
-Many follow-up variables are sparse, such as \emph{medication of diabetes} or \emph{reason of early retirement}.
+Many interdependent variables are sparse, such as \emph{medication of diabetes} or \emph{reason of early retirement}.
 %%
@@
 %%
-Many previously described results are confounded by \emph{subject size}.
+Many previously described results are influenced by \emph{subject size}.
 %%
@@
 %%
-Dividing subjects into \emph{body size} groups potentially highlights classifications not confounded by \emph{body size}.
+Dividing subjects into \emph{body size} groups potentially highlights classification not influenced by \emph{body size}.
+%%
+%Dividing subjects into \emph{body size} groups potentially highlights classifications not confounded by \emph{body size}.
 %%
@@
 
+The methods presented herein may be applied to comprehensive epidemiological data sets to learn more about mutual dependencies among variables and to generate hypotheses on potential associations and subgroups.
+%
+These hypotheses, however, have to be substantiated by dedicated statistical analyses and replication in independent cohorts.
+
 %In our future work, we will focus on more precise models for extracting measures described in Section~\ref{sec:ProposedFeatures}.

### 2014-09-29 draft.tex
Further changes using Henrys feedback

@@
 % \abstract{The abstract should summarize the contents of the paper and should contain at least 70 and at most 200 words.}
+% TODO: Auf 200 Wörter runterkürzen
 \abstract{ 
-The goal of epidemiology is to provide insight into disease causations.
+%The goal of epidemiology is to provide insight into disease causations.
+Epidemiology aims to provide insight into disease causations.
 %%
@@
 %We assess the predictive quality of image-derived variables for lower back pain.
-We assess the relation between image-derived variables of the lumbar spine with lower back pain in a cross-sectional study.
+We assess potential relations between image-derived variables of the lumbar spine with lower back pain in a cross-sectional study.
 %We assess the predictive quality of image-derived variables for lower back pain.
@@
 %%
-\item The identification of lumbar spine shape properties associated with back pain,
+\item The identification of lumbar spine shape properties potentially associated with back pain,
 %\item The identification and evaluation of lumbar spine shape properties associated with back pain,
@@
 %%
-%Every information reduction can introduce a bias to the data, since it proposes an assumption about the data.
-Every information reduction proposes an assumption about the data and can therefore introduce a bias.
-%%
-For example, using age to divide subjects into \emph{young} and \emph{old} categories can distort statistical results.
-%%
-This distortion is reduced with increasing number of discretization steps.
+%Every information reduction proposes an assumption about the data and can therefore introduce a bias.
+%%
+%For example, using age to divide subjects into \emph{young} and \emph{old} categories can distort statistical results.
+%%
+%This distortion is reduced with increasing number of discretization steps.
 %%

### 2014-09-30 draft.tex
weakened some parts of the Results section

@@
 \label{sec:Conclusion}
-\noindent The presented results indicate that the image based variables \emph{torsion}, \emph{curvature} and \emph{curvature angle} are not enough to characterize pathological changes in the lumbar spine.
-%%
-They are well suited to characterize subject \emph{gender} due to their \emph{body size} difference using \emph{mean curvature} and \emph{mean curvature angle}.
+\noindent The presented results indicate that the image based variables \emph{torsion}, \emph{curvature} and \emph{curvature angle} are not enough to characterize pathological changes in the lumbar spine in the given quality.
+%%
+They are associated with subject \emph{gender} due to their \emph{body size} difference using \emph{mean curvature} and \emph{mean curvature angle}.
 %%
@@
 %%
-\emph{Gender} was characterized well in the analysis of all subjects as well as subjects divided by \emph{BMI}.
+\emph{Gender} was detected well in the analysis of all subjects as well as subjects divided by \emph{BMI}.
 
@@
 %%
-For some \emph{body size} groups, certain \emph{nutrition} or \emph{psychological} variables can be predicted with a comparatively high error.
+Associations with certain \emph{nutrition} or \emph{psychological} variables can be observed with a comparatively high error for some \emph{body size} groups.
+%For some \emph{body size} groups, certain \emph{nutrition} or \emph{psychological} variables can be predicted with a comparatively high error.
 %%
@@
 \\\\
-Combining the power of statistical analysis, visual analytics and data mining techniques is essential for analyzing increasingly complex heterogenous population data.
+Combining the power of statistical analysis, visual analytics and data mining techniques are essential for analyzing increasingly complex heterogenous population data.
 %%

### 2014-09-30 draft.tex
Notes of Kais revisions, revised formula
Revision of Kai

@@
 \usepackage{xcolor}
-\usepackage[]{algorithm2e}
+\usepackage[ruled,vlined]{algorithm2e}
 \usepackage[utf8]{inputenc}
@@
 % $\sum \frac{curvature_i}{curvature_all}$.
-	\item \emph{Normalized Curvature} is calculated as weighted sum of curvature between all adjacent points describing the centerline: $\sum_{i=1}^I \frac{curvature_i}{i}$.
+	\item \emph{Mean Curvature} is calculated as weighted sum of curvature between all adjacent points describing the centerline: $\sum_{i=1}^I \frac{curvature_i}{I}$.
 We refer to total curvature as \emph{curvature}.
 %%
-	\item \emph{Normalized Torsion} (deviation of a curve from its current course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum_{i=1}^I \frac{torsion_i}{i}$.
+	\item \emph{Mean Torsion} (deviation of a curve from its current course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum_{i=1}^I \frac{torsion_i}{I}$.
 We refer to total torsion from now on as \emph{torsion}.
@@
 	%%
-	The shape color encodes the distance to the overall male (top chart) or female (bottom chart) mean shape.
+	The shape color encodes the distance (red for x axis, blue for y axis, green for z axis) to the overall male (top chart) or female (bottom chart) mean shape.
 	%%
@@
 \paragraph{The Error Term. }
-Normally the error for a classification is calculated with $error = \frac{correctlyClassified_{n}} {total_{n}}$ where $n$ is the number of subjects.
+\com{Kai}
+Normally the error for a classification is calculated with $error = \frac{correctlyClassified_{n}} {N}$ where $n$ is the number of subjects.
 %%
@@
 \begin{equation}
-error = 1 - \frac{\sum_{m=0}^M \frac{correctlyClassified_{m}}{total_{m}}}{number_{dimensions}}
+error = 1 - \frac{\sum_{m=1}^M \frac{correctlyClassified_{m}}{M}}{number_{dimensions}}
 \end{equation}
@@
 %%
-The vast majority of non-image variables can not be described well using the classifier.
+The vast majority of non-image variables can not be automatically classified based on the image-variables.
 %%

### 2014-10-01 draft.tex
Minor Revisions at the Introduction

@@
 %%
-Furthermore, we provide an IVA technique to compare the classifiers with interactive scatterplot visualizations.
+We provide an IVA technique to compare classifiers with interactive scatterplot visualizations.
 %%
@@
 %%
-\item The identification of the most important variables via data mining methods (including a novel semi-quantitative evaluation concept) embedded in our IVA-framework
+%\item The identification of the most important variables via data mining methods (including a novel semi-quantitative evaluation concept) embedded in our IVA-framework
+\item The identification of the most important variables via data mining methods using a novel semi- quantitative overview visualization concept embedded in our IVA-framework
 %%

### 2014-10-01 draft.tex
First round of Steffis revision

@@
 
-\keywords{Epidemiology, Interactive Visual Analysis, Classification, Multi Modal Data}
+\keywords{Epidemiology, Interactive Visual Analysis, Classification, Multi-Modal Data}
 
@@
 %%
-We provide an IVA technique to compare classifiers with interactive scatterplot visualizations.
+We provide an IVA technique to compare classifiers with interactive scatter plot visualizations.
 %%
@@
 %%
-Back pain is one of the most frequent complaints in the western civilization and gains importance due to the population aging \cite{Hoy2010}.
+Back pain is one of the most frequent complaints in the Western civilization and gains importance due to the population aging \cite{Hoy2010}.
 %%
@@
 
-We present a analysis of the image-derived data of a cohort study including lumbar spine datasets.
+We present an analysis of the image-derived data of a cohort study including lumbar spine datasets.
 %%
@@
 %Then, Interactive Visual Analysis (IVA) provides insight into the quality of the image-derived data as well as their potential as back pain risk factors.
-Then, Interactive Visual Analysis (IVA) highlights mutual dependencies between image-derived data and back pain related variables.
-%%
-The workflow is part of an IVA framework, constructed to enhance the epidemiological workflow with iterative data driven techniques.
+Then, Interactive Visual Analysis (IVA) highlights mutual dependencies between image-derived data and back pain-related variables.
+%%
+The workflow is part of an IVA framework, constructed to enhance the epidemiological workflow with iterative data-driven techniques.
 %%
@@
 %\item The identification of the most important variables via data mining methods (including a novel semi-quantitative evaluation concept) embedded in our IVA-framework
-\item The identification of the most important variables via data mining methods using a novel semi- quantitative overview visualization concept embedded in our IVA-framework
+\item The identification of the most important variables via data mining methods using a novel semi- quantitative overview visualization concept embedded in our IVA-framework.
 %%
@@
 %%
-The analysis, as well as the presented IVA tool is available as supplementary material in a interactive \texttt{R Markdown}\footnote{Developed by RStudio, Inc; \texttt{rmarkdown.rstudio.com}} document (\url{blind.dnsalias.com}).
+The analysis, as well as the presented IVA tool is available as supplementary material in an interactive \texttt{R Markdown}\footnote{Developed by RStudio, Inc; \texttt{rmarkdown.rstudio.com}} document (\url{blind.dnsalias.com}).
 %%
@@
 %%
-For example, the effect of nutrition on prostate cancer requires age-normalization due to the increased incidence for older subjects.
+For example, the effect of nutrition on prostate cancer requires age normalization due to the increased incidence for older subjects.
 %%
@@
 
-Epidemiological data acquisition includes a variety of techniques, such as medical examinations, self reported questionnaires or genetic examinations.
+Epidemiological data acquisition includes a variety of techniques, such as medical examinations, self-reported questionnaires or genetic examinations.
 %Epidemiologists acquire data using a wide range of techniques, such as medical examinations, self reported questionnaires or genetic examinations.
@@
 %%
-Lower back pain is one of the most frequent complaints in the western civilization \cite{Hoy2010}.
+Lower back pain is one of the most frequent complaints in the Western civilization \cite{Hoy2010}.
 %%
@@
 %%
-Their proposed \emph{deviation plot} shows distribution-specific measures of a variable, such as skewness or inter-quantile-range, making variables comparable.
+Their proposed \emph{deviation plot} shows distribution-specific measures of a variable, such as skewness or inter-quantile range, making variables comparable.
 %%
@@
 %%
-The plot depends on the type of variables, which are pairwise visualized.
+The plot depends on the type of variables, which are visualized pairwise.
 %%
@@
 %%
-They present an interactive data mining tool, which can analyze association rules and highlight interesting relations, which may trigger new hypothesis about the data.
+They present an interactive data mining tool, which can analyze association rules and highlight interesting relations, which may trigger new hypotheses about the data.
 %%
@@
 \\\\
-Unique for our work is the combination of data mining techniques with an IVA-approach by observing interesting feature relations in the context of image-derived features using multiple decision trees for one visualization.
+Unique in our work is the combination of data mining techniques with an IVA approach by observing interesting feature relations in the context of image-derived features using multiple decision trees for one visualization.
 %%
@@
 %%
-%Without the 9 image-derived parameter, the data set comprises of 134 variables:
-The data set comprises of 134 variables (and 9 additional image-derived variables described in the next subsection):
+The data set comprises 134 variables (and 9 additional image-derived variables described in the next subsection):
 \begin{itemize}
@@
 %%
-A click on the L3 vertebra center initializes the model, two clicks on the top and bottom of the vertebra describes an initial height estimation (Fig.~\ref{fig:centerline}~a).
+A click on the L3 vertebra center initializes the model, two clicks on the top and bottom of the vertebra describe an initial height estimation (Fig.~\ref{fig:centerline}~a).
 %%
@@
 %%
-Each subjects' model is mapped to a common reference frame using procrustes analysis to compensate for differences in translation, 
+Each subject's model is mapped to a common reference frame using procrustes analysis to compensate for differences in translation, 
 rotation and scaling.
@@
 	\item \emph{Mean Torsion} (deviation of a curve from its current course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum_{i=1}^I \frac{torsion_i}{I}$.
-We refer to total torsion from now on as \emph{torsion}.
+From now on, we refer to total torsion as \emph{torsion}.
 %%
@@
 %%
-Women have a higher live expectance than men, hence their higher share in the old age group.
+Women have a higher life expectancy than men, hence their higher share in the old age group.
 %%
@@
 %%
-Large body weight increases the spine load, resulting in a bend shape.
+Large body weight increases the spine load, resulting in a bent shape.
 %%
@@
 %%
-Since our image-derived variables are metric, their pairwise combinations are visualized using scatterplots.
+Since our image-derived variables are metric, their pairwise combinations are visualized using scatter plots.
 %%
@@
 %%
-Their correlation with back pain is denoted right of the matrix diagonal.
-%%
-The box plots (right) and histograms (bottom) display the distribution of each image-variable encoded with back pain.
+Their correlation with back pain is denoted to the right of the matrix diagonal.
+%%
+The box plots (right) and histograms (bottom) display the distribution of each image variable encoded with back pain.
 %%
@@
 %%
-The complete analysis can be found under the experiments section of the supplementary material (\url{blind.dnsalias.com})
+The complete analysis can be found in the experiments section of the supplementary material (\url{blind.dnsalias.com})
 %% ------------------------------------------
@@
 %%
-We calculated the contingency matrix for all size groups and looked for correlations between image- and non-image derived variables.
+We calculated the contingency matrix for all size groups and searched for correlations between image- and non-image derived variables.
 %%
@@
 %%
-One surprising result was the small correlation of \emph{torsion} with \emph{parkinson} (0.24).
+One surprising result was the small correlation of \emph{torsion} with \emph{Parkinson's disease} (0.24).
 %%
@@
 %%
-For example, if a variable indicating a disease is negative for 90\% of the subjects and the classifier simply assigns all subjects to \emph{not ill} the error metric would yield an error of 10\%, even though it is very bad.
+For example, if a variable indicating a disease is negative for 90\% of the subjects and the classifier simply assigns all subjects to \emph{not ill}, the error metric would yield an error of 10\%, even though it is very bad.
 %%
@@
 %%
-The scatter plot axis are defined by tree size and the previously described error metric.
+The scatter plot axes are defined by tree size and the previously described error metric.
 %%
@@
 %We removed variables with less than 100 subjects, since they have too few subjects to be statistically significant.
-Therefore we provide a way to adjust the minimal number of subjects for each variable using a slider input.
+Therefore, we provide a way to adjust the minimal number of subjects for each variable using a slider input.
 %%
@@
 %%
-Selecting a variable using a drop down menu allows the user to select the variable used for subject subdivisions, e.g. \emph{gender} or \emph{employment status}.
+Selecting a variable using a drop-down menu allows the user to select the variable used for subject subdivisions, e.g. \emph{gender} or \emph{employment status}.
 %%
@@
 \subsubsection{Implementation}
-All analysis are carried out using \texttt{R}, a widely used programming language for statistical calculations and visualizations.
+All analyses are carried out using \texttt{R}, a widely used programming language for statistical calculations and visualizations.
 %%
@@
 %%
-The prototype is available under \url{blind.dnsalias.com}.
+The prototype is available at \url{blind.dnsalias.com}.
 %%---------
@@
 %%
-Interactive scatterplot visualization of all C5.0 classification results.
+Interactive scatter plot visualization of all C5.0 classification results.
 %%
@@
 %%
-The left scatterplot shows the results for all variables, either metrics expressed via their quantiles, or categorical.
-%%
-The right scatterplot displays the dummy variables derived from the original variables.
+The left scatter plot shows the results for all variables, either metrics expressed via their quantiles, or categorical.
+%%
+The right scatter plot displays the dummy variables derived from the original variables.
 %%
@@
 	%%
-	\item subdivision into \emph{males} and \emph{females},
-	%%
-	\item subdivision by \emph{Body-Mass-Index quantiles} ($BMI = \frac{m}{l^2}$ where $m$ is the \emph{body mass} in kilogram and $l$ is the \emph{body size} in meter), yielding the groups (17, 24.7] (24.7, 27.4] (27.4, 30.5] (30.5, 48],
+	\item subdivision into \emph{male} and \emph{female},
+	%%
+	\item subdivision by \emph{Body Mass Index quantiles} ($BMI = \frac{m}{l^2}$ where $m$ is the \emph{body mass} in kilogram and $l$ is the \emph{body size} in meter), yielding the groups (17, 24.7] (24.7, 27.4] (27.4, 30.5] (30.5, 48],
 	%%
@@
 %%
-The vast majority of non-image variables can not be automatically classified based on the image-variables.
+The vast majority of non-image variables cannot be automatically classified based on the image-variables.
 %%
@@
 %
-It can be discriminated with an error of $0.31$ using 7 rules and incorporates only \emph{curvature} and \emph{curvature angle} related variables.
+It can be discriminated with an error of $0.31$ using 7 rules and incorporates only \emph{curvature-} and \emph{curvature angle}-related variables.
 %%
@@
 %%
-Almost all subjects who are medicated ($796$ of $1,058$) were correctly classified, the vast majority of non-medicated subjects ($262$ of $1,058$) are false-positive classified, yielding a poor quality of the classifier w.r.t epidemiological research.
+Almost all subjects who are medicated ($796$ of $1,058$) were correctly classified.
+%%
+The vast majority of non-medicated subjects ($262$ of $1,058$) are false-positive classified, yielding a poor quality of the classifier w.r.t epidemiological research.
 %%This is however due to the high amount of people, who are medicated (68\%) and who are almost all correctly class
@@
 %%
-The four \emph{body size} groups could be characterized with an error $0.48$, but the decision tree comprises 71 rules and imposes overfitting.
+The four \emph{body size} groups could be characterized with an error of $0.48$, but the decision tree comprises 71 rules and imposes overfitting.
 
@@
 %%
-Variables, such as subjects sized 139-164 cm, between 64 and 90 years of \emph{age} or \emph{nutrition} related variables are dominantly populated by one manifestation.
+Variables, such as subjects sized 139-164 cm, between 64 and 90 years of \emph{age} or \emph{nutrition}-related variables are dominantly populated by one manifestation.
 %%
@@
 %\com{Potentially skip these due to the confounding of body size}
-\subsection{Body-Mass-Index Groups}
+\subsection{Body Mass Index Groups}
 %%
@@
 %%
-The error varies between $0.31$ (\emph{BMI} of \emph{$30.5-48$}. 4 decision rules) to $0.39$ (\emph{BMI} of \emph{$24.7-27.4$}, 5 decision rules).
-%%
-The starting age of smoking could be characterized well with an error between $0.25$ to $0.32$ for all \emph{BMI} groups except of subjects with \emph{BMI} of \emph{$30.5-48$}.
+The error varies between $0.31$ (\emph{BMI} of \emph{$30.5-48$}, 4 decision rules) to $0.39$ (\emph{BMI} of \emph{$24.7-27.4$}, 5 decision rules).
+%%
+The starting age of smoking could be characterized well with an error between $0.25$ and $0.32$ for all \emph{BMI} groups except for subjects with a \emph{BMI} of \emph{$30.5-48$}.
 %%

### 2014-10-01 draft.tex
Included Steffis revision, cleaned up old parts

@@
 \paragraph{The Error Term. }
-\com{Kai}
 Normally the error for a classification is calculated with $error = \frac{correctlyClassified_{n}} {N}$ where $n$ is the number of subjects.
@@
 %%
-%Therefore the discriminative power of each dimension in the equation
-%error = 1 - \frac{\[ \sum_{m=0}^M \] \frac{M_{classifiedCorrectly}}{M_all}}{number of dimensions}
 \begin{equation}
@@
 The error is scaled to denote perfect classification with 0; 1 is equal to random selection.
-% The error rate has to be judged depending on the number of classes of the target variable.
-% %%
-% An error rate of 50\% for a target variable with two manifestations makes the classifier equal to throwing a coin.
-% %%
-% The same error rate is acceptable a target variable with more than two manifestations.
-% %%
-% Picking a value for a variable of four manifestations with an error rate of 50\% is better than the random selection (error rate of 75\%).
-% %%
-% The following scales the error rate so that 0 denotes perfect classification and 1 is equal to random selection.
-% %%
-% \begin{equation}
-% error_{normalized} = \frac{error}{100 - \frac{100}{class number}}
-% \end{equation}
-%%
-%%
-%We consider a result rated below $0.5$ an result of interest, a value below $0.25$ represents a good classification.
+%%
 We consider a result rated $0.5$ or below good enough to display it in the visualization, a value below $0.25$ represents a good classification.
@@
 %%
-%If we for example classify and compare the same variables for different subject groups, for example male and female subjects.
 Classification and comparison of variables for subject groups (e.g. male and female subjects) in one plot can be achieved by color coding group affiliation on the data points.
-%Classification of the same variable 
-%%
-%We can support this by color coding the data points according to group affiliation.
 
@@
 %%
-%We removed variables with less than 100 subjects, since they have too few subjects to be statistically significant.
 Therefore, we provide a way to adjust the minimal number of subjects for each variable using a slider input.
@@
 Outlier results with very large decision trees would otherwise distort the resulting plot.
-%%
-
-%\com{ToDo: SQRT Skalierung der Treesize-Dimension, da hauptsaechlich geringe Groessen von interesse sind, ausreisser aber diesen Bereich schwer durchschaubar machen}
-% %%
-%
-% %%
-% Abstracting the classification results into two continuos variables allows us to visualize the result of 134 classifications in one scatter plot.
-% %%
-% By mapping the group affiliation (in our case body size) to color or shape, we can even plot all 402 results in a single visualization (Fig.~\ref{fig:decision-tree} a).
 %%
@@
 %%
-For example, a pain indicator variable ranging from \emph{1 - no pain} to \emph{4 - large pain} is subdivided into four dichotomous variables. %(\emph{No pain - Yes/No} to \emph{Large Pain - Yes/no}).
+For example, a pain indicator variable ranging from \emph{1 - no pain} to \emph{4 - large pain} is subdivided into four dichotomous variables.
 %%
@@
 This is useful for our classification, because it allows to determine which manifestation of a variable can be described best using the image data variable.
-%%\com{ToDo: Dummy-Variables}
 %%------------
@@
 %%
-%Categorization using metric variables, such as \emph{body size}, is done using their quantiles.
 Metric variables, such as \emph{body size} are discretized using their quantiles.
@@
 It allows to combine the power of static \texttt{R Markdown} reports with dynamically parameterized \texttt{ggvis} plots.
-%It allows the combination of the ability to create comprehensible reports as well as adjusting parameters using user interface controls, to adjust visible data and variables in the plot to allow the viewer to analyze the data.
 %%
 Furthermore, calculations based on a prior data selection can be redone within the report.
-%We implemented our application as web based prototype using \texttt{R Shiny}\footnote{R Shiny footnote} in combination with \texttt{RMarkdown}.
-%%
-%The package combines the power of \texttt{R} with advantages of web-based applications.
 %%
@@
 %%
-%no group (a), subdivision into \emph{male and female subjects} (b), quartiles of \emph{age} (c) and quartiles of \emph{body size} (d).
-no group (a), subdivision into \emph{male} and \emph{female} subjects (b), quartiles of \emph{Body-Mass-Index} (\emph{BMI}) (c) and quartiles of \emph{body size} (d).
-%%
-%(a) displays all subjects, (b) is divided into male and female subjects, (c) is categorized into quantiles of age and (d) shows the classification of subjects divided by quantiles of body size.
+no group (a), subdivision into \emph{male} and \emph{female} subjects (b), quartiles of \emph{Body Mass Index} (\emph{BMI}) (c) and quartiles of \emph{body size} (d).
 %%
@@
 %%
-%A interactive versions of the plot (see supplemental material) makes all data points clickable, showing the corresponding decision tree in a tool tip.
 The interactive plot (see supplemental material) has clickable data points, displaying corresponding decision tree in a tool tip.
-%A interactive versions of the plot (see supplemental material) makes all data points clickable, showing the corresponding decision tree in a tool tip.
 %%
@@
 %%
-%The \emph{BMI} is calculated using the formula $BMI = \frac{m}{l^2}$ where $m$ is the \emph{body mass} in kilogram and $l$ is the \emph{body size} in meter.
-%%
 We plotted each group twice.
@@
 %%
-%The first impression of the resulting scatter plots quickly shows a very high amount of variables above an error rate of $0.6$.
-%%
-
-% None of the pain indicators can be reliably described using the image-based parameter.
+
 None of the pain indicators can be described reliably using the image-based variables.
@@
 The distinctness lies in the average difference in \emph{body size} between \emph{male} and \emph{female} subjects.
-%
-%It mostly incorporates \emph{mean curvature}, \emph{sagittal curvature angle} und sagittal \emph{mean curvature}.
-%The perfect categorization of the \emph{dialysis} parameter was conspicuous.
-%%
-%This is due to the subject distribution among the manifestation.
-%%
-%Only two subjects receive a dialysis.
-%%
-%The good discrimination is not meaningful.
 %%
@@
 %%
-%1,058 subjects are classified with an error rate of $0,47$ solely based on coronal mean curvature.
-%%
 Almost all subjects who are medicated ($796$ of $1,058$) were correctly classified.
@@
 The vast majority of non-medicated subjects ($262$ of $1,058$) are false-positive classified, yielding a poor quality of the classifier w.r.t epidemiological research.
-%%This is however due to the high amount of people, who are medicated (68\%) and who are almost all correctly class
-%%
-%Other features well classified include marital status, foot pain and number of born children.
 %%
@@
 The classifier neglects the other group and yields an error below $0.5$.
-%The dummy variable analysis yields fewer results.
-%%
-%It shows that subjects between 64 and 90 years of age can be detected with 23\% error rate by using the curvature angle and coronal mean curvature.
-%%
-%Subjects with body size between 177 cm and 202 cm can classified with an error rate of 22,1\%. solely using mean curvature.
 %%
@@
 The dummy variable analysis showed that female subjects of \emph{139-164~cm body size} could be discriminated using the \emph{mean curvature} and \emph{curvature angle}, with an error of $0.38$.
-
-%\com{Potentially skip these due to the confounding of body size}
+%%
 \subsection{Body Mass Index Groups}
@@
 %%
-The result however is probably overfitted to the data due to tree sizes between $14$ and $16$.
-%%
-
-Some variables, such as \emph{body size} can be described with an error of $0.3$ to $0.36$ but only using large decision trees with over 20 rules.
-%%
-%Notable is also the decreasing error with increasing decision tree size.
+The result however, is probably overfitted to the data due to tree sizes between $14$ and $16$.
+%%
+
+Some variables, such as \emph{body size}, can be described with an error of $0.3$ to $0.36$, but only using large decision trees with over 20 rules.
 %%
@@
 %%
-The stress induced spine deformation seems to influence the pain levels for these subjects directly.
-%How much the spine is deformed under the heavy stress seems to directly influence the pain levels for these subjects.
+The stress-induced spine deformation seems to directly influence the pain levels for these subjects.
 
@@
 %%
-%It shows that mean curvature can be used to predict dependencies to variables, such as \emph{high blood pressure}, \emph{hypothyroidism} or \emph{nutrition}.
-
-% \subsubsection{Age Groups}
-% %%
-% Gender could be classified for each age group using \emph{mean curvature} and \emph{curvature angle}.
-% %%
-% The accuracy varies between $0.35$ (\emph{$43.8-54$ years of age}, 3 decision rules) to $0.27$ (\emph{$21-43.8$ years of age}, 6 decision rules).
-% %%
-% The starting age of smoking could be characterized well with an error rate between $0.25$ to $0.3$ for all \emph{age} groups except of \emph{$21.1-43.8$ years of age}.
-% %%
-% The result however is probably over fitted to the data due to tree sizes between $16$ and $23$.
-% %%
-% \com{Warum?}
-% %%
-%
-% Many variables, such as \emph{body size} can be described with an error rate of $0.35$ to $0.45$ but only using large decision trees with over 20 rules.
-% %%
-% Notable is also the increasing accuracy with increasing decision tree size.
-% %%
-% For subjects between \emph{$43.8-54$ years of age}, the classifier discriminated subjects with thyroid nodules with an error rate of $0.32$.
-%
-% The dummy variable analysis shows many results using a decision tree with one rule based on \emph{mean curvature} with accuracy between $0.4$ and $0.5$.
-% %%
-% It shows that mean curvature can be used to predict dependencies to variables, such as \emph{high blood pressure}, \emph{hypothyroidism} or \emph{nutrition}.
-
 \subsection{Size Groups}
@@
 %
-The question is, whether the inter-group spine variability parameter is enough for predicting other variables or not.
-%%
-Dividing subjects into \emph{body size} groups potentially highlights classification not influenced by \emph{body size}.
-%%
-%Dividing subjects into \emph{body size} groups potentially highlights classifications not confounded by \emph{body size}.
+The question is whether the inter-group spine variability parameter is sufficient for predicting other variables or not.
+%%
+Dividing subjects into \emph{body size} groups potentially highlights classifications not influenced by \emph{body size}.
 %%
 \paragraph{Large Decision Trees. }
-\emph{Back pain}-associated variables can be explained for various \emph{size}-groups, but we could not extract universal rules.
-%%
-\emph{Radiating back pain} could be described with error of $0.39$ using 23 rules for subjects between \emph{$171-177~cm$ body size} using \emph{torsion} and \emph{mean curvature}.
+\emph{Back pain}-associated variables can be explained for various \emph{size} groups, but we could not extract universal rules.
+%%
+\emph{Radiating back pain} could be described with an error of $0.39$ using 23 rules for subjects between \emph{$171-177~cm$ body size} with \emph{torsion} and \emph{mean curvature}.
 %%
@@
 %%
-The dummy variables show several variables described using only one decision rule with an error between $0.42$ to $0.47$.
+The dummy variables show several variables described using only one decision rule with an error between $0.42$ and $0.47$.
 %%
@@
 These variables include \emph{nutrition}, \emph{thyroid disorder} and \emph{social problems induced by back pain}.
-
-%The resulting trees consist either only one rule or a rich set of them (>15).
-%Either they divide the variables using only one decision rule, or a large set of rules (> 15).
-
-% \com{Manchmal kann es passieren, dass die Quantile fuer eine Variable gleich sind - z.B.
-bei Dauer Jahre Schwerarbeit, weil viele Probanden hier eben eine 0 haben.
-Deswegen kann das Ding nicht in vier Qunatile unterteilt werden, die ersten 3 sind 0.
-Deswegen wurden nur die einzigartigen Unterscheidungen genommen (statt \emph{quantile(myVar)} wurde \emph{unique(quantile(myVar))} verwendet.)}
-
-%TODO: Muss das wirklich noch rein?
-%\com{Auswertung der Anteile der Parameter an den Ergebnissen (Torsion, Curvature, etc. - welches ist der dominanteste Parameter?)}
-
+%%
+%TODO: Muss das rein? Auswertung der Anteile der Parameter an den Ergebnissen (Torsion, Curvature, etc. - welches ist der dominanteste Parameter?)
+%%
 \section{\uppercase{Conclusion}}
 \label{sec:Conclusion}
-\noindent The presented results indicate that the image based variables \emph{torsion}, \emph{curvature} and \emph{curvature angle} are not enough to characterize pathological changes in the lumbar spine in the given quality.
+\noindent The presented results indicate that the image-based variables \emph{torsion}, \emph{curvature} and \emph{curvature angle} are not sufficient to characterize pathological changes in the lumbar spine in the given quality.
 %%
@@
 %%
-This is due to the higher curvature of smaller subjects and the significant difference of average \emph{body size} between males and females.
+This is due to the higher curvature of smaller subjects and the significant difference of average \emph{body size} between male and female.
 %%
@@
 %%
-The remaining information within these groups is not enough to characterize back pain-related variables.
+The remaining information within these groups is not sufficient to characterize back pain-related variables.
 %%
 Associations with certain \emph{nutrition} or \emph{psychological} variables can be observed with a comparatively high error for some \emph{body size} groups.
-%For some \emph{body size} groups, certain \emph{nutrition} or \emph{psychological} variables can be predicted with a comparatively high error.
 %%
@@
 %%
-Their complexity impose overfitting to the data and they are not suitable for extracting universal rules.
+Their complexity imposes overfitting to the data and they are not suitable for extracting universal rules.
 %%
@@
 %%
-Their ability to consider variable combination makes them more powerful than correlation coefficients calculated for each variable.
+Their ability to consider variable combinations makes them more powerful than correlation coefficients calculated for each variable.
 %%
@@
 %%
-Therefore the method is applicable not only for deriving information based on image data, but on all potential target variables.
+Therefore, the method is applicable not only for deriving information based on image data, but on all potential target variables.
 %%
@@
 %%
-This allows both to assess the discriminative power of the data set regarding the target variables, as well as the most important variables for the classification.
-%Except for gender in the age-plot, no variable could be explained with reasonable accuracy for age or body size.
+This allows to assess both the discriminative power of the data set regarding the target variables, and as the most important variables for the classification.
+%%
 \paragraph{Predictive Power of Image-Derived Variables.}
-We showed, that \emph{torsion}, \emph{curvature} and \emph{curvature angle} of the lumbar spine at the presented precision are not enough to characterize lumbar back pain in the SHIP data set.
-%%
-Our method made it possible to assess their discriminative power, which is largely limited for separating male and female subjects, \emph{nutrition} variables, as well as different disease indicators.
-%%
-The C4.5 algorithm showed to be an effective tool for evaluating a set of derived metrics regarding their suitability to classify non-image variables.
-%%
-Over-fitting to the data indicated by complex decision trees have to be taken into account as well.
-%%
-% \subsection{Proposed Features}
-% \label{sec:ProposedFeatures}
-% \begin{figure}[!h]
-%   %\vspace{-0.2cm}
-%   \centering
-%   \includegraphics[width=0.475\textwidth]{figures/proposed-features}
-%   \caption{
-% 	Proposed features.
-% 	%%
-% 	(a) dented vertebraes are a sign of heavy stress and bone absorption.
-% 	%%
-% 	(b, c) the spine canal shape can be used to characterize scoliosis (curvature sagittal) and lordosis (curvature coronal) as well as the \emph{Cobb-Angles} \cite{Cobb}.
-% 	%%
-% 	(c) Spine canal thickness is associated with herniated disks.
-% 	%%
-% 	}
-%   \label{fig:proposedFeatures}
-% \end{figure}
-% %%
-% \noindent Pathological deformation of the lumbar spine is usually the last stage and is associated with very strong lower back pain.
-% %%
-% We have to capture earlier signs of pathological change in order to derive better predictive models.
-% %%
-%
-% A very early sign of pathological deformation is the \emph{bone resorption in the center of a vertebra}.
-% %%
-% They change from being block-shaped to be dented in the center (Fig~\ref{fig:proposedFeatures}~a).
-% %%
-% This information can be obtained by segmenting the whole vertebra or the top and bottom point if each vertebra center.
-% %%
-% Another valuable variable would be the \emph{spine canal thickness} (Fig~\ref{fig:proposedFeatures}~c).
-% %%
-% Low spine canal thickness can be an indicator for an impending herniated disk.
-% %%
-% Both surface texture of the vertebra and thickness of the spine canal are used to diagnose herniated disks.
-% %%
-% The \emph{overall spine canal shape} is also of interest, since scoliosis and lordosis can be characterized more precise by deriving the Cobb angles from this shape (Fig~\ref{fig:proposedFeatures}~b,~c).
-%% ------------------------------------
+We showed that \emph{torsion}, \emph{curvature} and \emph{curvature angle} of the lumbar spine at the presented precision are not sufficient to characterize lumbar back pain in the SHIP data set.
+%%
+Our method allowed to assess their discriminative power, which is largely limited for separating male and female subjects, \emph{nutrition} variables, as well as different disease indicators.
+%%
+The C4.5 algorithm proved to be an effective tool for evaluating a set of derived metrics regarding their suitability to classify non-image variables.
+%%
+Over-fitting to the data indicated by complex decision trees has to be taken into account as well.
+%%
 \paragraph{Future Work.}
-%\com{ToDo: Passen die Contributions hier wirklich?}
 %% Contributions
@@
 %%
-We applied the technique to derive insight into the predictive power of 9 image-derived variables for 134 non-image variables.
+We applied the technique to gain insight into the predictive power of 9 image-derived variables for 134 non-image variables.
 %%
@@
 
-%In our future work, we will focus on more precise models for extracting measures described in Section~\ref{sec:ProposedFeatures}.
 In our future work, we will focus on more precise models for extracting measures.
 %%
-Dented vertebras are a early sign of a pathological deformation, so we want to capture this by segmenting the top and bottom point of each vertebra center.
+Dented vertebras are an early sign of a pathological deformation, so we want to capture this by segmenting the top and bottom point of each vertebra center.
 %%

### 2014-10-01 draft.tex
Final Revision

@@
 % \abstract{The abstract should summarize the contents of the paper and should contain at least 70 and at most 200 words.}
-% TODO: Auf 200 Wörter runterkürzen
 \abstract{ 
-%The goal of epidemiology is to provide insight into disease causations.
 Epidemiology aims to provide insight into disease causations.
@@
 %%
-%We assess the predictive quality of image-derived variables for lower back pain.
 We assess potential relations between image-derived variables of the lumbar spine with lower back pain in a cross-sectional study.
-%We assess the predictive quality of image-derived variables for lower back pain.
 %%
@@
 %%
-%We analyze the predictive power of shape-describing variables for non-image variables, e.g. pain indicators.
 We analyze mutual dependencies among shape-describing variables and non-image variables, e.g. pain indicators.
@@
 %%
-As a first result, we conclude that image-based variables are only sufficient for classification of lifestyle factors within the data set.
+As a first result, we conclude that image-based variables are only sufficient for classification of lifestyle factors within the data.
 %%
@@
 The workflow is part of an IVA framework, constructed to enhance the epidemiological workflow with iterative data-driven techniques.
+%%
+We focus on highlighting new unexpected correlations and trigger \emph{hypotheses generation}, rather than statistically validate complex epidemiological correlations.
 %%
@@
 %%
-These include nutrition, physical activity and body posture combined with physical stress at work.
+These include \emph{nutrition}, \emph{physical activity} and \emph{body posture combined with physical stress at work}.
 %%
@@
 %%
-\cite{Glasser2013} use variables derived from DCE-MRI data capturing the perfusion in the tumorous tissue.
+\cite{Glasser2013} use variables derived from DCE-MRI data capturing the perfusion in tumorous tissue.
 %%
@@
 	\begin{itemize}
-		\item 43 dichotomous (binary) variables, mostly indicating the presence of a disease, e.g. pancreatitis or high blood pressure
+		\item 43 dichotomous (binary) variables, mostly indicating the presence of a disease, e.g. \emph{pancreatitis} or \emph{high blood pressure}
 		\item 70 variables with more than two levels, indicating pain levels, nutrition and social factors, such as marital or retirement status
@@
 %%
-Others are exclusive to subgroups, such as women-specific variables, e.g. menstrual status.
+Others are exclusive to subgroups, such as women-specific variables, e.g. \emph{menstrual status}.
 %%
@@
 %%
-The spine protocol consisted of a sagittal $T_1$-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) and a sagittal $T_2$-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
+The spine protocol consisted of a sagittal $T_1$- and $T_2$-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
 %%
@@
 	\item \emph{Mean Torsion} (deviation of a curve from its current course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum_{i=1}^I \frac{torsion_i}{I}$.
-From now on, we refer to total torsion as \emph{torsion}.
+We refer to total torsion as \emph{torsion}.
 %%
@@
 	%%
-	Correlation of age and gender regarding the lumbar spine size.
-	%%
-	The bar chart shows subjects divided into four different age groups (x-axis) and their subject number (y-axis).
+	Correlation of \emph{age} and \emph{gender} regarding the lumbar spine size.
+	%%
+	The bar chart shows subjects divided into four different \emph{age} groups (x-axis) and their subject count (y-axis).
 	%%
@@
 %%
-Since men are on average taller than women and people of old age shrink due to bone erosion, gender and age are also risk factors (Fig.~\ref{fig:age-gender}).
+Since men are on average taller than women and people of old age shrink due to \emph{bone erosion}, \emph{gender} and \emph{age} are also risk factors (Fig.~\ref{fig:age-gender}).
 %%
@@
 %%
-Large body weight increases the spine load, resulting in a bent shape.
-%%
-To assess their influence, we take them into account when spine curvature and torsion is correlated with non-image variables.
+Large \emph{body weight} increases the spine load, resulting in a bent shape.
+%%
+To assess their influence, we take them into account when spine \emph{curvature} and \emph{torsion} is correlated with non-image variables.
 %All these variables need to be taken into account, when spine curvature and torsion is correlated with non-image variables.
 %%
-Since the gender encodes body size, we decided to divide subjects into body size groups.
+Since the \emph{gender} encodes \emph{body size}, we decided to divide subjects into \emph{body size} groups.
 %%
@@
 %%
-We conclude that curvature is not as reliable on the transversal plane as it is on the other planes.
+We conclude that \emph{curvature} is not as reliable on the transversal plane as it is on the other planes.
 %%--------------------------------
@@
   \caption{%%
-A generalized pairs plot of all image-derived variables colored by presence (red) or absence (turquoise) of back pain.
+A Generalized Pairs Plot of all image-derived variables colored by presence (red) or absence (turquoise) of back pain.
 %%
@@
 \paragraph{Correlation of Image Variables With Lumbar Back Pain.}
-Figure~\ref{fig:image-parameter-range} shows the distribution of all image-variables as generalized pairs plot.
+Figure~\ref{fig:image-parameter-range} shows the distribution of all image-variables as Generalized Pairs Plot.
 %%
@@
 %%
-The box plots show no difference between subjects with and without back pain.
+The box plots show no difference between subjects with or without back pain.
 
@@
 %%
-The first principal component explains 47\% of the variance and weights primarily \emph{mean curvature}, \emph{curvature sagittal}, \emph{curvature angle} and \emph{curvature angle sagittal}.
-%%
-The second component, adding 16\% of the variance, weigths  \emph{curvature coronal} and \emph{curvature angle coronal}.
-%%
-The third component explains 12\% of the variance and weights \emph{torsion} and \emph{curvature transverse}.
+The first principal component explains 47\% of the variance and weights primarily \emph{mean curvature}, \emph{sagittal curvature}, \emph{curvature angle} and \emph{sagittal curvature angle}.
+%%
+The second component, adding 16\% of the variance, weigths \emph{coronal curvature} and \emph{coronal curvature angle}.
+%%
+The third component explains 12\% of the variance and weights \emph{torsion} and \emph{transverse curvature}.
 %%
@@
 %%
-This is useful for our classification, because it allows to determine which manifestation of a variable can be described best using the image data variable.
+This is useful for our classification, because it allows to determine which manifestation of a variable can be described best using the image data variables.
 %%------------
@@
 The second plot shows all categorical variables transformed into dichotomous dummy variables (recall subsection~\ref{sec:C5.0}).
-
+%%
+The shown mutual dependencies aim to amplify \emph{hypothesis generation}.
+%%
+Dedicated statistical analysis of these results solely focussed on epidemiological research is not the scope of this paper.
+%%
 \subsection{All Variables}
@@
 %%
-The only variable reliably classified in this group is gender. % with an error rate of 31.5\% using a decision tree of seven rules.
+The only variable reliably classified in this group is \emph{gender}. % with an error rate of 31.5\% using a decision tree of seven rules.
 %
@@
 \paragraph{Large Decision Trees. }
-\emph{Back pain}-associated variables can be explained for various \emph{size} groups, but we could not extract universal rules.
+Back pain-associated variables can be explained for various \emph{size} groups, but we could not extract universal rules.
 %%
@@
 %%
-These methods do not aim to replace the traditional epidemiological workflow, but rather intervenes at the weak points of standard statistical methods.
+These methods do not aim to replace the traditional epidemiological workflow, but rather intervene at the weak points of standard statistical methods.
 %%

### 2014-10-01 draft.tex
Replaced Associations

@@
 %%
-Hence, associations between non-image variables were successfully detected and characterized.
+Hence, relations between non-image variables were successfully detected and characterized.
 }
@@
 %%
-We extract associations between spine shape and back pain characteristics.
+We extract possible associations between spine shape and back pain characteristics.
 %We analyze their discriminative power for different back pain characteristics.
@@
 %%
-	\item Statistical methods, such as regression analysis, assess the association of selected variables with the investigated disease.
+	\item Statistical methods, such as regression analysis, assess the correlation of selected variables with the investigated disease.
 %%
@@
 %%
-Epidemiologists assume associations between lumbar back pain and lifestyle factors.
+Epidemiologists assume correlations between lumbar back pain and lifestyle factors.
 %%
@@
 \end{figure*}
-We calculated an association matrix to assess correlations between the image variables.
+We calculated an correlation matrix to assess correlations between the image variables.
 %%

### 2014-10-02 draft.tex
Minor revision of Bernhard

@@
 \end{figure*}
-We calculated an correlation matrix to assess correlations between the image variables.
+We calculated a correlation matrix to assess correlations between the image variables.
 %%
@@
 %%
-The first three principal components explain 75\% of the variance in the image-derived variables.
-%%
-The first principal component explains 47\% of the variance and weights primarily \emph{mean curvature}, \emph{sagittal curvature}, \emph{curvature angle} and \emph{sagittal curvature angle}.
+The first three principal components explain 75\% of the variance in the nine image-derived variables.
+%%
+The first principal component explains 47\% of the variance and comprises primarily \emph{mean curvature}, \emph{sagittal curvature}, \emph{curvature angle} and \emph{sagittal curvature angle}.
 %%
@@
 %%
-The third component explains 12\% of the variance and weights \emph{torsion} and \emph{transverse curvature}.
+The third component explains 12\% of the variance and comprises \emph{torsion} and \emph{transverse curvature}.
 %%
@@
 %%
-We applied a heterogenous correlation technique to derive correlations between all variables in the data set.
+%We applied a heterogenous correlation technique to derive correlations between all variables in the data set.
+%We applied a heterogenous correlation technique to derive correlations between all variables in the data set.
+Different correlation metrics depending on the type of the individual variables are used to derive correlations between all variables in the data set.
 %%
@@
 %%
-Other than that, \emph{torsion} correlated with almost no variables (p values between 0 and 0.05).
+Other than that, \emph{torsion} correlated with almost no variables (values between 0 and 0.05).
 %%
@@
 %%
-Decision trees are a popular classification method in data mining to create predictive models.
+Decision trees are a established classification method in data mining to create predictive models.
 %%
@@
 Readability is only granted for small trees. %, complex structures with many branches are not desirable.
+%%
+Large, complex trees are difficult to trace and to understand.
 %%
@@
  Create a decision node that splits on $a_{best}$\;
- Recurse on the sublists obtained by splitting on $a_{best}$, and add those nodes as children of node\;
+ Iterate on the sublists obtained by splitting on $a_{best}$, and add those nodes as children of node\;
  \caption{Building a decision tree using the C4.5 Algorithm}
@@
 %%
-Our error term therefore incorporates the discriminative power of each manifestation and is denoted as follows:
+Based on this we use a summary error, which incorporates the discriminative power of each manifestation and is denoted as follows:
+%Our error term therefore incorporates the discriminative power of each manifestation and is denoted as follows:
 %%
@@
 
-\paragraph{Attribute Mapping}
+\paragraph{Attribute Mapping.}
 %%
@@
 
-We apply a square root scale for the tree size axis to highlight data points with few decision rules.
+We apply a square root scale for the tree size axis to highlight decision trees with few decision rules.
 %%
@@
 %%
-This is realized by clicking on an entry on the visualization, which then interactively displays the corresponding decision tree in detail.
+This is realized by clicking on an entry in the visualization, which then interactively displays the corresponding decision tree in detail.
 %%
@@
 \end{figure*}
-\noindent In this section, we show which non-image variables can be described using the 9 image derived variables.
+\noindent In this section, we show which non-image variables can be explained using the 9 image derived variables.
 %%
@@
 %%
-The only variable reliably classified in this group is \emph{gender}. % with an error rate of 31.5\% using a decision tree of seven rules.
+The only variable reliably classified in this group is \emph{gender}, which can be predicted with an error of $0.31$ using 7 rules and incorporates only \emph{curvature-} and \emph{curvature angle}-related variables.
 %
-It can be discriminated with an error of $0.31$ using 7 rules and incorporates only \emph{curvature-} and \emph{curvature angle}-related variables.
+%It can be predicted with an error of $0.31$ using 7 rules and incorporates only \emph{curvature-} and \emph{curvature angle}-related variables.
 %%

### 2014-12-04 draft.tex
Minor changes at Formula
removed blind status

@@
 
-% TODO: BLIND Version
-\author{Blind}
-% \author{\authorname{Paul Klemm\sup{1}, Sylvia Glaßer\sup{1}, Kai Lawonn\sup{1}, Marko Rak\sup{1} Henry Völzke\sup{2}, Katrin Hegenscheid\sup{3} and Bernhard Preim\sup{1}}
-% \affiliation{\sup{1} Department of Simulation and Graphics, University of Magdeburg, Germany}
-% \affiliation{\sup{2} Institute for Community Medicine, University of Greifswald, Germany}
-% \affiliation{\sup{3} Institute for Diagnostic Radiology and Neuro-Radiology, University of Greifswald, Germany}
-% \email{\{klemm, glasser\}@isg.cs.uni-magdeburg.de}
-% }
+% \author{Blind}
+\author{\authorname{Paul Klemm\sup{1}, Sylvia Glaßer\sup{1}, Kai Lawonn\sup{1}, Marko Rak\sup{1} Henry Völzke\sup{2}, Katrin Hegenscheid\sup{3} and Bernhard Preim\sup{1}}
+\affiliation{\sup{1} Department of Simulation and Graphics, University of Magdeburg, Germany}
+\affiliation{\sup{2} Institute for Community Medicine, University of Greifswald, Germany}
+\affiliation{\sup{3} Institute for Diagnostic Radiology and Neuro-Radiology, University of Greifswald, Germany}
+\email{\{klemm, glasser\}@isg.cs.uni-magdeburg.de}
+}
 
@@
 %%
-%% TODO BLIND "We incorporate their" -> We incorporate the ...
-We incorporate their IVA approach by concurrently displaying complex relationships between variables via decision trees.
+We incorporate the IVA approach by concurrently displaying complex relationships between variables via decision trees.
 %%
@@
 \begin{equation}
-error = 1 - \frac{\sum_{m=1}^M \frac{correctlyClassified_{m}}{M}}{number_{dimensions}}
+totalError = 1 - \sum_{m=1}^M \frac{correctlyClassified_{m}}{M\cdot N_m}
 \end{equation}
@@
 %%
-%TODO: Muss das rein? Auswertung der Anteile der Parameter an den Ergebnissen (Torsion, Curvature, etc. - welches ist der dominanteste Parameter?)
+%TODO: [medium] Muss das rein? Auswertung der Anteile der Parameter an den Ergebnissen (Torsion, Curvature, etc. - welches ist der dominanteste Parameter?)
 %%

### 2014-12-04 draft.tex
cleaned comments

@@
 
+% modify floating spacings 
+% \addtolength{\abovecaptionskip}{-0.0mm}
+% \addtolength{\belowcaptionskip}{-0.0mm}
+% \addtolength{\dbltextfloatsep}{-5.0mm}
+% \addtolength{\textfloatsep}{-4mm}
+% \addtolength{\intextsep}{-5.0mm}
+%
+% % modify paragraph spacings
+% \addtolength{\parskip}{-5.0mm}
+
 \begin{document}
 
-%\title{Insight Into Lumbar Back Pain  \subtitle{What the Lumbar Spine Tells About Your Life} }
 \title{Interactive Visual Analysis of Lumbar Back Pain  \subtitle{What the Lumbar Spine Tells About Your Life} }
@@
 
-% \abstract{The abstract should summarize the contents of the paper and should contain at least 70 and at most 200 words.}
 \abstract{ 
@@
 \label{sec:Introduction}
-%%Epidemiology is the study of causation of diseases.
+
 \noindent Epidemiology is the study of dissemination, causes and results of health-related states and events.
@@
 
-%For epidemiological research, cohort studies are acquired.
 Cohort studies are an instrument of epidemiological research.
-%For epidemiological research, cohort studies are carried out.
 %%
@@
 Epidemiological analysis of back pain aims to determine risk factors.
-%Many epidemiological-motivated approaches aim at improved characterization of this disease with respect to patient-specific attributes.
-%Since back pain is one of the most frequent diseases in the western civilization, many epidemiological-motivated approaches aim at improved characterization of this disease with respect to patient-specific attributes.
 %%
@@
 We extract possible associations between spine shape and back pain characteristics.
-%We analyze their discriminative power for different back pain characteristics.
 %%
@@
 %%
-%Then, Interactive Visual Analysis (IVA) provides insight into the quality of the image-derived data as well as their potential as back pain risk factors.
 Then, Interactive Visual Analysis (IVA) highlights mutual dependencies between image-derived data and back pain-related variables.
@@
 \item The identification of lumbar spine shape properties potentially associated with back pain,
-%\item The identification and evaluation of lumbar spine shape properties associated with back pain,
 %%
@@
 %%
-%\item The identification of the most important variables via data mining methods (including a novel semi-quantitative evaluation concept) embedded in our IVA-framework
 \item The identification of the most important variables via data mining methods using a novel semi- quantitative overview visualization concept embedded in our IVA-framework.
@@
 %%
-%The IVA tool and all shown plots presented in this paper, except of Fig.~\ref{fig:age-gender}, are available as supplementary material under the following link: \url{blind.dnsalias.com}.
 %% ---------------------------------------------
@@
 %%
-%When for example analyzing influences of nutrition to prostate cancer, the results need to be age-normalized.
-%%
 Age acts as a \emph{risk factor} for prostate cancer.
-%%
-%These \emph{confounding} variables are often hard to find.
 %%
@@
 %%------------------------------
-%\subsection{Epidemiological Data}
 \subsection{Challenges of Epidemiological Data Analysis}
@@
 %%
-%For example, women-specific questions or data about a disease treatment only affects subjects suffering from this condition.
 For example, data about a disease treatment only affects subjects suffering from this condition.
@@
 Epidemiological data acquisition includes a variety of techniques, such as medical examinations, self-reported questionnaires or genetic examinations.
-%Epidemiologists acquire data using a wide range of techniques, such as medical examinations, self reported questionnaires or genetic examinations.
 %%
@@
 %%
-%Every information reduction proposes an assumption about the data and can therefore introduce a bias.
-%%
-%For example, using age to divide subjects into \emph{young} and \emph{old} categories can distort statistical results.
-%%
-%This distortion is reduced with increasing number of discretization steps.
-%%
 
@@
 The analysis of these data is challenging.
-%These data are hard to analyze.
 %%
@@
 %%
-%The lower (\emph{lumbar}) spine is the most stressed spine part .
-%%
 Lower back pain is one of the most frequent complaints in the Western civilization \cite{Hoy2010}.
@@
 %%
-%Potential \emph{confounding} effects are also subject of current research.
-%%
-
-%Epidemiologists want to characterize the healthy aging process of the spine.
-Epidemiologists want to characterize the relation between the aging and degenerative process of the spine \cite{Szpalski2005}. %healthy aging process of the spine.
+
+Epidemiologists want to characterize the relation between the aging and degenerative process of the spine \cite{Szpalski2005}.
 %%
@@
 %%
-%Klemm et al. propose an extension to the previously described epidemiological workflow using \emph{Interactive Visual Analysis} (IVA) \cite{Klemm2014VIS}.
+% TODO: BLIND
 \cite{Klemm2014VIS} propose an extension to the previously described epidemiological workflow using IVA.
@@
 %%
-%Our work, however, does not contain geographical information of any kind.
 %%----------------------------
@@
 %%
-%The results can be easily communicated to domain experts with little knowledge about the algorithm behind the results.
-%%
 The effort of analyzing decision trees increases with their complexity.
@@
 %%
-%Clustering techniques applied to non-image data are strongly dependent on the chosen variables and distance measures \cite{Klemm2014BVM}.
 Clustering techniques are strongly dependent on the chosen variables and distance measures \cite{Klemm2014BVM}.
-%%
-%Klemm et al. found \texttt{k-Prototypes} and \texttt{DBSCAN} appropriate for this task.
 %%
@@
 %%
-%It is an active research area to determine these correlations.
-%%
-
-%%
 \subsection{Non-Image Data}
-%\com{Analysis Using Heterogeneous Correlation}
+%%
 The non-image variables range from somatometric variables describing body measures to medical examinations, such as laboratory tests as well as lifestyle factors, e.g. sporting activity or nutrition.
@@
 	\item 21 metric variables, mostly describing somatometric variables and markers retrieved from blood analysis
+	%%
 	\item 113 categorical variables divided into
+	%%
 	\begin{itemize}
+		%%
 		\item 43 dichotomous (binary) variables, mostly indicating the presence of a disease, e.g. \emph{pancreatitis} or \emph{high blood pressure}
+		%%
 		\item 70 variables with more than two levels, indicating pain levels, nutrition and social factors, such as marital or retirement status
+		%%
 	\end{itemize}
+%%
 \end{itemize}
@@
 	We also extracted the variables for each projection axis to assess their information gain.
-	%(b) Using the Frenet formulas, we extract the curvature, torsion and the curvature angle from the lumbar spine canal center line.
 	}
@@
 %%
-% \com{From VIS'14 Paper}
-% The lumbar spine was detected in the image data using a hierarchical finite element method by Rak et al. [32].
-% %%
-% This semiautomatic method requires the user to initialize the tetrahedron-based finite element models (FEM) with a click on the L3 vertebra.
-% %%
-% Two user-defined landmarks on the top and bottom of the L3 vertebra describe an initial model height estimation.
-% %%
-% The model uses a weighted sum of T1 and T2-weighted MR images to detect the lumbar spine shape.
-% %%
-% Once registered, it captures information about the shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae [21].
-% %%
-% Due to incorrect initialization, strongly deformed spines, contrast differences and artifacts, the model was not able to detect lumbar spines for all subjects.
-% %%
-% We obtained and worked with 2,540 tetrahedron models of the lumbar spine.
-% %%
-% For clustering, we extracted the centerline of the lumbar spine canal, which captures information about lordosis and scoliosis (the medical terms for spine curvature) [21].
-% %%
-
-%\com{Marko, kannst du mal bitte schauen, ob das hier so korrekt ist, bzw. ob man das konkreter formulieren kann?}
+
 We have to assess the model accuracy to extract key figures from it.
@@
 It captures reliable information about spine canal curvature.
-%% BLIND
-%In a previous work (\cite{Klemm2013VMV}), we extracted a centerline representation of the lumbar spine canal from the detection model (Fig.~\ref{fig:centerline}~a).
-The centerline representation of the lumbar spine canal was extracted from the detection model (Fig.~\ref{fig:centerline}~a) using the approach of \cite{Klemm2013VMV}.
+%%
+In a previous work (\cite{Klemm2013VMV}), we extracted a centerline representation of the lumbar spine canal from the detection model (Fig.~\ref{fig:centerline}~a).
 
@@
 %%
-%Thus, we extracted the following metrics from the model:
-%%
-%\paragraph{Curvature}
 \begin{itemize}
-	% \item The curvature of the centerline is calculated using the summed mean curvature between adjacent points:
-% %%
-% $\sum \frac{curvature_i}{curvature_all}$.
 	\item \emph{Mean Curvature} is calculated as weighted sum of curvature between all adjacent points describing the centerline: $\sum_{i=1}^I \frac{curvature_i}{I}$.
@@
 In the next section, we present the experiments we conducted to assess the influence of the lumbar spine shape to lower back pain.
-
-%\subsection{Creating a Decision Tree using C5.0}
-%\com{C4.5 Algorithm for creating decision trees; }
-
-%\subsection{Parameter Assessment}
-%Based on first results
-
+%%
 \section{\uppercase{Experiments and Preliminary Results}}
 \label{sec:Experiments}
-\begin{figure}[!h]
-  %\vspace{-0.2cm}
-  \centering
-  \includegraphics[width=0.475\textwidth]{figures/age-gender}
-  \caption{
-	%%
-	Correlation of \emph{age} and \emph{gender} regarding the lumbar spine size.
-	%%
-	The bar chart shows subjects divided into four different \emph{age} groups (x-axis) and their subject count (y-axis).
-	%%
-	Each bar contains the mean lumbar spine of the respective group.
-	%%
-	The shape color encodes the distance (red for x axis, blue for y axis, green for z axis) to the overall male (top chart) or female (bottom chart) mean shape.
-	%%
-	The dark gray share of each bar encodes the portion of male (top chart) or female (bottom chart) subjects.
-	%%
-%	Women have a higher live expectance than men, hence their higher share in the old age group.
-	%%
-%	Also, women are on average smaller than men, hence the larger shape similarity with older subjects.
-	%%
-	%Due to bone erosion, older people are also smaller on average.
-	}
-  \label{fig:age-gender}
-\end{figure}
+% \begin{figure}[!h]
+%   %\vspace{-0.2cm}
+%   \centering
+%   \includegraphics[width=0.475\textwidth]{figures/age-gender}
+%   \caption{
+% 	%%
+% 	Correlation of \emph{age} and \emph{gender} regarding the lumbar spine size.
+% 	%%
+% 	The bar chart shows subjects divided into four different \emph{age} groups (x-axis) and their subject count (y-axis).
+% 	%%
+% 	Each bar contains the mean lumbar spine of the respective group.
+% 	%%
+% 	The shape color encodes the distance (red for x axis, blue for y axis, green for z axis) to the overall male (top chart) or female (bottom chart) mean shape.
+% 	%%
+% 	The dark gray share of each bar encodes the portion of male (top chart) or female (bottom chart) subjects.
+% 	%%
+% %	Women have a higher live expectance than men, hence their higher share in the old age group.
+% 	%%
+% %	Also, women are on average smaller than men, hence the larger shape similarity with older subjects.
+% 	%%
+% 	%Due to bone erosion, older people are also smaller on average.
+% 	}
+%   \label{fig:age-gender}
+% \end{figure}
 %%
@@
 %%
-Since men are on average taller than women and people of old age shrink due to \emph{bone erosion}, \emph{gender} and \emph{age} are also risk factors (Fig.~\ref{fig:age-gender}).
+Since men are on average taller than women and people of old age shrink due to \emph{bone erosion}, \emph{gender} and \emph{age} are also risk factors.
 %%
@@
 %%
-%Due to bone erosion, older people are also smaller on average.
-%%
 Large \emph{body weight} increases the spine load, resulting in a bent shape.
@@
 To assess their influence, we take them into account when spine \emph{curvature} and \emph{torsion} is correlated with non-image variables.
-%All these variables need to be taken into account, when spine curvature and torsion is correlated with non-image variables.
 %%
@@
 Discretizing metric variables using quartiles avoids small outlier groups.
-%To avoid small outlier groups, epidemiologists recommend using quartiles to discretize metric variables.
 %%
@@
 As a first experiment we correlated the shape variable with the dichotomous back pain indicator using \texttt{GPLOM's} \cite{GPLOMS}.
-%%
-%Pairs plots visualize continuous and categorical variables pairwise in a matrix view.
-%%
-%Scatterplots display pairs of continuous variables, bar charts for continuous variable as a function of a categorical variable and box plots pairs of categorical variables.
 %%
@@
   \includegraphics[width=1.0\textwidth]{figures/image-parameter-range}
-  \caption{%%
+  \caption{
+%%
 A Generalized Pairs Plot of all image-derived variables colored by presence (red) or absence (turquoise) of back pain.
@@
 %%
-%Torsion and curvature/curvature angle do not correlate.
 \emph{Torsion} does not correlate with any other image-derived variable.
@@
 %%
-%We applied a heterogenous correlation technique to derive correlations between all variables in the data set.
-%We applied a heterogenous correlation technique to derive correlations between all variables in the data set.
 Different correlation metrics depending on the type of the individual variables are used to derive correlations between all variables in the data set.
@@
 These observations brought us to the decision to incorporate sophisticated data mining techniques to assess the influence of the image-derived variables.
-
-% \subsubsection{Diese Themen hier unterbringen}
-% \begin{itemize}
-% 	\item Direkter Vergleich
-% 	\item Verteilung des Means
-% 	\item Korrelationsmatrix
-% 	\item PCA
-% \end{itemize}
-% Weil das nicht ausreichend war, haben wir einen C5.0 Baum erstellt, um alle potentiellen Zusammenhänge zwischen bildbezogenen und Nichtbildbezogenen Parametern herzustellen - Was kann man mit den Bilddaten erklären und was nicht?
-%
-% HetCor hier noch mit unterbringen.
-Damit kann man eine Korrelationsmatrix unabhängig vom Datentyp berechnen, um eine große Korrellationsmatrix zu erhalten (Supplementary Material).
-Hier können die Einträge in den Zeilen der Bildparameter angeschaut werden, um zu schauen, ob man eine Korrellation mit bestimmten Parametern finden kann!
-Das müsste auch wieder für alle Größengruppen passieren!
-(https://stats.stackexchange.com/questions/101016/pca-and-component-scores-based-on-a-mix-of-continuous-binary-and-categorical-va)
-% Creating a Decision Tree using C5.0
-
-%\com{C4.5 Algorithm for creating decision trees; }
-
+%%
 \section{\uppercase{Evaluation of Decision Trees}}
@@
 %%
-%Leafs of a decision tree represent class labels, branches represent decision trees.
 Leafs of a decision tree represent class labels, branches represent feature conjunctions leading to the class labels.
@@
 %%
-%Epidemiologists can interpret the results without having deep knowledge about the classification algorithm.
-%%
-Readability is only granted for small trees. %, complex structures with many branches are not desirable.
+Readability is only granted for small trees.
 %%
@@
 %%
-%The algorithm then tries to find a decision tree, which divides the samples using the input variables just like the target variables.
-%%
-%This means, that every node in the tree represents the attribute which splits the data most efficiently into the target subset.
-%%
-The pseudocode for the algorithm is defined in Algorithm~\ref{alg:c4.5} \cite{C4.5Algorithm}.
-\begin{algorithm}[]
- Check for base cases\;
- \For{each attribute $a$}{
-  Find the normalized information gain ratio from splitting on $a$\;
- }
- Let $a_{best}$ be the attribute with the highest normalized information gain\;
- Create a decision node that splits on $a_{best}$\;
- Iterate on the sublists obtained by splitting on $a_{best}$, and add those nodes as children of node\;
- \caption{Building a decision tree using the C4.5 Algorithm}
- \label{alg:c4.5}
-\end{algorithm}
+% The pseudocode for the algorithm is defined in Algorithm~\ref{alg:c4.5} \cite{C4.5Algorithm}.
+% \begin{algorithm}[]
+%  Check for base cases\;
+%  \For{each attribute $a$}{
+%   Find the normalized information gain ratio from splitting on $a$\;
+%  }
+%  Let $a_{best}$ be the attribute with the highest normalized information gain\;
+%  Create a decision node that splits on $a_{best}$\;
+%  Iterate on the sublists obtained by splitting on $a_{best}$, and add those nodes as children of node\;
+%  \caption{Building a decision tree using the C4.5 Algorithm}
+%  \label{alg:c4.5}
+% \end{algorithm}
 C5.0 is developed to produce smaller decision trees than C4.5 and improve the execution time.
@@
 %%
-% \begin{enumerate}
-% %%
-% \item Check for base cases
-% %%
-% \item For each attribute $a$
-% %%
-% \begin{enumerate}
-% %%
-% \item Find the normalized information gain ratio from splitting on $a$
-% %%
-% \end{enumerate}
-% %%
-% \item Let $a_{best}$ be the attribute with the highest normalized information gain
-% %%
-% \item Create a decision node that splits on $a_{best}$
-% %%
-% \item Recurse on the sublists obtained by splitting on $a_{best}$, and add those nodes as children of node
-% %%
-% \end{enumerate}
-%%
 Categorical attributes with more levels are biased with more information gain in a decision tree \cite{deng2011bias}.
 %%
-%Creating dummy variables bypasses this problem.
 Creating a dummy variable by converting each manifestation of a categorical variable into a dichotomous variable bypasses this problem.
-%%
-%
-%The actual use for the resulting tree is the classification of new observations (subjects).
-%%
-%Yet, we are interested in the decision rules and the classification accuracy.
 %%
@@
 Further subdivision, e.g. by quantiles of \emph{body size} increases the number to 402 trees.
-%Dividing the subjects into four groups using body size, the number increases to 402 trees.
-%%
-%To make the results cognitive feasible, we have to abstract the results of the classification.
+%%
 We have to abstract the results of the classification to keep the mental effort of interpreting the data low.
@@
 %%
-%We follow the visual information-seeking mantra and want to provide an overview first, then details-on-demand \cite{shneiderman1996}.
 We follow the Visual Analytics mantra of analyzing first, show the important and analyze further \cite{Keim}.
@@
 Based on this we use a summary error, which incorporates the discriminative power of each manifestation and is denoted as follows:
-%Our error term therefore incorporates the discriminative power of each manifestation and is denoted as follows:
 %%
@@
 %
-%It can be predicted with an error of $0.31$ using 7 rules and incorporates only \emph{curvature-} and \emph{curvature angle}-related variables.
 %%
@@
 \section*{\uppercase{Acknowledgements}}
-\noindent Omitted due to blind review.
-%\noindent SHIP is part of the Community Medicine Research net of the University of Greifswald, Germany, which is funded by the Federal Ministry of Education and Research (grant no.
+\noindent SHIP is part of the Community Medicine Research net of the University of Greifswald, Germany, which is funded by the Federal Ministry of Education and Research (grant no.
 03ZIK012), the Ministry of Cultural Affairs as well as the Social Ministry of the Federal State of Mecklenburg-West Pomerania.

### 2014-12-04 draft.tex
shortened first two sections, created new paragraphs to save some space

@@
 %%
-So far, epidemiological research is strongly hypothesis-driven.
-%%
-More precisely, observations made by clinicians are translated into hypotheses, which are then statistically evaluated using data variables from epidemiological studies.
-%%
-As a result, the hypothesis will be accepted or rejected.
-%%
-
+% So far, epidemiological research is strongly hypothesis-driven.
+% %%
+% More precisely, observations made by clinicians are translated into hypotheses, which are then statistically evaluated using data variables from epidemiological studies.
+% %%
+% As a result, the hypothesis will be accepted or rejected.
+% %%
+%%
 Cohort studies are an instrument of epidemiological research.
 %%
-Nowadays, these studies often comprise medical image data, e.g., tomographic image data sets, as well as general patient-specific attributes, such as \emph{gender}, \emph{age} and \emph{weight}.
-%%
-Back pain is one of the most frequent complaints in the Western civilization and gains importance due to the population aging \cite{Hoy2010}.
+Nowadays, these studies often comprise medical image data, e.g., tomographic image data sets. %, as well as general patient-specific attributes, such as \emph{gender}, \emph{age} and \emph{weight}.
+%%
+We analyze back pain, one of the most frequent complaints in the Western civilization. %\cite{Hoy2010}. % and gains importance due to the population aging \cite{Hoy2010}.
 %%
@@
 %%
-Furthermore, previous analyses mainly focused on a single variable (e.g. \emph{BMI}) or a small subset of variables.
-%%
-
-We present an analysis of the image-derived data of a cohort study including lumbar spine datasets.
+%Furthermore, previous analyses mainly focused on a single variable (e.g. \emph{BMI}) or a small subset of variables.
+%%
+
+We present an analysis of the image-derived data from a cohort study lumbar spine dataset.
 %%
@@
 %%
-The workflow is part of an IVA framework, constructed to enhance the epidemiological workflow with iterative data-driven techniques.
-%%
-We focus on highlighting new unexpected correlations and trigger \emph{hypotheses generation}, rather than statistically validate complex epidemiological correlations.
+% The workflow is part of an IVA framework, constructed to enhance the epidemiological workflow with iterative data-driven techniques.
+%%
+We focus on highlighting new correlations and trigger \emph{hypotheses generation}, rather than statistically validate complex epidemiological correlations.
 %%
@@
 %%
-\item The identification of the most important variables via data mining methods using a novel semi- quantitative overview visualization concept embedded in our IVA-framework.
+\item The identification of the most important variables via data mining methods using a novel semi- quantitative overview visualization concept. % embedded in our IVA-framework.
 %%
@@
 %%
-The analysis, as well as the presented IVA tool is available as supplementary material in an interactive \texttt{R Markdown}\footnote{Developed by RStudio, Inc; \texttt{rmarkdown.rstudio.com}} document (\url{blind.dnsalias.com}).
+%The analysis, as well as the presented IVA tool is available as supplementary material in an interactive \texttt{R Markdown}\footnote{Developed by RStudio, Inc; \texttt{rmarkdown.rstudio.com}} document (\url{blind.dnsalias.com}).
 %%
@@
 %%
-For example, the effect of nutrition on prostate cancer requires age normalization due to the increased incidence for older subjects.
-%%
-Age acts as a \emph{risk factor} for prostate cancer.
-%%
-Statistical correlation does not imply causation--epidemiologists need to assess the medical soundness of the statistical results.
+% For example, the effect of nutrition on prostate cancer requires age normalization due to the increased incidence for older subjects.
+% %%
+% Age acts as a \emph{risk factor} for prostate cancer.
+% %%
+% Statistical correlation does not imply causation--epidemiologists need to assess the medical soundness of the statistical results.
 %%------------------------------
-\subsection{Challenges of Epidemiological Data Analysis}
-Epidemiological data originates from a wide range of studies.
-%%
-The study type depends on the condition of interest.
-%%
-Most common are case-control studies analyzing one specific disease and its influence on the human body.
-%%
-We focus on large scale cohort study data.
+\paragraph{Challenges of Epidemiological Data Analysis.}
+% Epidemiological data originates from a wide range of studies.
+% %%
+% The study type depends on the condition of interest.
+% %%
+% Most common are case-control studies analyzing one specific disease and its influence on the human body.
+%%
+We focus on epidemiological large scale cohort study data.
 %%
@@
 %%
-
 Epidemiological data are heterogenous and incomplete.
@@
 %%
-
 Epidemiological data acquisition includes a variety of techniques, such as medical examinations, self-reported questionnaires or genetic examinations.
@@
 %%
-To compare these data, information reduction techniques are necessary.
+Information reduction techniques are necessary to compare these data.
 %%
 For example, continuous data, such as age, is often discretized into quantile bins for visualizing age-dependent disease frequencies.
-%%
 
@@
 %%
-Segmentation data is usually analyzed by abstracting it into key figures, such as diameters or distances.
-%%
-These numeric values can be compared with non-image variables to retrieve correlations.
+%Segmentation data is usually analyzed by abstracting it into key figures, such as diameters or distances.
+%%
+%These numeric values can be compared with non-image variables to retrieve correlations.
 %%--------------------------------
-\subsection{Lower Back Pain}
+\\\\
+\noindent \textbf{Lower Back Pain.}
 %%
@@
 %%
-Epidemiologists assume correlations between lumbar back pain and lifestyle factors.
-%%
-These include \emph{nutrition}, \emph{physical activity} and \emph{body posture combined with physical stress at work}.
+% Epidemiologists assume correlations between lumbar back pain and lifestyle factors.
+% %%
+% These include \emph{nutrition}, \emph{physical activity} and \emph{body posture combined with physical stress at work}.
 %%
@@
 %%
-
 Epidemiologists want to characterize the relation between the aging and degenerative process of the spine \cite{Szpalski2005}.
 %%
-To achieve this, they have to analyze the lumbar spine shape as well as the mentioned lifestyle factors.
+They have to analyze the lumbar spine shape as well as lifestyle factors.
 %% -------------------------------
@@
 %%
-\paragraph{Visual Analysis of Medical Image and Non-Image Data}
-%%
-% TODO: BLIND
-\cite{Klemm2014VIS} propose an extension to the previously described epidemiological workflow using IVA.
+\noindent \textbf{Visual Analysis of Medical Image and Non-Image Data}
+%%
+In a previous work, we proposed an extension to the previously described epidemiological workflow using IVA \cite{Klemm2014VIS}.
 %%
@@
 %%
-We incorporate the IVA approach by concurrently displaying complex relationships between variables via decision trees.
+%We incorporate the IVA approach by concurrently displaying complex relationships between variables via decision trees.
 %%
@@
 %%
-\paragraph{Visual Analysis of Heterogenous Non-Image Data.}
+\\\\
+\noindent \textbf{Visual Analysis of Heterogenous Non-Image Data.}
 %%
@@
 %%
-\cite{Dai} define a \emph{Concept Map}, linking cancer-associated features by incorporating choropleth maps with bar charts, parallel coordinates and scatter plots with regression lines.
-%%
-The \emph{Concept Map} is iteratively refined using primarily the underlying geographical data.
+% \cite{Dai} define a \emph{Concept Map}, linking cancer-associated features by incorporating choropleth maps with bar charts, parallel coordinates and scatter plots with regression lines.
+% %%
+% The \emph{Concept Map} is iteratively refined using primarily the underlying geographical data.
 %%
 %%----------------------------
-\paragraph{Decision Rule-Driven Analysis of Medical Data.}
+\\\\
+\noindent \textbf{Decision Rule-Driven Analysis of Medical Data.}
 Related work in this field often focuses on clinical diagnosis.
@@
 %%----------------------------
-
+%%
 \cite{Pinheiro} applied association rule mining to create high-confidence decision rules based on variables, such as age, gender and geographic location about the mortality rate in liver cancer data.
 %%
-\cite{Sekhavat} propose an interactive visualization technique for decision trees, where all decisions are displayed in a heat map.
-%%
-The visualization translates the decision tree into a heat map, where each cell represents the connectivity between an initial and a target value.
-%%
-The resulting interactive visualization is well suited for analyzing single, but not multiple decision trees.
-%%
-\paragraph{SHIP Data Analysis.}
+% \cite{Sekhavat} propose an interactive visualization technique for decision trees, where all decisions are displayed in a heat map.
+% %%
+% The visualization translates the decision tree into a heat map, where each cell represents the connectivity between an initial and a target value.
+% %%
+% The resulting interactive visualization is well suited for analyzing single, but not multiple decision trees.
+%%
+\\\\
+\noindent \textbf{SHIP Data Analysis.}
 \cite{Klemm2013VMV} analyzed the lumbar spine variability of 490 \texttt{SHIP-2} subjects.
@@
 %%
-\subsection{Non-Image Data}
+\\\\
+\noindent \textbf{Non-Image Data}.
 %%
@@
 Sparse variables are statistically less resilient and need to be treated with care during the analysis.
-
-\subsection{Image-Derived Data}
-\begin{figure}[!h]
+\\\\
+\noindent \textbf{Image-Derived Data.}
+\begin{figure}[!t]
   %\vspace{-0.2cm}
@@
 %%
-\paragraph{GPLOM Analysis.}
+\\\\
+\noindent \textbf{GPLOM Analysis.}
 As a first experiment we correlated the shape variable with the dichotomous back pain indicator using \texttt{GPLOM's} \cite{GPLOMS}.
@@
 %%--------------------------------
-\paragraph{Correlation Matrix.}
+\\\\
+\noindent \textbf{Correlation Matrix.}
 \begin{figure*}[htb]
@@
 %%--------------------------------------------------------
-\paragraph{Correlation of Image Variables With Lumbar Back Pain.}
+\\\\
+\noindent \textbf{Correlation of Image Variables With Lumbar Back Pain.}
 Figure~\ref{fig:image-parameter-range} shows the distribution of all image-variables as Generalized Pairs Plot.
@@
 The box plots show no difference between subjects with or without back pain.
-
-\paragraph{Assessing the Information Gain Using the Principal Component Analysis.}
+\\\\
+\noindent \textbf{Assessing the Information Gain Using the Principal Component Analysis.}
 %%
@@
 %% ------------------------------------------
-\paragraph{Heterogenous Correlations.}
+\\\\
+\noindent \textbf{Heterogenous Correlations.}
 %%
@@
 
-\subsection{The C4.5/C5.0 Algorithm}
-\label{sec:C5.0}
-The C4.5 algorithm builds decision trees based on information entropy on a data set.
-%%
-Such a calculation requires a numeric or categorical target variable.
+We use the C4.5 algorithm, which builds decision trees based on information entropy on a data set.
+%%
+Categorical attributes with more levels are biased with more information gain in a decision tree \cite{deng2011bias}.
+%%
+Creating a dummy variable by converting each manifestation of a categorical variable into a dichotomous variable bypasses this problem.
+%%
+In the following analysis, we strongly focus on the complexity of decision trees and the classification accuracy.
+
+%\subsection{The C4.5/C5.0 Algorithm}
+%\label{sec:C5.0}
+%The C4.5 algorithm builds decision trees based on information entropy on a data set.
+%%
+%Such a calculation requires a numeric or categorical target variable.
 %%
@@
 % \end{algorithm}
-C5.0 is developed to produce smaller decision trees than C4.5 and improve the execution time.
-%%
-We use the \texttt{R} implementation of C5.0 \cite{c5.0classification}.
-%%
-Categorical attributes with more levels are biased with more information gain in a decision tree \cite{deng2011bias}.
-%%
-Creating a dummy variable by converting each manifestation of a categorical variable into a dichotomous variable bypasses this problem.
-%%
-In the following analysis, we strongly focus on the complexity of decision trees and the classification accuracy.
+%C5.0 is developed to produce smaller decision trees than C4.5 and improve the execution time.
+%%
+%We use the \texttt{R} implementation of C5.0 \cite{c5.0classification}.
+%%
+% Categorical attributes with more levels are biased with more information gain in a decision tree \cite{deng2011bias}.
+% %%
+% Creating a dummy variable by converting each manifestation of a categorical variable into a dichotomous variable bypasses this problem.
+% %%
+% In the following analysis, we strongly focus on the complexity of decision trees and the classification accuracy.
 
@@
 The scatter plot is our central element for the interactive analysis of decision trees.
-
-\paragraph{The Error Term. }
+\\\\
+\noindent \textbf{The Error Term. }
 Normally the error for a classification is calculated with $error = \frac{correctlyClassified_{n}} {N}$ where $n$ is the number of subjects.
@@
 It allows for comparability of error rates between variables with different manifestation count.
-
-\paragraph{Attribute Mapping.}
+\\\\
+\noindent \textbf{Attribute Mapping.}
 %%
@@
 %%
-The second plot shows all categorical variables transformed into dichotomous dummy variables (recall subsection~\ref{sec:C5.0}).
+The second plot shows all categorical variables transformed into dichotomous dummy variables. % (recall subsection~\ref{sec:C5.0}).
 %%
@@
 %%
-\paragraph{Large Decision Trees. }
+\\\\
+\noindent \textbf{Large Decision Trees.}
 Back pain-associated variables can be explained for various \emph{size} groups, but we could not extract universal rules.
@@
 %%
-\paragraph{Small Decision Trees. }
+\\\\
+\noindent \textbf{Small Decision Trees.}
 %%
@@
 %%
-\paragraph{Applicability.}
+\\\\
+\noindent \textbf{Applicability.}
 %%
@@
 %%
-\paragraph{Predictive Power of Image-Derived Variables.}
+\\\\
+\noindent \textbf{Predictive Power of Image-Derived Variables.}
 We showed that \emph{torsion}, \emph{curvature} and \emph{curvature angle} of the lumbar spine at the presented precision are not sufficient to characterize lumbar back pain in the SHIP data set.
@@
 %%
-\paragraph{Future Work.}
+\\\\
+\noindent \textbf{Future Work.}
 %% Contributions
@@
 \section*{\uppercase{Acknowledgements}}
+\begin{small}
 \noindent SHIP is part of the Community Medicine Research net of the University of Greifswald, Germany, which is funded by the Federal Ministry of Education and Research (grant no.
@@
 This work was supported by the DFG Priority Program 1335: Scalable Visual Analytics.
+\end{small}
 

### 2014-12-04 draft.tex
shortened 'data set' section

@@
 %%
-\noindent Our approach allows the simultaneous analysis of many variables.
-%%
-Therefore, epidemiologists compiled the data set with a wide range of variables possibly correlating with lumbar back pain.
-%%
-The data set comprises 6,753 subjects from two cohorts (4,420 from \texttt{SHIP-Trend-0} and 2,333 from \texttt{SHIP-2}) including non-image and image-derived data.
+% \noindent Our approach allows the simultaneous analysis of many variables.
+%%
+%Therefore, epidemiologists compiled the data set with a wide range of variables possibly correlating with lumbar back pain.
+\noindent Epidemiologists compiled a data set with a wide range of variables possibly correlating with lumbar back pain, comprising non-image and image-derived data for 6,753 subjects
+%%
+%The data set comprises 6,753 subjects from two cohorts (4,420 from \texttt{SHIP-Trend-0} and 2,333 from \texttt{SHIP-2}) including non-image and image-derived data.
 %%
@@
 %%
-The data set comprises 134 variables (and 9 additional image-derived variables described in the next subsection):
+The data set comprises 134 variables: %(and 9 additional image-derived variables described in the next subsection):
 \begin{itemize}
-	\item 21 metric variables, mostly describing somatometric variables and markers retrieved from blood analysis
+	\item 21 metric variables, describing somatometric variables and markers retrieved from blood analysis
 	%%
@@
 %%
-Many variables comprise missing data.
-%%
-Some variables are follow-up questions, covering the treatment of a disease.
-%%
-Others are exclusive to subgroups, such as women-specific variables, e.g. \emph{menstrual status}.
-%%
-Sparse variables are statistically less resilient and need to be treated with care during the analysis.
-\\\\
+% Many variables comprise missing data.
+% %%
+% Some variables are follow-up questions, covering the treatment of a disease.
+% %%
+% Others are exclusive to subgroups, such as women-specific variables, e.g. \emph{menstrual status}.
+% %%
+% Sparse variables are statistically less resilient and need to be treated with care during the analysis.
 \noindent \textbf{Image-Derived Data.}
@@
   \caption{
-	(a) left: Tetrahedron-based finite element model (FEM) of the lumbar spine, capturing the L1-L5 vertebrae as well as the lumbar spine canal.
-	%%
-	(a) right: FEM aligned to an MRI data set.
-	%%
-	The purple dashed line represents the centerline, describing the lumbar spine canal with 92 points.
-	%%
-	(b) Using the Frenet frame, we extract the weighted sum of curvature and torsion for all 92 points (white dashes).
-	%%
-	The curvature angle ($\alpha$) is calculated using the top, bottom and middle point.
-	%%
-	We also extracted the variables for each projection axis to assess their information gain.
+	(a) Finite element model (FEM) of the lumbar spine (left), capturing the L1-L5 vertebrae and the lumbar spine canal (right).
+	%%
+%	(a) right: FEM aligned to an MRI data set.
+	%%
+	%The purple dashed line represents the centerline, describing the lumbar spine canal with 92 points.
+	The purple dashed line describes the lumbar spine canal centerline with 92 points.
+	%%
+	(b) We extract the weighted sum of \emph{curvature} and \emph{torsion} for all 92 points (white dashes).
+	%%
+	The \emph{curvature angle} ($\alpha$) is calculated using the top, bottom and middle point.
+	%%
+	We extracted the variables for each projection axis to assess their information gain.
 	}
@@
 \end{figure}
-Four trained radiologists acquired the medical image data for each subject in a standardized way using a 1.5~Tesla Magnetic Resonance Imaging (MRI) scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany).
-%%
-The spine protocol consisted of a sagittal $T_1$- and $T_2$-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
-%%
-
+Magnetic Resonance Imaging (MRI) scans were obtained for each subject \cite{Hegenscheid2013}.
+%%
+% Four trained radiologists acquired the medical image data for each subject in a standardized way using a 1.5~Tesla Magnetic Resonance Imaging (MRI) scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany).
+% %%
+% The spine protocol consisted of a sagittal $T_1$- and $T_2$-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
+% %%
+%%
 A hierarchical finite element method was used to detect the lumbar spine in the MRI data \cite{Rak2013}.
 %%
-The tetrahedron-based finite element model is initialized using three user-defined landmarks.
-%%
-A click on the L3 vertebra center initializes the model, two clicks on the top and bottom of the vertebra describe an initial height estimation (Fig.~\ref{fig:centerline}~a).
-%%
-The registered model captures information about the lumbar spine canal shape and the position of the L1-L5 vertebrae.
-%%
-Each subject's model is mapped to a common reference frame using procrustes analysis to compensate for differences in translation, 
-rotation and scaling.
-%%
-The detection fails for several subjects due to imprecise initialization, imaging artifacts or strongly deformed spines.
-%%
-$2,540$ out of $6,753$ lumbar spine models were obtained and constitute the foundation of this work.
-%%
-
-We have to assess the model accuracy to extract key figures from it.
-%%
-The detection model depicts the vertebrae positions, but lacks detailed information about their volume.
-%%
-It captures reliable information about spine canal curvature.
-%%
-In a previous work (\cite{Klemm2013VMV}), we extracted a centerline representation of the lumbar spine canal from the detection model (Fig.~\ref{fig:centerline}~a).
-
-%%
-Using the Frenet frame, we calculated the following metrics from the model (Fig.~\ref{fig:centerline}~b) \cite{Frenet}:
+%The tetrahedron-based finite element model (Fig.~\ref{fig:centerline}~a) is initialized using three user-defined landmarks.
+%%
+%A click on the L3 vertebra center initializes the model, two clicks on the top and bottom of the vertebra describe an initial height estimation (Fig.~\ref{fig:centerline}~a).
+%%
+The tetrahedron-based finite element model (Fig.~\ref{fig:centerline}~a) captures information about the lumbar spine canal shape and the position of the L1-L5 vertebrae.
+%The registered model captures information about the lumbar spine canal shape and the position of the L1-L5 vertebrae.
+%%
+%Each subject's model is mapped to a common reference frame using procrustes analysis to compensate for differences in translation, rotation and scaling.
+%%
+The detection fails for several subjects due imaging artifacts or strongly deformed spines, yielding models for $2,540$ out of $6,753$ subjects.
+%%
+%$2,540$ out of $6,753$ lumbar spine models were obtained and constitute the foundation of this work.
+%%
+%We have to assess the model accuracy to extract key figures from it.
+%%
+The detection model depicts the vertebrae positions, spine canal curvature, but lacks detailed information about their volume.
+%%
+%It captures reliable information about spine canal curvature.
+%%
+In \cite{Klemm2013VMV}, we extracted a centerline representation of the lumbar spine canal from the detection model (Fig.~\ref{fig:centerline}~a).
+
+%%
+Using the Frenet frame, we calculated the following metrics from the model (Fig.~\ref{fig:centerline}~b): %\cite{Frenet}:
 %%
@@
 %%
-These metrics are also extracted in the sagittal, coronal and transversal projection of the model.
-%%
-We assess the information gain of each dimension using these projections.
-%%
-This yields a total of 9 image-derived variables.
+% These metrics are also extracted in the sagittal, coronal and transversal projection of the model.
+% %%
+% We assess the information gain of each dimension using these projections.
+% %%
+% This yields a total of 9 image-derived variables.
+These metrics are also extracted in the sagittal, coronal and transversal projection of the model, yielding 9 image-derived variables.
 %%

### 2014-12-05 draft.tex
shortened 'related work' section

@@
 %%
-In a previous work, we proposed an extension to the previously described epidemiological workflow using IVA \cite{Klemm2014VIS}.
-%%
-The workflow consists of an iterative sequence of \emph{group selection}, \emph{variable selection} and \emph{visualization}.
-%%
-It aims to trigger \emph{hypothesis generation} by providing visualizations able to concurrently analyze multiple heterogenous variables at once using correlation measures.
+In a previous work, we proposed an extension to the previously described epidemiological workflow using IVA, consisting of an iterative sequence of \emph{group selection}, \emph{variable selection} and \emph{visualization} \cite{Klemm2014VIS}.
+%%
+%The workflow consists of an iterative sequence of \emph{group selection}, \emph{variable selection} and \emph{visualization}.
+%%
+\emph{Hypothesis generation} is amplified by concurrently analyzing heterogenous variables at once using correlation measures.
 %%
@@
 %%
-\cite{Steenwijk} propose a coordinated linked view system for both image-related and non-image data.
-%%
-It incorporates multiple plot types, such as scatter plots, parallel coordinates and time plots with brushing and linking facilities.
-%%
-Similar to our work, they quantify image data and project it into the non-image data information space.
-%%
-\cite{Turkay} follow a similar approach by deriving descriptive data metrics from image data.
+% \cite{Steenwijk} propose a coordinated linked view system for both image-related and non-image data.
+% %%
+% It incorporates multiple plot types, such as scatter plots, parallel coordinates and time plots. %with brushing and linking facilities.
+% %%
+% Similar to our work, they quantify image data and project it into the non-image data information space.
+%%
+\cite{Turkay} derive descriptive data metrics from image data.
 %%
@@
 %%
-Generalized Pairs Plots (\texttt{GPLOM'S}) visualize heterogenous variables in a plot matrix \cite{GPLOMS,Francois}.
+Generalized Pairs Plots (\texttt{GPLOM'S}) visualize heterogenous variables in a plot matrix \cite{GPLOMS}. %\cite{GPLOMS,Francois}.
 %%
@@
 \noindent \textbf{Decision Rule-Driven Analysis of Medical Data.}
-Related work in this field often focuses on clinical diagnosis.
+%%
+% Related work in this field often focuses on clinical diagnosis.
 %%
@@
 %%
-The effort of analyzing decision trees increases with their complexity.
+% The effort of analyzing decision trees increases with their complexity.
 %%
@@
 %%
-They present an interactive data mining tool, which can analyze association rules and highlight interesting relations, which may trigger new hypotheses about the data.
+They present an interactive data mining tool, which can analyze association rules and highlight interesting relations. %, which may trigger new hypotheses about the data.
 %%
 We combine both ideas by validating the significance of image-derived variables.
-We aim at deriving new relationships by training decision trees to explain target variables.
+% %%
+% We aim at deriving new relationships by training decision trees to explain target variables.
 %%----------------------------
 %%
-\cite{Pinheiro} applied association rule mining to create high-confidence decision rules based on variables, such as age, gender and geographic location about the mortality rate in liver cancer data.
+%\cite{Pinheiro} applied association rule mining to create high-confidence decision rules based on variables, such as age, gender and geographic location about the mortality rate in liver cancer data.
 %%
@@
 %%
-Clustering techniques are strongly dependent on the chosen variables and distance measures \cite{Klemm2014BVM}.
+%Clustering techniques are strongly dependent on the chosen variables and distance measures \cite{Klemm2014BVM}.
 %%
@@
 %%
+% TODO: [high] Unterschied zu VMV Publikation herausarbeiten
 \section{\uppercase{The Lumbar Spine Data Set}}

### 2014-12-05 draft.tex
shortened 'Interactive Decision Tree Visualization' section

@@
 %%
-They present an interactive data mining tool, which can analyze association rules and highlight interesting relations. %, which may trigger new hypotheses about the data.
+They present an interactive data mining tool, which can analyze association rules and highlight relations. %, which may trigger new hypotheses about the data.
 %%
@@
 	\end{itemize}
-%%
 \end{itemize}
 %%
+All categorical variables are converted into binary \emph{dummy variables}, indicate the presence or absence of a categorical variable manifestation.
+%with multiple manifestations into several dichotomous variables.
+%%
+%Each dichotomous variable encodes the presence of a manifestation.
+%%
+For example, a pain indicator variable ranging from \emph{1 - no pain} to \emph{4 - large pain} is subdivided into four dichotomous variables.
+%%
+%One subject can only have one of these variables set to true.
+%%
+This is useful for our classification, because it allows to determine which manifestation of a variable can be described best using the image data variables.
+%%
 % Many variables comprise missing data.
@@
 % Sparse variables are statistically less resilient and need to be treated with care during the analysis.
+\\\\
 \noindent \textbf{Image-Derived Data.}
@@
 In \cite{Klemm2013VMV}, we extracted a centerline representation of the lumbar spine canal from the detection model (Fig.~\ref{fig:centerline}~a).
-
+%%
 %%
@@
 \noindent Spine shape is possibly associated with several somatometric variables.
-%%
+%% TODO: Shorten
 Larger people have also a longer spine with a straighter shape.
@@
 %%
-\section{\uppercase{Evaluation of Decision Trees}}
+% \section{\uppercase{Evaluation of Decision Trees}}
+\section{\uppercase{Interactive Decision Tree Visualization}}
 \label{sec:DecisionTrees}
@@
 %%
-They work with numerical as well as categorical data.
+%They work with numerical as well as categorical data.
 %%
@@
 %%
-Large, complex trees are difficult to trace and to understand.
-%%
-Too many branches also impose overfitting to the data \cite{DecisionTree}.
-
+%Large, complex trees are difficult to trace and to understand.
+%%
+Too many branches impose overfitting to the data \cite{DecisionTree}.
+%%
+\\\\
+\noindent \textbf{Creating Decision Trees.}
 We use the C4.5 algorithm, which builds decision trees based on information entropy on a data set.
@@
 % In the following analysis, we strongly focus on the complexity of decision trees and the classification accuracy.
-
-\subsection{Interactive Display of Decision Trees}
+\subsection{Visualizing Decision Trees}
+\label{subsec:VisualizingDecisionTrees}
 We have to create a decision tree for every non-image variable to analyze which one can be explained by image-derived variables.
@@
 %%------------
-\subsubsection{Visualization of Classification Results}
-\label{subsec:VisualizationOfClassificationResults}
+\\\\
+\noindent \textbf{Visualization of Classification Results.}
 %%
@@
 \\\\
-\noindent \textbf{The Error Term. }
-Normally the error for a classification is calculated with $error = \frac{correctlyClassified_{n}} {N}$ where $n$ is the number of subjects.
-%%
-The metric usually works well for variables with uniform distribution.
-%%
-It distorts the result for other distribution types.
-%%
-For example, if a variable indicating a disease is negative for 90\% of the subjects and the classifier simply assigns all subjects to \emph{not ill}, the error metric would yield an error of 10\%, even though it is very bad.
+\noindent \textbf{The Error Term.}
+Calculating the mean classification error is imprecise for non-uniform distributions.
+%Normally the error for a classification is calculated with $error = \frac{correctlyClassified_{n}} {N}$ where $n$ is the number of subjects.
+%%
+%The metric usually works well for variables with uniform distribution.
+%%
+%It distorts the result for other distribution types.
+%%
+For example, if a variable indicating a disease is negative for 90\% of the subjects and the classifier simply assigns all subjects to \emph{not ill}, we get an mean error of 10\%, even though it is very bad.
 %%
@@
 \end{equation}
+%%
 $M$ represents the set of manifestation of each variable.
@@
 \\\\
-\noindent \textbf{Attribute Mapping.}
-%%
-The scatter plot axes are defined by tree size and the previously described error metric.
+\textbf{Attribute Mapping.}
+%%
+The scatter plot axes are defined by \emph{tree size} and the previously described \emph{error metric}.
 %%
@@
 Classification and comparison of variables for subject groups (e.g. male and female subjects) in one plot can be achieved by color coding group affiliation on the data points.
-
+%%
 Many interdependent variables are sparse, such as \emph{medication of diabetes} or \emph{reason of early retirement}.
@@
 %%
-
 We apply a square root scale for the tree size axis to highlight decision trees with few decision rules.
@@
 Outlier results with very large decision trees would otherwise distort the resulting plot.
-%%
-\subsubsection{Dummy Variables}
-Dummy variables convert a categorical variable with multiple manifestations into several dichotomous variables.
-%%
-Each dichotomous variable encodes the presence of a manifestation.
-%%
-For example, a pain indicator variable ranging from \emph{1 - no pain} to \emph{4 - large pain} is subdivided into four dichotomous variables.
-%%
-One subject can only have one of these variables set to true.
-%%
-This is useful for our classification, because it allows to determine which manifestation of a variable can be described best using the image data variables.
 %%------------
-\subsubsection{Interaction With the Visualization}
+\subsection{Interaction With the Decision Tree Visualization}
 %%
@@
 %%------------
-\subsubsection{Implementation}
-All analyses are carried out using \texttt{R}, a widely used programming language for statistical calculations and visualizations.
+\\\\
+\noindent \textbf{Implementation.}
+All analyses are available at \url{blind.dnsalias.com} and are carried out using \texttt{R}, a widely used programming language for statistical calculations and visualizations.
 %%
@@
 %%
-As opposed to the standard \texttt{R} plots, \texttt{ggvis} allows to adjust visualization variables using user interface controls, such as sliders.
-%%
-In order to make the train of thought comprehensible, we used \texttt{RMarkdown}, which allows to create reports by combining \texttt{R} with the \texttt{Markdown} syntax.
-%%
-We used \texttt{R Shiny}\footnote{Developed by RStudio, Inc; \texttt{shiny.rstudio.com}} to make the report available as dynamic web application.
-%%
-It allows to combine the power of static \texttt{R Markdown} reports with dynamically parameterized \texttt{ggvis} plots.
-%%
-Furthermore, calculations based on a prior data selection can be redone within the report.
-%%
 The web-based approach allows us to quickly exchange results with our collaborating epidemiological experts.
@@
 %%
-Exchanging the prototype becomes as easy as exchanging a hyperlink.
-%%
-The prototype is available at \url{blind.dnsalias.com}.
+% As opposed to the standard \texttt{R} plots, \texttt{ggvis} allows to adjust visualization variables using user interface controls, such as sliders.
+% %%
+% In order to make the train of thought comprehensible, we used \texttt{RMarkdown}, which allows to create reports by combining \texttt{R} with the \texttt{Markdown} syntax.
+% %%
+% We used \texttt{R Shiny}\footnote{Developed by RStudio, Inc; \texttt{shiny.rstudio.com}} to make the report available as dynamic web application.
+% %%
+% It allows to combine the power of static \texttt{R Markdown} reports with dynamically parameterized \texttt{ggvis} plots.
+% %%
+% Furthermore, calculations based on a prior data selection can be redone within the report.
+%%
+% The web-based approach allows us to quickly exchange results with our collaborating epidemiological experts.
+% %%
+% They can use the technique without installing any software.
+% %%
+% Exchanging the prototype becomes as easy as exchanging a hyperlink.
+% %%
+% The prototype is available at \url{blind.dnsalias.com}.
 %%---------
@@
 %%
-The x-axis shows the number of decisions of the underlying model, the y-axis the classification error (see Section~\ref{subsec:VisualizationOfClassificationResults}).
+The x-axis shows the number of decisions of the underlying model, the y-axis the classification error (see Section~\ref{subsec:VisualizingDecisionTrees}).
 %%

### 2014-12-05 draft.tex
shortened 'Interactive Decision Tree Visualization' section

@@
 In the following analysis, we strongly focus on the complexity of decision trees and the classification accuracy.
-
+%
 %\subsection{The C4.5/C5.0 Algorithm}
@@
 % In the following analysis, we strongly focus on the complexity of decision trees and the classification accuracy.
-\subsection{Visualizing Decision Trees}
-\label{subsec:VisualizingDecisionTrees}
+%\subsection{Visualizing Decision Trees}
+
 We have to create a decision tree for every non-image variable to analyze which one can be explained by image-derived variables.
@@
 %%------------
-\subsection{Interaction With the Decision Tree Visualization}
+\\\\
+\noindent \textbf{Interaction With the Decision Tree Visualization.}
 %%
@@
 %%
-The x-axis shows the number of decisions of the underlying model, the y-axis the classification error (see Section~\ref{subsec:VisualizingDecisionTrees}).
+The x-axis shows the number of decisions of the underlying model, the y-axis the classification error (see Section~\ref{sec:DecisionTrees}).
 %%

### 2014-12-05 draft.tex
shortened 'Experiments and Preliminary Results' section

@@
 %%
-\noindent Spine shape is possibly associated with several somatometric variables.
-%% TODO: Shorten
+\noindent Spine shape is influenced by several somatometric variables.
+%% TODO: Introduce this better! Go into detail on the train of thought
 Larger people have also a longer spine with a straighter shape.
 %%
-Since men are on average taller than women and people of old age shrink due to \emph{bone erosion}, \emph{gender} and \emph{age} are also risk factors.
-%%
-Women have a higher life expectancy than men, hence their higher share in the old age group.
-%%
-Also, women are on average smaller than men, hence the larger shape similarity with older subjects.
-%%
-Large \emph{body weight} increases the spine load, resulting in a bent shape.
+%Since men are on average taller than women and people of old age shrink due to \emph{bone erosion}, \emph{gender} and \emph{age} are also risk factors.
+%%
+%Women have a higher life expectancy than men, hence their higher share in the old age group.
+%%
+%Also, women are on average smaller than men, hence the larger shape similarity with older subjects.
+%%
+High \emph{body weight} increases the spine load, resulting in a bent shape.
 %%
@@
 \noindent \textbf{GPLOM Analysis.}
-As a first experiment we correlated the shape variable with the dichotomous back pain indicator using \texttt{GPLOM's} \cite{GPLOMS}.
-%%
-Since our image-derived variables are metric, their pairwise combinations are visualized using scatter plots.
+As a first experiment we correlated the shape variable with the dichotomous back pain indicator using a \texttt{GPLOM}.
+%%
+The metric image-derived variables are pairwise visualized using scatter plots.
+%Since our image-derived variables are metric, their pairwise combinations are visualized using scatter plots.
 %%
@@
 The projections to the transversal planes attract attention as they have many outliers.
-%%
-We conclude that \emph{curvature} is not as reliable on the transversal plane as it is on the other planes.
+%% % TODO: [high] Add Link
+We conclude that \emph{curvature} is not as reliable on the transversal plane as it is on the other planes, which was also confirmed through a Principal Component Analysis (see supplementary material at TODO:LINK).
+%%
+The \texttt{GPLOM} shows very similar distributions of subjects with or without back pain with respect to the shape variables in all sub plots.
+%%
 %%--------------------------------
-\\\\
-\noindent \textbf{Correlation Matrix.}
+% \\\\
+% \noindent \textbf{Correlation Matrix.}
 \begin{figure*}[htb]
@@
 \end{figure*}
-We calculated a correlation matrix to assess correlations between the image variables.
-%%
-The pearson correlation coefficients between the numeric variables are depicted right of the matrix diagonal (Fig.~\ref{fig:image-parameter-range}).
-%%
-\emph{Curvature}, \emph{curvature angle} and \emph{torsion} strongly correlate with their planar projections.
-%%
-Also the \emph{mean curvature} and \emph{curvature angle} correlate by a factor of $-0.89$.
-%%
-\emph{Torsion} does not correlate with any other image-derived variable.
+% We calculated a correlation matrix to assess correlations between the image variables.
+% %%
+% The pearson correlation coefficients between the numeric variables are depicted right of the matrix diagonal (Fig.~\ref{fig:image-parameter-range}).
+% %%
+% \emph{Curvature}, \emph{curvature angle} and \emph{torsion} strongly correlate with their planar projections.
+% %%
+% Also the \emph{mean curvature} and \emph{curvature angle} correlate by a factor of $-0.89$.
+% %%
+% \emph{Torsion} does not correlate with any other image-derived variable.
 %%--------------------------------------------------------
-\\\\
-\noindent \textbf{Correlation of Image Variables With Lumbar Back Pain.}
-Figure~\ref{fig:image-parameter-range} shows the distribution of all image-variables as Generalized Pairs Plot.
-%%
-No statistically significant correlation could be observed through all subject groups.
-%%
-The box plots show no difference between subjects with or without back pain.
-\\\\
-\noindent \textbf{Assessing the Information Gain Using the Principal Component Analysis.}
-%%
-To determine the information gain per image-derived variable, we calculated a Principal Component Analysis (PCA) and compared the dimension loadings.
-%%
-The PCA results in a linear projection of the variable space ordering their components by maximum variance.
-%%
-The first three principal components explain 75\% of the variance in the nine image-derived variables.
-%%
-The first principal component explains 47\% of the variance and comprises primarily \emph{mean curvature}, \emph{sagittal curvature}, \emph{curvature angle} and \emph{sagittal curvature angle}.
-%%
-The second component, adding 16\% of the variance, weigths \emph{coronal curvature} and \emph{coronal curvature angle}.
-%%
-The third component explains 12\% of the variance and comprises \emph{torsion} and \emph{transverse curvature}.
-%%
-This supports our prior conclusion about the low information gain of the transverse planes.
-%%
-\emph{Torsion} also adds little variance to the information space.
-%%
-The complete analysis can be found in the experiments section of the supplementary material (\url{blind.dnsalias.com})
+% \\\\
+% \noindent \textbf{Correlation of Image Variables With Lumbar Back Pain.}
+% Figure~\ref{fig:image-parameter-range} shows the distribution of all image-variables as Generalized Pairs Plot.
+% %%
+% No statistically significant correlation could be observed through all subject groups.
+% %%
+% The box plots show no difference between subjects with or without back pain.
+% \\\\
+% \noindent \textbf{Assessing the Information Gain Using the Principal Component Analysis.}
+% A Principal Component Analysis (PCA) of the image data yielded that \emph{mean curvature}, \emph{sagittal curvature}, \emph{curvature angle} and \emph{sagittal curvature angle} explain almost 50\% of the image variance.
+% %%
+% \emph{Torsion} and metrics of the \emph{transverse} projection planes contribute very little variance, supporting our conclusion about their low information gain.
+%%
+%%
+% To determine the information gain per image-derived variable, we calculated a Principal Component Analysis (PCA) and compared the dimension loadings.
+% %%
+% The PCA results in a linear projection of the variable space ordering their components by maximum variance.
+% %%
+% The first three principal components explain 75\% of the variance in the nine image-derived variables.
+% %%
+% The first principal component explains 47\% of the variance and comprises primarily \emph{mean curvature}, \emph{sagittal curvature}, \emph{curvature angle} and \emph{sagittal curvature angle}.
+% %%
+% The second component, adding 16\% of the variance, weigths \emph{coronal curvature} and \emph{coronal curvature angle}.
+% %%
+% The third component explains 12\% of the variance and comprises \emph{torsion} and \emph{transverse curvature}.
+% %%
+% This supports our prior conclusion about the low information gain of the transverse planes.
+% %%
+% \emph{Torsion} also adds little variance to the information space.
+% %%
+% The complete analysis can be found in the experiments section of the supplementary material (\url{blind.dnsalias.com})
 %% ------------------------------------------
@@
 %%
-The method uses the following correlation metrics for the different type combinations:
-%%
-\begin{itemize}
-%%
-\item \emph{Pearson product-moment} for two continuous variables, 
-\item \emph{Polyserial} correlation for one continuous and one categorical variable, and
-\item \emph{Polychoric} correlations for two categorical variables.
-\end{itemize}
-%%
-All correlation values are scaled between \texttt{0 - no correlation} and \texttt{1 - perfect correlation}.
-%%
-Some variables are too sparse for calculating correlations, for example \emph{treatment of diabetes}, or \emph{medication against high blood pressure}.
-%%
-These variables have to be omitted in the analysis, since they are not statistically resilient.
-%%
-We display the resulting \emph{contingency matrix} using a heat map, encoding correlation values with color brightness with white for $0$ and dark blue for $1$, as presented in \cite{Klemm2014VIS}.
+The method uses \emph{pearson product-moment} for two continuous variables, \emph{polyserial} correlation for one continuous and one categorical variable, and \emph{Polychoric} correlations for two categorical variables.
+%The method uses the following correlation metrics for the different type combinations:
+%%
+% \begin{itemize}
+% %%
+% \item \emph{Pearson product-moment} for two continuous variables,
+% \item \emph{Polyserial} correlation for one continuous and one categorical variable, and
+% \item \emph{Polychoric} correlations for two categorical variables.
+% \end{itemize}
+%%
+All values are scaled between \texttt{0 - no correlation} and \texttt{1 - perfect correlation}.
+%%
+Some variables are too sparse for calculating correlations, for example \emph{treatment of diabetes}, or \emph{medication against high blood pressure} and are omitted, since they are not statistically resilient.
+%%
+We display the resulting \emph{contingency matrix} using a heat map, encoding correlation values with color brightness with white for $0$ and dark blue for $1$.
 %%
@@
 
-\vfill
+% \vfill
 \bibliographystyle{style/apalike}

### 2014-12-08 draft.tex
included Bernhards revision

@@
 %%
-As a first result, we conclude that image-based variables are only sufficient for classification of lifestyle factors within the data.
-%%
-A correlation between lumbar spine shape and lower back pain was not existent in the automatically trained classifiers.
+As a first result, we conclude that image-based variables are only sufficient to predict lifestyle factors within the data.
+%%
+A correlation between lumbar spine shape and lower back pain could not be found with the automatically trained classifiers.
 %%
@@
 %%
-For this purpose, large population studies, such as the \emph{Study of Health in Pomerania} (SHIP) \cite{SHIP}, gather as much information as possible about participants to be assessed towards different diseases.
-%%
-This information is used to determine risk factors for diseases, help people to live healthier or to support the diagnosis of widespread diseases.
+Large population studies, such as the \emph{Study of Health in Pomerania} (SHIP) \cite{SHIP}, gather as much information as possible about participants to be assessed towards different diseases.
+%%
+This information is used to determine risk factors for diseases, inform people about healthier lifestyles or to support the diagnosis of widespread diseases.
 %%
@@
 %%
-Nowadays, these studies often comprise medical image data, e.g., tomographic image data sets. %, as well as general patient-specific attributes, such as \emph{gender}, \emph{age} and \emph{weight}.
+%Nowadays, these studies often comprise medical image data, e.g., tomographic image data sets. %, as well as general patient-specific attributes, such as \emph{gender}, \emph{age} and \emph{weight}.
 %%
@@
 %%
-Epidemiological analysis of back pain aims to determine risk factors.
-%%
-Although the shape and constitution of the spine, especially the lumbar spine, plays an important role for back pain, an automatic classification approach for characterization of back pain based on lumbar back spine attributes is still missing.
+%Epidemiological analysis of back pain aims to determine risk factors.
+%%
+Although the shape and constitution of the spine, especially the lumbar spine, plays an important role for back pain, an automatic classification approach for characterization of back pain based on lumbar spine attributes is still missing.
 %%
@@
 %%
-These studies aim at collecting as much data as possible for each subject.
+% These studies aim at collecting as much data as possible for each subject.
 %%
@@
 %%
-Epidemiologists want to characterize the relation between the aging and degenerative process of the spine \cite{Szpalski2005}.
+Epidemiologists want to characterize the relation between aging and degenerative process of the spine \cite{Szpalski2005}.
 %%
@@
 %%
-\noindent \textbf{Visual Analysis of Medical Image and Non-Image Data}
-%%
-In a previous work, we proposed an extension to the previously described epidemiological workflow using IVA, consisting of an iterative sequence of \emph{group selection}, \emph{variable selection} and \emph{visualization} \cite{Klemm2014VIS}.
+\noindent \textbf{Visual Analysis of Medical Image and Non-Image data}
+%%
+In a previous work, we proposed an epidemiological workflow using IVA, consisting of an iterative sequence of \emph{group selection}, \emph{variable selection} and \emph{visualization} \cite{Klemm2014VIS}.
 %%
@@
 %%
-A survey on image-centric cohort studies and strategies to analyze the resulting data is given in \cite{Preim2014}.
+A survey on image-centric cohort studies and strategies to analyze the resulting data is given in \cite{Preim2015}.
 %%
@@
 %%
-\texttt{GPLOM'S} are well suited for an overview visualization, but take up much screen space and are therefore only suitable for few variables at once (see example in Fig.~\ref{fig:image-parameter-range} later on).
+\texttt{GPLOM'S} provide an overview, but take up much screen space and are therefore only suitable for few variables at once (see example in Fig.~\ref{fig:image-parameter-range} later on).
 %%
@@
 %%
-Decision trees are easily readable and are well suited for classifying medical data.
+Decision trees are easily readable and are frequently used for classifying medical data.
 %%
@@
 %%
-\cite{Niemann2014} assess hepatic steatosis (fatty liver disease) risk factors using decision trees.
+\cite{Niemann2014} assess risk factors for hepatic steatosis (fatty liver disease) risk factors using decision trees.
 %%
@@
 %%
-The detection fails for several subjects due imaging artifacts or strongly deformed spines, yielding models for $2,540$ out of $6,753$ subjects.
+The detection fails for several subjects due to imaging artifacts or strongly deformed spines, yielding models for $2,540$ out of $6,753$ subjects.
 %%
@@
 
-These observations brought us to the decision to incorporate sophisticated data mining techniques to assess the influence of the image-derived variables.
+These observations brought us to the decision to incorporate data mining techniques to assess the influence of the image-derived variables.
 %%
@@
 %%
-Decision trees are a established classification method in data mining to create predictive models.
+Decision trees are used in data mining to create predictive models.
 %%
@@
 %%
-For example, if a variable indicating a disease is negative for 90\% of the subjects and the classifier simply assigns all subjects to \emph{not ill}, we get an mean error of 10\%, even though it is very bad.
+For example, if a variable indicating a disease is negative for 90\% of the subjects and the classifier simply assigns all subjects to \emph{not ill}, we get a mean error of 10\%, even though it is very bad.
 %%
@@
 \\\\
-\noindent \textbf{Interaction With the Decision Tree Visualization.}
-%%
-The described visualization provides a good overview over the classification results.
+\noindent \textbf{Interaction with the Decision Tree Visualization.}
+%%
+The described visualization provides a good overview of the classification results.
 %%
@@
 %%
-Selecting a variable using a drop-down menu allows the user to select the variable used for subject subdivisions, e.g. \emph{gender} or \emph{employment status}.
+The subject subdivision is controlled by the selection of variables, such as \emph{gender} or \emph{employment status}.
+%Selecting a variable using a drop-down menu allows the user to select the variable used for subject subdivisions, e.g. \emph{gender} or \emph{employment status}.
 %%
@@
 The stress-induced spine deformation seems to directly influence the pain levels for these subjects.
-
+%%
 The dummy variable analysis shows many results using a decision tree with one rule based on \emph{mean curvature} or \emph{curvature angle} with an error between $0.35$ and $0.47$.
@@
 %
-The question is whether the inter-group spine variability parameter is sufficient for predicting other variables or not.
+The question is whether the inter-group spine variability parameter is sufficient for predicting other variables.
 %%
@@
 
-Another observation during the analysis is that large decision trees achieve low error scores.
-%%
-Their complexity imposes overfitting to the data and they are not suitable for extracting universal rules.
-%%
-\\\\
+% Another observation during the analysis is that large decision trees achieve low error scores.
+%%
+% Their complexity imposes overfitting to the data and they are not suitable for extracting universal rules.
+%%
+%\\\\
 \noindent \textbf{Applicability.}
@@
 We provide a method for comparative analysis of decision trees independent of the variable manifestation count using interactive scatter plots.
-%%
+%% TODO: [high] This is no related work
 We applied the technique to gain insight into the predictive power of 9 image-derived variables for 134 non-image variables.
@@
 
+%% TODO: [high] Here starts the future work
 In our future work, we will focus on more precise models for extracting measures.
@@
 \\\\
-Combining the power of statistical analysis, visual analytics and data mining techniques are essential for analyzing increasingly complex heterogenous population data.
+%% TODO: [high] This is a summary, not future work
+Combining the power of statistical analysis, visual analytics and data mining techniques is essential for analyzing increasingly complex heterogenous population data.
 %%

### 2014-12-08 Rebuttal.md
added Rebuttal file for tracking changes due to reviews

# Rebuttal

## Reviewer 1.
- Applicability: Comparison to standard workflow is not possible. The classic epidemiological workflow is strictly hypothesis driven and therefore requires a small set of 2 - 3 variables, which are correlated with a target variable. IVA methods amplify *hypothesis generation* and therefore aim to provide insight into unexpected relationships in the data.

### 2014-12-08 draft.tex
revised results figure
revised Summary and Conclusion section

@@
 %%
-Since the \emph{gender} encodes \emph{body size}, we decided to divide subjects into \emph{body size} groups.
+%Since the \emph{gender} encodes \emph{body size}, we decided to divide subjects into \emph{body size} groups.
 %%
@@
 %%
-Further subdivision, e.g. by quantiles of \emph{body size} increases the number to 402 trees.
+Further subdivision, e.g. by quantiles of \emph{body mass index} increases the number to 402 trees.
 %%
@@
 \section{\uppercase{Results}}
-\begin{figure*}[p!]
+% \begin{figure*}[p!]
+\begin{figure*}[!t]
   %\vspace{-0.2cm}
   \centering
-  \includegraphics[width=0.95\textwidth]{figures/results}
+  \includegraphics[width=1\textwidth]{figures/results}
   \caption{
@@
 %%
-no group (a), subdivision into \emph{male} and \emph{female} subjects (b), quartiles of \emph{Body Mass Index} (\emph{BMI}) (c) and quartiles of \emph{body size} (d).
+no group (a), subdivision into \emph{male} and \emph{female} subjects (b) and quartiles of \emph{Body Mass Index} (\emph{BMI}) (c). % and quartiles of \emph{body size} (d).
 %%
@@
 %%
-We also create subject groups to assess the influence of variables affecting the lumbar spine shape.
-%%
-We carry out the analysis using different subject groups:
-\begin{enumerate}
-	%%
-	\item All subjects,
-	%%
-	\item subdivision into \emph{male} and \emph{female},
-	%%
-	\item subdivision by \emph{Body Mass Index quantiles} ($BMI = \frac{m}{l^2}$ where $m$ is the \emph{body mass} in kilogram and $l$ is the \emph{body size} in meter), yielding the groups (17, 24.7] (24.7, 27.4] (27.4, 30.5] (30.5, 48],
-	%%
-	\item subdivision by \emph{size quantiles}, yielding the groups (139, 164], (164, 171], (171, 177], (177, 202].
-	%%
-\end{enumerate}
+We create subject groups to assess the influence of variables affecting the lumbar spine shape.
+%%
+The groups are: all subjects; \emph{males} and \emph{females}; and a subdivision by \emph{Body Mass Index quantiles} ($BMI = \frac{m}{l^2}$ where $m$ is the \emph{body mass} in kilogram and $l$ is the \emph{body size} in meter), yielding the groups (17, 24.7] (24.7, 27.4] (27.4, 30.5] (30.5, 48].
+% We carry out the analysis using different subject groups:
+% \begin{enumerate}
+% 	%%
+% 	\item All subjects,
+% 	%%
+% 	\item subdivision into \emph{male} and \emph{female},
+% 	%%
+% 	\item subdivision by \emph{Body Mass Index quantiles} ($BMI = \frac{m}{l^2}$ where $m$ is the \emph{body mass} in kilogram and $l$ is the \emph{body size} in meter), yielding the groups (17, 24.7] (24.7, 27.4] (27.4, 30.5] (30.5, 48],
+% 	%%
+% 	\item subdivision by \emph{size quantiles}, yielding the groups (139, 164], (164, 171], (171, 177], (177, 202].
+% 	%%
+% \end{enumerate}
 %%
@@
 %%
-The first plot shows all original variables.
-%%
-The second plot shows all categorical variables transformed into dichotomous dummy variables. % (recall subsection~\ref{sec:C5.0}).
+The first plot shows all original variables, the second all categorical variables transformed into dichotomous \emph{dummy variables}.
+%%
+%The second plot shows all categorical variables transformed into dichotomous dummy variables. % (recall subsection~\ref{sec:C5.0}).
 %%
@@
 %%
-\subsection{All Variables}
-%%
-The vast majority of non-image variables cannot be automatically classified based on the image-variables.
+\\\\
+\noindent \textbf{All Variables.}
+%%
+The majority of non-image variables cannot be automatically classified based on the image-variables.
 %%
@@
 %%
-Almost all subjects who are medicated ($796$ of $1,058$) were correctly classified.
-%%
-The vast majority of non-medicated subjects ($262$ of $1,058$) are false-positive classified, yielding a poor quality of the classifier w.r.t epidemiological research.
+Almost all medicated subjects were correctly classified ($796$ of $1,058$).
+%%
+The majority of non-medicated subjects are false-positive classified ($262$ of $1,058$), yielding a poor quality of the classifier w.r.t epidemiological research.
 %%
@@
 %%
-\subsection{Gender Groups}
+\\\\
+\noindent \textbf{Gender Groups.}
 %%
@@
 %%
-\subsection{Body Mass Index Groups}
+\\\\
+\noindent \textbf{Body Mass Index Groups.}
 %%
@@
 %%
-\subsection{Size Groups}
-%%
-Many previously described results are influenced by \emph{subject size}.
-%%
-Differences between \emph{male} and \emph{female} subjects can be explained by the average \emph{body size} difference.
-%%
-For example, large subjects are already characterized by their rather straight spine.
-%
-The question is whether the inter-group spine variability parameter is sufficient for predicting other variables.
-%%
-Dividing subjects into \emph{body size} groups potentially highlights classifications not influenced by \emph{body size}.
-%%
-\\\\
-\noindent \textbf{Large Decision Trees.}
-Back pain-associated variables can be explained for various \emph{size} groups, but we could not extract universal rules.
-%%
-\emph{Radiating back pain} could be described with an error of $0.39$ using 23 rules for subjects between \emph{$171-177~cm$ body size} with \emph{torsion} and \emph{mean curvature}.
-%%
-For subjects sized \emph{$177-202~cm$} the error increases to $0.47$ using 20 decision rules.
-%%
-There are several decision trees for laboratory values, e.g.
+% \subsection{Size Groups}
+% %%
+% Many previously described results are influenced by \emph{subject size}.
+% %%
+% Differences between \emph{male} and \emph{female} subjects can be explained by the average \emph{body size} difference.
+% %%
+% For example, large subjects are already characterized by their rather straight spine.
+% %
+% The question is whether the inter-group spine variability parameter is sufficient for predicting other variables.
+% %%
+% Dividing subjects into \emph{body size} groups potentially highlights classifications not influenced by \emph{body size}.
+% %%
+% \\\\
+% \noindent \textbf{Large Decision Trees.}
+% Back pain-associated variables can be explained for various \emph{size} groups, but we could not extract universal rules.
+% %%
+% \emph{Radiating back pain} could be described with an error of $0.39$ using 23 rules for subjects between \emph{$171-177~cm$ body size} with \emph{torsion} and \emph{mean curvature}.
+% %%
+% For subjects sized \emph{$177-202~cm$} the error increases to $0.47$ using 20 decision rules.
+% %%
+% There are several decision trees for laboratory values, e.g.
 \emph{alanine aminotransferase} value (relevant for diagnosis of liver or gallbladder illness) in the blood can be described with an error of $0.4$ (\emph{$139-164~cm$}) to $0.36$ (\emph{$164 - 171~cm$}).
-%%
-Similar values can be observed for \emph{cholesterol} or \emph{age}.
-%%
-Due to the large decision trees, these results are not usable and impose overfitting to the data.
-%%
-\\\\
-\noindent \textbf{Small Decision Trees.}
-%%
-The dummy variables show several variables described using only one decision rule with an error between $0.42$ and $0.47$.
-%%
-Most of these variables have a dominant manifestation and the classifier shows a low detection precision for the second manifestation.
-%%
-These variables include \emph{nutrition}, \emph{thyroid disorder} and \emph{social problems induced by back pain}.
-%%
-%TODO: [medium] Muss das rein? Auswertung der Anteile der Parameter an den Ergebnissen (Torsion, Curvature, etc. - welches ist der dominanteste Parameter?)
-%%
-\section{\uppercase{Conclusion}}
+% %%
+% Similar values can be observed for \emph{cholesterol} or \emph{age}.
+% %%
+% Due to the large decision trees, these results are not usable and impose overfitting to the data.
+% %%
+% \\\\
+% \noindent \textbf{Small Decision Trees.}
+% %%
+% The dummy variables show several variables described using only one decision rule with an error between $0.42$ and $0.47$.
+% %%
+% Most of these variables have a dominant manifestation and the classifier shows a low detection precision for the second manifestation.
+% %%
+% These variables include \emph{nutrition}, \emph{thyroid disorder} and \emph{social problems induced by back pain}.
+% %%
+\section{\uppercase{Summary \& Conclusion}}
 \label{sec:Conclusion}
-\noindent The presented results indicate that the image-based variables \emph{torsion}, \emph{curvature} and \emph{curvature angle} are not sufficient to characterize pathological changes in the lumbar spine in the given quality.
-%%
-They are associated with subject \emph{gender} due to their \emph{body size} difference using \emph{mean curvature} and \emph{mean curvature angle}.
-%%
-This is due to the higher curvature of smaller subjects and the significant difference of average \emph{body size} between male and female.
-%%
-\emph{Gender} was detected well in the analysis of all subjects as well as subjects divided by \emph{BMI}.
-
-Subjects grouped using \emph{body size} have a similar spine shape and a lower variance.
-%%
-The remaining information within these groups is not sufficient to characterize back pain-related variables.
-%%
-Associations with certain \emph{nutrition} or \emph{psychological} variables can be observed with a comparatively high error for some \emph{body size} groups.
-%%
-
-% Another observation during the analysis is that large decision trees achieve low error scores.
-%%
-% Their complexity imposes overfitting to the data and they are not suitable for extracting universal rules.
-%%
-%\\\\
-\noindent \textbf{Applicability.}
-%%
-Classification methods based on decision trees have proven to be useful for assessing the discriminative power of a variable set.
-%%
-Their ability to consider variable combinations makes them more powerful than correlation coefficients calculated for each variable.
-%%
-This advantage comes with a much more complex output, the results are more challenging to assess and to abstract.
-%%
-Our method to plot derived metrics and custom-tailored error measures proved to be effective.
-%%
-Huge result spaces could be navigated fast using our interactive visualization.
-%%
-Therefore, the method is applicable not only for deriving information based on image data, but on all potential target variables.
-%%
-
-% BLIND
-%As described in a prior work (\cite{Klemm2014VIS}), the suggestion of potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
-Suggesting potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
-%%
-We can achieve this also using the presented method by trying to describe several (dichotomous) target values with a decision tree constructed from all available data.
-%%
-This allows to assess both the discriminative power of the data set regarding the target variables, and as the most important variables for the classification.
-%%
-\\\\
-\noindent \textbf{Predictive Power of Image-Derived Variables.}
-We showed that \emph{torsion}, \emph{curvature} and \emph{curvature angle} of the lumbar spine at the presented precision are not sufficient to characterize lumbar back pain in the SHIP data set.
-%%
-Our method allowed to assess their discriminative power, which is largely limited for separating male and female subjects, \emph{nutrition} variables, as well as different disease indicators.
-%%
-The C4.5 algorithm proved to be an effective tool for evaluating a set of derived metrics regarding their suitability to classify non-image variables.
-%%
-Over-fitting to the data indicated by complex decision trees has to be taken into account as well.
-%%
-\\\\
-\noindent \textbf{Future Work.}
 %% Contributions
-We provide a method for comparative analysis of decision trees independent of the variable manifestation count using interactive scatter plots.
+\noindent We provide a method for comparative analysis of decision trees independent of the variable manifestation count using interactive scatter plots.
 %% TODO: [high] This is no related work
@@
 These hypotheses, however, have to be substantiated by dedicated statistical analyses and replication in independent cohorts.
-
-%% TODO: [high] Here starts the future work
-In our future work, we will focus on more precise models for extracting measures.
-%%
-Dented vertebras are an early sign of a pathological deformation, so we want to capture this by segmenting the top and bottom point of each vertebra center.
-%%
-Another interesting metric is the spine canal thickness over the whole spine, capturing early signs of herniated disc disease.
-%%
-As another focus, we want to include the method into existing visual analytics methods designed for analyzing shape information for epidemiological data \cite{Klemm2014VIS}.
-%%
-\\\\
-%% TODO: [high] This is a summary, not future work
+% \noindent The presented results indicate that the image-based variables \emph{torsion}, \emph{curvature} and \emph{curvature angle} are not sufficient to characterize pathological changes in the lumbar spine in the given quality.
+% %%
+% They are associated with subject \emph{gender} due to their \emph{body size} difference using \emph{mean curvature} and \emph{mean curvature angle}.
+% %%
+% This is due to the higher curvature of smaller subjects and the significant difference of average \emph{body size} between male and female.
+% %%
+% \emph{Gender} was detected well in the analysis of all subjects as well as subjects divided by \emph{BMI}.
+%%
+% Subjects grouped using \emph{body size} have a similar spine shape and a lower variance.
+% %%
+% The remaining information within these groups is not sufficient to characterize back pain-related variables.
+% %%
+% Associations with certain \emph{nutrition} or \emph{psychological} variables can be observed with a comparatively high error for some \emph{body size} groups.
+% %%
+%%
+% Another observation during the analysis is that large decision trees achieve low error scores.
+%%
+% Their complexity imposes overfitting to the data and they are not suitable for extracting universal rules.
+%%
+\\\\
+\noindent \textbf{Predictive Power of Image-Derived Variables.}
+The presented results indicate that \emph{torsion}, \emph{curvature} and \emph{curvature angle} of the lumbar spine at the presented precision are not sufficient to characterize lumbar back pain in the SHIP data set.
+%%
+Our method allowed to assess their discriminative power, which is largely limited for separating male and female subjects, \emph{nutrition} variables, as well as different disease indicators.
+%%
+The C4.5 algorithm proved to be an effective tool for evaluating a set of derived metrics regarding their suitability to classify non-image variables.
+%%
+Over-fitting to the data indicated by complex decision trees has to be taken into account as well.
+\\\\
+\noindent \textbf{Applicability.}
+%%
+Methods supporting \emph{hypothesis generation} based on image information are new to the application domain.
+%%
+They are an addition to the standard epidemiological workflow as they highlight new possibly unknown relationships.
+%%
+Classification methods based on decision trees have proven to be useful for assessing the discriminative power of a variable set.
+%%
+Their ability to consider variable combinations makes them more powerful than correlation coefficients calculated for each variable.
+%%
+This advantage comes with a much more complex output, the results are more challenging to assess and to abstract.
+%%
+Our method to plot derived metrics and custom-tailored error measures proved to be effective.
+%%
+Huge result spaces could be navigated fast using our interactive visualization.
+%%
+Therefore, the method is applicable not only for deriving information based on image data, but on all \emph{potential target variables}.
+%%
+% BLIND
+%As described in a prior work (\cite{Klemm2014VIS}), the suggestion of potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
+% Suggesting potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
+% %%
+% This can be achieved by creating a describe several (dichotomous) target values with a decision tree constructed from all available data
+% We can achieve this also using the presented method by trying to describe several (dichotomous) target values with a decision tree constructed from all available data.
+% %%
+% This allows to assess both the discriminative power of the data set regarding the target variables, and as the most important variables for the classification.
+%%
+\\\\
+\noindent \textbf{Future Work.}
+We will focus on more precise models for extracting measures.
+%%
+Dented vertebras hint a pathological deformation, and can be capture by segmenting the top and bottom point of each vertebra center.
+%%
+Spine canal thickness indicates signs of herniated disc disease and is also of interest.
+%Another interesting metric is the spine canal thickness over the whole spine, capturing early signs of herniated disc disease.
+%%
+We aim to include the method into existing visual analytics methods designed for analyzing shape information for epidemiological data \cite{Klemm2014VIS}.
+%%
+\\\\
+\noindent \textbf{Outlook.}
+%%
 Combining the power of statistical analysis, visual analytics and data mining techniques is essential for analyzing increasingly complex heterogenous population data.
 %%
-These methods do not aim to replace the traditional epidemiological workflow, but rather intervene at the weak points of standard statistical methods.
+These methods do not aim to replace the traditional epidemiological workflow, but rather complement the weak points of standard statistical methods.
 %%
 Our method provides a novel way to gain insight into these complex data sets and amplifies hypotheses generation.
-
+%%
 \section*{\uppercase{Acknowledgements}}

### 2014-12-08 Rebuttal.md
minor revisions at draft and added comments to the Rebuttal

@@
-# Rebuttal
+Rebuttal
+========
 
-## Reviewer 1.
+This rebuttal is written in markdown.
+
+Reviewer 1
+----------
+
+- Figure 1 was completely removed since it does not supports the train of thought as we intended and due to lack of space
 - Applicability: Comparison to standard workflow is not possible.
@@
 IVA methods amplify *hypothesis generation* and therefore aim to provide insight into unexpected relationships in the data.
+- Exploration process is unclear
+	- at the beginning of Section 5 - Experiments and Preliminary Results, all subjects

### 2014-12-08 draft.tex
minor revisions at draft and added comments to the Rebuttal

@@
 \noindent \textbf{SHIP Data Analysis.}
-\cite{Klemm2013VMV} analyzed the lumbar spine variability of 490 \texttt{SHIP-2} subjects.
-%%
-They incorporate hierarchical agglomerative clustering to derive shape groups, yielding several groups of average shape and several outlier clusters.
+% TODO: [high] This is currently only one paper!
+In a prior work, we analyzed the lumbar spine variability of 490 \texttt{SHIP-2} subjects \cite{Klemm2013VMV}.
+%%
+We incorporate hierarchical agglomerative clustering to derive shape groups, yielding several groups of average shape and several outlier clusters.
 %%
@@
 %%
-Our data extraction process is based on the detection algorithm and lumbar spine canal extraction presented by \cite{Klemm2013VMV}.
+We use the same detection algorithm and lumbar spine canal extraction as foundation of this work.
+%Our data extraction process is based on the detection algorithm and lumbar spine canal extraction presented by \cite{Klemm2013VMV}.
 %%
@@
 %%
-\noindent Spine shape is influenced by several somatometric variables.
+%The first step is to analyze if the dichotomous back pain variable can be discriminated using single shape variables or sets of shape variables.
+%%
+\noindent In this section, the image-derived variables are analyzed towards the dichotomous \emph{back pain} indicator using a \texttt{GPLOM} and all non-image variables using heterogenous correlations.
+%In this section, we aim to retrieve correlations between the dichotomous \emph{back pain} indicator and the image-derived variables using a \texttt{GPLOM} as well as a assessment, which non-image variables
+Spine shape is influenced by several somatometric variables.
 %% TODO: Introduce this better! Go into detail on the train of thought
@@
 \noindent \textbf{GPLOM Analysis.}
-As a first experiment we correlated the shape variable with the dichotomous back pain indicator using a \texttt{GPLOM}.
+As first experiment we correlated the shape variable with the dichotomous back pain indicator using a \texttt{GPLOM}.
 %%

### 2014-12-09 Rebuttal.md
minor changes to make train of thought comprehensible; added rebuttal comments

@@
 This rebuttal is written in markdown.
+
+General Changes
+---------------
+
+- substantial reductions had to take place, since we were given a `short paper` slot with an 8-page limit.
+	- redundancy between `Introduction` and `Epidemiological Background` was minimized as much as possible
+	- `Related Work` section was shortened and covers less papers to meet the page limit.
+	- `The Lumbar Spine Data Set` section was shortened, `Figure 2` was removed since we found it did not support the train of thought as we intended
+	- `Experiments and Preliminary Results` section was shortened and changed due to the Reviewer feedback. It should be more precise and comprehensible.
+	- `Results` section now covers *all subjects*, subjects categorized by *gender* and *body mass index*
+		- `Figure 3` (former `Figure 4`) was adapted accordingly
+	- `Summary & Conclusion` was completely revised and is now structured properly
+		- Due to the lack of time until the camera ready deadline as well as the page limit, we are not able to conduct a detailed user test.
+We did, however, discuss our approach in more detail in the last section
 
@@
 - Exploration process is unclear
+	- Role of Decision Trees and why they are chosen is now captured in the introduction of section 6 - Interactive Decision Tree Visualization.
 	- at the beginning of Section 5 - Experiments and Preliminary Results, all subjects
+	- Experiments and Preliminary Results completely restructured to make it comprehensible
+- Response to Minor Comments:
+	- *GPLOM Analysis Section.
+“The combination of the image variables with back pain is visualized as histogram at the left side…” left should be bottom*: Back pain is encoded with presence (red) and absence (turquoise) in all plots left of the matrix diagonal.
+We clarified this in the text.

### 2014-12-09 draft.tex
minor changes to make train of thought comprehensible; added rebuttal comments

@@
 %%
-The combination of the image variables with back pain is visualized as histogram at the left side of the matrix and as box plot on the right side.
+The combination of the image variables with back pain is visualized as histogram at the left side of the matrix diagonal and as box plot on the right side.
 %%
@@
 %%
-We rely on predictive classification trained to obtain a complex rule set on how combinations of the image-variable explain non-image variables.
+We rely on predictive classification trained to obtain a complex rule set on how combinations of the image-variables explain non-image variables.
+%%
+%The presented method aims to support an explorative analysis by trying to identify relationships with all non-image variables and also 
+%The nature of this analysis is more explorative as it includes all 
 %%
 Decision trees are used in data mining to create predictive models.
+%%
+These models are build w.r.t all input variables and capture more complex relationships than correlation coefficients.
 %%

### 2014-12-09 Rebuttal.md
removed SHIP subsection in the Related Work section; added information about differences towards the VMV

@@
 	- `Related Work` section was shortened and covers less papers to meet the page limit.
+		- as requested by [Reviewer 3], we elaborate on the differences towards the VMV paper in the `Related Work` section
 	- `The Lumbar Spine Data Set` section was shortened, `Figure 2` was removed since we found it did not support the train of thought as we intended

### 2014-12-09 Reviews & Rebuttal.md
combined Rebuttal and Reviews in one markdown file

Reviewer Feedback (formatted into Markdown for better readability)

# Reviews
Paper Title: Interactive Visual Analysis of Lumbar Back Pain – What the Lumbar Spine Tells About Your Life
Paper #: 3
Paper Selection State: Short Paper (20 minutes oral presentation)		
Number of Reviews: 3
Rating Scale: 1: Lowest Value, 6: Highest Value

## Reviewer 1

### Rating

- Abstract and Introduction are adequate?: Yes
- Conclusions/Future Work are convincing?: No
- Figures are adequate ? in number and quality: No
- Improve critical discussion ? validation: Yes
- Improve English?: No
- Needs comparative evaluation?: No
- Needs more experimental results?: No
- Originality Newness of the ideas expressed: 2
- Overall Rating Weighted value of above items: 3
- Paper formatting needs adjustment?: No
- Presentation Structure/Length/English: 4
- References are up-to-date and appropriate?: Yes
- Relevance Paper fits one or more of the topic areas? 2
- Significance Is the problem worth the given attention? 3
- Technical Quality/Theoretical soundness/methodology: 3

### Observations for Author
The paper describes an interactive visual analysis aimed at finding relationships between lumbar spine geometrical properties and lower back pain. Lumbar spine geometry is captured by extracting 9 image related metrics from more than 2,000 magnetic resonances. Such numerical information is explored looking for a possible correlation with the binary lower back pain variable, using a scatterplot matrix (on the 9 metrics) and plotting 2 histograms and two box-plots of the binary lower back pain for each numerical variable. Because no statistically significant correlation could be observed the paper proceeds using the principal component analysis and an heterogeneous correlation of more than 100 patient-specific attributes (e.g., age, gender, etc.), finding weak correlations. The third analysis phase uses decision trees, generating a tree for every patient specific attribute in order to locate the attributes that can be explained by image derived variables, producing up to 402 trees that are filtered according to their size and quality and plotted as points on scatterplots, arranged on four classification of the data. Some results are, eventually, discussed (eg., gender can be discriminated with an error of 0.31 using 7 rules and using only two image based variables). I’m not satisfied with the paper for three mains reasons:

- 1) The exploration strategy is, in my opinion, not clear: there are not clear goals or net steps. The analysis is conducted using a brute force attack (scatterplot matrix, correlation of all 134x134 patient-related attributes, exhaustive decision tree generation). Results are collected without a general strategy: e.g., what is the relevance of high correlation between curvature angle and mean curvature with the main goal of understanding how the 9 numerical variables and the 134 patient-based attributes relates to the lower back pain? What is the role of decision trees explaining how numerical variables explains patient-based attributes in explaining the source of lower back pain? And, again, what is the role of PCA? The general impression I get from the paper contains the application of a set of standard and well known statistical techniques without a clear relationship of the main goal of understanding the lower back pain; 
- 2)Visualizations are inserted in the discussion without making clear how they contribute to the overall process. How images on Figure 3 drive the next step? Moreover, some of them are confusing. What is the purpose of Figure 2 w.r.t. the main process? Moreover, no part of an image is the source for the next step: where is the evidence of :”show the important and analyze further”? Where is the interactive part of the images? Just browsing on dots on scatterplots and getting the trees? 
- 3) Weak or no clue is provided about how images allowed for getting the final conclusions. How do the authors get the insights from the Figure 4? Focusing on the origin zone, with low error rate and little size tree?

On the other hand the amount of work described in the paper is relevant, and likely useful for the medical domain. For this reason, I rate the paper as borderline. Minor comments: Figure 2. The caption is confusing: if the top part refers to male and the bottom part refers to female, what is the meaning of: “The dark gray share of each bar encodes the portion of male (top chart) or female (bottom chart) subjects”? GPLOM Analysis Section. “The combination of the image variables with back pain is visualized as histogram at the left side…” left should be bottom Correlatiob Matrix “an correlation” Text on figure 3 is hard to read.

The prototype is not available at blind.dnsalias.com

## Reviewer 2

### Rating

- Abstract and Introduction are adequate?: Yes
- Conclusions/Future Work are convincing?: Yes
- Figures are adequate ? in number and quality: Yes
- Improve critical discussion ? validation: Yes
- Improve English?: No
- Needs comparative evaluation?: Yes
- Needs more experimental results?: Yes
- Originality Newness of the ideas expressed: 3
- Overall Rating Weighted value of above items: 4
- Paper formatting needs adjustment?: No
- Presentation Structure/Length/English: 4
- References are up-to-date and appropriate?: Yes
- Relevance Paper fits one or more of the topic areas?: 4
- Significance Is the problem worth the given attention? 3
- Technical Quality Theoretical soundness/methodology: 3

### Observations for Author
The paper proposes a new way to visually explore decision tree results. It is used to analyze lumbar spine data and understand relationship between 9 image features and 134 non-image features. The strengths of the paper:
- the paper is well-written and gives detailed introduction into the epidemiology domain
- the visual analytics approach is well-motivated and intuitive The weaknesses of the paper:
- the explanation of the domain is probably too lengthy
- the approach was not evaluated with user tests or at least with some inspection methods Suggestions:å
- the paper would be strengthened if we could see some illustration where the proposed approach is uniquely capable of providing some actionable insights as compared to the existing approaches (such as looking at the decision trees or performing feature selection)

## Reviewer 3

### Rating
- Abstract and Introduction are adequate?: Too long
- Conclusions/Future Work are convincing?: No
- Figures are adequate ? in number and quality: Yes
- Improve critical discussion ? validation: No
- Improve English?: No
- Needs comparative evaluation?: No
- Needs more experimental results?: No
- Originality Newness of the ideas expressed: 3
- Overall Rating Weighted value of above items: 3
- Paper formatting needs adjustment?: Yes
- Presentation Structure/Length/English: 2
- References are up-to-date and appropriate?: Yes
- Relevance Paper fits one or more of the topic areas?: 6
- Significance Is the problem worth the given attention?: 3
- Technical Quality Theoretical soundness/methodology: 3

### Observations for Author
This paper is an extension of the authors' previous VMV paper. The IVA appears quite the same with the addition of a decision tree that the authors claim is data mining. The paper is very long and does not need to be. The authors should be specific about their contributions (the IVA is not new) and condense the paper to about 8 pages maximum. Be specific about how this is an extension of the VMV paper and what is new and different. Compare and contrast with that paper rather than obfuscate the two. The future work has a paragraph ending in '(' I guess completing the section is future work? I would argue for a weak accept of this paper though it needs a rewrite to be publishable.

Rebuttal
========

This rebuttal is written in markdown. A graphical summary of the reviews can be found [here](https://docs.google.com/spreadsheets/d/1DTPt5eGe6PpMiNh38p4YsQrKLI3bqKhylZQKZKiZg4w/edit?usp=sharing).

General Changes
---------------

- substantial reductions had to take place, since we were given a `short paper` slot with an 8-page limit.
	- redundancy between `Introduction` and `Epidemiological Background` was minimized as much as possible
	- `Related Work` section was shortened and covers less papers to meet the page limit.
		- as requested by [Reviewer 3], we elaborate on the differences towards the VMV paper in the `Related Work` section
	- `The Lumbar Spine Data Set` section was shortened, `Figure 2` was removed since we found it did not support the train of thought as we intended
	- `Experiments and Preliminary Results` section was shortened and changed due to the Reviewer feedback. It should be more precise and comprehensible.
	- `Results` section now covers *all subjects*, subjects categorized by *gender* and *body mass index*
		- `Figure 3` (former `Figure 4`) was adapted accordingly
	- `Summary & Conclusion` was completely revised and is now structured properly
	- due to the lack of time until the camera ready deadline as well as the page limit, we are not able to conduct a detailed user test. We did, however, discuss our approach and its drawbacks in more detail in the last section
	- Figure 1 was completely removed since it does not supports the train of thought as we intended and due to lack of space

Reviewer 1
----------

- *"The exploration strategy is, in my opinion, not clear: there are not clear goals or net steps."*
	- role of Decision Trees and why they are chosen is now captured in the introduction of section 6 - Interactive Decision Tree Visualization.
	- at the beginning of Section 5 - Experiments and Preliminary Results, all subjects
	- experiments and Preliminary Results completely restructured to make it comprehensible
- *"What is the role of decision trees explaining how numerical variables explains patient-based attributes in explaining the source of lower back pain?"*
	- the first step is always finding associations with back pain. As we conclude, that we could not extract such associations at the given model precision, the focus shifts towards assessing, which variables can be described through the shape variables.
- Applicability: Comparison to standard workflow is not possible. The classic epidemiological workflow is strictly hypothesis driven and therefore requires a small set of 2 - 3 variables, which are correlated with a target variable. IVA methods amplify *hypothesis generation* and therefore aim to provide insight into unexpected relationships in the data.
- *"Where is the interactive part of the images? Just browsing on dots on scatterplots and getting the trees?"*
	- Browsing the decision tree results is one interactive part. More important, however, is the ability to select target variables, which are used to subdivide the subjects, to assess their influence towards the result. Also, depending on how many subjects are included in a certain sub group (e.g. people suffering from a rare condition), the minimum subject count for a classification and the maximum error can be adjusted.
- *"Weak or no clue is provided about how images allowed for getting the final conclusions. How do the authors get the insights from the Figure 4? Focusing on the origin zone, with low error rate and little size tree?"*
	- We clarified this in the Figure caption.
- response to Minor Comments:
	- *"GPLOM Analysis Section. “The combination of the image variables with back pain is visualized as histogram at the left side…” left should be bottom"*: Back pain is encoded with presence (red) and absence (turquoise) in all plots left of the matrix diagonal. We clarified this in the text.
- *"The prototype is not available at blind.dnsalias.com"*
	- Regarding to our server logs we had a 100% uptime through the time of the reviewing process, we cannot comprehend whats gone wrong there

Reviewer 2
----------

- *"the explanation of the domain is probably too lengthy"*
	- we removed redundancies and shortened the `Epidemiological Background` section
- *"the approach was not evaluated with user tests or at least with some inspection methods"*
	- due to the lack of time until the camera ready deadline as well as the page limit, we are not able to conduct a detailed user test. We did, however, discuss our approach and its drawbacks in more detail in the last section
- *"the paper would be strengthened if we could see some illustration where the proposed approach is uniquely capable of providing some actionable insights as compared to the existing approaches"*
	- the standard workflow of the epidemiologists is deriving hypotheses from the clinical day to day practice. The data are analyzed towards 2-3 independent variables at maximum towards a dependent variable, such as back pain. To our knowledge, there are no existing approaches used by epidemiologists in the day to day practice allowing for an explorative analysis using decision trees, so it is difficult to conduct a comparison as suggested.

Reviewer 3
----------

- *"The authors should be specific about their contributions (the IVA is not new) and condense the paper to about 8 pages maximum."*
	- we elaborate on the differences towards the VMV paper in more detailed in the `Related Work` section.
	- the paper was shortened as requested

### 2014-12-09 draft.tex
removed SHIP subsection in the Related Work section; added information about differences towards the VMV
minor revisions

@@
 %%
-For this purpose, we combine data mining algorithms with data visualization techniques.
+For this purpose, we combine classification algorithms with data visualization techniques.
 %%
@@
 %%
-\item The identification of the most important variables via data mining methods using a novel semi- quantitative overview visualization concept. % embedded in our IVA-framework.
+\item The identification of the most important variables via classification methods using a novel semi- quantitative overview visualization concept. % embedded in our IVA-framework.
 %%
@@
 %%
+In a prior work, we analyzed the lumbar spine variability of 490 \texttt{SHIP-2} subjects \cite{Klemm2013VMV}.
+%%
+We incorporate hierarchical agglomerative clustering to derive shape groups, yielding several groups of average shape and several outlier clusters.
+%%
+The extracted lumbar spine shape was correlated with subject size.
+%%
+%We use the same detection algorithm and lumbar spine canal extraction as foundation of this work.
+%%
+As opposed to clustering the spine shape, we analyze the discriminative power of the resulting data towards back pain and other non-image variables in this work.
+%Clustering techniques are strongly dependent on the chosen variables and distance measures \cite{Klemm2014BVM}.
+%%
+%%
 \\\\
@@
 \\\\
-\noindent \textbf{SHIP Data Analysis.}
-% TODO: [high] This is currently only one paper!
-In a prior work, we analyzed the lumbar spine variability of 490 \texttt{SHIP-2} subjects \cite{Klemm2013VMV}.
-%%
-We incorporate hierarchical agglomerative clustering to derive shape groups, yielding several groups of average shape and several outlier clusters.
-%%
-The extracted lumbar spine shape was correlated with subject size.
-%%
-We use the same detection algorithm and lumbar spine canal extraction as foundation of this work.
-%Our data extraction process is based on the detection algorithm and lumbar spine canal extraction presented by \cite{Klemm2013VMV}.
-%%
-%Clustering techniques are strongly dependent on the chosen variables and distance measures \cite{Klemm2014BVM}.
-%%
-\\\\
-Unique in our work is the combination of data mining techniques with an IVA approach by observing interesting feature relations in the context of image-derived features using multiple decision trees for one visualization.
+Unique in our work is the combination of classification techniques with an IVA approach by observing interesting feature relations in the context of image-derived features using multiple decision trees for one visualization.
 %%
@@
 %%
-% TODO: [high] Unterschied zu VMV Publikation herausarbeiten
 \section{\uppercase{The Lumbar Spine Data Set}}
@@
 
-These observations brought us to the decision to incorporate data mining techniques to assess the influence of the image-derived variables.
+These observations brought us to the decision to incorporate classification techniques to assess the influence of the image-derived variables.
 %%
@@
 %%
-Decision trees are used in data mining to create predictive models.
+Decision trees are used to create predictive models.
 %%
@@
 %%
+Results of high interest are located close to the axis origins.
 }
@@
 Over-fitting to the data indicated by complex decision trees has to be taken into account as well.
+%%
+The presented method only captures linear relationships between variables.
+%%
+To take more complex associations into account, methods such as regression analysis have to be incorporated.
 \\\\
@@
 %%
-Combining the power of statistical analysis, visual analytics and data mining techniques is essential for analyzing increasingly complex heterogenous population data.
+Combining the power of statistical analysis, visual analytics and classification techniques is essential for analyzing increasingly complex heterogenous population data.
 %%

### 2014-12-09 draft.tex
shortened various sections

@@
 %%
-\begin{itemize}
-%%
-	\item Physicians formulate hypotheses based on observations made in their clinical practice.
-%%
-	\item Epidemiologists compile a list of variables depicting a hypothesis to prove it.
-%%
-	\item Statistical methods, such as regression analysis, assess the correlation of selected variables with the investigated disease.
-%%
-\end{itemize}
+(1) Physicians formulate hypotheses based on observations made in their clinical practice; (2) Epidemiologists compile a list of variables depicting a hypothesis to prove it; (3) Statistical methods, such as regression analysis, assess the correlation of selected variables with the investigated disease.
+%%
+% \begin{enumerate}
+% %%
+% 	\item Physicians formulate hypotheses based on observations made in their clinical practice.
+% %%
+% 	\item Epidemiologists compile a list of variables depicting a hypothesis to prove it.
+% %%
+% 	\item Statistical methods, such as regression analysis, assess the correlation of selected variables with the investigated disease.
+% %%
+% \end{enumerate}
 %%
@@
 %%
-Many diseases, such as different cancer types, are more likely with increasing age \cite{Fletcher}.
+Many diseases, such as different cancer types, are for example more likely with increasing age \cite{Fletcher}.
 %%
@@
 Spine shape is influenced by several somatometric variables.
-%% TODO: Introduce this better! Go into detail on the train of thought
+%%
 Larger people have also a longer spine with a straighter shape.
@@
 %%
-To assess their influence, we take them into account when spine \emph{curvature} and \emph{torsion} is correlated with non-image variables.
+To assess their influence, we take them into account when spine \emph{curvature} and \emph{torsion} are correlated with non-image variables.
 %%
@@
 %%
-Discretizing metric variables using quartiles avoids small outlier groups.
+Discretizing metric variables using quartiles avoids outlier groups.
 %%
@@
 %%
-We calculated the contingency matrix for all size groups and searched for correlations between image- and non-image derived variables.
-%%
-The resulting contingency matrices are available as supplementary material and show no strong correlation with any of the variable (see \url{blind.dnsalias.com}).
+We calculated the contingency matrix for all size groups and searched for correlations between image- and non-image variables.
+%%
+The resulting contingency matrices show no strong correlation with image variables (see experiments page at \url{blind.dnsalias.com}).
 %%
@@
 %%
-The web-based approach allows us to quickly exchange results with our collaborating epidemiological experts.
+The web-based approach allows for quick exchange with our collaborating epidemiological experts.
 %%
@@
 The four \emph{body size} groups could be characterized with an error of $0.48$, but the decision tree comprises 71 rules and imposes overfitting.
-
-The analysis of the dummy variables yields a result similar to the \emph{blood pressure medication}.
+%%
+The dummy variable analysis yields a result similar to the \emph{blood pressure medication}.
 %%
@@
 %%
-
 Some variables, such as \emph{body size}, can be described with an error of $0.3$ to $0.36$, but only using large decision trees with over 20 rules.
@@
 \noindent We provide a method for comparative analysis of decision trees independent of the variable manifestation count using interactive scatter plots.
-%% TODO: [high] This is no related work
+%%
 We applied the technique to gain insight into the predictive power of 9 image-derived variables for 134 non-image variables.

### 2014-12-10 draft.tex
updated supplementary material URL

@@
 %%
-%The analysis, as well as the presented IVA tool is available as supplementary material in an interactive \texttt{R Markdown}\footnote{Developed by RStudio, Inc; \texttt{rmarkdown.rstudio.com}} document (\url{blind.dnsalias.com}).
+%The analysis, as well as the presented IVA tool is available as supplementary material in an interactive \texttt{R Markdown}\footnote{Developed by RStudio, Inc; \texttt{rmarkdown.rstudio.com}} document (\url{ivapp15.dnsalias.com}).
 %%
@@
 % %%
-% The complete analysis can be found in the experiments section of the supplementary material (\url{blind.dnsalias.com})
+% The complete analysis can be found in the experiments section of the supplementary material (\url{ivapp15.dnsalias.com})
 %% ------------------------------------------
@@
 %%
-The resulting contingency matrices show no strong correlation with image variables (see experiments page at \url{blind.dnsalias.com}).
+The resulting contingency matrices show no strong correlation with image variables (see experiments page at \url{ivapp15.dnsalias.com}).
 %%
@@
 \noindent \textbf{Implementation.}
-All analyses are available at \url{blind.dnsalias.com} and are carried out using \texttt{R}, a widely used programming language for statistical calculations and visualizations.
+All analyses are available at \url{ivapp15.dnsalias.com} and are carried out using \texttt{R}, a widely used programming language for statistical calculations and visualizations.
 %%
@@
 % %%
-% The prototype is available at \url{blind.dnsalias.com}.
+% The prototype is available at \url{ivapp15.dnsalias.com}.
 %%---------
@@
 %\section*{\uppercase{Appendix}}
-%\noindent The IVA tool and all shown plots presented in this paper are available as supplementary material under the following link: \url{blind.dnsalias.com}.
+%\noindent The IVA tool and all shown plots presented in this paper are available as supplementary material under the following link: \url{ivapp15.dnsalias.com}.
 

### 2014-12-10 draft.tex
first draft with 8 pages

@@
 %%
-In a prior work, we analyzed the lumbar spine variability of 490 \texttt{SHIP-2} subjects \cite{Klemm2013VMV}.
-%%
-We incorporate hierarchical agglomerative clustering to derive shape groups, yielding several groups of average shape and several outlier clusters.
-%%
-The extracted lumbar spine shape was correlated with subject size.
+In a prior work, we analyzed the lumbar spine variability of 490 \texttt{SHIP} subjects \cite{Klemm2013VMV}.
+%%
+We incorporate hierarchical agglomerative clustering to derive shape groups, yielding average shape groups and outliers.
+%%
+%The extracted lumbar spine shape was correlated with subject size.
 %%
@@
 	%%
-	\item 113 categorical variables divided into
-	%%
-	\begin{itemize}
-		%%
-		\item 43 dichotomous (binary) variables, mostly indicating the presence of a disease, e.g. \emph{pancreatitis} or \emph{high blood pressure}
-		%%
-		\item 70 variables with more than two levels, indicating pain levels, nutrition and social factors, such as marital or retirement status
-		%%
+	\item 113 categorical variables divided into 43 dichotomous (binary) variables, mostly indicating the presence of a disease, e.g.
+\emph{pancreatitis} or \emph{high blood pressure} and 70 variables with more than two levels, indicating pain levels, nutrition and social factors, such as marital or retirement status
+	% %%
+% 	\begin{itemize}
+% 		%%
+% 		\item 43 dichotomous (binary) variables, mostly indicating the presence of a disease, e.g. \emph{pancreatitis} or \emph{high blood pressure}
+% 		%%
+% 		\item 70 variables with more than two levels, indicating pain levels, nutrition and social factors, such as marital or retirement status
+% 		%%
+% 	\end{itemize}
 \end{itemize}
-\end{itemize}
-%%
-All categorical variables are converted into binary \emph{dummy variables}, indicate the presence or absence of a categorical variable manifestation.
+%%
+All categorical variables are converted into binary \emph{dummy variables}, indicating the presence or absence of a categorical variable manifestation.
 %with multiple manifestations into several dichotomous variables.
@@
 %%
-For example, a pain indicator variable ranging from \emph{1 - no pain} to \emph{4 - large pain} is subdivided into four dichotomous variables.
+For example, a pain indicator variable ranging from \emph{1 - no pain} to \emph{4 - large pain} is subdivided into four dichotomous variables, allowing to determine which manifestation can be described best using the image-based variables
 %%
@@
 %%
-This is useful for our classification, because it allows to determine which manifestation of a variable can be described best using the image data variables.
+%This allows to determine which manifestation of a variable can be described best using the image data variables.
+%This is useful for our classification, because it allows to determine which manifestation of a variable can be described best using the image data variables.
 %%
@@
 The projections to the transversal planes attract attention as they have many outliers.
-%% % TODO: [high] Add Link
-We conclude that \emph{curvature} is not as reliable on the transversal plane as it is on the other planes, which was also confirmed through a Principal Component Analysis (see supplementary material at TODO:LINK).
-%%
-The \texttt{GPLOM} shows very similar distributions of subjects with or without back pain with respect to the shape variables in all sub plots.
+%%
+We conclude that \emph{curvature} is not as reliable on the transversal plane as it is on the other planes, which was also confirmed through a Principal Component Analysis (see supplementary material at \url{ivapp15.dnsalias.com}).
+%%
+The \texttt{GPLOM} shows similar distributions of subjects with or without back pain with respect to the shape variables in all sub plots.
 %%
@@
 %%
-The described visualization provides a good overview of the classification results.
-%%
-We still want to be able to display \emph{details-on-demand} \cite{shneiderman1996} and examine a decision tree in detail.
-%%
-This is realized by clicking on an entry in the visualization, which then interactively displays the corresponding decision tree in detail.
+The visualization provides a good overview of the classification results.
+%%
+%We still want to be able to display \emph{details-on-demand} and examine a decision tree in detail.
+%%
+\emph{Details-on-demand} are displayed by clicking on an entry in the visualization, which then displays the corresponding decision tree in detail.
 %%
 This allows to sequentially analyze the classifications.
-
+%%
 We also provide controls for adjusting the maximum classification error and minimum subject count for a variable.
@@
 %%
-The result however, is probably overfitted to the data due to tree sizes between $14$ and $16$.
-%%
-Some variables, such as \emph{body size}, can be described with an error of $0.3$ to $0.36$, but only using large decision trees with over 20 rules.
+The result is overfitted to the data due to tree sizes between $14$ and $16$.
+%%
+Some variables, such as \emph{body size}, can be described with an error of $0.3$ to $0.36$ using large decision trees with over 20 rules.
 %%
@@
 %%
-The dichotomous variable whether the subject has felt \emph{pain in the last seven days} can also be predicted for this group using the same features.
-%%
-The resulting tree consists of $8$ rules and has an error of $0.35$.
+Subjects experiencing \emph{pain in the last seven days} can also be predicted for this group using the same features with a tree consisting of $8$ rules and has an error of $0.35$
+%The dichotomous variable whether the subject has felt \emph{pain in the last seven days} can also be predicted for this group using the same features.
+%%
+%The resulting tree consists of $8$ rules and has an error of $0.35$.
 %%

### 2014-12-10 draft.tex
incorporated Sylvias revision

@@
 %%
-We analyze mutual dependencies among shape-describing variables and non-image variables, e.g. pain indicators.
+We analyze mutual dependencies among shape-describing variables and non-image variables, e.g., pain indicators.
 %%
@@
 %%
-The exact causes as well as particularly vulnerable risk groups are unknown.
+The exact causes as well as particular vulnerable risk groups are unknown.
 %%
@@
 %%
-Generalized Pairs Plots (\texttt{GPLOM'S}) visualize heterogenous variables in a plot matrix \cite{GPLOMS}. %\cite{GPLOMS,Francois}.
+Generalized Pairs Plots (\texttt{GPLOMs}) visualize heterogenous variables in a plot matrix \cite{GPLOMS}. %\cite{GPLOMS,Francois}.
 %%
@@
 %%
-\texttt{GPLOM'S} provide an overview, but take up much screen space and are therefore only suitable for few variables at once (see example in Fig.~\ref{fig:image-parameter-range} later on).
+\texttt{GPLOMs} provide an overview, but take up much screen space and are therefore only suitable for few variables at once (see example in Fig.~\ref{fig:image-parameter-range} later on).
 %%
@@
 %%
-They present an interactive data mining tool, which can analyze association rules and highlight relations. %, which may trigger new hypotheses about the data.
+They present an interactive data mining tool, that can analyze association rules and highlight relations. %, which may trigger new hypotheses about the data.
 %%
@@
 %%
-The non-image variables range from somatometric variables describing body measures to medical examinations, such as laboratory tests as well as lifestyle factors, e.g. sporting activity or nutrition.
+The non-image variables range from somatometric variables describing body measures to medical examinations, such as laboratory tests as well as lifestyle factors, e.g., sporting activity or nutrition.
 %%
@@
 	%%
-	\item 113 categorical variables divided into 43 dichotomous (binary) variables, mostly indicating the presence of a disease, e.g.
-\emph{pancreatitis} or \emph{high blood pressure} and 70 variables with more than two levels, indicating pain levels, nutrition and social factors, such as marital or retirement status
+	\item 113 categorical variables divided into 43 dichotomous (binary) variables, mostly indicating the presence of a disease, e.g., \emph{pancreatitis} or \emph{high blood pressure} and 70 variables with more than two levels, indicating pain levels, nutrition and social factors, such as marital or retirement status
 	% %%
@@
 \begin{itemize}
-	\item \emph{Mean Curvature} is calculated as weighted sum of curvature between all adjacent points describing the centerline: $\sum_{i=1}^I \frac{curvature_i}{I}$.
-We refer to total curvature as \emph{curvature}.
-%%
-	\item \emph{Mean Torsion} (deviation of a curve from its current course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum_{i=1}^I \frac{torsion_i}{I}$.
-We refer to total torsion as \emph{torsion}.
-%%
-	\item \emph{Curvature angle $\alpha$} is defined by the middle point of the spine canal centerline as \emph{vertex} and the line between middle point and top/bottom point as \emph{sides}.
+	\item \emph{Mean Curvature} is calculated as weighted sum of curvature between all adjacent points describing the centerline: $\sum_{i=1}^I \frac{\textit{curvature}_i}{I}$.
+We refer to the total curvature as \emph{curvature}.
+%%
+	\item \emph{Mean Torsion} (deviation of a curve from its current course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum_{i=1}^I \frac{\textit{torsion}_i}{I}$.
+We refer to the total torsion as \emph{torsion}.
+%%
+	\item \emph{Curvature angle $\alpha$} is defined by the middle point of the spine canal centerline as \emph{vertex} and the line connecting middle point and top/bottom point as \emph{sides}.
 %%
@@
 %%
-These models are build w.r.t all input variables and capture more complex relationships than correlation coefficients.
+These models are build w.r.t. all input variables and capture more complex relationships than correlation coefficients.
 %%
@@
 %%
-Further subdivision, e.g. by quantiles of \emph{body mass index} increases the number to 402 trees.
+Further subdivision, e.g., by quantiles of \emph{body mass index} increases the number to 402 trees.
 %%
@@
 %%
-We follow the Visual Analytics mantra of analyzing first, show the important and analyze further \cite{Keim}.
+We follow the Visual Analytics mantra of \emph{analyzing first}, \emph{show the important} and \emph{analyze further} \cite{Keim}.
 %%
@@
 \begin{equation}
-totalError = 1 - \sum_{m=1}^M \frac{correctlyClassified_{m}}{M\cdot N_m}
+\textit{totalError} = 1 - \sum_{m=1}^M \frac{\textit{correctlyClassified}_{m}}{M\cdot N_m}
 \end{equation}
 %%
-$M$ represents the set of manifestation of each variable.
+$M$ represents the set of manifestations of each variable.
 %%
@@
 %%
-Classification and comparison of variables for subject groups (e.g. male and female subjects) in one plot can be achieved by color coding group affiliation on the data points.
+Classification and comparison of variables for subject groups (e.g., male and female subjects) in one plot can be achieved by color coding group affiliation on the data points.
 %%
@@
 %%
-Furthermore, we map the number of subjects associated with a variable to point diameter in the scatter plot.
+Furthermore, we map the number of subjects associated with a variable to point the diameter in the scatter plot.
 %%
@@
 %%
-The shown mutual dependencies aim to amplify \emph{hypothesis generation}.
+The shown mutual dependencies aim to amplify \emph{hypotheses generation}.
 %%
@@
 %%
-The majority of non-medicated subjects are false-positive classified ($262$ of $1,058$), yielding a poor quality of the classifier w.r.t epidemiological research.
+The majority of non-medicated subjects are false-positive classified ($262$ of $1,058$), yielding a poor quality of the classifier w.r.t. epidemiological research.
 %%
@@
 %%
-The dummy variable analysis showed that female subjects of \emph{139-164~cm body size} could be discriminated using the \emph{mean curvature} and \emph{curvature angle}, with an error of $0.38$.
+The dummy variable analysis showed that female subjects of 139-164~cm \emph{body size} could be discriminated using the \emph{mean curvature} and \emph{curvature angle}, with an error of $0.38$.
 %%
@@
 %%
-Our method allowed to assess their discriminative power, which is largely limited for separating male and female subjects, \emph{nutrition} variables, as well as different disease indicators.
+Our method allowed to assess their discriminative power, which is largely limited to separating male and female subjects, \emph{nutrition} variables, as well as different disease indicators.
 %%

### 2014-12-10 Reviews & Rebuttal.md
revised first three sections, including an error in Figure 1

@@
 
-Rebuttal
-========
+Rebuttal & Coverletter
+======================
 
-This rebuttal is written in markdown. A graphical summary of the reviews can be found [here](https://docs.google.com/spreadsheets/d/1DTPt5eGe6PpMiNh38p4YsQrKLI3bqKhylZQKZKiZg4w/edit?usp=sharing).
+This rebuttal/coverletter is written in markdown.
+A graphical summary of the reviews can be found [here](https://docs.google.com/spreadsheets/d/1DTPt5eGe6PpMiNh38p4YsQrKLI3bqKhylZQKZKiZg4w/edit?usp=sharing).
 
@@
 - *"The exploration strategy is, in my opinion, not clear: there are not clear goals or net steps."*
-	- role of Decision Trees and why they are chosen is now captured in the introduction of section 6 - Interactive Decision Tree Visualization.
-	- at the beginning of Section 5 - Experiments and Preliminary Results, all subjects
+	- role of Decision Trees and why they are chosen is now captured in the introduction of Section 6 - Interactive Decision Tree Visualization.
 	- experiments and Preliminary Results completely restructured to make it comprehensible
@@
 	- the first step is always finding associations with back pain.
-As we conclude, that we could not extract such associations at the given model precision, the focus shifts towards assessing, which variables can be described through the shape variables.
-- Applicability: Comparison to standard workflow is not possible.
-The classic epidemiological workflow is strictly hypothesis driven and therefore requires a small set of 2 - 3 variables, which are correlated with a target variable.
-IVA methods amplify *hypothesis generation* and therefore aim to provide insight into unexpected relationships in the data.
+As we conclude, that no such associations could be extracted at the given model precision, the focus shifts towards assessing, which variables can be described through the shape variables.
 - *"Where is the interactive part of the images? Just browsing on dots on scatterplots and getting the trees?"*
@@
 To our knowledge, there are no existing approaches used by epidemiologists in the day to day practice allowing for an explorative analysis using decision trees, so it is difficult to conduct a comparison as suggested.
+IVA methods amplify *hypotheses generation* and therefore aim to provide insight into unexpected relationships in the data.
 

### 2014-12-10 draft.tex
revised first three sections, including an error in Figure 1

@@
 %%
-\item The detection of associations between image-based, socio-demographic and medical variables for hypothesis generation,
+\item The detection of associations between image-based, socio-demographic and medical variables for \emph{hypotheses generation},
 %%
@@
 %%
-(1) Physicians formulate hypotheses based on observations made in their clinical practice; (2) Epidemiologists compile a list of variables depicting a hypothesis to prove it; (3) Statistical methods, such as regression analysis, assess the correlation of selected variables with the investigated disease.
+(1) physicians formulate hypotheses based on observations made in their clinical practice; (2) epidemiologists compile a list of variables depicting a hypothesis to prove it; (3) statistical methods, such as regression analysis, assess the correlation of selected variables with the investigated disease.
 %%
@@
 %%
-\emph{Hypothesis generation} is amplified by concurrently analyzing heterogenous variables at once using correlation measures.
+\emph{Hypotheses generation} is amplified by concurrently analyzing heterogenous variables at once using correlation measures.
 %%
@@
 %%
-This approach aims to trigger \emph{hypothesis generation} by outlining tendencies between these variables.
+This approach aims to trigger \emph{hypotheses generation} by outlining tendencies between these variables.
 %%
@@
 %%
-In a prior work, we analyzed the lumbar spine variability of 490 \texttt{SHIP} subjects \cite{Klemm2013VMV}.
+In a prior work, we analyzed the lumbar spine variability of 490 subjects \cite{Klemm2013VMV}.
 %%
@@
 %%
-To classify breast tumors, they train a decision-tree classifier, concluding that the extracted kinetic and morphological features alone are not sufficient for tumor type classification.
-%%
-\cite{Niemann2014} assess risk factors for hepatic steatosis (fatty liver disease) risk factors using decision trees.
-%%
-They present an interactive data mining tool, that can analyze association rules and highlight relations. %, which may trigger new hypotheses about the data.
+To classify breast tumors, they train a decision-tree classifier, concluding that the extracted kinetic and morphological variables alone are not sufficient for tumor type classification.
+%%
+\cite{Niemann2014} assess risk factors for hepatic steatosis (fatty liver disease) using decision trees.
+%%
+Their interactive data mining tool can analyze association rules and highlight relations.
+%They present an interactive data mining tool, that can analyze association rules and highlight relations. %, which may trigger new hypotheses about the data.
 %%
@@
 \\\\
-Unique in our work is the combination of classification techniques with an IVA approach by observing interesting feature relations in the context of image-derived features using multiple decision trees for one visualization.
+Unique in our work is the combination of classification techniques with an IVA approach by observing interesting variable relations in the context of image-derived variables using multiple decision trees for one visualization.
 %%
@@
 %%
-Leafs of a decision tree represent class labels, branches represent feature conjunctions leading to the class labels.
+Leafs of a decision tree represent class labels, branches represent variable conjunctions leading to the class labels.
 %%
@@
 %%
-Subjects experiencing \emph{pain in the last seven days} can also be predicted for this group using the same features with a tree consisting of $8$ rules and has an error of $0.35$
+Subjects experiencing \emph{pain in the last seven days} can also be predicted for this group using the same variables with a tree consisting of $8$ rules and has an error of $0.35$
 %The dichotomous variable whether the subject has felt \emph{pain in the last seven days} can also be predicted for this group using the same features.
@@
 %%
-Methods supporting \emph{hypothesis generation} based on image information are new to the application domain.
+Methods supporting \emph{hypotheses generation} based on image information are new to the application domain.
 %%

### 2014-12-10 draft.tex
revised rest of the sections

@@
 %%
-We rely on predictive classification trained to obtain a complex rule set on how combinations of the image-variables explain non-image variables.
+We rely on predictive classification to obtain a complex rule set on how combinations of the image-variables explain non-image variables.
 %%
@@
 %%
-Further subdivision, e.g., by quantiles of \emph{body mass index} increases the number to 402 trees.
+Further subdivision, e.g., by quantiles of \emph{body mass index}, increases the number to 402 trees.
 %%
@@
 %%
-This is included in our IVA framework.
+%This is included in our IVA framework.
 %%------------
@@
 %%
-Interactive scatter plot visualization of all C5.0 classification results.
+Interactive scatter plot visualization of all classification results.
 %%
@@
 %%
-The groups are: all subjects; \emph{males} and \emph{females}; and a subdivision by \emph{Body Mass Index quantiles} ($BMI = \frac{m}{l^2}$ where $m$ is the \emph{body mass} in kilogram and $l$ is the \emph{body size} in meter), yielding the groups (17, 24.7] (24.7, 27.4] (27.4, 30.5] (30.5, 48].
+The groups are: \emph{all subjects}; \emph{males} and \emph{females}; and a subdivision by \emph{Body Mass Index quantiles} ($BMI = \frac{m}{l^2}$ where $m$ is the \emph{body mass} in kilogram and $l$ is the \emph{body size} in meter), yielding the groups (17, 24.7] (24.7, 27.4] (27.4, 30.5] (30.5, 48].
 % We carry out the analysis using different subject groups:
@@
 %%
-Subjects experiencing \emph{pain in the last seven days} can also be predicted for this group using the same variables with a tree consisting of $8$ rules and has an error of $0.35$
+Subjects experiencing \emph{pain in the last seven days} can also be predicted for this group using the same variables with a tree consisting of $8$ rules and an error of $0.35$
 %The dichotomous variable whether the subject has felt \emph{pain in the last seven days} can also be predicted for this group using the same features.
@@
 %%
-We applied the technique to gain insight into the predictive power of 9 image-derived variables for 134 non-image variables.
-%%
-The analysis was carried out for subjects divided by \emph{gender}, \emph{BMI}, and \emph{body size} to assess their influence on the lumbar spine shape.
+We applied the method to gain insight into the predictive power of 9 image-derived variables for 134 non-image variables with focus on \emph{back pain}.
+%%
+The analysis was performed for subjects groups of \emph{gender}, \emph{BMI}, and \emph{body size} to assess their influence on the lumbar spine shape.
 

### 2014-12-12 draft.tex
added correct classification metric description

@@
 %%
-The error is scaled to denote perfect classification with 0; 1 is equal to random selection.
-%%
-We consider a result rated $0.5$ or below good enough to display it in the visualization, a value below $0.25$ represents a good classification.
+%The error is scaled to denote perfect classification with 0; 1 equals always wrong; 0.5 equals random selection.
+The error represents the share of incorrect classifications and denotes perfect classification with 0; always wrong with 1; and random selection with 0.5.
+%%
+We display results below $0.5$ in the visualization, a value below $0.25$ represents a good classification.
+%We consider a result rated  or below good enough to display it in the visualization, a value below $0.25$ represents a good classification.
 %%

### 2014-12-12 draft.tex
included Bernhards revision

@@
 \affiliation{\sup{3} Institute for Diagnostic Radiology and Neuro-Radiology, University of Greifswald, Germany}
-\email{\{klemm, glasser\}@isg.cs.uni-magdeburg.de}
+\email{klemm@isg.cs.uni-magdeburg.de}
 }
@@
 %%
-The segmentation results are evaluated and quantified by employing shape-describing variables, such as the spine canal course, its curvature and torsion.
+The segmentation results are evaluated and quantified by employing shape-describing variables, such as spine canal curvature and torsion.
 %%
@@
 %%
-As a first result, we conclude that image-based variables are only sufficient to predict lifestyle factors within the data.
+As a first result, we conclude that image-based variables are only sufficient to describe lifestyle factors within the data.
 %%
@@
 %%
-Hence, relations between non-image variables were successfully detected and characterized.
+Hence, relations between non-image variables were successfully detected and described.
 }
@@
 %%
-Therefore, statistical analysis has to take missing data into account.
+Statistical analysis has to take missing data into account.
 %%
@@
 %%
-For example, continuous data, such as age, is often discretized into quantile bins for visualizing age-dependent disease frequencies.
+For example, continuous data, such as body height, is often discretized into quantile bins for visualizing age-dependent disease frequencies.
 
@@
 %%
-Epidemiologists want to characterize the relation between aging and degenerative process of the spine \cite{Szpalski2005}.
+Epidemiologists want to describe the relation between aging and degenerative process of the spine \cite{Szpalski2005}.
 %%
@@
 %%
-\noindent \textbf{Visual Analysis of Medical Image and Non-Image data}
+\noindent \textbf{Visual Analysis of Medical Image and Non-Image Data.}
 %%
@@
 %%
-\texttt{GPLOMs} provide an overview, but take up much screen space and are therefore only suitable for few variables at once (see example in Fig.~\ref{fig:image-parameter-range} later on).
+\texttt{GPLOMs} provide an overview, but take up much screen space and are therefore only suitable for a few variables at once (see example in Fig.~\ref{fig:image-parameter-range}).
 %%
@@
 %Therefore, epidemiologists compiled the data set with a wide range of variables possibly correlating with lumbar back pain.
-\noindent Epidemiologists compiled a data set with a wide range of variables possibly correlating with lumbar back pain, comprising non-image and image-derived data for 6,753 subjects
+\noindent Epidemiologists compiled a data set with a wide range of variables possibly correlating with lumbar back pain, comprising non-image and image-derived data for 6,753 subjects.
 %%
@@
 \begin{itemize}
-	\item 21 metric variables, describing somatometric variables and markers retrieved from blood analysis
-	%%
-	\item 113 categorical variables divided into 43 dichotomous (binary) variables, mostly indicating the presence of a disease, e.g., \emph{pancreatitis} or \emph{high blood pressure} and 70 variables with more than two levels, indicating pain levels, nutrition and social factors, such as marital or retirement status
+	\item 21 metric variables, describing somatometric variables and markers retrieved from blood analysis.
+	%%
+	\item 113 categorical variables divided into 43 dichotomous (binary) variables, mostly indicating the presence of a disease, e.g., \emph{pancreatitis} or \emph{high blood pressure} and 70 variables with more than two levels, indicating pain levels, nutrition and social factors, such as marital or retirement status.
 	% %%
@@
 %%
-For example, a pain indicator variable ranging from \emph{1 - no pain} to \emph{4 - large pain} is subdivided into four dichotomous variables, allowing to determine which manifestation can be described best using the image-based variables
+For example, a pain indicator variable ranging from \emph{1 - no pain} to \emph{4 - large pain} is subdivided into four dichotomous variables, allowing to determine which manifestation can be described best using the image-based variables.
 %%
@@
 \begin{itemize}
-	\item \emph{Mean Curvature} is calculated as weighted sum of curvature between all adjacent points describing the centerline: $\sum_{i=1}^I \frac{\textit{curvature}_i}{I}$.
-We refer to the total curvature as \emph{curvature}.
-%%
-	\item \emph{Mean Torsion} (deviation of a curve from its current course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum_{i=1}^I \frac{\textit{torsion}_i}{I}$.
-We refer to the total torsion as \emph{torsion}.
-%%
-	\item \emph{Curvature angle $\alpha$} is defined by the middle point of the spine canal centerline as \emph{vertex} and the line connecting middle point and top/bottom point as \emph{sides}.
+	\item \emph{Mean Curvature} is defined as the average curvature of all points describing the centerline: $\sum_{i=1}^I \frac{\textit{curvature}_i}{I}$.
+We refer to the mean curvature as \emph{curvature}.
+%%
+	\item \emph{Mean Torsion} (deviation of a curve from its current course) is defined as the average torsion of all points describing the centerline: $\sum_{i=1}^I \frac{\textit{torsion}_i}{I}$.
+We refer to the mean torsion as \emph{torsion}.
+%%
+	\item \emph{Curvature angle $\alpha$} is the angle defined by the middle point of the spine canal centerline as \emph{vertex} and the line connecting middle point and top/bottom point as \emph{sides}.
 %%
@@
 \noindent \textbf{GPLOM Analysis.}
-As first experiment we correlated the shape variable with the dichotomous back pain indicator using a \texttt{GPLOM}.
-%%
-The metric image-derived variables are pairwise visualized using scatter plots.
+As first experiment we correlated the shape variable with the dichotomous back pain indicator using a \texttt{GPLOM} (Fig.~\ref{fig:image-parameter-range}).
+%%
+The metric image-derived variables are pairwise visualized using scatter plots on the left side of the matrix diagonal.
 %Since our image-derived variables are metric, their pairwise combinations are visualized using scatter plots.
 %%
-The combination of the image variables with back pain is visualized as histogram at the left side of the matrix diagonal and as box plot on the right side.
-%%
-Figure~\ref{fig:image-parameter-range} (right) shows the range of each variable as box plot.
+%The combination of the image variables with back pain is visualized as histogram at the left side of the matrix diagonal and as box plot on the right side.
+The combination of the image variables with back pain is visualized as histogram in the last row and as box plot in the last column.
 %%
@@
 %%
-The method uses \emph{pearson product-moment} for two continuous variables, \emph{polyserial} correlation for one continuous and one categorical variable, and \emph{Polychoric} correlations for two categorical variables.
+The method uses \emph{Pearson product-moment} for two continuous variables, \emph{polyserial} correlation for one continuous and one categorical variable, and \emph{polychoric} correlations for two categorical variables.
 %The method uses the following correlation metrics for the different type combinations:
@@
 %%
-Some variables are too sparse for calculating correlations, for example \emph{treatment of diabetes}, or \emph{medication against high blood pressure} and are omitted, since they are not statistically resilient.
+Some variables are too sparse for calculating correlations, for example \emph{treatment of diabetes}, or \emph{medication against high blood pressure} are omitted, since they are not statistically resilient.
 %%
@@
 %%
-The optimal classification uses a few rules to precisely characterize the target variable.
+The optimal classification uses a few rules to precisely describe the target variable.
 %%
@@
 %%
-For example, if a variable indicating a disease is negative for 90\% of the subjects and the classifier simply assigns all subjects to \emph{not ill}, we get a mean error of 10\%, even though it is very bad.
+For example, if a variable indicating a disease is negative for 90\% of the subjects and the classifier simply assigns all subjects to \emph{not ill}, we get a mean error of 10\%, even though it is very imprecise.
 %%
@@
 %%
+$N_m$ denotes the number of subjects in manifestation $m$.
+%%
 %The error is scaled to denote perfect classification with 0; 1 equals always wrong; 0.5 equals random selection.
@@
 %%
-Many interdependent variables are sparse, such as \emph{medication of diabetes} or \emph{reason of early retirement}.
+Many variables are sparse, such as \emph{medication of diabetes} or \emph{reason of early retirement}.
 %%
@@
 %%
-Therefore, we provide a way to adjust the minimal number of subjects for each variable using a slider input.
+Therefore, we provide a way to adjust the minimal number of subjects for each variable using a slider.
 %%
@@
 %%
-Furthermore, we map the number of subjects associated with a variable to point the diameter in the scatter plot.
+Furthermore, we map the number of subjects associated with a variable to the diameter in the scatter plot.
 %%
@@
 
-None of the pain indicators can be described reliably using the image-based variables.
-%%
-The only variable reliably classified in this group is \emph{gender}, which can be predicted with an error of $0.31$ using 7 rules and incorporates only \emph{curvature-} and \emph{curvature angle}-related variables.
+None of the pain indicators can be reliably described using the image-based variables.
+%%
+The only variable reliably classified in this group is \emph{gender}, which can be described with an error of $0.31$ using 7 rules and incorporates only \emph{curvature-} and \emph{curvature angle}-related variables.
 %
@@
 %%
-The four \emph{body size} groups could be characterized with an error of $0.48$, but the decision tree comprises 71 rules and imposes overfitting.
+The four \emph{body size} groups could be described with an error of $0.48$, but the decision tree comprises 71 rules and imposes overfitting.
 %%
@@
 %%
-Variables, such as subjects sized 139-164 cm, between 64 and 90 years of \emph{age} or \emph{nutrition}-related variables are dominantly populated by one manifestation.
+Variables, such as subject size 139-164 cm, between 64 and 90 years of \emph{age} or \emph{nutrition}-related variables are dominantly populated by one manifestation.
 %%
@@
 %%
-Classification using groups divided by \emph{gender} do not produce satisfying results.
+Classifications using groups divided by \emph{gender} do not produce satisfying results.
 %%
@@
 %%
-The starting age of smoking could be characterized well with an error between $0.25$ and $0.32$ for all \emph{BMI} groups except for subjects with a \emph{BMI} of \emph{$30.5-48$}.
+The starting age of smoking could be described well with an error between $0.25$ and $0.32$ for all \emph{BMI} groups except for subjects with a \emph{BMI} of \emph{$30.5-48$}.
 %%
@@
 %%
-Using mostly \emph{mean curvature} and \emph{curvature angle}, the \emph{leg pain level} can be predicted using $14$ rules with an error of $0.46$ for obese subjects (\emph{BMI} higher than $30$).
-%%
-Subjects experiencing \emph{pain in the last seven days} can also be predicted for this group using the same variables with a tree consisting of $8$ rules and an error of $0.35$
+Using mostly \emph{mean curvature} and \emph{curvature angle}, the \emph{leg pain level} can be described using $14$ rules with an error of $0.46$ for obese subjects (\emph{BMI} higher than $30$).
+%%
+Subjects experiencing \emph{pain in the last seven days} can also be described for this group using the same variables with a tree consisting of $8$ rules and an error of $0.35$
 %The dichotomous variable whether the subject has felt \emph{pain in the last seven days} can also be predicted for this group using the same features.
@@
 %%
-The analysis was performed for subjects groups of \emph{gender}, \emph{BMI}, and \emph{body size} to assess their influence on the lumbar spine shape.
-
-The methods presented herein may be applied to comprehensive epidemiological data sets to learn more about mutual dependencies among variables and to generate hypotheses on potential associations and subgroups.
+The analysis was performed for subjects groups of \emph{gender}, \emph{BMI} and \emph{body size} to assess their influence on the lumbar spine shape.
+
+The methods presented herein may be applied to comprehensive epidemiological data sets to investigate mutual dependencies among variables and to generate hypotheses on potential associations and subgroups.
 %
@@
 \noindent \textbf{Predictive Power of Image-Derived Variables.}
-The presented results indicate that \emph{torsion}, \emph{curvature} and \emph{curvature angle} of the lumbar spine at the presented precision are not sufficient to characterize lumbar back pain in the SHIP data set.
+The presented results indicate that \emph{torsion}, \emph{curvature} and \emph{curvature angle} of the lumbar spine at the presented precision are not sufficient to describe lumbar back pain in the SHIP data set.
 %%

### 2014-12-15 draft.tex
first set of Sylvias revision

@@
 % \author{Blind}
-\author{\authorname{Paul Klemm\sup{1}, Sylvia Glaßer\sup{1}, Kai Lawonn\sup{1}, Marko Rak\sup{1} Henry Völzke\sup{2}, Katrin Hegenscheid\sup{3} and Bernhard Preim\sup{1}}
+\author{\authorname{Paul Klemm\sup{1}, Sylvia Glaßer\sup{1}, Kai Lawonn\sup{1}, Marko Rak\sup{1}, Henry Völzke\sup{2}, Katrin Hegenscheid\sup{3} and Bernhard Preim\sup{1}}
 \affiliation{\sup{1} Department of Simulation and Graphics, University of Magdeburg, Germany}
@@
 %%
-In a previous work, we proposed an epidemiological workflow using IVA, consisting of an iterative sequence of \emph{group selection}, \emph{variable selection} and \emph{visualization} \cite{Klemm2014VIS}.
+In a previous work, we proposed an epidemiological IVA workflow, consisting of an iterative sequence of \emph{group selection}, \emph{variable selection} and \emph{visualization} \cite{Klemm2014VIS}.
 %%
@@
 %%
-\emph{Hypotheses generation} is amplified by concurrently analyzing heterogenous variables at once using correlation measures.
+\emph{Hypotheses generation} amplified by concurrently analyzing heterogenous variables at once using correlation measures.
 %%
@@
 %%
-Their interactive data mining tool can analyze association rules and highlight relations.
+Their interactive data mining tool can analyze association rules and highlights relations.
 %They present an interactive data mining tool, that can analyze association rules and highlight relations. %, which may trigger new hypotheses about the data.
@@
 %%
-For example, a pain indicator variable ranging from \emph{1 - no pain} to \emph{4 - large pain} is subdivided into four dichotomous variables, allowing to determine which manifestation can be described best using the image-based variables.
+For example, a pain indicator variable ranging from \emph{1 - no pain} to \emph{4 - large pain} is subdivided into four dichotomous variables to determine, which manifestation can be described best using the image-based variables.
 %%

### 2014-12-15 draft.tex
revised version from Sylvia

@@
 %%
-Therefore, we automatically extract a decision tree classifier for each non-image variable.
+Therefore, we automatically train a decision tree classifier for each non-image variable.
 %%
@@
 %%
-Many diseases, such as different cancer types, are for example more likely with increasing age \cite{Fletcher}.
+For example, the incidence rate of many diseases, such as different cancer types increases with age \cite{Fletcher}.
+%Many diseases, such as different cancer types, are for example more likely with increasing age \cite{Fletcher}.
 %%
@@
 %%
-\emph{Hypotheses generation} amplified by concurrently analyzing heterogenous variables at once using correlation measures.
+\emph{Hypotheses generation} was amplified by concurrently analyzing heterogenous variables at once using correlation measures.
 %%
@@
 %%
-\cite{Turkay} derive descriptive data metrics from image data.
+\cite{Turkay} derived descriptive data metrics from image data.
 %%
@@
 %%
-This approach aims to trigger \emph{hypotheses generation} by outlining tendencies between these variables.
+This approach aimed to trigger \emph{hypotheses generation} by outlining tendencies between these variables.
 %%
@@
 %%
-We incorporate hierarchical agglomerative clustering to derive shape groups, yielding average shape groups and outliers.
+We incorporated hierarchical agglomerative clustering to derive shape groups, yielding average shape groups and outliers.
 %%
@@
 %%
-\cite{Zhang} analyze subject groups in a web-based linked view system.
+\cite{Zhang} analyzed subject groups in a web-based linked view system.
 %%
@@
 %%
-They define a cohort as variable-divided subject group, differing from the epidemiological understanding of the term.
+They defined a cohort as variable-divided subject group, differing from the epidemiological understanding of the term.
 %%
@@
 %%
-Both use decision trees for their analyses.
+Both used decision trees for their analyses.
 %%
@@
 %%
-\cite{Glasser2013} use variables derived from DCE-MRI data capturing the perfusion in tumorous tissue.
-%%
-To classify breast tumors, they train a decision-tree classifier, concluding that the extracted kinetic and morphological variables alone are not sufficient for tumor type classification.
-%%
-\cite{Niemann2014} assess risk factors for hepatic steatosis (fatty liver disease) using decision trees.
+\cite{Glasser2013} used variables derived from DCE-MRI data capturing the perfusion in tumorous tissue.
+%%
+To classify breast tumors, they trained a decision-tree classifier, concluding that the extracted kinetic and morphological variables alone are not sufficient for tumor type classification.
+%%
+\cite{Niemann2014} assessed risk factors for hepatic steatosis (fatty liver disease) using decision trees.
 %%
@@
 %Therefore, epidemiologists compiled the data set with a wide range of variables possibly correlating with lumbar back pain.
-\noindent Epidemiologists compiled a data set with a wide range of variables possibly correlating with lumbar back pain, comprising non-image and image-derived data for 6,753 subjects.
+\noindent Our work is based on a data set compiled by Epidemiologists with a wide range of variables possibly correlating with lumbar back pain, comprising non-image and image-derived data for 6,753 subjects.
+%\noindent Epidemiologists compiled a data set with a wide range of variables possibly correlating with lumbar back pain, comprising non-image and image-derived data for 6,753 subjects.
 %%
@@
 	%%
-	\item 113 categorical variables divided into 43 dichotomous (binary) variables, mostly indicating the presence of a disease, e.g., \emph{pancreatitis} or \emph{high blood pressure} and 70 variables with more than two levels, indicating pain levels, nutrition and social factors, such as marital or retirement status.
+	\item 113 categorical variables divided into 43 dichotomous (binary) variables, mostly indicating the presence of a disease, e.g., \emph{pancreatitis} or \emph{high blood pressure}, and 70 variables with more than two levels, indicating pain levels, nutrition and social factors, such as marital or retirement status.
 	% %%
@@
 	%%
-	(b) We extract the weighted sum of \emph{curvature} and \emph{torsion} for all 92 points (white dashes).
-	%%
-	The \emph{curvature angle} ($\alpha$) is calculated using the top, bottom and middle point.
+	(b) We extract the weighted sum of \emph{curvature} and \emph{torsion} for all 92 points (white dashes) and the \emph{curvature angle} ($\alpha$).
+%is calculated using the top, bottom and middle point.
 	%%
@@
 %%
-Larger people have also a longer spine with a straighter shape.
+Larger people also have a longer spine with a straighter shape.
 %%
@@
 %%
-We conclude that \emph{curvature} is not as reliable on the transversal plane as it is on the other planes, which was also confirmed through a Principal Component Analysis (see supplementary material at \url{ivapp15.dnsalias.com}).
-%%
-The \texttt{GPLOM} shows similar distributions of subjects with or without back pain with respect to the shape variables in all sub plots.
+We conclude that \emph{curvature} is not as reliable on the transversal plane as it is on the other planes, which was also confirmed through a principal component analysis (see supplementary material at \url{ivapp15.dnsalias.com}).
+%%
+The \texttt{GPLOM} shows similar distributions of subjects with or without back pain with respect to the shape variables in all sub-plots.
 %%
@@
 %%
-A Generalized Pairs Plot of all image-derived variables colored by presence (red) or absence (turquoise) of back pain.
+A \texttt{GPLOM} of all image-derived variables colored by presence (red) or absence (turquoise) of back pain.
 %%
@@
 %%
-No correlations with back pain are present as it can be seen in this plot.
+No correlations with back pain can be identified in this plot.
 }
@@
 %%
-The method uses \emph{Pearson product-moment} for two continuous variables, \emph{polyserial} correlation for one continuous and one categorical variable, and \emph{polychoric} correlations for two categorical variables.
+The method uses the \emph{Pearson product-moment} for two continuous variables, \emph{polyserial} correlation for one continuous and one categorical variable and \emph{polychoric} correlations for two categorical variables.
 %The method uses the following correlation metrics for the different type combinations:
@@
 %%
-These models are build w.r.t. all input variables and capture more complex relationships than correlation coefficients.
+These models are built w.r.t. all input variables and capture more complex relationships than correlation coefficients.
 %%
@@
 %%
-Readability is only granted for small trees.
+%Readability is only granted for small trees.
 %%
@@
 %%
-The classification algorithm may produce higher accuracy for variables with less subjects due to the small sample size.
-%%
-This makes these results less reliable.
+The classification algorithm may produce higher accuracy for variables with less subjects due to the small sample size, making these results less reliable.
 %%
@@
 %%
-Metric variables, such as \emph{body size} are discretized using their quantiles.
-%%
-This allows to assess the influences of a variable to the classification process.
+Metric variables, such as \emph{body size}, are discretized using their quantiles.
+%%
+This allows to assess the influence of a variable to the classification process.
 %%------------
@@
 %%
-The left scatter plot shows the results for all variables, either metrics expressed via their quantiles, or categorical.
-%%
-The right scatter plot displays the dummy variables derived from the original variables.
+The upper scatter plot shows the results for all variables, either metrics expressed via their quantiles, or categorical.
+%%
+The lower scatter plot displays the dummy variables derived from the original variables.
 %%
@@
 %%
-We only display variables with an error below $0.5$, results above this threshold are inaccurate.
-%%
-The interactive plot (see supplemental material) has clickable data points, displaying corresponding decision tree in a tool tip.
+We only display variables with an error below $0.5$, results above this threshold are dismissed.
+%%
+The interactive plot (see supplemental material) has clickable data points, displaying the corresponding decision tree in a tool tip.
 %%
@@
 %%
-Almost all medicated subjects were correctly classified ($796$ of $1,058$).
+A high share of medicated subjects were correctly classified ($796$ of $1,058$).
 %%
@@
 %%
-The classifier neglects the other group and yields an error below $0.5$.
+The classifier neglects the other groups and yields an error below $0.5$.
 %%
@@
 %%
-Subjects experiencing \emph{pain in the last seven days} can also be described for this group using the same variables with a tree consisting of $8$ rules and an error of $0.35$
+This result also imposes overfitting.
+%%
+Subjects experiencing \emph{pain in the last seven days} can also be described for this group using the same variables with a tree consisting of $8$ rules and an error of $0.35$.
 %The dichotomous variable whether the subject has felt \emph{pain in the last seven days} can also be predicted for this group using the same features.
@@
 %%
-They are an addition to the standard epidemiological workflow as they highlight new possibly unknown relationships.
+They are an addition to the standard epidemiological workflow as they highlight new and possibly unknown relationships.
 %%
@@
 %%
-Dented vertebras hint a pathological deformation, and can be capture by segmenting the top and bottom point of each vertebra center.
+Dented vertebrae indicate pathological deformation, and can be capture by segmenting the top and bottom point of each vertebra center.
 %%

### 2014-12-15 draft.tex
named the visualization decision tree quality plot

@@
 %%
-We provide an IVA technique to compare classifiers with interactive scatter plot visualizations.
+We provide an IVA technique to compare classifiers with interactive \emph{decision tree quality plot}. %scatter plot visualizations.
 %%
@@
 %%
-\item The identification of the most important variables via classification methods using a novel semi- quantitative overview visualization concept. % embedded in our IVA-framework.
+\item The identification of the most important variables via classification methods using a novel \emph{decision tree quality plot}.
 %%
@@
 % \section{\uppercase{Evaluation of Decision Trees}}
-\section{\uppercase{Interactive Decision Tree Visualization}}
+%\section{\uppercase{Interactive Decision Tree Visualization}}
+\section{\uppercase{Interactive Decision Tree Quality Plot}}
 \label{sec:DecisionTrees}
@@
 \noindent \textbf{Creating Decision Trees.}
-We use the C4.5 algorithm, which builds decision trees based on information entropy on a data set.
+We use the C4.5 algorithm, which builds decision trees based on information entropy.
 %%
@@
 %%
-The scatter plot is our central element for the interactive analysis of decision trees.
+This \emph{decision tree quality plot} is our central element for the interactive analysis of decision trees.
 \\\\
@@
 %The error is scaled to denote perfect classification with 0; 1 equals always wrong; 0.5 equals random selection.
-The error represents the share of incorrect classifications and denotes perfect classification with 0; always wrong with 1; and random selection with 0.5.
+The error represents the share of incorrect classifications and denotes perfect classification with 0 and always wrong with 1. %; and random selection with 0.5.
 %%
@@
 %%
-Outlier results with very large decision trees would otherwise distort the resulting plot.
+Outlier results with large decision trees would otherwise distort the resulting plot.
 %%------------
 \\\\
-\noindent \textbf{Interaction with the Decision Tree Visualization.}
+\noindent \textbf{Decision Tree Visualization Quality Plot Interaction.}
 %%
@@
 %%
-We also provide controls for adjusting the maximum classification error and minimum subject count for a variable.
+We provide controls for adjusting the maximum classification error and minimum subject count for a variable.
 %%
@@
 %%
-Interactive scatter plot visualization of all classification results.
+Decision tree quality plot of all classification results.
+%%Interactive scatter plot visualization of all classification results.
 %%
@@
 %% Contributions
-\noindent We provide a method for comparative analysis of decision trees independent of the variable manifestation count using interactive scatter plots.
+\noindent We provide a method for comparative analysis of decision trees independent of the variable manifestation count using a novel \emph{decision tree quality plot}.
 %%

### 2014-12-16 draft.tex
revised version

@@
 %%
-We provide an IVA technique to compare classifiers with interactive \emph{decision tree quality plot}. %scatter plot visualizations.
+We provide an IVA technique to compare classifiers with a \emph{decision tree quality plot}. %scatter plot visualizations.
 %%
@@
 %%
-(1) physicians formulate hypotheses based on observations made in their clinical practice; (2) epidemiologists compile a list of variables depicting a hypothesis to prove it; (3) statistical methods, such as regression analysis, assess the correlation of selected variables with the investigated disease.
+(1) physicians formulate hypotheses based on observations made in their clinical practice; (2) epidemiologists compile a list of variables depicting a hypothesis to prove it; (3) statistical methods, such as regression analysis, assess the correlation of selected variables with the investigated condition.
 %%
@@
 %%
-For example, continuous data, such as body height, is often discretized into quantile bins for visualizing age-dependent disease frequencies.
+For example, continuous data, such as age, is often discretized into quantile bins for visualizing age-dependent disease frequencies.
 
@@
 	%%
-	(b) We extract the weighted sum of \emph{curvature} and \emph{torsion} for all 92 points (white dashes) and the \emph{curvature angle} ($\alpha$).
+	(b) We extract the weighted sum of \emph{curvature} and \emph{torsion} for all 92 points (white dashes) and the \emph{curvature angle} ($\alpha$) for each projection axis to assess their information gain.
 %is calculated using the top, bottom and middle point.
 	%%
-	We extracted the variables for each projection axis to assess their information gain.
+%	We extract the variables for each projection axis to assess their information gain.
 	}
@@
 \\\\
-\noindent \textbf{Visualization of Classification Results.}
+\noindent \textbf{Decision Tree Quality Plot.}
 %%
@@
 %%
-Based on this we use a summary error, which incorporates the discriminative power of each manifestation and is denoted as follows:
+Based on this we use a summary error based on the weighted mean, which incorporates the discriminative power of each manifestation and is denoted as follows:
 %%
@@
 %%
-Furthermore, we map the number of subjects associated with a variable to the diameter in the scatter plot.
+Furthermore, we map the number of subjects associated with a variable to the diameter in the decision tree quality plot.
 %%
@@
 \\\\
-\noindent \textbf{Decision Tree Visualization Quality Plot Interaction.}
+\noindent \textbf{Decision Tree Quality Plot Interaction.}
 %%
@@
 %%
-The analysis was performed for subjects groups of \emph{gender}, \emph{BMI} and \emph{body size} to assess their influence on the lumbar spine shape.
+The analysis was performed for subject groups of \emph{gender}, \emph{BMI} and \emph{body size} to assess their influence on the lumbar spine shape.
 
@@
 %%
-Our method allowed to assess their discriminative power, which is largely limited to separating male and female subjects, \emph{nutrition} variables, as well as different disease indicators.
+Our method allows to assess their discriminative power, which is largely limited to separating male and female subjects, \emph{nutrition} variables, as well as different disease indicators.
 %%
@@
 %%
-Huge result spaces could be navigated fast using our interactive visualization.
+Huge result spaces could be navigated fast using our decision tree quality plot.
 %%
@@
 %%
-Dented vertebrae indicate pathological deformation, and can be capture by segmenting the top and bottom point of each vertebra center.
+Dented vertebrae indicate pathological deformation, and can be captured by segmenting the top and bottom point of each vertebra center.
 %%
@@
 %%
-Combining the power of statistical analysis, visual analytics and classification techniques is essential for analyzing increasingly complex heterogenous population data.
+Combining the power of statistical analyses, visual analytics and classification techniques is essential for analyzing increasingly complex heterogenous population data.
 %%
@@
 %%
-Our method provides a novel way to gain insight into these complex data sets and amplifies hypotheses generation.
+Our method provides a novel way to gain insight into these complex data sets and amplifies \emph{hypotheses generation}.
 %%
@@
 This work was supported by the DFG Priority Program 1335: Scalable Visual Analytics.
+The work was partly funded by the German Ministry of Education and Research (BMBF) within the STIMULATE project (grant no.
+03FO16101A).
 \end{small}

### 2014-12-16 draft.tex
references results figure in the text; revised acknowledgements

@@
 %%
-The upper scatter plot shows the results for all variables, either metrics expressed via their quantiles, or categorical.
-%%
-The lower scatter plot displays the dummy variables derived from the original variables.
+The upper plot shows the results for all variables, either metrics expressed via their quantiles, or categorical.
+%%
+The lower plot displays the dummy variables derived from the original variables.
 %%
@@
 Dedicated statistical analysis of these results solely focussed on epidemiological research is not the scope of this paper.
+%%
+The resulting plots can be seen in Figure~\ref{fig:results}.
 %%
@@
 This work was supported by the DFG Priority Program 1335: Scalable Visual Analytics.
-The work was partly funded by the German Ministry of Education and Research (BMBF) within the STIMULATE project (grant no.
-03FO16101A).
+This work was supported by the federal state of Saxony-Anhalt under grant number 'I 60' within the Forschungscampus STIMULATE.
 \end{small}

### 2014-12-16 Reviews & Rebuttal.md
Added new plot name to rebuttal, removed old images

@@
 - Figure 1 was completely removed since it does not supports the train of thought as we intended and due to lack of space
+- We named our decision tree visualization *decision tree quality plot*
 

### 2014-12-17 draft.tex
removed comments from the LaTeX source
fixed incorrect subject count for image detection

@@
 
-% modify floating spacings 
-% \addtolength{\abovecaptionskip}{-0.0mm}
-% \addtolength{\belowcaptionskip}{-0.0mm}
-% \addtolength{\dbltextfloatsep}{-5.0mm}
-% \addtolength{\textfloatsep}{-4mm}
-% \addtolength{\intextsep}{-5.0mm}
-%
-% % modify paragraph spacings
-% \addtolength{\parskip}{-5.0mm}
-
 \begin{document}
@@
 
-% \author{Blind}
 \author{\authorname{Paul Klemm\sup{1}, Sylvia Glaßer\sup{1}, Kai Lawonn\sup{1}, Marko Rak\sup{1}, Henry Völzke\sup{2}, Katrin Hegenscheid\sup{3} and Bernhard Preim\sup{1}}
@@
 %%
-We provide an IVA technique to compare classifiers with a \emph{decision tree quality plot}. %scatter plot visualizations.
+We provide an IVA technique to compare classifiers with a \emph{decision tree quality plot}.
 %%
@@
 %%
-% So far, epidemiological research is strongly hypothesis-driven.
-% %%
-% More precisely, observations made by clinicians are translated into hypotheses, which are then statistically evaluated using data variables from epidemiological studies.
-% %%
-% As a result, the hypothesis will be accepted or rejected.
-% %%
-%%
 Cohort studies are an instrument of epidemiological research.
 %%
-%Nowadays, these studies often comprise medical image data, e.g., tomographic image data sets. %, as well as general patient-specific attributes, such as \emph{gender}, \emph{age} and \emph{weight}.
-%%
-We analyze back pain, one of the most frequent complaints in the Western civilization. %\cite{Hoy2010}. % and gains importance due to the population aging \cite{Hoy2010}.
-%%
-%Epidemiological analysis of back pain aims to determine risk factors.
+We analyze back pain, one of the most frequent complaints in the Western civilization.
 %%
 Although the shape and constitution of the spine, especially the lumbar spine, plays an important role for back pain, an automatic classification approach for characterization of back pain based on lumbar spine attributes is still missing.
-%%
-%Furthermore, previous analyses mainly focused on a single variable (e.g. \emph{BMI}) or a small subset of variables.
 %%
@@
 Then, Interactive Visual Analysis (IVA) highlights mutual dependencies between image-derived data and back pain-related variables.
-%%
-% The workflow is part of an IVA framework, constructed to enhance the epidemiological workflow with iterative data-driven techniques.
 %%
@@
 \end{itemize}
-%%
-%The analysis, as well as the presented IVA tool is available as supplementary material in an interactive \texttt{R Markdown}\footnote{Developed by RStudio, Inc; \texttt{rmarkdown.rstudio.com}} document (\url{ivapp15.dnsalias.com}).
-%%
 %% ---------------------------------------------
@@
 %%
-% \begin{enumerate}
-% %%
-% 	\item Physicians formulate hypotheses based on observations made in their clinical practice.
-% %%
-% 	\item Epidemiologists compile a list of variables depicting a hypothesis to prove it.
-% %%
-% 	\item Statistical methods, such as regression analysis, assess the correlation of selected variables with the investigated disease.
-% %%
-% \end{enumerate}
-%%
 Mutually dependent variables make this analysis challenging.
@@
 For example, the incidence rate of many diseases, such as different cancer types increases with age \cite{Fletcher}.
-%Many diseases, such as different cancer types, are for example more likely with increasing age \cite{Fletcher}.
-%%
-% For example, the effect of nutrition on prostate cancer requires age normalization due to the increased incidence for older subjects.
-% %%
-% Age acts as a \emph{risk factor} for prostate cancer.
-% %%
-% Statistical correlation does not imply causation--epidemiologists need to assess the medical soundness of the statistical results.
 %%------------------------------
 \paragraph{Challenges of Epidemiological Data Analysis.}
-% Epidemiological data originates from a wide range of studies.
-% %%
-% The study type depends on the condition of interest.
-% %%
-% Most common are case-control studies analyzing one specific disease and its influence on the human body.
 %%
 We focus on epidemiological large scale cohort study data.
-%%
-% These studies aim at collecting as much data as possible for each subject.
 %%
@@
 Segmentation algorithms are not generally available and need to be carefully adapted for each body structure.
-%%
-%Segmentation data is usually analyzed by abstracting it into key figures, such as diameters or distances.
-%%
-%These numeric values can be compared with non-image variables to retrieve correlations.
 %%--------------------------------
@@
 Lower back pain is one of the most frequent complaints in the Western civilization \cite{Hoy2010}.
-%%
-% Epidemiologists assume correlations between lumbar back pain and lifestyle factors.
-% %%
-% These include \emph{nutrition}, \emph{physical activity} and \emph{body posture combined with physical stress at work}.
 %%
@@
 %%
-%The workflow consists of an iterative sequence of \emph{group selection}, \emph{variable selection} and \emph{visualization}.
-%%
 \emph{Hypotheses generation} was amplified by concurrently analyzing heterogenous variables at once using correlation measures.
-%%
-%We incorporate the IVA approach by concurrently displaying complex relationships between variables via decision trees.
-%%
-% \cite{Steenwijk} propose a coordinated linked view system for both image-related and non-image data.
-% %%
-% It incorporates multiple plot types, such as scatter plots, parallel coordinates and time plots. %with brushing and linking facilities.
-% %%
-% Similar to our work, they quantify image data and project it into the non-image data information space.
 %%
@@
 %%
-%The extracted lumbar spine shape was correlated with subject size.
-%%
-%We use the same detection algorithm and lumbar spine canal extraction as foundation of this work.
-%%
 As opposed to clustering the spine shape, we analyze the discriminative power of the resulting data towards back pain and other non-image variables in this work.
-%Clustering techniques are strongly dependent on the chosen variables and distance measures \cite{Klemm2014BVM}.
-%%
 %%
@@
 \texttt{GPLOMs} provide an overview, but take up much screen space and are therefore only suitable for a few variables at once (see example in Fig.~\ref{fig:image-parameter-range}).
-%%
-% \cite{Dai} define a \emph{Concept Map}, linking cancer-associated features by incorporating choropleth maps with bar charts, parallel coordinates and scatter plots with regression lines.
-% %%
-% The \emph{Concept Map} is iteratively refined using primarily the underlying geographical data.
-%%
 %%----------------------------
@@
 \noindent \textbf{Decision Rule-Driven Analysis of Medical Data.}
-%%
-% Related work in this field often focuses on clinical diagnosis.
 %%
@@
 Decision trees are easily readable and are frequently used for classifying medical data.
-%%
-% The effort of analyzing decision trees increases with their complexity.
 %%
@@
 Their interactive data mining tool can analyze association rules and highlights relations.
-%They present an interactive data mining tool, that can analyze association rules and highlight relations. %, which may trigger new hypotheses about the data.
 %%
 We combine both ideas by validating the significance of image-derived variables.
-% %%
-% We aim at deriving new relationships by training decision trees to explain target variables.
-%%----------------------------
-%%
-%\cite{Pinheiro} applied association rule mining to create high-confidence decision rules based on variables, such as age, gender and geographic location about the mortality rate in liver cancer data.
-%%
-% \cite{Sekhavat} propose an interactive visualization technique for decision trees, where all decisions are displayed in a heat map.
-% %%
-% The visualization translates the decision tree into a heat map, where each cell represents the connectivity between an initial and a target value.
-% %%
-% The resulting interactive visualization is well suited for analyzing single, but not multiple decision trees.
 %%
@@
 %%
-% \noindent Our approach allows the simultaneous analysis of many variables.
-%%
-%Therefore, epidemiologists compiled the data set with a wide range of variables possibly correlating with lumbar back pain.
-\noindent Our work is based on a data set compiled by Epidemiologists with a wide range of variables possibly correlating with lumbar back pain, comprising non-image and image-derived data for 6,753 subjects.
-%\noindent Epidemiologists compiled a data set with a wide range of variables possibly correlating with lumbar back pain, comprising non-image and image-derived data for 6,753 subjects.
-%%
-%The data set comprises 6,753 subjects from two cohorts (4,420 from \texttt{SHIP-Trend-0} and 2,333 from \texttt{SHIP-2}) including non-image and image-derived data.
+\noindent Our work is based on a data set compiled by Epidemiologists with a wide range of variables possibly correlating with lumbar back pain, comprising non-image and image-derived data for 3,234 subjects.
 %%
@@
 	\item 113 categorical variables divided into 43 dichotomous (binary) variables, mostly indicating the presence of a disease, e.g., \emph{pancreatitis} or \emph{high blood pressure}, and 70 variables with more than two levels, indicating pain levels, nutrition and social factors, such as marital or retirement status.
-	% %%
-% 	\begin{itemize}
-% 		%%
-% 		\item 43 dichotomous (binary) variables, mostly indicating the presence of a disease, e.g. \emph{pancreatitis} or \emph{high blood pressure}
-% 		%%
-% 		\item 70 variables with more than two levels, indicating pain levels, nutrition and social factors, such as marital or retirement status
-% 		%%
-% 	\end{itemize}
+	%%
 \end{itemize}
@@
 All categorical variables are converted into binary \emph{dummy variables}, indicating the presence or absence of a categorical variable manifestation.
-%with multiple manifestations into several dichotomous variables.
-%%
-%Each dichotomous variable encodes the presence of a manifestation.
 %%
@@
 %%
-%One subject can only have one of these variables set to true.
-%%
-%This allows to determine which manifestation of a variable can be described best using the image data variables.
-%This is useful for our classification, because it allows to determine which manifestation of a variable can be described best using the image data variables.
-%%
-% Many variables comprise missing data.
-% %%
-% Some variables are follow-up questions, covering the treatment of a disease.
-% %%
-% Others are exclusive to subgroups, such as women-specific variables, e.g. \emph{menstrual status}.
-% %%
-% Sparse variables are statistically less resilient and need to be treated with care during the analysis.
 \\\\
@@
 	%%
-%	(a) right: FEM aligned to an MRI data set.
-	%%
-	%The purple dashed line represents the centerline, describing the lumbar spine canal with 92 points.
 	The purple dashed line describes the lumbar spine canal centerline with 92 points.
@@
 	%%
-%	We extract the variables for each projection axis to assess their information gain.
 	}
@@
 %%
-% Four trained radiologists acquired the medical image data for each subject in a standardized way using a 1.5~Tesla Magnetic Resonance Imaging (MRI) scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany).
-% %%
-% The spine protocol consisted of a sagittal $T_1$- and $T_2$-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
-% %%
-%%
 A hierarchical finite element method was used to detect the lumbar spine in the MRI data \cite{Rak2013}.
 %%
-%The tetrahedron-based finite element model (Fig.~\ref{fig:centerline}~a) is initialized using three user-defined landmarks.
+The tetrahedron-based finite element model (Fig.~\ref{fig:centerline}~a) captures information about the lumbar spine canal shape and the position of the L1-L5 vertebrae.
 %%
-%A click on the L3 vertebra center initializes the model, two clicks on the top and bottom of the vertebra describe an initial height estimation (Fig.~\ref{fig:centerline}~a).
-%%
-The tetrahedron-based finite element model (Fig.~\ref{fig:centerline}~a) captures information about the lumbar spine canal shape and the position of the L1-L5 vertebrae.
-%The registered model captures information about the lumbar spine canal shape and the position of the L1-L5 vertebrae.
-%%
-%Each subject's model is mapped to a common reference frame using procrustes analysis to compensate for differences in translation, rotation and scaling.
-%%
-The detection fails for several subjects due to imaging artifacts or strongly deformed spines, yielding models for $2,540$ out of $6,753$ subjects.
-%%
-%$2,540$ out of $6,753$ lumbar spine models were obtained and constitute the foundation of this work.
-%%
-%We have to assess the model accuracy to extract key figures from it.
+The detection fails for several subjects due to imaging artifacts or strongly deformed spines, yielding models for $2,540$ out of $3,234$ subjects.
 %%
@@
 %%
-%It captures reliable information about spine canal curvature.
-%%
 In \cite{Klemm2013VMV}, we extracted a centerline representation of the lumbar spine canal from the detection model (Fig.~\ref{fig:centerline}~a).
-%%
 %%
@@
 %%
-% These metrics are also extracted in the sagittal, coronal and transversal projection of the model.
-% %%
-% We assess the information gain of each dimension using these projections.
-% %%
-% This yields a total of 9 image-derived variables.
 These metrics are also extracted in the sagittal, coronal and transversal projection of the model, yielding 9 image-derived variables.
@@
 \label{sec:Experiments}
-% \begin{figure}[!h]
-%   %\vspace{-0.2cm}
-%   \centering
-%   \includegraphics[width=0.475\textwidth]{figures/age-gender}
-%   \caption{
-% 	%%
-% 	Correlation of \emph{age} and \emph{gender} regarding the lumbar spine size.
-% 	%%
-% 	The bar chart shows subjects divided into four different \emph{age} groups (x-axis) and their subject count (y-axis).
-% 	%%
-% 	Each bar contains the mean lumbar spine of the respective group.
-% 	%%
-% 	The shape color encodes the distance (red for x axis, blue for y axis, green for z axis) to the overall male (top chart) or female (bottom chart) mean shape.
-% 	%%
-% 	The dark gray share of each bar encodes the portion of male (top chart) or female (bottom chart) subjects.
-% 	%%
-% %	Women have a higher live expectance than men, hence their higher share in the old age group.
-% 	%%
-% %	Also, women are on average smaller than men, hence the larger shape similarity with older subjects.
-% 	%%
-% 	%Due to bone erosion, older people are also smaller on average.
-% 	}
-%   \label{fig:age-gender}
-% \end{figure}
-%%
-%The first step is to analyze if the dichotomous back pain variable can be discriminated using single shape variables or sets of shape variables.
 %%
 \noindent In this section, the image-derived variables are analyzed towards the dichotomous \emph{back pain} indicator using a \texttt{GPLOM} and all non-image variables using heterogenous correlations.
-%In this section, we aim to retrieve correlations between the dichotomous \emph{back pain} indicator and the image-derived variables using a \texttt{GPLOM} as well as a assessment, which non-image variables
+%%
 Spine shape is influenced by several somatometric variables.
@@
 %%
-%Since men are on average taller than women and people of old age shrink due to \emph{bone erosion}, \emph{gender} and \emph{age} are also risk factors.
-%%
-%Women have a higher life expectancy than men, hence their higher share in the old age group.
-%%
-%Also, women are on average smaller than men, hence the larger shape similarity with older subjects.
-%%
 High \emph{body weight} increases the spine load, resulting in a bent shape.
@@
 To assess their influence, we take them into account when spine \emph{curvature} and \emph{torsion} are correlated with non-image variables.
-%%
-%Since the \emph{gender} encodes \emph{body size}, we decided to divide subjects into \emph{body size} groups.
 %%
@@
 The metric image-derived variables are pairwise visualized using scatter plots on the left side of the matrix diagonal.
-%Since our image-derived variables are metric, their pairwise combinations are visualized using scatter plots.
 %%
-%The combination of the image variables with back pain is visualized as histogram at the left side of the matrix diagonal and as box plot on the right side.
 The combination of the image variables with back pain is visualized as histogram in the last row and as box plot in the last column.
@@
 The \texttt{GPLOM} shows similar distributions of subjects with or without back pain with respect to the shape variables in all sub-plots.
-%%
 %%--------------------------------
-% \\\\
-% \noindent \textbf{Correlation Matrix.}
 \begin{figure*}[htb]
-  %\vspace{-0.2cm}
   \centering
@@
 \end{figure*}
-% We calculated a correlation matrix to assess correlations between the image variables.
-% %%
-% The pearson correlation coefficients between the numeric variables are depicted right of the matrix diagonal (Fig.~\ref{fig:image-parameter-range}).
-% %%
-% \emph{Curvature}, \emph{curvature angle} and \emph{torsion} strongly correlate with their planar projections.
-% %%
-% Also the \emph{mean curvature} and \emph{curvature angle} correlate by a factor of $-0.89$.
-% %%
-% \emph{Torsion} does not correlate with any other image-derived variable.
-%%--------------------------------------------------------
-% \\\\
-% \noindent \textbf{Correlation of Image Variables With Lumbar Back Pain.}
-% Figure~\ref{fig:image-parameter-range} shows the distribution of all image-variables as Generalized Pairs Plot.
-% %%
-% No statistically significant correlation could be observed through all subject groups.
-% %%
-% The box plots show no difference between subjects with or without back pain.
-% \\\\
-% \noindent \textbf{Assessing the Information Gain Using the Principal Component Analysis.}
-% A Principal Component Analysis (PCA) of the image data yielded that \emph{mean curvature}, \emph{sagittal curvature}, \emph{curvature angle} and \emph{sagittal curvature angle} explain almost 50\% of the image variance.
-% %%
-% \emph{Torsion} and metrics of the \emph{transverse} projection planes contribute very little variance, supporting our conclusion about their low information gain.
 %%
-%%
-% To determine the information gain per image-derived variable, we calculated a Principal Component Analysis (PCA) and compared the dimension loadings.
-% %%
-% The PCA results in a linear projection of the variable space ordering their components by maximum variance.
-% %%
-% The first three principal components explain 75\% of the variance in the nine image-derived variables.
-% %%
-% The first principal component explains 47\% of the variance and comprises primarily \emph{mean curvature}, \emph{sagittal curvature}, \emph{curvature angle} and \emph{sagittal curvature angle}.
-% %%
-% The second component, adding 16\% of the variance, weigths \emph{coronal curvature} and \emph{coronal curvature angle}.
-% %%
-% The third component explains 12\% of the variance and comprises \emph{torsion} and \emph{transverse curvature}.
-% %%
-% This supports our prior conclusion about the low information gain of the transverse planes.
-% %%
-% \emph{Torsion} also adds little variance to the information space.
-% %%
-% The complete analysis can be found in the experiments section of the supplementary material (\url{ivapp15.dnsalias.com})
-%% ------------------------------------------
 \\\\
@@
 The method uses the \emph{Pearson product-moment} for two continuous variables, \emph{polyserial} correlation for one continuous and one categorical variable and \emph{polychoric} correlations for two categorical variables.
-%The method uses the following correlation metrics for the different type combinations:
-%%
-% \begin{itemize}
-% %%
-% \item \emph{Pearson product-moment} for two continuous variables,
-% \item \emph{Polyserial} correlation for one continuous and one categorical variable, and
-% \item \emph{Polychoric} correlations for two categorical variables.
-% \end{itemize}
 %%
@@
 %%
-% \section{\uppercase{Evaluation of Decision Trees}}
-%\section{\uppercase{Interactive Decision Tree Visualization}}
 \section{\uppercase{Interactive Decision Tree Quality Plot}}
@@
 We rely on predictive classification to obtain a complex rule set on how combinations of the image-variables explain non-image variables.
-%%
-%The presented method aims to support an explorative analysis by trying to identify relationships with all non-image variables and also 
-%The nature of this analysis is more explorative as it includes all 
 %%
@@
 Decision trees are easy to understand and to read.
-%%
-%They work with numerical as well as categorical data.
-%%
-%Readability is only granted for small trees.
-%%
-%Large, complex trees are difficult to trace and to understand.
 %%
@@
 In the following analysis, we strongly focus on the complexity of decision trees and the classification accuracy.
-%
-%\subsection{The C4.5/C5.0 Algorithm}
-%\label{sec:C5.0}
-%The C4.5 algorithm builds decision trees based on information entropy on a data set.
 %%
-%Such a calculation requires a numeric or categorical target variable.
-%%
-% The pseudocode for the algorithm is defined in Algorithm~\ref{alg:c4.5} \cite{C4.5Algorithm}.
-% \begin{algorithm}[]
-%  Check for base cases\;
-%  \For{each attribute $a$}{
-%   Find the normalized information gain ratio from splitting on $a$\;
-%  }
-%  Let $a_{best}$ be the attribute with the highest normalized information gain\;
-%  Create a decision node that splits on $a_{best}$\;
-%  Iterate on the sublists obtained by splitting on $a_{best}$, and add those nodes as children of node\;
-%  \caption{Building a decision tree using the C4.5 Algorithm}
-%  \label{alg:c4.5}
-% \end{algorithm}
-%C5.0 is developed to produce smaller decision trees than C4.5 and improve the execution time.
-%%
-%We use the \texttt{R} implementation of C5.0 \cite{c5.0classification}.
-%%
-% Categorical attributes with more levels are biased with more information gain in a decision tree \cite{deng2011bias}.
-% %%
-% Creating a dummy variable by converting each manifestation of a categorical variable into a dichotomous variable bypasses this problem.
-% %%
-% In the following analysis, we strongly focus on the complexity of decision trees and the classification accuracy.
-%\subsection{Visualizing Decision Trees}
 
@@
 We have to abstract the results of the classification to keep the mental effort of interpreting the data low.
-%%
-%This is included in our IVA framework.
 %%------------
@@
 Calculating the mean classification error is imprecise for non-uniform distributions.
-%Normally the error for a classification is calculated with $error = \frac{correctlyClassified_{n}} {N}$ where $n$ is the number of subjects.
-%%
-%The metric usually works well for variables with uniform distribution.
-%%
-%It distorts the result for other distribution types.
 %%
@@
 %%
-%The error is scaled to denote perfect classification with 0; 1 equals always wrong; 0.5 equals random selection.
 The error represents the share of incorrect classifications and denotes perfect classification with 0 and always wrong with 1. %; and random selection with 0.5.
@@
 We display results below $0.5$ in the visualization, a value below $0.25$ represents a good classification.
-%We consider a result rated  or below good enough to display it in the visualization, a value below $0.25$ represents a good classification.
 %%
@@
 %%
-%We still want to be able to display \emph{details-on-demand} and examine a decision tree in detail.
-%%
 \emph{Details-on-demand} are displayed by clicking on an entry in the visualization, which then displays the corresponding decision tree in detail.
@@
 The subject subdivision is controlled by the selection of variables, such as \emph{gender} or \emph{employment status}.
-%Selecting a variable using a drop-down menu allows the user to select the variable used for subject subdivisions, e.g. \emph{gender} or \emph{employment status}.
 %%
@@
 They can use the technique without installing any software.
-%%
-% As opposed to the standard \texttt{R} plots, \texttt{ggvis} allows to adjust visualization variables using user interface controls, such as sliders.
-% %%
-% In order to make the train of thought comprehensible, we used \texttt{RMarkdown}, which allows to create reports by combining \texttt{R} with the \texttt{Markdown} syntax.
-% %%
-% We used \texttt{R Shiny}\footnote{Developed by RStudio, Inc; \texttt{shiny.rstudio.com}} to make the report available as dynamic web application.
-% %%
-% It allows to combine the power of static \texttt{R Markdown} reports with dynamically parameterized \texttt{ggvis} plots.
-% %%
-% Furthermore, calculations based on a prior data selection can be redone within the report.
-%%
-% The web-based approach allows us to quickly exchange results with our collaborating epidemiological experts.
-% %%
-% They can use the technique without installing any software.
-% %%
-% Exchanging the prototype becomes as easy as exchanging a hyperlink.
-% %%
-% The prototype is available at \url{ivapp15.dnsalias.com}.
 %%---------
 \section{\uppercase{Results}}
-% \begin{figure*}[p!]
 \begin{figure*}[!t]
-  %\vspace{-0.2cm}
   \centering
@@
 Decision tree quality plot of all classification results.
-%%Interactive scatter plot visualization of all classification results.
 %%
@@
 %%
-no group (a), subdivision into \emph{male} and \emph{female} subjects (b) and quartiles of \emph{Body Mass Index} (\emph{BMI}) (c). % and quartiles of \emph{body size} (d).
+no group (a), subdivision into \emph{male} and \emph{female} subjects (b) and quartiles of \emph{Body Mass Index} (\emph{BMI}) (c).
 %%
@@
 The groups are: \emph{all subjects}; \emph{males} and \emph{females}; and a subdivision by \emph{Body Mass Index quantiles} ($BMI = \frac{m}{l^2}$ where $m$ is the \emph{body mass} in kilogram and $l$ is the \emph{body size} in meter), yielding the groups (17, 24.7] (24.7, 27.4] (27.4, 30.5] (30.5, 48].
-% We carry out the analysis using different subject groups:
-% \begin{enumerate}
-% 	%%
-% 	\item All subjects,
-% 	%%
-% 	\item subdivision into \emph{male} and \emph{female},
-% 	%%
-% 	\item subdivision by \emph{Body Mass Index quantiles} ($BMI = \frac{m}{l^2}$ where $m$ is the \emph{body mass} in kilogram and $l$ is the \emph{body size} in meter), yielding the groups (17, 24.7] (24.7, 27.4] (27.4, 30.5] (30.5, 48],
-% 	%%
-% 	\item subdivision by \emph{size quantiles}, yielding the groups (139, 164], (164, 171], (171, 177], (177, 202].
-% 	%%
-% \end{enumerate}
 %%
@@
 The first plot shows all original variables, the second all categorical variables transformed into dichotomous \emph{dummy variables}.
-%%
-%The second plot shows all categorical variables transformed into dichotomous dummy variables. % (recall subsection~\ref{sec:C5.0}).
 %%
@@
 The only variable reliably classified in this group is \emph{gender}, which can be described with an error of $0.31$ using 7 rules and incorporates only \emph{curvature-} and \emph{curvature angle}-related variables.
-%
 %%
@@
 Subjects experiencing \emph{pain in the last seven days} can also be described for this group using the same variables with a tree consisting of $8$ rules and an error of $0.35$.
-%The dichotomous variable whether the subject has felt \emph{pain in the last seven days} can also be predicted for this group using the same features.
-%%
-%The resulting tree consists of $8$ rules and has an error of $0.35$.
 %%
@@
 %%
-% \subsection{Size Groups}
-% %%
-% Many previously described results are influenced by \emph{subject size}.
-% %%
-% Differences between \emph{male} and \emph{female} subjects can be explained by the average \emph{body size} difference.
-% %%
-% For example, large subjects are already characterized by their rather straight spine.
-% %
-% The question is whether the inter-group spine variability parameter is sufficient for predicting other variables.
-% %%
-% Dividing subjects into \emph{body size} groups potentially highlights classifications not influenced by \emph{body size}.
-% %%
-% \\\\
-% \noindent \textbf{Large Decision Trees.}
-% Back pain-associated variables can be explained for various \emph{size} groups, but we could not extract universal rules.
-% %%
-% \emph{Radiating back pain} could be described with an error of $0.39$ using 23 rules for subjects between \emph{$171-177~cm$ body size} with \emph{torsion} and \emph{mean curvature}.
-% %%
-% For subjects sized \emph{$177-202~cm$} the error increases to $0.47$ using 20 decision rules.
-% %%
-% There are several decision trees for laboratory values, e.g.
-\emph{alanine aminotransferase} value (relevant for diagnosis of liver or gallbladder illness) in the blood can be described with an error of $0.4$ (\emph{$139-164~cm$}) to $0.36$ (\emph{$164 - 171~cm$}).
-% %%
-% Similar values can be observed for \emph{cholesterol} or \emph{age}.
-% %%
-% Due to the large decision trees, these results are not usable and impose overfitting to the data.
-% %%
-% \\\\
-% \noindent \textbf{Small Decision Trees.}
-% %%
-% The dummy variables show several variables described using only one decision rule with an error between $0.42$ and $0.47$.
-% %%
-% Most of these variables have a dominant manifestation and the classifier shows a low detection precision for the second manifestation.
-% %%
-% These variables include \emph{nutrition}, \emph{thyroid disorder} and \emph{social problems induced by back pain}.
-% %%
 \section{\uppercase{Summary \& Conclusion}}
@@
 The methods presented herein may be applied to comprehensive epidemiological data sets to investigate mutual dependencies among variables and to generate hypotheses on potential associations and subgroups.
-%
+%%
 These hypotheses, however, have to be substantiated by dedicated statistical analyses and replication in independent cohorts.
-% \noindent The presented results indicate that the image-based variables \emph{torsion}, \emph{curvature} and \emph{curvature angle} are not sufficient to characterize pathological changes in the lumbar spine in the given quality.
-% %%
-% They are associated with subject \emph{gender} due to their \emph{body size} difference using \emph{mean curvature} and \emph{mean curvature angle}.
-% %%
-% This is due to the higher curvature of smaller subjects and the significant difference of average \emph{body size} between male and female.
-% %%
-% \emph{Gender} was detected well in the analysis of all subjects as well as subjects divided by \emph{BMI}.
-%%
-% Subjects grouped using \emph{body size} have a similar spine shape and a lower variance.
-% %%
-% The remaining information within these groups is not sufficient to characterize back pain-related variables.
-% %%
-% Associations with certain \emph{nutrition} or \emph{psychological} variables can be observed with a comparatively high error for some \emph{body size} groups.
-% %%
-%%
-% Another observation during the analysis is that large decision trees achieve low error scores.
-%%
-% Their complexity imposes overfitting to the data and they are not suitable for extracting universal rules.
 %%
@@
 %%
-% BLIND
-%As described in a prior work (\cite{Klemm2014VIS}), the suggestion of potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
-% Suggesting potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
-% %%
-% This can be achieved by creating a describe several (dichotomous) target values with a decision tree constructed from all available data
-% We can achieve this also using the presented method by trying to describe several (dichotomous) target values with a decision tree constructed from all available data.
-% %%
-% This allows to assess both the discriminative power of the data set regarding the target variables, and as the most important variables for the classification.
-%%
 \\\\
@@
 Spine canal thickness indicates signs of herniated disc disease and is also of interest.
-%Another interesting metric is the spine canal thickness over the whole spine, capturing early signs of herniated disc disease.
 %%
@@
 
-% \vfill
 \bibliographystyle{style/apalike}
@@
 
-%\section*{\uppercase{Appendix}}
-%\noindent The IVA tool and all shown plots presented in this paper are available as supplementary material under the following link: \url{ivapp15.dnsalias.com}.
-
 \vfill
